### 2022-06-16 main.tex
pre-mmseqs2

\documentclass[aps,rmp,reprint,superscriptaddress,notitlepage,10pt]{revtex4-1}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsthm,amsfonts,amssymb,amscd}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{enumerate}
\usepackage[final]{hyperref}

\graphicspath{{../figs/}{./figs}}

\begin{document}
%\title{PanGraph: scalable multiple genome alignment for pan-genome analysis}
\title{PanGraph: scalable bacterial pan-genome graph construction}
\author{Nicholas Noll}
\affiliation{Kavli Institute for Theoretical Physics, University of California, Santa Barbara \looseness=-5}
\author{Marco Molari}
\affiliation{Swiss Institute of Bioinformatics, Basel, Switzerland \looseness=-5}
\affiliation{Biozentrum, University of Basel, Basel, Switzerland \looseness=-5}
\author{Richard A.~Neher}
\affiliation{Swiss Institute of Bioinformatics, Basel, Switzerland \looseness=-5}
\affiliation{Biozentrum, University of Basel, Basel, Switzerland \looseness=-5}

\begin{abstract}
The genomic diversity of microbes is commonly parameterized as population genetic polymorphisms relative to a reference genome of a well-characterized, but arbitrary, isolate. 
Reference genomes contain a fraction of the microbial \emph{pangenome}, the set of genes observed within \emph{all} isolates of a given species, and are thus blind to both the dynamics of the accessory genome, as well as variation within gene order and copy number.
With the wide-spread usage of long-read sequencing, the number of high-quality, complete genome assemblies has increased dramatically.
Traditional computational approaches towards whole-genome analysis either scale poorly, or treat genomes as dissociated bags of genes, and thus are not suited for this new era.
Here, we present \emph{PanGraph}, a Julia based library and command line interface for aligning whole genomes into a graph, wherein each genome is represented as an undirected path along vertices, which in turn, encapsulate homologous multiple sequence alignments.
The resultant data structure succinctly summarizes population-level nucleotide and structural polymorphisms and can be exported into a several common formats for either downstream analysis or immediate visualization.
\end{abstract}

\maketitle

%\section{Introduction}
During evolution, microbial genomes change by both local modifications and large-scale alterations \cite{arnold2021horizontal}.
Local mutations conserve sequence architecture and only change a few nucleotides by substitution, insertion or deletion.
Conversely, large-scale mutations reorganize the sequence, and involve either the homologous recombination of large segments or wholesale gene loss or gain from the environment.
The accumulation of such polymorphisms over time complicates comparative genomic analyses of present day isolates; microbial genomes are mosaics of homologous sequence, locally related by an asexual phylogenetic tree \cite{sakoparnig2021whole}.
%Scalable inference techniques of this complicated relatedness structure would, in part, spur the advance of critical epidemiological surveillance tools, analogous to NextStrain, for bacterial pathogens.

Recent advances in long-read sequencing have enabled the low-cost assembly of isolated genomes at the quality of reference databases \cite{whibley2021changing}.
The accumulation of many complete genomes promises to rapidly improve our ability to quantify the evolutionary dynamics that drive microbial diversity in natural populations.
Unfortunately, comparative genomic analyses have not kept pace with the technological advance, opting either for reference-based approaches that only partially capture microbial diversity \cite{tettelin2008comparative}  or to construct a \emph{pangenome} that accurately captures nucleotide polymorphisms within genes but approximates structural polymorphisms as simple gene presence-absence relationships \cite{page2015roary,ding2018panx} .
This new era of pangenomics demands novel data structures to encapsulate the \emph{complete} diversity of a given genomic sample set.

In recent years, efforts have focused on generalizing the \emph{pangenome} framework of microbial diversity to graphical models \cite{eizenga2020pangenome}.
At a high level, pangenome graphs generalize the reference sequence coordinate system conventionally used; microbial diversity is, instead, encapsulated by a graph-based data structure, with vertices or edges labelled by homologous DNA sequences, in which a walk recapitulates a subset of mosaic genomes.
As such, pangenome graphs can be thought of as an analogous construct to the well-known assembly graph, in which whole genomes are individual ``reads" of the underlying microbial structure \cite{myers2005fragment}.

While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging.
Colored generalizations of the de Bruijin graph-based assemblers have been successively used to build graphs from large sequence sets, however the underlying efficiency derives from a fixed kmer size which prevents modelling long-range homology \cite{iqbal2012novo,muggli2017succinct}.
An orthogonal approach has been to formulate the inference of the pangenome graph as a multiple genome alignment.
However, current methods either scale poorly to large sets of genomes \cite{darling2010progressivemauve}, focus on comparisons across diverse sets of species across the tree of life at the cost of memory \cite{armstrong2020progressive}, or utilize a reference-guided approach by partitioning genomes first into annotated genes \cite{gautreau2020ppanggolin,colquhoun2021pandora}.

Here we present PanGraph, a Julia \cite{bezanson2017julia} library and command line interface, designed to efficiently align large sets of closely related genomes into a pangenome graph on personal computers.
The resulting graph both compresses the input sequence set and succinctly captures the population diversity at multiple scales ranging from nucleotide mutations and indels to structural polymorphisms driven by inversions, rearrangements, and gene gain/loss.
The underlying graph data structure can be exported into numerous formats for downstream analysis and visualization in software such as Bandage \cite{wick2015bandage}.
%The algorithm, along with extensive documentation, examples, and instructions for installation, are open source and can be found on GitHub.
%We anticipate PanGraph to be immediately useful in understanding microbial evolution in natural populations.

\section{Algorithms and implementation}
\emph{PanGraph} transforms an arbitrary set of genomes into a \emph{graph} that simultaneously compresses the collection of sequences and exhaustively summarizes both the structural and nucleotide-level polymorphisms.
The graph is composed of \emph{pancontigs}, which represent linear multiple-sequence alignments of homologous sequence found within one or more input genomes.
\emph{Pancontigs} are connected by an edge if they are syntenic on at least one input sequence; individual sequences are then recapitulated by contiguous \emph{paths} through the graph.

\emph{PanGraph's} overarching strategy is to approximate multiple-genome alignment by iterative pairwise alignment, in the spirit of progressive alignment tools \cite{darling2010progressivemauve,armstrong2020progressive}.
A guide tree is utilized to linearize the problem's complexity by approximating multiple-sequence alignment as a quenched order of successive pairwise alignments \cite{feng1987progressive}.
Pairwise graph alignment is performed by an all-to-all alignment of the \emph{pancontigs} between both graphs.

\subsection{Guide tree construction}
The alignment guide tree is constructed subject to three design constraints: (i) sequences are aligned sequentially based upon similarity, (ii) the similarity computation scales sub-quadratically with the number of input sequences, and (iii) the resultant tree is balanced.
To this end, we formulate the algorithm as a two step process.
The initial guide tree is constructed by neighbor-joining \cite{saitou1987neighbor}; the pairwise distance between sequences is approximated by the Jaccard distance between sequence minimizers \cite{roberts2004reducing}.
Computationally, each sequence can be sketched into its set of minimizers in linear time while the cardinality of all pairwise intersections can be computed by sorting the list of all minimizers to efficiently count overlaps.
Hence, the pairwise distance matrix is estimated in log-linear time.
The final guide tree is constructed as the balanced binary tree constrained to reproduce the topological ordering of leaves found initially.
See Fig \ref{fig:visualization}A for a graphical depiction of the guide tree.

\subsection{Iterative graph alignment}
The iterative pangraph construction is illustrated in Fig.~\ref{fig:visualization}.
The pangraph is initialized as one subgraph per input genome, each representing its respective genome as a single \emph{pancontig}.
Pairs of subgraphs are aligned in a postorder traversal of the constructed guide tree.
The identified homologous intervals of the pancontig are then merged, thereby creating shorter contigs that represent homologous sections of multiple input genomes, see Fig. \ref{fig:visualization}B for a graphical deption. 
These steps of pairwise graph alignment and merging of homologous intervals are repeated until the root of the guide tree is reached.

\paragraph{Pairwise alignment.}
Full genome alignment between two closely related isolates is a well-studied problem with many high-quality tools available \cite{li2018minimap2,marccais2018mummer4}.
We chose to use \emph{minimap2} as the core pairwise genome aligner for its proven speed, sensitivity, and easy-to-use, exposed library API \cite{li2018minimap2}.
However, we note that \emph{PanGraph} is written to be modular; in the future, additional alignment kernels can be added with ease.
The \emph{minimap2} alignment kernel is included within a custom Julia wrapper, available at \url{github.com/nnoll/minimap2_jll.jl}.
The kernel interface expects two lists of \emph{pancontigs} to align as input and will output a list of putative overlapping alignments between both input lists.

\paragraph{Merging of homologous sequence.}
\emph{Pancontigs} encapsulate linear multiple-sequence alignments which are modelled internally by a star phylogeny, i.e. are assumed to be well-described by a reference sequence augmented by independent SNPs and indels for each contained isolate.
Importantly, during the all-to-all alignment phase, \emph{pancontigs} are aligned based \emph{solely} upon their consensus sequence.
All putative homologous alignments found by \emph{pancontig} alignment are ranked according to the pseudo-energy
\begin{equation}
    E = -\ell + \alpha N_c + \beta N_m
\end{equation}
where $\ell$, $N_c$, and $N_m$ denote the alignment length, number of \emph{pancontigs} created by the merger, and number of polymorphisms per genome in the newly created {pancontig} respectively.
See Fig \ref{fig:visualization}B for a graphical definition for each term.
Additionally, $\alpha$ and $\beta$ are hyperparameters of the algorithm, respectively controlling the tradeoff between fragmentation of the graph and the maximum diversity within each block.
Only alignments with negative pseudo-energy are performed.

\begin{figure}[htb]
    \includegraphics[width=.5\textwidth]{algorithm.png}
    \caption{{\bf Overview of PanGraph algorithm}.
    (A) The alignment graph is constructed progressively by aligning graphs pairwise up a guide tree constructed from neighbor-joining the minimizer overlap between strains. 
    (B) During pairwise alignment, \emph{pancontigs} (blue and green)
    are merged by identifying homologous intervals (shown in yellow).
    If the underlying alignments are viewed compatible, i.e. the energy is less than 0, the pancontigs are merged.
    }
    \label{fig:visualization}
\end{figure}

At the graph level, the merger of two \emph{pancontigs} defines a new \emph{pancontig}, connected on both sides by edges to the neighboring \emph{pancontigs} of both inputs, and thus locally collapses the two graphs under consideration.
At the nucleotide level, the pairwise alignment of two \emph{pancontigs} maps the reference of one onto the other; the merger of two \emph{pancontigs} requires the application of the map onto the underlying multiple-sequence alignment.
Once both sets of sequences are placed onto a common coordinate system, the resultant consensus sequence, and thus polymorphisms, are recomputed.
This procedure can be viewed as an online multiple sequence alignment algorithm.

The above procedure is repeated until no alignments with negative energy remain.
Upon completion, transitive edges within the graph are removed by merging adjacent \emph{pancontigs}.

\subsection{Parallelism}
\emph{PanGraph} is designed with a message-passing architecture to enable scalable parallelism, as shown in Fig \ref{fig:visualization}A for a cartoon example.
Each internal node of the guide tree represents a job that performs a single pairwise graph alignment between its two children.
The process will block until both of its children processes have completed and subsequently pass the result of their pairwise graph alignment up to the parent.
All jobs run concurrently from the start of the algorithm; the Julia scheduler resolves the order of dependencies naturally \cite{bezanson2017julia}.
As such, the number of parallel computations is automatically scaled to the number of available threads allocated by the user at the onset of the alignment.

\subsection{Graph Export and Availability}
The constructed pangenome graph can be exported in a variety of file formats for downstream analysis and visualization.
In addition to a custom JSON schema, PanGraph can export the alignment as a GFA file, where each \emph{pancontig} is represented as a segment and each genome as a path.
This allows for visualization in software such as Bandage \cite{wick2015bandage}.
Lastly, we provide functionality to export as a conventional pangenome, however wherein \emph{pancontigs} supplant putative gene clusters, that can be visualized directly by the PanX toolkit \cite{ding2018panx}.

PanGraph is published under an MIT license with source code, extensive documentation, examples, and instructions for installation available at \url{github.com/neherlab/pangraph}.
Additionally, pre-built binaries are available as GitHub releases.
All data and scripts used to validate PanGraph are available within the same repository.

\section{Validation and performance}
In order to assess the accuracy of the pangenome graph inferred by PanGraph, and to quantify its performance characteristics as a function of input size, we generated synthetic data.
Specifically, we simulated populations of size $N=100$ and sequence length $L=50000$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (each with independent rates set to $1e-2$ per generation), and horizontal sequence transfer that occurs with tunable rate $r$ per generation per genome.
The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
The simulation framework is distributed within the PanGraph command line toolsuite for external use.

\begin{figure}[htb]
    \includegraphics[width=.5\textwidth]{paper-accuracy-20.png}
    \caption{{\bf Accuracy against synthetic data}. Cumulative density functions for breakpoint accuracy comparing simulated and reconstructed pangenome graphs for various SNP rates, $\mu$.}
    \label{fig:toy-accuracy}
\end{figure}

We quantified the accuracy of PanGraph algorithm by computing the distance, $\delta$, of the inferred breakpoints relative to their known locus stored from the evolutionary simulation.
Critically, we found that the accuracy was \emph{independent} of the rate of horizontal gene transfer $r$ and thus underlying graph complexity.
The predominant determinant of accuracy was the sample diversity, shown in Fig. \ref{fig:toy-performance} which displays the distributions of $\delta$ for $100$ independent simulations ran for each mutation rate $\mu$.
For approximately clonal populations, $\sim 80\%$ of breakpoints were estimated to be within $10$bp away from the ``true" location.
Conversely, for samples with average pairwise diversity above $2\%$, $\sim 80\%$ of breakpoints are estimated only within $100$ bp.
This limitation is primarily driven by lack of sensitivity in the pairwise alignment phase performed using \emph{minimap2} \cite{li2018minimap2}.

\begin{figure}[htb]
    \includegraphics[width=.5\textwidth]{benchmark.png}
    \caption{{\bf Algorithm performance}.
        PanGraph scales linearly with the number of input genomes.
        This is a direct result of the guide tree simplification.
        The solid line and ribbons display the mean and standard deviation respectively.
        All runs were performed on an AMD Ryzen 7 2700X utilizing 8 cores.
    }
    \label{fig:toy-performance}
\end{figure}

The algorithmic complexity was measured empirically by simulating populations of increasing size and average sequence length.
The mean and standard deviation of results obtained from 25 iterations are shown in Fig. \ref{fig:toy-performance}.
Importantly, PanGraph's computational complexity grows \emph{linearly} with the number of input genomes.
We note that the PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of Minimap2 \cite{li2018minimap2}.

We additionally validated PanGraph on natural populations sampled from RefSeq \cite{o2016reference}.
Obviously the ``true" breakpoints are not known \emph{a priori} and thus the direct analog of Fig. \ref{fig:toy-accuracy} is not possible.
However, we postulate that most \emph{bona fide} breakpoints generated from evolutionary events will happen at or near gene boundaries; there is preliminary evidence for this assumption \cite{oliveira2017chromosomal}.
We compared the empirical distribution of distances to gene boundaries to a null distribution generated from the data, in which breakpoints are sampled uniformly from each genome, constrained to produce equivalent number of partitions as was measured by PanGraph.
For all tested cases, we find our boundaries cluster closer to genes than would be expected by chance.
The median distance from a Pangraph boundary to annotated gene ranges between 50-100 bp, depending on the species, while the random breakpoint is, on average, 600 bp off.

\begin{figure}[htb]
    \includegraphics[width=.5\textwidth]{panx-compare.png}
    \caption{{\bf Pancontigs cluster towards gene boundaries}.
    The distribution of the nearest gene boundary for each pancontig breakpoint, shown for five species, with samples obtained from RefSeq \cite{o2016reference}.
    (Left) Empirical distribution obtained over all genomes.
    (Right) Null distribution constructed by randomly sampling positions on each genome, constrained to conserve the number of breakpoints.
    }
    \label{fig:panx-accuracy}
\end{figure}

The computational time and maximum memory required to build each alignment is summarized in Table \ref{table:panx-performance}.

\begin{table}[hb]
    \caption{{\bf Real-world Performance}.
    Elapsed time and maximum RAM usage needed to compute the pangraph for the following collections of RefSeq genomes.
    All runs were performed on an AMD Ryzen 7 2700X utilizing 8 cores.
    }
    \begin{tabular}{l r r r }
    \hline\hline
    Species & Sample size & Wall time (min) & Max memory (GB) \\
    \hline
    E.~coli         & 307 & 426 & 11.7 \\
    H.~pylori       & 85 & 102 & 1.2 \\
    K.~pneumoniae   & 109 & 132 & 2.1 \\
    M.~tuberculosis & 51 & 91 & 0.4 \\
    P.~marinus      & 10 & 12 & 0.1 \\
    \hline
    \end{tabular}
    \label{table:panx-performance}
\end{table}

\section{Discussion}
While single nucleotide differences in the core genome are straightforward to analyze with existing tools, these analysis miss the great majority of genetic diversity. 
The ability to rapidly align large sets of closely related, fully assembled genomes is a critical tool for the investigation of the processes governing microbial evolution.
We hope PanGraph will spur novel insight microbial diversity and the processes by which bacteria adapt and change.

Currently, PanGraph is limited to collections of genomes that are closely related with typical sequence divergence smaller that than 3\%. 
This limitation originates from the core tool used for pairwise alignment, \texttt{minimap2}, that was determined to lose sensitivity if sequences are sufficiently diverged under the chosen parameters.
Homologous sequence that is more diverged than this cut-off will not be reliably merged by pangraph.
This effectively limits the TMRCA of sequences within a block. 
More diverged sequence, for example originating from a cross-species transfer event, will thus be represented as alternative path.
In addition to the intrinsic limit of \texttt{minimap2}, this cut-off can be tuned with the parameter $\alpha$ and can for example be used to restrict merging to very similar sequences and thus capture all recent horizontal events as divergent paths in the graph.

We emphasize that PanGraph was designed as a modular tool.
In the future, we anticipate utilizing other pairwise ``alignment" kernels in addition to minimap2 that will allow the user to trade-off speed for sensitivity.

\section*{Acknowledgement}
We are grateful to Boris Shraiman for stimulating discussions.
This study was funded by the University of Basel and the NSF.

\bibliography{cite}{}

\end{document}

### 2022-09-05 main.tex
Update on Overleaf.

@@
 Full genome alignment between two closely related isolates is a well-studied problem with many high-quality tools available \cite{li2018minimap2,marccais2018mummer4}.
-We chose to use \emph{minimap2} as the core pairwise genome aligner for its proven speed, sensitivity, and easy-to-use, exposed library API \cite{li2018minimap2}.
-However, we note that \emph{PanGraph} is written to be modular; in the future, additional alignment kernels can be added with ease.
-The \emph{minimap2} alignment kernel is included within a custom Julia wrapper, available at \url{github.com/nnoll/minimap2_jll.jl}.
-The kernel interface expects two lists of \emph{pancontigs} to align as input and will output a list of putative overlapping alignments between both input lists.
+We chose to use \emph{minimap2} as the core pairwise genome aligner for its proven speed, sensitivity, and easy-to-use exposed library API \cite{li2018minimap2}.
+This alignment kernel is included within a custom Julia wrapper, available at \url{github.com/nnoll/minimap2_jll.jl}.
+However, we note that \emph{PanGraph} is written to be modular, and additional alignment kernels can be added with ease.
+In particular we decided to include the option to use \emph{mmseqs2} \cite{steinegger2017mmseqs2} as an alternative alignment kernel, because of its sensitivity on highly diverged sequences at the cost of higher computational time.
+In general, the kernel interface expects two lists of \emph{pancontigs} to align as input and will output a list of putative overlapping alignments between both input lists.
 
@@
 \section{Validation and performance}
+
+\subsection{Validation on synthetic data}
+
 In order to assess the accuracy of the pangenome graph inferred by PanGraph, and to quantify its performance characteristics as a function of input size, we generated synthetic data.
 Specifically, we simulated populations of size $N=100$ and sequence length $L=50000$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
-In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (each with independent rates set to $1e-2$ per generation), and horizontal sequence transfer that occurs with tunable rate $r$ per generation per genome.
+In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (each with independent rates set to $10^{-2}$ per generation), and horizontal sequence transfer that occurs with tunable rate $r$ per generation per genome.
 The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
@@
 \begin{figure}[htb]
-    \includegraphics[width=.5\textwidth]{paper-accuracy-20.png}
-    \caption{{\bf Accuracy against synthetic data}. Cumulative density functions for breakpoint accuracy comparing simulated and reconstructed pangenome graphs for various SNP rates, $\mu$.}
+    \includegraphics[width=.4\textwidth]{figs/median_misplacement_vs_divergence.pdf}
+    \caption{{\bf Accuracy against synthetic data}}
     \label{fig:toy-accuracy}
@@
 The predominant determinant of accuracy was the sample diversity, shown in Fig.
-\ref{fig:toy-performance} which displays the distributions of $\delta$ for $100$ independent simulations ran for each mutation rate $\mu$.
+\ref{fig:toy-accuracy} which displays the distributions of $\delta$ for $100$ independent simulations ran for each mutation rate $\mu$.
 For approximately clonal populations, $\sim 80\%$ of breakpoints were estimated to be within $10$bp away from the ``true" location.
@@
 \begin{figure}[htb]
-    \includegraphics[width=.5\textwidth]{benchmark.png}
+    \includegraphics[width=.4\textwidth]{benchmark.pdf}
     \caption{{\bf Algorithm performance}.
@@
 The algorithmic complexity was measured empirically by simulating populations of increasing size and average sequence length.
-The mean and standard deviation of results obtained from 25 iterations are shown in Fig. \ref{fig:toy-performance}.
+The mean and standard deviation of results obtained from 50 iterations are shown in Fig. \ref{fig:toy-performance}.
 Importantly, PanGraph's computational complexity grows \emph{linearly} with the number of input genomes.
 We note that the PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of Minimap2 \cite{li2018minimap2}.
+
+\subsection{Validation on real data}
 
@@
 
-\begin{figure}[htb]
-    \includegraphics[width=.5\textwidth]{panx-compare.png}
-    \caption{{\bf Pancontigs cluster towards gene boundaries}.
-    The distribution of the nearest gene boundary for each pancontig breakpoint, shown for five species, with samples obtained from RefSeq \cite{o2016reference}.
-    (Left) Empirical distribution obtained over all genomes.
-    (Right) Null distribution constructed by randomly sampling positions on each genome, constrained to conserve the number of breakpoints.
-    }
-    \label{fig:panx-accuracy}
-\end{figure}
+\begin{figure*}[htb]
+    \includegraphics[width=.8\textwidth]{figs/incr_size.pdf}
+    \caption{{\bf Pangenome graph properties vs. dataset size}}
+    \label{fig:panx-size}
+\end{figure*}
+
+\begin{figure*}[htb]
+    \includegraphics[width=.7\textwidth]{figs/panx_benchmark.pdf}
+    \caption{{\bf Benchmark on real data}.}
+    \label{fig:panx-benchmark}
+\end{figure*}
 
@@
 \end{table}
+
+\begin{figure*}[htb]
+    \includegraphics[width=.8\textwidth]{figs/proj_single_escherichia_coli.pdf}
+    \caption{{\bf Test of graph marginalization}.}
+    \label{fig:marginalization}
+\end{figure*}
 

### 2022-09-05 supplementary.tex
Update on Overleaf.

\documentclass[aps,rmp,reprint,superscriptaddress,notitlepage,10pt,onecolumn]{revtex4-1}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsthm,amsfonts,amssymb,amscd}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{enumerate}
\usepackage[final]{hyperref}

\newcommand{\avg}[1]{\langle #1 \rangle}

\graphicspath{{../figs_suppl/}{./figs_suppl}}

\begin{document}
\title{PanGraph: scalable bacterial pan-genome graph construction \\ Supplementary Information}
\author{Nicholas Noll}
\affiliation{Kavli Institute for Theoretical Physics, University of California, Santa Barbara \looseness=-5}
\author{Marco Molari}
\affiliation{Swiss Institute of Bioinformatics, Basel, Switzerland \looseness=-5}
\affiliation{Biozentrum, University of Basel, Basel, Switzerland \looseness=-5}
\author{Richard A.~Neher}
\affiliation{Swiss Institute of Bioinformatics, Basel, Switzerland \looseness=-5}
\affiliation{Biozentrum, University of Basel, Basel, Switzerland \looseness=-5}

\maketitle

\section{Generation of synthetic data}

As a first test we benchmark the performances and accuracy of \textit{PanGraph} on synthetic data, obtained by simulating the evolution of a population in which individuals can loose part of the sequence from deletions, or gain new genetic material via Horizontal Gene Transfer (HGT) from other members of the population.\\
In the simulation a population of size $N$ is evolved for $T$ generations using a Wright-Fisher model \cite{hudson2002generating}. Each individual in the population has a circular genome of initial size $L$. At each generation mutations occur at a rate $\mu$ per position per generation.\\
Deletions can occur at a rate $d$ per genome per generation. When a deletion is suggested on isolate $n$, having a genome length $L_n$, a new desired length $L'_n$ is extracted from a Gaussian distribution with mean $L$ and variance $\sigma^2$. If $\Delta = L'_n - L_n < 0$ then a random chunk of length $|\Delta|$ is removed from the genome, reducing it to size $L'_n$. If $\Delta > 0$ no deletion is performed. This ensures that the length distribution of genomes in the population remains close to the desired length $L$, with $\sigma^2$ being a proxy for the variance.
HGT occurs at a rate $h$ per genome per generation. Similarly to deletions, when a HGT event is suggested a new desired length $L'_n$ is extracted from the same Gaussian distribution, and this time the event is performed only if $\Delta = L'_n - L_n > 0$. In this case a random chunk of sequence of length $\Delta$ is extracted from a random individual in the parent population (possibly the parent of isolate $n$ itself) and inserted in a random position of the genome of isolate $n$.
Finally, inversions occur at a rate $i$ per genome per generation. When an inversion is proposed a random length $\Delta$ is chosen in the same way, and a random chunk of length $|\Delta|$ is inverted.
Standard values of the simulation parameters are reported in Table~\ref{table:sim-params}.

\begin{table}[hb]
    \begin{tabular}{c l l}
    \hline\hline
    Parameter & Description & Standard Value \\
    \hline
    $N$ & population size & 100 \\
    $T$ & n. of simulated generations & 50 \\
    $L$ & average genome size & $5 \cdot 10^5$ [bp] \\
    $\sigma$ & s.t.d. of genome size & $L/10$ [bp]  \\
    $\mu$ & mutation rate & 0.005 [1/bp $\cdot$ generation] \\
    $d$ & deletion rate & 0.05 [1/generation] \\
    $i$ & inversion rate & 0.01 [1/generation] \\
    $h$ & HGT rate & 0.1 [1/generation] \\
    \hline
    \end{tabular}
    \caption{{\bf Simulation parameters}. Description of all simulation parameters used for the generation of synthetic data. Unless otherwise specified, the standard value is used.}
    \label{table:sim-params}
\end{table}

At the end of the simulation the resulting mosaic genomes are collected, along with the real underlying pangenome graph that can be used as the ground truth against which to check the performances of \textit{PanGraph}. The code to perform these simulations is shipped with \textit{PanGraph} in the \verb|generate| command.

\section{Benchmark of \textit{PanGraph} on synthetic data}

We test the performance of \textit{PanGraph} on synthetic data generated with the simulation described above. We focus on two aspects: the computational performances of the algorithm (time and memory requirements) as a function of the dataset size, and the accuracy of the algorithm in reconstructing the real pangenome graph of the population as a function of sequence divergence.

\subsection{Computational performance}

To test the computational performances of \textit{PanGraph} we generated data using the model described in the previous section, with standard value of the parameters but varying the average genome length $L=[1,5,10,50,100,500]$ kbp and the population size $N=[10,20,50,100,200,500,100]$. For each pair of $N,L$ values we generated 50 different datasets. On each dataset we ran the \textit{PanGraph} \verb|build| command, using \textit{minimap2} as alignment kernel with \textit{asm20} option. We used standard value for the energy parameters $\alpha=100$, $\beta=10$. Runs were performed using 8 cores. For each run we measured the wall-time of the command, the maximum memory requirements (maximum resident size) and the average cpu percent. Results are displayed in Fig.~\ref{fig:benchmark-perf-suppl}.
Thanks to the guide tree architecture of \textit{PanGraph} the run time scales almost linearly with the number of isolates.

\begin{figure}[h]
    \includegraphics[width=.9\textwidth]{figs_suppl/benchmark_suppl.pdf}
    \caption{{\bf Computational performance of PanGraph algorightm on synthetic data, as a function of dataset size}. We run PanGraph on artificially generated datasets consisting of a variable number of isolates with genomes of variable length (colorscale). For each condition we display the average and standard deviation over 50 different runs of: (A) algorithm wall-time, (B) maximum memory requirement and (C) average percentage of cpu used on a total of 8 cores.}
    \label{fig:benchmark-perf-suppl}
\end{figure}

\subsection{Accuracy in reconstructing the pangenome graph}

To test the accuracy of \textit{PanGraph} in reconstructing the pangenome graph of a set of isolates we generated artificial data using the procedure described above, obtaining both a set of genome sequences and the underlying real pangenome graph. This was done using standard value of the parameters, but varying the rate of HGT in the interval $h=[0.01,1]$ and the mutation rate $\mu=[0,0.01]$. For each $(h,\mu)$ pair 25 different set of data were generated.\\
On each set of data we executed \textit{PanGraph} with values for the energy parameters $\alpha=0$ and $\beta=0$, and with three different alignment kernels:
\begin{itemize}
    \item \textit{minimap2} with option \textit{asm10}
    \item \textit{minimap2} with option \textit{asm20}
    \item \textit{mmseqs2}
\end{itemize}
Thus obtaining three different pangenome graphs per dataset.\\
To link the mutation rate parameter $\mu$ with the average pairwise SNPs distance $\avg{d}$ between homologous segments in the population we evaluated the average pairwise distance for every pancontig in the pangenome graph comprising more than a single path. For each graph we then perform the weighted average of the divergences, using as weight the length of the panconting. Finally, for every pair of parameters $(\mu,h)$ we average these numbers over the 25 different simulations. Results are displayed in Fig.~\ref{fig:snps-suppl}.\\
While the rate of HGT does not influence the average pairwise divergence of pancontigs, the choice of alignment kernel controls their maximum divergence. \textit{Minimap2} is able to correctly merge pangenome graphs with up to $\avg{d} \sim 7\%$ with the \textit{asm20} option, while \textit{mmseqs2} reaches $\avg{d} \sim 10\%$ at the expense of higher time requirements.\\
By performing a linear fit on the datapoints with $h \leq 0.002$ we are able to recover the conversion factor between the HGT rate and average pairwise divergence of pancontigs in our simulations: $\avg{d} \sim 20.8 \, h$.

\begin{figure}[h]
    \includegraphics[width=0.5\textwidth]{figs_suppl/snps_rate_vs_divergence.pdf}
    \caption{{\bf mutation rate vs divergence on synthetic data}. We compare the mutation rate $\mu$ of our simulations with the average pairwise divergence of sequences in pancontigs of the resulting pangenome graphs. This is done for different values of the rate of HGT $h$ and for three different alignment kernels. As expected, the divergence is only marginally influenced by the HGT rate, but the saturating value of the divergence depends strongly on the choice of alignment kernel, with \textit{mmseqs2} being able to merge sequences with average divergence higher than 10\%. To find the relationship between the mutation rate $\mu$ and the average pairwise divergence of the sequences we peform a linear fit on data points for $\mu \leq 0.002$. This provides the conversion factor $\avg{d} \sim 20.8 \, \mu$.}
    \label{fig:snps-suppl}
\end{figure}

Once this link has been established we group simulations by the value of $h$ and check for accuracy comparing the reconstructed pangenome graphs with the ground truth provided by our simulation. In particular for each isolate we consider the breakpoints between different pancontigs that tile the genome, and measure the displacement between the real position of these breakpoints and the position reconstructed by pangraph.\\
To evaluate this displacement it is first necessary to link breakpoints on the real and reconstructed pangenome graph. We formulate this problem as an assignment problem, and solve it numerically using the Hungarian method \cite{kuhn1955hungarian}. For each isolate we call $b_i$ for $i=1,\ldots,I$ the panconting breakpoints on the real pangenome graph, and $b_j$ for $j=1,\ldots,J$ the ones on the reconstructed pangenome graph.\footnote{To avoid artifacts generated by the default threshold distance $d=100$ bp of pangraph, if on any of the two pangenome graphs a pair of breakpoints sits at a distance smaller than this threshold, we remove one of them from the set of breakpoints.} We define a cost matrix $D_{ij} = d(b_i,b_j)$ whose elements are distances on the genome of the isolate of pairs of breakpoints on the two graphs. We then numerically solve the assignment problem on this matrix, and obtain a set of $K = \min\{ I,J \}$ pairings between breakpoints such that the distance between the pairs is minimal and all the breakpoints of the graph with minimum number of breakpoints have been paired. The average distance between breakpoints in these pairs is the average breakpoint distance of the graph.\\
For each simulation we obtain a number of average breakpoint distances equal to the number of isolates in the simulation. In Fig.~\ref{fig:benchmark-accuracy-suppl} we plot the cumulative distribution of these distances, considering only distances smaller than 1 kbp, stratifying simulations according to the average pairwise divergence of pancontigs inferred from the value of $\mu$. Each panel correspond to a different alignment kernel. As the sequence diversity increases we observe a clear transition. For low diversity most of the breakpoints are inferred to be only a few bp away from their real position, while for highly diverged sequences the position of breakpoints is not precisely inferred and can be hundreds of bp away from their real position. For different alignment kernels this transition occurs at different values of the divergence, with \textit{mmseqs2} being accurate at higher divergence values, as can be visualize from Fig.~XX in the main text.

\begin{figure}[h]
    \includegraphics[width=\textwidth]{figs_suppl/accuracy_comparison.pdf}
    \caption{{\bf Accuracy of PanGraph algorithm on synthetic data}.}
    \label{fig:benchmark-accuracy-suppl}
\end{figure}

\section{Benchmark of \textit{PanGraph} on real data}

\begin{figure}[h]
    \includegraphics[width=\textwidth]{figs_suppl/panx_benchmark_suppl.pdf}
    \caption{{\bf Benchmark on real data}.}
    \label{fig:panx-benchmark-suppl}
\end{figure}

\section{Graph marginalization}

Given a graph composed of many different genomes, PanGraph is capable of marginalizing this graph on a subset of strains with the \verb|marginalize| command. To test this feature we built pangenome graphs on genomes from 5 different species: \textit{Klebsiella Pneumoniae} (KP), \textit{Helicobacter Pylori} (HP), \textit{Prochlorococcus Marinus} (PM), \textit{Mycobacterium Tuberculosis} (MT) and \textit{Escherichia Coli} (EC), extracted from RefSeq.\footnote{The list of accession numbers for each species and pairs is available on the repository: \url{https://github.com/neherlab/pangraph/blob/feat/finalize-paper/script/config/projection_strains.json}} For each strain we took 50 different strains (10 for PM) and built an unique pangenome graph using the \verb|build| command with standard values of the parameters. For each species we then randomly picked 50 different pairs of strains (45 for PM) and extracted the corresponding sub-graphs using the \verb|marginalize| command. We compared each of these graphs to their counterpart generated simply by using the \verb|build| command on the corresponding pair of strains.\\
The comparison between the marginalized (MG) and pairwise (PW) graph consists in splitting each genome using the overlap of the breakpoints given by the two different graphs. Then for each resulting segment we check whether it is shared with the other strain of the pair on either of the graphs. This gives raise to four different possibilities, corresponding to the segment being shared (1) or not (0) on either of the two graphs (MG/PW). We consider four different classes of segments: \textit{agree on sharing} (1-1 or 0-0), \textit{shared on both} (1-1), \textit{private on both} (0-0) and \textit{disagree on sharing} (1-0 or 0-1). For each class we measure the total and average length of the segments belonging to that class, and display the results in Fig.~\ref{fig:marginalize-suppl}.\\
For all species except for PM we find that MG graphs are compatible with their PW counterpart: segments on which the graphs \textit{disagree on sharing} represent only a minor fraction of the genome (much smaller than the fraction of shared or private segments) and their size is compatible with the default block length threshold of PanGraph (100 bp). PM represents and exception, with the great majority of segments being private on each graph and having large size. This expected since for PM the strains considered are highly diverged, above the sensitivity of minimap2 ($\sim 27\%$ core divergence). This causes the majority of blocks not to be merged, and generates a strong order dependence. The size of segments with \textit{disagreement on sharing} is also the largest, some segments having size of 10s of kbp.

\begin{figure}[h]
    \includegraphics[width=.8\textwidth]{figs_suppl/proj_fig.pdf}
    \caption{{\bf Marginalized graphs are compatible with pairwise graphs}. Comparison of marginalized (MG) and pairwise (PW) built graphs for 5 different species. For each pair of strains we measure the length of segments in which MG and PW graphs agree on whether the segments are shared on both strains (\textit{agree on sharing}). These can either be both shared (\textit{shared on both}) or both private (\textit{private on both}). We also measure the length of segments on which the MG and PW graphs disagree (\textit{disagree on sharing}). For each of these for cases we report the total length as fraction of the average genome (left column) and the average size of each segment (right column). For each category we display the distribution of average results for 50 different pairs (45 for PM).
    }
    \label{fig:marginalize-suppl}
\end{figure}

\bibliography{cite}{}

\end{document}

### 2022-09-06 supplementary.tex
feat: continued supplementary

@@
 
-\section{Benchmark of \textit{PanGraph} on synthetic data}
+\section{Benchmark of PanGraph on synthetic data}
 
@@
 
-\begin{figure}[h]
+\begin{figure}[htb]
     \includegraphics[width=.9\textwidth]{figs_suppl/benchmark_suppl.pdf}
@@
 
-\begin{figure}[h]
+\begin{figure}[htb]
     \includegraphics[width=0.5\textwidth]{figs_suppl/snps_rate_vs_divergence.pdf}
@@
 
-\begin{figure}[h]
+\begin{figure}[htb]
     \includegraphics[width=\textwidth]{figs_suppl/accuracy_comparison.pdf}
-    \caption{{\bf Accuracy of PanGraph algorithm on synthetic data}.}
+    \caption{{\bf Accuracy of PanGraph algorithm on synthetic data}.
+For each of our generated datasets we evaluate the average displacement of inferred pancontig breakpoints.
+The figure represents the cumulative distribution of average misplacements, stratified by the average pairwise divergence of sequences in pancontigs (see colorbar on the right).
+Each column corresponds to a different alignment kernel used to infer the pangenome graph.
+As divergence increases we observe a transition in the precision of breakpoint position estimation.
+At low divergence the position of most breakpoints is correctly estimated withing a few bps, while at high divergence most breakpoints can be misplaced of multiple hundreds of bps.
+The transition value for the divergence depends on the alignemnt kernel used, with \textit{mmseqs2} having the highest accuracy at the expense of higher computational time.}
     \label{fig:benchmark-accuracy-suppl}
@@
 
-\section{Benchmark of \textit{PanGraph} on real data}
+\section{Benchmark of PanGraph on real data}
 
-\begin{figure}[h]
+We benchmarked \textit{PanGraph} on real data, focusing on the computational performances and the properties of the resulting pangenome graphs.
+
+\subsection{Dataset selection}
+
+We downloaded from RefSeq \cite{o2016reference} complete chromosomes from 5 different bacterial species, listed in Table~\ref{table:panx-dataset}.\footnote{A complete list of accession numbers can be found in...}.
+These genomes had previously been analyzed with PanX \cite{ding2018panx}.
+For each species we also report the average genome lenght and average pairwise divergence on core genome.
+
+% TODO: also report average core genome size?
+% TODO: avoid talking about PanX?
+% TODO: add link with accesion numbers and download?
+
+\begin{table}[hb]
+    \begin{tabular}{l l l l}
+        \hline\hline
+        Species         & N. isolates & Avg. chromosome length [Mbp] & Avg. pairwise divergence (core genes) \\
+        \hline
+        E.~Coli         & 307         & 4.98                         &                                       \\
+        K.~Pneumoniae   & 109         & 5.35                         &                                       \\
+        H.~Pylori       & 85          & 1.62                         &                                       \\
+        M.~Tuberculosis & 51          & 4.40                         &                                       \\
+        P.~Marinus      & 10          & 1.80                         &                                       \\
+        \hline
+    \end{tabular}
+    \caption{{\bf Dataset overview}.
+We downloaded complete chromosome sequences from RefSeq for 5 different bacterial species.
+For each species we report the number of isolates downloaded, along with the average chromosome length and average pairwise divergence on core genes.}
+    \label{table:panx-dataset}
+\end{table}
+
+\subsection{Pangenome graph construction}
+
+We built pangenome graphs using five different options for the alignment kernel:
+\begin{itemize}
+    \item kernel \textit{minimap2} with option \textit{asm10} and default values for the energy parameters $\alpha=100$, $\beta=10$.
+    \item kernel \textit{minimap2} with option \textit{asm20} and default values for the energy parameters $\alpha=100$, $\beta=10$.
+    \item kernel \textit{mmseqs2} and default values for the energy parameters $\alpha=100$, $\beta=10$.
+    \item kernel \textit{minimap2} with option \textit{asm10} and null energy parameters $\alpha=0$, $\beta=0$.
+    \item kernel \textit{mmseqs2} and null energy parameters $\alpha=0$, $\beta=0$.
+\end{itemize}
+% TODO: describe expectations and reasons for the choice
+
+\subsection{Benchmark results}
+
+\begin{figure}[htb]
     \includegraphics[width=\textwidth]{figs_suppl/panx_benchmark_suppl.pdf}
@@
 Given a graph composed of many different genomes, PanGraph is capable of marginalizing this graph on a subset of strains with the \verb|marginalize| command.
-To test this feature we built pangenome graphs on genomes from 5 different species: \textit{Klebsiella Pneumoniae} (KP), \textit{Helicobacter Pylori} (HP), \textit{Prochlorococcus Marinus} (PM), \textit{Mycobacterium Tuberculosis} (MT) and \textit{Escherichia Coli} (EC), extracted from RefSeq.\footnote{The list of accession numbers for each species and pairs is available on the repository: \url{https://github.com/neherlab/pangraph/blob/feat/finalize-paper/script/config/projection_strains.json}} For each strain we took 50 different strains (10 for PM) and built an unique pangenome graph using the \verb|build| command with standard values of the parameters.
+To test this feature we built pangenome graphs on genomes from 5 different species: \textit{Klebsiella Pneumoniae} (KP), \textit{Helicobacter Pylori} (HP), \textit{Prochlorococcus Marinus} (PM), \textit{Mycobacterium Tuberculosis} (MT) and \textit{Escherichia Coli} (EC), extracted from RefSeq.\footnote{The list of selected accession numbers for each species and pair is available on the repository: \url{https://github.com/neherlab/pangraph/blob/feat/finalize-paper/script/config/projection_strains.json}} For each strain we took 50 different strains (10 for PM) and built an unique pangenome graph using the \verb|build| command with standard values of the parameters.
 For each species we then randomly picked 50 different pairs of strains (45 for PM) and extracted the corresponding sub-graphs using the \verb|marginalize| command.
@@
 
-\begin{figure}[h]
+\begin{figure}[htb]
     \includegraphics[width=.8\textwidth]{figs_suppl/proj_fig.pdf}

### 2022-09-06 supplementary.tex
feat: continue supplementary

@@
 We formulate this problem as an assignment problem, and solve it numerically using the Hungarian method \cite{kuhn1955hungarian}.
-For each isolate we call $b_i$ for $i=1,\ldots,I$ the panconting breakpoints on the real pangenome graph, and $b_j$ for $j=1,\ldots,J$ the ones on the reconstructed pangenome graph.\footnote{To avoid artifacts generated by the default threshold distance $d=100$ bp of pangraph, if on any of the two pangenome graphs a pair of breakpoints sits at a distance smaller than this threshold, we remove one of them from the set of breakpoints.} We define a cost matrix $D_{ij} = d(b_i,b_j)$ whose elements are distances on the genome of the isolate of pairs of breakpoints on the two graphs.
+For each isolate we call $b_i$ for $i=1,\ldots,I$ the panconting breakpoints on the real pangenome graph, and $b_j$ for $j=1,\ldots,J$ the ones on the reconstructed pangenome graph.\footnote{To avoid artifacts generated by the default threshold distance $d=100$ bp of pangraph, if on any of the two pangenome graphs a pair of breakpoints sits at a distance smaller than this threshold, we remove one of them from the set of breakpoints.} We define a cost matrix $D_{ij} = d(b_i,b_j)$ whose elements are distances on the genome of pairs of breakpoints from the two graphs.
 We then numerically solve the assignment problem on this matrix, and obtain a set of $K = \min\{ I,J \}$ pairings between breakpoints such that the distance between the pairs is minimal and all the breakpoints of the graph with minimum number of breakpoints have been paired.
@@
 
-We downloaded from RefSeq \cite{o2016reference} complete chromosomes from 5 different bacterial species, listed in Table~\ref{table:panx-dataset}.\footnote{A complete list of accession numbers can be found in...}.
+We downloaded from RefSeq \cite{o2016reference} complete chromosomes from 5 different bacterial species:\footnote{A complete list of accession numbers can be found in...} \textit{Klebsiella Pneumoniae} (KP), \textit{Helicobacter Pylori} (HP), \textit{Prochlorococcus Marinus} (PM), \textit{Mycobacterium Tuberculosis} (MT) and \textit{Escherichia Coli} (EC).
+The number of isolates selected per species, average chromosome length and average pairwise divergence on core genes are reported in Table~\ref{table:panx-dataset}.
 These genomes had previously been analyzed with PanX \cite{ding2018panx}.
-For each species we also report the average genome lenght and average pairwise divergence on core genome.
 
@@
         \hline
-        E.~Coli         & 307         & 4.98                         &                                       \\
-        K.~Pneumoniae   & 109         & 5.35                         &                                       \\
-        H.~Pylori       & 85          & 1.62                         &                                       \\
-        M.~Tuberculosis & 51          & 4.40                         &                                       \\
-        P.~Marinus      & 10          & 1.80                         &                                       \\
+        E.~Coli         & 307         & 5.0                          &                                       \\
+        K.~Pneumoniae   & 109         & 5.3                          &                                       \\
+        H.~Pylori       & 85          & 1.6                          &                                       \\
+        M.~Tuberculosis & 51          & 4.4                          &                                       \\
+        P.~Marinus      & 10          & 1.8                          &                                       \\
         \hline
@@
 \end{itemize}
-% TODO: describe expectations and reasons for the choice
+From the results described in the previous section we expect \textit{mmseqs2} to be able to merge sequences with higher divergence than \textit{minimap2}.
+Moreover from the definition of the pseudo-energy \textbf{eq.~1} in the main text, the value of the energy parameters $\beta$ defines an upper threshold for sequence divergence $d \sim 1/\beta$.
+Homologous sequences with divergence higher than this value will result in positive pseudo-energy and will not be merged in a single pancontig.
+For $\beta=10$ the critical divergence threshold is $d \sim 10\%$.
+Executing pangraph with the option $\alpha=0$, $\beta=0$ will result in the merging of more diverged sequences, at the cost of a more fragmented pangenome graph (more but shorter pancontigs).
 
 \subsection{Benchmark results}
+
+In Fig.~\ref{fig:panx-benchmark-suppl} we display the results of the benchmakr.
+On every panel we display on the x-axis results for different species, and indicate with different colors the five possible choices for the alignment kernel.
 
@@
 
+% TODO: simplify dataset descritpion (is already described above)
+
 Given a graph composed of many different genomes, PanGraph is capable of marginalizing this graph on a subset of strains with the \verb|marginalize| command.
-To test this feature we built pangenome graphs on genomes from 5 different species: \textit{Klebsiella Pneumoniae} (KP), \textit{Helicobacter Pylori} (HP), \textit{Prochlorococcus Marinus} (PM), \textit{Mycobacterium Tuberculosis} (MT) and \textit{Escherichia Coli} (EC), extracted from RefSeq.\footnote{The list of selected accession numbers for each species and pair is available on the repository: \url{https://github.com/neherlab/pangraph/blob/feat/finalize-paper/script/config/projection_strains.json}} For each strain we took 50 different strains (10 for PM) and built an unique pangenome graph using the \verb|build| command with standard values of the parameters.
+To test this feature we built pangenome graphs on a subset of genomes from the 5 different species we previously analyzed.
+For each species we took 50 different strains (10 for PM)\footnote{The list of selected accession numbers for each species and pair is available on the repository: \url{https://github.com/neherlab/pangraph/blob/feat/finalize-paper/script/config/projection_strains.json}} and built an unique pangenome graph using the \verb|build| command with standard values of the parameters.
 For each species we then randomly picked 50 different pairs of strains (45 for PM) and extracted the corresponding sub-graphs using the \verb|marginalize| command.

### 2022-09-06 supplementary.tex
feat: continue supplementary

@@
 \usepackage{enumerate}
+\usepackage{xr} % reference to other external latex documents
 \usepackage[final]{hyperref}
+
+\externaldocument{main}
 
@@
 This was done using standard value of the parameters, but varying the rate of HGT in the interval $h=[0.01,1]$ and the mutation rate $\mu=[0,0.01]$.
-For each $(h,\mu)$ pair 25 different set of data were generated.\\
+For each $(h,\mu)$ pair 25 different sets of data were generated.\\
 On each set of data we executed \textit{PanGraph} with values for the energy parameters $\alpha=0$ and $\beta=0$, and with three different alignment kernels:
@@
 For low diversity most of the breakpoints are inferred to be only a few bp away from their real position, while for highly diverged sequences the position of breakpoints is not precisely inferred and can be hundreds of bp away from their real position.
-For different alignment kernels this transition occurs at different values of the divergence, with \textit{mmseqs2} being accurate at higher divergence values, as can be visualize from Fig.~XX in the main text.
+For different alignment kernels this transition occurs at different values of the divergence, with \textit{mmseqs2} being accurate at higher divergence values, as can be visualize from Fig.~\ref{fig:toy-accuracy} in the main text.
 
@@
 % TODO: also report average core genome size?
+% TODO: evaluate core genome divergence? Or average mash triangle distance?
 % TODO: avoid talking about PanX?
@@
 From the results described in the previous section we expect \textit{mmseqs2} to be able to merge sequences with higher divergence than \textit{minimap2}.
-Moreover from the definition of the pseudo-energy \textbf{eq.~1} in the main text, the value of the energy parameters $\beta$ defines an upper threshold for sequence divergence $d \sim 1/\beta$.
+Moreover from the definition of the pseudo-energy eq.~\ref{eq:pseudo-energy} in the main text, the value of the energy parameters $\beta$ defines an upper threshold for sequence divergence $d \sim 1/\beta$.
 Homologous sequences with divergence higher than this value will result in positive pseudo-energy and will not be merged in a single pancontig.
@@
     \includegraphics[width=\textwidth]{figs_suppl/panx_benchmark_suppl.pdf}
-    \caption{{\bf Benchmark on real data}.}
+    \caption{{\bf Benchmark on real data}.
+We run \textit{PanGraph} on datasets consisting of 5 different bacterial species, and using 5 different options for the alignment kernel as described in the text.
+        \textbf{A}: number of isolates for each species.
+        \textbf{B}: algoritm wall-time.
+        \textbf{C}: maximum memory usage.
+        \textbf{D}: n. pancontigs in the resulting pangenome graph.
+        \textbf{E}: minimum n. of pancontigs that can add up to 50\% of the total pagenome graph length.
+        \textbf{F}: threshold length such that if all pancontigs longer than this threshold are added they cover more than 50\% of the full pangenome graph.
+        \textbf{G}: total length of the pangenome graph (sum of all pancontig lengths).
+        \textbf{H}: fraction of the pangenome graph that is composed of core pancontigs, i.e. pancontigs that occur exactly once per isolate.
+        \textbf{I}: ratio between the length of the pangenome graph and the total length of all chromosomes included in the graph.
+    }
     \label{fig:panx-benchmark-suppl}
 \end{figure}
+
+% TODO: explain the sense of L50 and N50
 

### 2022-09-06 supplementary.tex
feat: continue supplementary

@@
 
-In Fig.~\ref{fig:panx-benchmark-suppl} we display the results of the benchmakr.
-On every panel we display on the x-axis results for different species, and indicate with different colors the five possible choices for the alignment kernel.
+In Fig.~\ref{fig:panx-benchmark-suppl} we report the results of the benchmark.
+Results are divided on every panel by species, with different colors indicating the five possible choices for the alignment kernel.\\
+Execution times (cf.
+panel B) vary from few minutes for PM to several hours for EC.
+This is consistent with dataset sizes.
+Aligning sequences with \textit{mmseqs2} consistently requires more time than \textit{minimap2}.
+The time required to build a pangraph for the 307 chromosomes of EC is only 5-6 h for \textit{minimap2}, and around 24h for \textit{mmseqs2}.
+The latter also has a higher minimum memory requirement (panel C, around 8 Gb when running on 8 cores) but this is not particularly restrictive when aligning several tens of isolates.\\
+% TODO: describe the properties of the graph (panel D-E-F)
+The total pangenome length (panel G) depends both on the number of isolates and the ability of the alignment kernel to merge homologous sequences.
+When setting the energy parameters to zero more diverged sequences are merged into the same pancontigs and the total pangenome graph size is consistently smaller.
+The total size of the pangenome graph is in general much smaller than the cumulative size of the genomes it contains (panel I), with compression of 1-2\% for graphs with a high number of isolates (EC) or with small accessory genome (MT).
+The fraction of core pancontigs (panel H) is consistent with the species expectation, with MT having the highest core fraction.
+For HP the highest core fraction is reached when using null energy parameters combined with \textit{mmseqs2} as alignment kernel, as this allows the merging of more diverged homologous sequences.
+For the PM dataset, which contains highly diverged isolates, only the latter alignment kernel manages to merge an appreciable fraction of core sequence.
 
@@
 We run \textit{PanGraph} on datasets consisting of 5 different bacterial species, and using 5 different options for the alignment kernel as described in the text.
+Runs were performed using 8 cores.
         \textbf{A}: number of isolates for each species.

### 2022-09-07 supplementary.tex
feat: updated table

@@
 
-% TODO: also report average core genome size?
-% TODO: evaluate core genome divergence? Or average mash triangle distance?
-% TODO: avoid talking about PanX?
+% TODO: describe incuded quantitites better, and the dataset characteristics.
+% TODO: decide what to include.
 % TODO: add link with accesion numbers and download?
+% TODO: double-check numbers
 
 \begin{table}[hb]
-    \begin{tabular}{l l l l}
+    \begin{tabular}{l l l l l l}
         \hline\hline
-        Species         & N. isolates & Avg. chromosome length [Mbp] & Avg. pairwise divergence (core genes) \\
+        Species         & N. isolates & $\avg{l}$ [Mbp] & $\avg{d}$ & $f_p$  & $f_g$  \\
         \hline
-        E.~Coli         & 307         & 5.0                          &                                       \\
-        K.~Pneumoniae   & 109         & 5.3                          &                                       \\
-        H.~Pylori       & 85          & 1.6                          &                                       \\
-        M.~Tuberculosis & 51          & 4.4                          &                                       \\
-        P.~Marinus      & 10          & 1.8                          &                                       \\
+        E.~Coli         & 307         & 5.0             & 1.6\%     & 3.7\%  & 15.4\% \\
+        K.~Pneumoniae   & 109         & 5.3             & 0.8\%     & 17.4\% & 46.5\% \\
+        H.~Pylori       & 85          & 1.6             & 4.2\%     & 30.2\% & 42.2\% \\
+        M.~Tuberculosis & 51          & 4.4             & 0.03\%    & 57.9\% & 61.7\% \\
+        P.~Marinus      & 10          & 1.8             & 24.3\%    & 21.0\% & 45.6\% \\
         \hline
@@
 We downloaded complete chromosome sequences from RefSeq for 5 different bacterial species.
-For each species we report the number of isolates downloaded, along with the average chromosome length and average pairwise divergence on core genes.}
+For each species we report the number of isolates downloaded, along with the average chromosome length $\avg{l}$, average pairwise divergence on core genes $\avg{d}$, fraction of core pangenome $f_p$ and fraction of core genome $f_g$.
+$f_p$ is the fraction of core gene cluster over the total (weighted by length), while $f_g$ is the average fraction of core genes in a genome (weighted by length).}
     \label{table:panx-dataset}

### 2022-09-08 supplementary.tex
Update on Overleaf.

--- 2022-09-06 supplementary.tex
@@
 
-We downloaded from RefSeq \cite{o2016reference} complete chromosomes from 5 different bacterial species:\footnote{A complete list of accession numbers can be found in...} \textit{Klebsiella Pneumoniae} (KP), \textit{Helicobacter Pylori} (HP), \textit{Prochlorococcus Marinus} (PM), \textit{Mycobacterium Tuberculosis} (MT) and \textit{Escherichia Coli} (EC).
-The number of isolates selected per species, average chromosome length and average pairwise divergence on core genes are reported in Table~\ref{table:panx-dataset}.
-These genomes had previously been analyzed with PanX \cite{ding2018panx}.
-
-% TODO: also report average core genome size?
-% TODO: evaluate core genome divergence? Or average mash triangle distance?
-% TODO: avoid talking about PanX?
-% TODO: add link with accesion numbers and download?
+We downloaded sets of annotated and complete chromosomes from RefSeq \cite{o2016reference} for 5 different bacterial species:\footnote{A complete list of accession numbers can be found in...} \textit{Klebsiella Pneumoniae} (KP), \textit{Helicobacter Pylori} (HP), \textit{Prochlorococcus Marinus} (PM), \textit{Mycobacterium Tuberculosis} (MT) and \textit{Escherichia Coli} (EC).
+In Table~\ref{table:panx-dataset} we report some summary statistics for these datasets.
+The number of isolates per species ranges from several hundred isolates for EC to 10 for PM.
+These data were previously analyzed using PanX \cite{ding2018panx}, a computational pipeline to cluster homologous genes together.
+This allowed us to identify core genes, and evaluate their average pairwise sequence divergence $\avg{d}$.
+Moreover we evaluate the fraction $f_p$ of core genes in the pangenome (weighted by gene length) and the average fraction $f_g$ of core gene sequence in a single genome.\\
+EC is the dataset with the highest number of isolates (307) and is also the one with the biggest accessory genome ($f_p \sim 4\%$).
+MT is the species with the smallest sequence divergence ($\avg{d} \sim 0.03\%$) and the biggest core genome ($f_p \sim 58\%$).
+The two datasets with highest sequence divergence are HP ($\avg{d} \sim 4\%$) and PM ($\avg{d} \sim 24\%$).
+For HP the average sequence divergence sits on the limit of what \textit{PanGraph} is able to accurately merge when using the \textit{minimap2} kernel (cf.
+Fig.~\ref{fig:benchmark-accuracy-suppl}), but is merged correctly by the \textit{mmseqs2} kernel.
+The sequence divergence of PM is instead too high and cannot be handled correctly by any of the kernels.
 
 \begin{table}[hb]
-    \begin{tabular}{l l l l}
+    \setlength{\tabcolsep}{9pt}
+    \begin{tabular}{l l c c c c}
         \hline\hline
-        Species         & N. isolates & Avg. chromosome length [Mbp] & Avg. pairwise divergence (core genes) \\
+        Species         & N. isolates & $\avg{l}$ [Mbp] & $\avg{d}$ & $f_p$  & $f_g$  \\
         \hline
-        E.~Coli         & 307         & 5.0                          &                                       \\
-        K.~Pneumoniae   & 109         & 5.3                          &                                       \\
-        H.~Pylori       & 85          & 1.6                          &                                       \\
-        M.~Tuberculosis & 51          & 4.4                          &                                       \\
-        P.~Marinus      & 10          & 1.8                          &                                       \\
+        E.~Coli         & 307         & 5.0             & 1.6\%     & 3.7\%  & 15.4\% \\
+        K.~Pneumoniae   & 109         & 5.3             & 0.8\%     & 17.4\% & 46.5\% \\
+        H.~Pylori       & 85          & 1.6             & 4.2\%     & 30.2\% & 42.3\% \\
+        M.~Tuberculosis & 51          & 4.4             & 0.03\%    & 57.9\% & 61.7\% \\
+        P.~Marinus      & 10          & 1.8             & 24.3\%    & 21.0\% & 45.6\% \\
         \hline
@@
 We downloaded complete chromosome sequences from RefSeq for 5 different bacterial species.
-For each species we report the number of isolates downloaded, along with the average chromosome length and average pairwise divergence on core genes.}
+For each species we report the number of isolates downloaded, along with the average chromosome length $\avg{l}$, average pairwise divergence on core genes $\avg{d}$, fraction of core gene sequence in the pangenome $f_p$ and fraction of core gene sequence in a single genome $f_g$.}
     \label{table:panx-dataset}
@@
 The latter also has a higher minimum memory requirement (panel C, around 8 Gb when running on 8 cores) but this is not particularly restrictive when aligning several tens of isolates.\\
-% TODO: describe the properties of the graph (panel D-E-F)
+We quantify the properties of pancontings by looking at their total number (panel D) the minimum number of pancontigs that include at least 50\% of all the pangenome graph sequence (L50 statistics, panel E) and the minimum length of pancontigs in this minimal set (N50 statistics, panel F).
+As expected, the use of null energy parameter results in general in a more fragmented graph, with a higher number of pancontigs that on average have a shorter length.
+HP is an exception, since it includes some highly-diverged sequences and the use of null energy parameters increases the default threshold divergence $d \sim 10\%$.
+This allows merging of sequences with higher divergence by \textit{mmseqs2} kernel, but not by the default \textit{minimap2} kernel, whose threshold average divergence is itself around 10\% (cf.
+Fig.~\ref{fig:benchmark-accuracy-suppl}).
+A second exception is PM.
+The average divergence of this dataset is far above the capabilities of the alignment kernels, and only the use of \textit{mmseqs2} with null energy parameters allows for a partial merging of some homologous sequences.
+The resulting graph is composed of few very long pancontigs, mostly containing sequence from a single isolate.
+If we exclude this exception, in general the pangenome graphs of our datasets have around $10^3$ / $10^4$ pancontigs, with 50\% of the sequence being contained in around 10\% of them, having minimum length of 1-10 kbp.\\
 The total pangenome length (panel G) depends both on the number of isolates and the ability of the alignment kernel to merge homologous sequences.
@@
 For HP the highest core fraction is reached when using null energy parameters combined with \textit{mmseqs2} as alignment kernel, as this allows the merging of more diverged homologous sequences.
-For the PM dataset, which contains highly diverged isolates, only the latter alignment kernel manages to merge an appreciable fraction of core sequence.
+For the PM dataset, which contains highly diverged isolates, only the latter alignment kernel manages to merge an appreciable fraction of core sequence, but is still unable to merge highly diverged sequences.
 
@@
 
-% TODO: explain the sense of L50 and N50
+% TODO: explain the meaning of L50 and N50
 
 \section{Graph marginalization}
-
-% TODO: simplify dataset descritpion (is already described above)
 
@@
 
-\bibliography{cite}{}
+\bibliography{cite}
 

### 2022-09-13 main.tex
feat: continue supplementary
feat: polished increasing size figure

@@
 All putative homologous alignments found by \emph{pancontig} alignment are ranked according to the pseudo-energy
-\begin{equation}
+\begin{equation}\label{eq:pseudo-energy}
     E = -\ell + \alpha N_c + \beta N_m
@@
 \begin{figure}[htb]
-    \includegraphics[width=.5\textwidth]{algorithm.png}
+    \includegraphics[width=.45\textwidth]{figs/algorithm.png}
     \caption{{\bf Overview of PanGraph algorithm}.
@@
 \begin{figure}[htb]
-    \includegraphics[width=.4\textwidth]{benchmark.pdf}
+    \includegraphics[width=.4\textwidth]{figs/benchmark.pdf}
     \caption{{\bf Algorithm performance}.
@@
 \begin{figure*}[htb]
-    \includegraphics[width=.8\textwidth]{figs/incr_size.pdf}
-    \caption{{\bf Pangenome graph properties vs. dataset size}}
+    \includegraphics[width=.9\textwidth]{figs/incr_size.pdf}
+    \caption{{\bf Pangenome graph properties vs.
+dataset size}.
+We build pangenome graphs with increasing number of isolates from the E.~Coli dataset and measure the scaling of different properties of the pangenome graph.
+To quantify variability, we repeat the pangraph construction 10 times for each size (except the complete graph containing all 307 isolates), with different random sets of isolates.
+Lines and shaded areas represent means and standard deviations over these 10 measurements.
+        \textbf{A}: n.
+of pancontigs in the graph.
+We count the total number of pacontigs (blue), the number of core pancontigs (orange) and the minimum number of pancontigs that contain more than 50\% of the pangenome (L50, black).
+        \textbf{B}: average size of pancontigs (blue), of only core pancontings (orange), and size of the smallest panconting in the minimal set that spans 50\% of the pangenome (N50, black).
+        \textbf{C}: cumulative size of all genomes in the pangenome graph (gray), total pangenome size (blue) and size of core pangenome (orange).
+    }
     \label{fig:panx-size}

### 2022-09-13 main.tex
feat: updated accuracy plot

@@
 \begin{figure}[htb]
-    \includegraphics[width=.4\textwidth]{figs/median_misplacement_vs_divergence.pdf}
-    \caption{{\bf Accuracy against synthetic data}}
+    \includegraphics[width=.4\textwidth]{figs/misplaced_fraction_vs_divergence.pdf}
+    \caption{{\bf Accuracy against synthetic data}.
+        We generate artificial data with varying degree of sequence divergence, and compare the real underlying pangenome graph with the one reconstructed by \textit{PanGraph}, for three different alignment kernels :\textit{minimap2} with \textit{asm10} and \textit{asm20} option, and \textit{mmseqs2}.
+In each comparison we evaluate the misplacement of breakpoints that we can pair on the two graphs within 1kbp.
+The plot displays the fraction of breakpoints that have misplacement greater than the standard \textit{PanGraph} precision threshold of 100bp, as a function of average pairwise sequence divergence.
+Line and shaded area represents mean and standard deviation over 25 repetitions.
+        Graph reconstruction with \textit{minimap2} and \textit{asm20} is accurate up to around 5\% average sequence divergence, while for \textit{mmseqs2} accuracy is maintained until $\sim$12\% divergence.
+    }
     \label{fig:toy-accuracy}
@@
     \includegraphics[width=.9\textwidth]{figs/incr_size.pdf}
-    \caption{{\bf Pangenome graph properties vs.
-dataset size}.
-We build pangenome graphs with increasing number of isolates from the E.~Coli dataset and measure the scaling of different properties of the pangenome graph.
-To quantify variability, we repeat the pangraph construction 10 times for each size (except the complete graph containing all 307 isolates), with different random sets of isolates.
-Lines and shaded areas represent means and standard deviations over these 10 measurements.
+    \caption{{\bf Pangenome graph properties vs. dataset size}.
+        We build pangenome graphs with increasing number of isolates from the E.~Coli dataset and measure the scaling of different properties of the graphs.
+Lines and shaded areas represent mean and standard deviation over 10 different repetitions on random subsets of isolates, except for the final point indicating the full graph (307 isolates).
+        The number and size of pancontigs scales sub-linearly with the number of isolates, with most of the pangenome being included in ~10\% of the pancontigs, and core pancontigs having higher-than-average size.
+\textit{PanGraph} is able to compress hundreds of genomes in a pangenome graph having only a fraction of the size.
         \textbf{A}: n.

### 2022-09-13 main.tex
feat: updated panx benchmark figs
feat: caption of real data benchmark

@@
 
+% TODO: talk more extensively about the meaning of the results?
 \begin{figure*}[htb]
-    \includegraphics[width=.7\textwidth]{figs/panx_benchmark.pdf}
-    \caption{{\bf Benchmark on real data}.}
+    \includegraphics[width=\textwidth]{figs/panx_benchmark.pdf}
+    \caption{{\bf Benchmark on real data}.
+        We build pangenome graphs from fully-assembled chromosomes from 5 different bacterial species.
+For each species we build graphs with three different alignment kernel options (\textit{minimap2} with \textit{asm10} or \textit{asm20} option and \textit{mmseqs2}) and two different settings for the pseudo-energy parameters (standard or null values).
+        \textbf{A}: \textit{PanGraph} wall-time when run in parallel on 8 cores.
+        \textbf{B}: fraction of core pangenome in the pangenome graph.
+        \textbf{C}: sequence compression, measured as the ratio between the pangenome graph size and the cumulative size of all the sequences contained in the graph.
+    }
     \label{fig:panx-benchmark}

### 2022-09-23 supplementary.tex
feat: added updated supplementary table
feat: improved kmer figure
feat: improved projection figures

@@
 
+\begin{table}[hb]
+    \setlength{\tabcolsep}{9pt}
+    \begin{tabular}{lccccc}
+        \hline\hline
+        Species         & N. isolates & $\avg{L_\text{gen}}$ [Mbp] & $L_\text{pang}$ [Mbp] & $L_\text{core}$ [Mbp] & $\avg{d_\text{core}}$ \\
+        \hline
+        E.~Coli         & 307         & 5.0                        & 17.7                  & 0.7                   & 1.6\%                 \\
+        K.~Pneumoniae   & 109         & 5.3                        & 13.0                  & 2.3                   & 0.8\%                 \\
+        H.~Pylori       & 85          & 1.6                        & 1.9                   & 0.6                   & 4.2\%                 \\
+        M.~Tuberculosis & 51          & 4.4                        & 4.1                   & 2.4                   & 0.03\%                \\
+        P.~Marinus      & 10          & 1.8                        & 3.3                   & 0.7                   & 26.9\%                \\
+        \hline
+    \end{tabular}
+    \caption{{\bf Dataset overview}.
+We downloaded complete chromosome sequences from RefSeq for 5 different bacterial species.
+For each species we report the number of isolates downloaded, along with the average chromosome length $\avg{l}$, total size of pangenome  $L_\text{tot}$ and core genome $L_\text{core}$, and average pairwise divergence on core genes $\avg{d}$.}
+    \label{table:panx-dataset-updated}
+\end{table}
+
 \subsection{Pangenome graph construction}
@@
 \begin{figure}[htb]
-    \includegraphics[width=.8\textwidth]{figs_suppl/proj_fig.pdf}
+    \includegraphics[width=.9\textwidth]{figs_suppl/proj_fig.pdf}
     \caption{{\bf Marginalized graphs are compatible with pairwise graphs}.

### 2022-09-27 supplementary.tex
Update on Overleaf.

--- 2022-09-07 supplementary.tex
@@
 
-\externaldocument{main}
-
 \newcommand{\avg}[1]{\langle #1 \rangle}
+\newcommand{\Lcore}{L_\text{core}}
+\newcommand{\Lpang}{L_\text{pang}}
+\newcommand{\Lgen}{L_\text{gen}}
+\newcommand{\dcore}{\langle d_\text{core} \rangle}
 
@@
 \section{Generation of synthetic data}
+\label{supp:data_generation}
 
-As a first test we benchmark the performances and accuracy of \textit{PanGraph} on synthetic data, obtained by simulating the evolution of a population in which individuals can loose part of the sequence from deletions, or gain new genetic material via Horizontal Gene Transfer (HGT) from other members of the population.\\
+As a first test we benchmark the performances and accuracy of \textit{PanGraph} on synthetic data, obtained by simulating the evolution of a population in which individuals accumulate mutations, can loose part of the sequence from deletions, or gain new genetic material via Horizontal Gene Transfer (HGT) from other members of the population.\\
 In the simulation a population of size $N$ is evolved for $T$ generations using a Wright-Fisher model \cite{hudson2002generating}.
 Each individual in the population has a circular genome of initial size $L$.
-At each generation mutations occur at a rate $\mu$ per position per generation.\\
+At each generation single-nucleotide mutations occur at a rate $\mu$ per position per generation.\\
 Deletions can occur at a rate $d$ per genome per generation.
@@
 
-We test the performance of \textit{PanGraph} on synthetic data generated with the simulation described above.
-We focus on two aspects: the computational performances of the algorithm (time and memory requirements) as a function of the dataset size, and the accuracy of the algorithm in reconstructing the real pangenome graph of the population as a function of sequence divergence.
+We test the performance of \textit{PanGraph} on synthetic data, focusing on two aspects: the computational performances of the algorithm (time and memory requirements) as a function of the dataset size, and the accuracy of the algorithm in reconstructing the real pangenome graph of the population as a function of sequence divergence.
 
@@
 
-To test the computational performances of \textit{PanGraph} we generated data using the model described in the previous section, with standard value of the parameters but varying the average genome length $L=[1,5,10,50,100,500]$ kbp and the population size $N=[10,20,50,100,200,500,100]$.
+To test the computational performances of \textit{PanGraph} we generated data using the model described in the previous section, with standard value of the parameters but varying the average genome length $L=[1,5,10,50,100,500]$ kbp and the population size $N=[10,20,50,100,200,500,1000]$.
 For each pair of $N,L$ values we generated 50 different datasets.
@@
 
-To test the accuracy of \textit{PanGraph} in reconstructing the pangenome graph of a set of isolates we generated artificial data using the procedure described above, obtaining both a set of genome sequences and the underlying real pangenome graph.
+To test the accuracy of \textit{PanGraph} in reconstructing the pangenome graph of a set of isolates we generated artificial data using the procedure described in Section~\ref{supp:data_generation}.
+For each simulation we obtain both a set of genome sequences and the underlying real pangenome graph.
 This was done using standard value of the parameters, but varying the rate of HGT in the interval $h=[0.01,1]$ and the mutation rate $\mu=[0,0.01]$.
@@
 \begin{itemize}
-    \item \textit{minimap2} with option \textit{asm10}
-    \item \textit{minimap2} with option \textit{asm20}
-    \item \textit{mmseqs2}
+    \itemsep1pt
+    \item \textit{minimap2} with option \textit{asm10},
+    \item \textit{minimap2} with option \textit{asm20},
+    \item \textit{mmseqs2},
 \end{itemize}
-Thus obtaining three different pangenome graphs per dataset.\\
-To link the mutation rate parameter $\mu$ with the average pairwise SNPs distance $\avg{d}$ between homologous segments in the population we evaluated the average pairwise distance for every pancontig in the pangenome graph comprising more than a single path.
-For each graph we then perform the weighted average of the divergences, using as weight the length of the panconting.
+thus obtaining three different pangenome graphs per dataset.\\
+To link the mutation rate parameter $\mu$ with the average pairwise SNPs distance $\avg{d}$ between homologous segments in the population we evaluated the average pairwise distance for every pancontig in the pangenome graph with depth greater than one.
+For each graph we then perform the weighted average of these divergences, using as weight the length of the panconting.
 Finally, for every pair of parameters $(\mu,h)$ we average these numbers over the 25 different simulations.
 Results are displayed in Fig.~\ref{fig:snps-suppl}.\\
-While the rate of HGT does not influence the average pairwise divergence of pancontigs, the choice of alignment kernel controls their maximum divergence.
-\textit{Minimap2} is able to correctly merge pangenome graphs with up to $\avg{d} \sim 7\%$ with the \textit{asm20} option, while \textit{mmseqs2} reaches $\avg{d} \sim 10\%$ at the expense of higher time requirements.\\
-By performing a linear fit on the datapoints with $h \leq 0.002$ we are able to recover the conversion factor between the HGT rate and average pairwise divergence of pancontigs in our simulations: $\avg{d} \sim 20.8 \, h$.
+While the rate of HGT does not influence the average pairwise divergence of pancontigs, the choice of alignment kernel controls their average divergence.
+With the \textit{asm20} option, \textit{Minimap2} is able to correctly merge pangenome graphs with up to populations with $\avg{d} \sim 7\%$ in our simulations, while \textit{mmseqs2} reaches $\avg{d} \sim 10\%$ at the expense of higher processing time.\\
+By performing a linear fit on the datapoints with $\mu \leq 0.002$ we are able to recover the conversion factor between the mutation rate and average pairwise divergence of pancontigs in our simulations: $\avg{d} \sim 20.8 \, \mu$.
 
@@
     \caption{{\bf mutation rate vs divergence on synthetic data}.
-We compare the mutation rate $\mu$ of our simulations with the average pairwise divergence of sequences in pancontigs of the resulting pangenome graphs.
+We compare the mutation rate $\mu$ of our simulations with the average pairwise divergence $\avg{d}$ of sequences in pancontigs of the resulting pangenome graphs.
 This is done for different values of the rate of HGT $h$ and for three different alignment kernels.
@@
 
-Once this link has been established we group simulations by the value of $h$ and check for accuracy comparing the reconstructed pangenome graphs with the ground truth provided by our simulation.
-In particular for each isolate we consider the breakpoints between different pancontigs that tile the genome, and measure the displacement between the real position of these breakpoints and the position reconstructed by pangraph.\\
+Once this link has been established we group simulations by the value of $h$ and measure accuracy by comparing the reconstructed pangenome graphs with the ground truth provided by our simulations.
+In particular, for each isolate we consider the breakpoints between different pancontigs that tile the genome, and measure the displacement between the real position of these breakpoints and the position reconstructed by pangraph.\\
 To evaluate this displacement it is first necessary to link breakpoints on the real and reconstructed pangenome graph.
 We formulate this problem as an assignment problem, and solve it numerically using the Hungarian method \cite{kuhn1955hungarian}.
-For each isolate we call $b_i$ for $i=1,\ldots,I$ the panconting breakpoints on the real pangenome graph, and $b_j$ for $j=1,\ldots,J$ the ones on the reconstructed pangenome graph.\footnote{To avoid artifacts generated by the default threshold distance $d=100$ bp of pangraph, if on any of the two pangenome graphs a pair of breakpoints sits at a distance smaller than this threshold, we remove one of them from the set of breakpoints.} We define a cost matrix $D_{ij} = d(b_i,b_j)$ whose elements are distances on the genome of pairs of breakpoints from the two graphs.
+For each isolate we call $b_i$ for $i=1,\ldots,I$ the panconting breakpoints on the real pangenome graph, and $b_j$ for $j=1,\ldots,J$ the ones on the reconstructed pangenome graph.\footnote{To avoid artifacts generated by the default threshold distance $d=100$ bp of
+\textit{PanGraph}, if any pair of breakpoints on the same graph sits at a distance smaller than this threshold, we remove one of them.} We define a cost matrix $D_{ij} = d(b_i,b_j)$ whose elements are distances on the genome of pairs of breakpoints from the two graphs.
 We then numerically solve the assignment problem on this matrix, and obtain a set of $K = \min\{ I,J \}$ pairings between breakpoints such that the distance between the pairs is minimal and all the breakpoints of the graph with minimum number of breakpoints have been paired.
@@
 For each simulation we obtain a number of average breakpoint distances equal to the number of isolates in the simulation.
-In Fig.~\ref{fig:benchmark-accuracy-suppl} we plot the cumulative distribution of these distances, considering only distances smaller than 1 kbp, stratifying simulations according to the average pairwise divergence of pancontigs inferred from the value of $\mu$.
-Each panel correspond to a different alignment kernel.
+In Fig.~\ref{fig:benchmark-accuracy-suppl} we plot the cumulative distribution of these distances, with a cutoff at 1kbp.
+We stratify simulations according to the average pairwise divergence of pancontigs inferred from the value of $\mu$.
+Each panel corresponds to a different alignment kernel.
 As the sequence diversity increases we observe a clear transition.
 For low diversity most of the breakpoints are inferred to be only a few bp away from their real position, while for highly diverged sequences the position of breakpoints is not precisely inferred and can be hundreds of bp away from their real position.
-For different alignment kernels this transition occurs at different values of the divergence, with \textit{mmseqs2} being accurate at higher divergence values, as can be visualize from Fig.~\ref{fig:toy-accuracy} in the main text.
+For different alignment kernels this transition occurs at different values of the divergence, with \textit{mmseqs2} being the most accurate, as can be visualize from Fig.~3 in the main text.
 
@@
 The figure represents the cumulative distribution of average misplacements, stratified by the average pairwise divergence of sequences in pancontigs (see colorbar on the right).
-Each column corresponds to a different alignment kernel used to infer the pangenome graph.
+Each panel corresponds to a different alignment kernel used to infer the pangenome graph (see panel title).
 As divergence increases we observe a transition in the precision of breakpoint position estimation.
-At low divergence the position of most breakpoints is correctly estimated withing a few bps, while at high divergence most breakpoints can be misplaced of multiple hundreds of bps.
-The transition value for the divergence depends on the alignemnt kernel used, with \textit{mmseqs2} having the highest accuracy at the expense of higher computational time.}
+At low divergence the position of most breakpoints is correctly estimated withing a few bps, while at high divergence most breakpoints can be misplaced by multiple hundreds of bps.
+The transition value for the divergence depends on the alignemnt kernel used, with \textit{mmseqs2} having the highest accuracy at the cost of higher computational time.}
     \label{fig:benchmark-accuracy-suppl}
@@
 
-We benchmarked \textit{PanGraph} on real data, focusing on the computational performances and the properties of the resulting pangenome graphs.
+We benchmark \textit{PanGraph} on real data, focusing on the computational performances and the properties of the resulting pangenome graphs.
 
@@
 
-We downloaded from RefSeq \cite{o2016reference} complete chromosomes from 5 different bacterial species:\footnote{A complete list of accession numbers can be found in...} \textit{Klebsiella Pneumoniae} (KP), \textit{Helicobacter Pylori} (HP), \textit{Prochlorococcus Marinus} (PM), \textit{Mycobacterium Tuberculosis} (MT) and \textit{Escherichia Coli} (EC).
-The number of isolates selected per species, average chromosome length and average pairwise divergence on core genes are reported in Table~\ref{table:panx-dataset}.
-These genomes had previously been analyzed with PanX \cite{ding2018panx}.
-
-% TODO: describe incuded quantitites better, and the dataset characteristics.
-% TODO: decide what to include.
-% TODO: add link with accesion numbers and download?
-% TODO: double-check numbers
+We downloaded sets of annotated and complete chromosomes from RefSeq \cite{o2016reference} for 5 different bacterial species: \textit{Klebsiella Pneumoniae} (KP), \textit{Helicobacter Pylori} (HP), \textit{Prochlorococcus Marinus} (PM), \textit{Mycobacterium Tuberculosis} (MT) and \textit{Escherichia Coli} (EC).
+In Table~\ref{table:panx-dataset} we report some summary statistics for these datasets.
+The number of isolates per species ranges from several hundreds for EC to 10 for PM.
+These data were previously analyzed using PanX \cite{ding2018panx}, a pipeline for the analysis of bacterial pangenome.
+The pipeline takes as input a set of annotated genomes, and it clusters and aligns homologous genes.
+This allowed us to identify core genes, and evaluate their average pairwise sequence divergence $\avg{d_\text{core}}$.
+We also evaluated the average genome length $
+\avg{\Lgen}$, the total pangenome length $\Lpang$ and core genome length $\Lcore$. These quantities are reported in Table~\ref{table:panx-dataset}.\\
+EC is the dataset with the highest number of isolates (307) and is also the one with the biggest pangenome genome.
+MT is the species with the smallest sequence divergence ($\dcore \sim 0.03\%$) and the biggest core genome.
+The two datasets with highest sequence divergence are HP ($\dcore \sim 4\%$) and PM ($\dcore \sim 24\%$).
+For HP the average sequence divergence sits on the limit of what \textit{PanGraph} is able to accurately merge when using the \textit{minimap2} kernel (cf.
+Fig.~\ref{fig:benchmark-accuracy-suppl}), but is merged correctly by the \textit{mmseqs2} kernel.
+The sequence divergence of PM is instead too high and cannot be handled correctly by any of the kernels.
 
 \begin{table}[hb]
-    \begin{tabular}{l l l l l l}
+    \setlength{\tabcolsep}{9pt}
+    \begin{tabular}{lccccc}
         \hline\hline
-        Species         & N. isolates & $\avg{l}$ [Mbp] & $\avg{d}$ & $f_p$  & $f_g$  \\
+        Species         & N. isolates & $\avg{\Lgen}$ [Mbp] & $\Lpang$ [Mbp] & $\Lcore$ [Mbp] & $\dcore$ \\
         \hline
-        E.~Coli         & 307         & 5.0             & 1.6\%     & 3.7\%  & 15.4\% \\
-        K.~Pneumoniae   & 109         & 5.3             & 0.8\%     & 17.4\% & 46.5\% \\
-        H.~Pylori       & 85          & 1.6             & 4.2\%     & 30.2\% & 42.2\% \\
-        M.~Tuberculosis & 51          & 4.4             & 0.03\%    & 57.9\% & 61.7\% \\
-        P.~Marinus      & 10          & 1.8             & 24.3\%    & 21.0\% & 45.6\% \\
+        E.~Coli         & 307         & 5.0                        & 17.7                  & 0.7                   & 1.6\%                 \\
+        K.~Pneumoniae   & 109         & 5.3                        & 13.0                  & 2.3                   & 0.8\%                 \\
+        H.~Pylori       & 85          & 1.6                        & 1.9                   & 0.6                   & 4.2\%                 \\
+        M.~Tuberculosis & 51          & 4.4                        & 4.1                   & 2.4                   & 0.03\%                \\
+        P.~Marinus      & 10          & 1.8                        & 3.3                   & 0.7                   & 26.9\%                \\
         \hline
@@
 We downloaded complete chromosome sequences from RefSeq for 5 different bacterial species.
-For each species we report the number of isolates downloaded, along with the average chromosome length $\avg{l}$, average pairwise divergence on core genes $\avg{d}$, fraction of core pangenome $f_p$ and fraction of core genome $f_g$.
-$f_p$ is the fraction of core gene cluster over the total (weighted by length), while $f_g$ is the average fraction of core genes in a genome (weighted by length).}
+For each species we report the number of isolates downloaded, along with the average chromosome length $\avg{\Lgen}$, total size of pangenome  $\Lpang$ and core genome $\Lcore$, and average pairwise divergence on core genes $\dcore$.}
     \label{table:panx-dataset}
@@
 \begin{itemize}
+    \itemsep1pt
     \item kernel \textit{minimap2} with option \textit{asm10} and default values for the energy parameters $\alpha=100$, $\beta=10$.
@@
 \end{itemize}
-From the results described in the previous section we expect \textit{mmseqs2} to be able to merge sequences with higher divergence than \textit{minimap2}.
-Moreover from the definition of the pseudo-energy eq.~\ref{eq:pseudo-energy} in the main text, the value of the energy parameters $\beta$ defines an upper threshold for sequence divergence $d \sim 1/\beta$.
+From the results described in the previous section (see Fig.~\ref{fig:benchmark-accuracy-suppl}) we expect \textit{mmseqs2} to be able to merge sequences with higher divergence than \textit{minimap2}.
+Moreover from the definition of the pseudo-energy eq.~(1) in the main text, the value of the energy parameters $\beta$ defines an upper threshold for sequence divergence $d \sim 1/\beta$.
 Homologous sequences with divergence higher than this value will result in positive pseudo-energy and will not be merged in a single pancontig.
 For $\beta=10$ the critical divergence threshold is $d \sim 10\%$.
-Executing pangraph with the option $\alpha=0$, $\beta=0$ will result in the merging of more diverged sequences, at the cost of a more fragmented pangenome graph (more but shorter pancontigs).
+Executing pangraph with the option $\alpha=0$, $\beta=0$ will result in the merging of more diverged sequences, at the cost of a more fragmented pangenome graph containing a higher number of shorter pancontigs.
 
@@
 
-In Fig.~\ref{fig:panx-benchmark-suppl} we report the results of the benchmark.
-Results are divided on every panel by species, with different colors indicating the five possible choices for the alignment kernel.\\
-Execution times (cf.
-panel B) vary from few minutes for PM to several hours for EC.
-This is consistent with dataset sizes.
+The results of the benchmark are displayed in Fig.~\ref{fig:panx-benchmark-suppl}.
+In each panel data are divided by species, with colors and texture identifying the five possible choices for the alignment kernel.\\
+Execution time (cf.
+panel B) depends on the dataset size and ranges from few minutes for PM to several hours for EC.
 Aligning sequences with \textit{mmseqs2} consistently requires more time than \textit{minimap2}.
-The time required to build a pangraph for the 307 chromosomes of EC is only 5-6 h for \textit{minimap2}, and around 24h for \textit{mmseqs2}.
-The latter also has a higher minimum memory requirement (panel C, around 8 Gb when running on 8 cores) but this is not particularly restrictive when aligning several tens of isolates.\\
-% TODO: describe the properties of the graph (panel D-E-F)
+The time required to build a pangenome graph for the 307 chromosomes of EC is only 5-6 h for \textit{minimap2}, and around 24h for \textit{mmseqs2}.
+The latter also has a higher minimum memory requirement (panel C, around 8 Gb when running on 8 cores) but this is comparable to \textit{minimap2} when aligning several tens of isolates.\\
+We quantify the properties of pancontings by looking at three main statistics: their total number (panel D) the size of the minimal set of pancontigs that includes at least 50\% of all the graph sequence (L50 statistics, panel E) and the minimum length of pancontigs in this minimal set (N50 statistics, panel F).
+As expected, the use of null energy parameters results on average in a more fragmented graph, having more and shorter pancontigs.
+In general the pangenome graphs of our datasets have around $10^3$ / $10^4$ pancontigs, with 50\% of the sequence being contained in around 10\% of them, having N50 length of around 1-10 kbp.\\
 The total pangenome length (panel G) depends both on the number of isolates and the ability of the alignment kernel to merge homologous sequences.
 When setting the energy parameters to zero more diverged sequences are merged into the same pancontigs and the total pangenome graph size is consistently smaller.
-The total size of the pangenome graph is in general much smaller than the cumulative size of the genomes it contains (panel I), with compression of 1-2\% for graphs with a high number of isolates (EC) or with small accessory genome (MT).
-The fraction of core pancontigs (panel H) is consistent with the species expectation, with MT having the highest core fraction.
-For HP the highest core fraction is reached when using null energy parameters combined with \textit{mmseqs2} as alignment kernel, as this allows the merging of more diverged homologous sequences.
-For the PM dataset, which contains highly diverged isolates, only the latter alignment kernel manages to merge an appreciable fraction of core sequence.
+The total size of the pangenome graph is in general much smaller than the cumulative size of the genomes it contains (panel I), with compression up to 1-2\% of the total size for graphs with a high number of isolates (EC) or with small accessory genome (MT).
+The fraction of core pancontigs (panel H) is consistent with the species expectation, with MT having the highest core fraction.\\
+
+Two species in our datasets hit the limits of the capabilities of \textit{PanGraph}: HP and PM.\\
+HP contains some homologous sequences with divergence higher than the default threshold $d = 1/\beta \sim 10\%$.
+Removing this threshold by using null energy parameters results in a graph with fewer pancontigs, smaller L50, greater fraction of core pangenome and higher level of sequence compression.
+These are all indications of more complete merging of homologous sequences.
+This effect is more marked for \textit{mmseqs2} than for \textit{minimap2}, in agreement with their different divergence limit (cf.
+Fig.~\ref{fig:benchmark-accuracy-suppl}).
+The HP datasets represents an example in which relaxing the default divergence threshold of pangraph and using a more accurate alignment kernel results in a more complete merging of homologous sequence.\\
+Conversely, the PM dataset contains highly diverged isolates, whose divergence go beyond the capabilities of both alignment kernels ($\dcore \sim 27 \%$, cf.
+Table~\ref{table:panx-dataset}).
+Only \textit{mmseqs2} manages to partially merge some of the least diverged core sequences, but is still unable to merge the majority of the core genes.
+This results in a small core genome fraction and poor sequence compression.
+In its current state \textit{PanGraph} cannot meaningfully merge genome with such high divergence into a pangenome graph.
 
@@
         \textbf{C}: maximum memory usage.
-        \textbf{D}: n. pancontigs in the resulting pangenome graph.
-        \textbf{E}: minimum n. of pancontigs that can add up to 50\% of the total pagenome graph length.
+        \textbf{D}: total number of pancontigs in the pangenome graph.
+        \textbf{E}: minimum number of pancontigs that can add up to 50\% of the total pagenome graph length.
         \textbf{F}: threshold length such that if all pancontigs longer than this threshold are added they cover more than 50\% of the full pangenome graph.

### 2022-09-28 supplementary.tex
Update on Overleaf.
feat: finished supplementary

@@
 
-% TODO: explain the meaning of L50 and N50
-
 \section{Graph marginalization}
 
-Given a graph composed of many different genomes, PanGraph is capable of marginalizing this graph on a subset of strains with the \verb|marginalize| command.
-To test this feature we built pangenome graphs on a subset of genomes from the 5 different species we previously analyzed.
-For each species we took 50 different strains (10 for PM)\footnote{The list of selected accession numbers for each species and pair is available on the repository: \url{https://github.com/neherlab/pangraph/blob/feat/finalize-paper/script/config/projection_strains.json}} and built an unique pangenome graph using the \verb|build| command with standard values of the parameters.
-For each species we then randomly picked 50 different pairs of strains (45 for PM) and extracted the corresponding sub-graphs using the \verb|marginalize| command.
-We compared each of these graphs to their counterpart generated simply by using the \verb|build| command on the corresponding pair of strains.\\
-The comparison between the marginalized (MG) and pairwise (PW) graph consists in splitting each genome using the overlap of the breakpoints given by the two different graphs.
-Then for each resulting segment we check whether it is shared with the other strain of the pair on either of the graphs.
-This gives raise to four different possibilities, corresponding to the segment being shared (1) or not (0) on either of the two graphs (MG/PW).
-We consider four different classes of segments: \textit{agree on sharing} (1-1 or 0-0), \textit{shared on both} (1-1), \textit{private on both} (0-0) and \textit{disagree on sharing} (1-0 or 0-1).
-For each class we measure the total and average length of the segments belonging to that class, and display the results in Fig.~\ref{fig:marginalize-suppl}.\\
-For all species except for PM we find that MG graphs are compatible with their PW counterpart: segments on which the graphs \textit{disagree on sharing} represent only a minor fraction of the genome (much smaller than the fraction of shared or private segments) and their size is compatible with the default block length threshold of PanGraph (100 bp).
-PM represents and exception, with the great majority of segments being private on each graph and having large size.
-This expected since for PM the strains considered are highly diverged, above the sensitivity of minimap2 ($\sim 27\%$ core divergence).
-This causes the majority of blocks not to be merged, and generates a strong order dependence.
-The size of segments with \textit{disagreement on sharing} is also the largest, some segments having size of 10s of kbp.
+Given a pangenome graph containing many different genomes, \textit{PanGraph} is capable of marginalizing it on a subset of strains with the \verb|marginalize| command.
+Marginalization removes undesired paths and merges any remaining transitive pancontigs, i.e.
+pancontigs that are co-linear in all of the remaining paths.
+The resulting graph should be equivalent to the one obtained by directly building a graph on the selected subset of strains, with the added advantage that marginalization is a fast operation.
+This makes it more convenient to build a pangenome graph for the full set of isolates considered, and then marginalizing it multiple times on interesting subsets, rather than building a new graph for each subset.
+
+We used genomes from the dataset described in the previous section to verify that marginalized graphs are compatible with graphs built directly from subset of strains.
+For each species we selected 50 different isolates (10 for PM) and built pangenome graphs using standard parameter values and the \textit{minimap2} kernel with \textit{asm20} option.
+For each species we then randomly picked 50 different pairs of isolates (all 45 for PM).\footnote{The list of selected accession numbers and selected pairs for each species is available on the repository: \url{https://github.com/neherlab/pangraph/blob/feat/finalize-paper/script/config/projection_strains.json}} For each pair we then built pairwise pangenome graphs using the same parameters, and compared them to the corresponding marginalized graph, obtained by projecting the full pangenome graph for the 50 isolates on the selected pair (cf.
+Fig.~6A, main text).\\
+The comparison was performed by considering that each graph partitions a genome into pancontigs.
+Each pancontig can either be shared with the other member of the pair, or can be private to the isolate (cf.
+Fig.~6A, main text).
+For each isolate we consider the two partitions defined by the marginalized and pairwise graph, and generate their intersection.
+Each block in this intersection can belong to two different categories: either the two original partitions agree or they disagree on whether the block is shared.
+Moreover, we can separate blocks on which we find agreement in two further categories: blocks that are shared and blocks that are private on both graphs.\\
+For each pair of genomes and each type of block we evaluate the fraction of the genome that they cover and their average size.
+In Fig.~\ref{fig:marginalize-suppl} we display the distribution of these quantities across all pairs for the different species considered.
+We consider the four categories described above: blocks on which we find agreement (\textit{agree}) or not (\textit{disagree}), and we separate agreement blocks in \textit{shared} and \textit{private} blocks.\\
+In all species except HP and PM we find very good agreement between pairwise and marginalized graphs: segments on which the graphs \textit{disagree} represent only a very minor fraction of the genome (much smaller than the fraction of shared or private blocks), and have size compatible with the default block length threhsold of \textit{PanGraph} (100 bp).
+As described in the previous section, the divergence of the HP dataset sit on the limit of what the \textit{minimap2} alignment kernel can successfully handle.
+In this case the results of merging different graphs together can depend on the order of merging, and pairwise and marginalized graphs can potentially show some disagreement.
+In this case the disagreement remains minor and only extends to few percents of the genome size, with disagreement blocks having size only slightly bigger than the 100 bp threshold of PanGraph.
+PM represents instead an example of a dataset with divergence much beyond the limit of the alignment kernel ($\dcore \sim 27 \%$).
+As a consequence, highly diverged homologous sequences are not merged, and the great majority of segments are private and have large size.
+Results also show a strong order-dependence, with some \textit{disagreement} blocks having size of 10s of kbp.
 
@@
     \includegraphics[width=.9\textwidth]{figs_suppl/proj_fig.pdf}
-    \caption{{\bf Marginalized graphs are compatible with pairwise graphs}.
-Comparison of marginalized (MG) and pairwise (PW) built graphs for 5 different species.
-For each pair of strains we measure the length of segments in which MG and PW graphs agree on whether the segments are shared on both strains (\textit{agree on sharing}).
-These can either be both shared (\textit{shared on both}) or both private (\textit{private on both}).
-We also measure the length of segments on which the MG and PW graphs disagree (\textit{disagree on sharing}).
-For each of these for cases we report the total length as fraction of the average genome (left column) and the average size of each segment (right column).
-For each category we display the distribution of average results for 50 different pairs (45 for PM).
+    \caption{
+        {\bf Marginalized graphs are compatible with pairwise graphs}.
+        We compare marginalized graphs with pairwise built graphs for different random pairs of strains from 5 different species, as described in the text.
+Each graph defines a partition of a genome into segments, that can either be shared in the pair or private to the genome.
+We intersect these two partitions, and categorize segments based on whether the two partitions \textit{agree} or \textit{disagree} on their sharing.
+Agreement segments are further split depending on whether the segment is \textit{shared} or \textit{private} on both graphs.
+The figure displays the distribution of the average size of these segments (right column) and the average fraction of genome that they cover (left column) for the 50 different pairs considered (45 for PM).
+        For all species with sequence divergence compatible with the limit of the alignment kernel used (KP, MT, EC), we find very good agreement between marginalized and pairwise graphs, with disagagreement segment only covering a very small fraction of the genome and having size compatible with \textit{PanGraph} threshold pancontig length (100 bp).
     }

### 2022-09-28 main.tex
feat: improved projection figures
feat: modified main

@@
 \maketitle
+
+% TODO: check consistency of numberings of figures and sections between SI and main text
 
@@
 Lastly, we provide functionality to export as a conventional pangenome, however wherein \emph{pancontigs} supplant putative gene clusters, that can be visualized directly by the PanX toolkit \cite{ding2018panx}.
+% TODO: add GFA version?
 
@@
 All data and scripts used to validate PanGraph are available within the same repository.
+% TODO: add data?
 
@@
 
-In order to assess the accuracy of the pangenome graph inferred by PanGraph, and to quantify its performance characteristics as a function of input size, we generated synthetic data.
-Specifically, we simulated populations of size $N=100$ and sequence length $L=50000$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
-In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (each with independent rates set to $10^{-2}$ per generation), and horizontal sequence transfer that occurs with tunable rate $r$ per generation per genome.
+As a first validation step we use generated synthetic data to quantify the performance characteristics of \textit{PanGraph} as a function of input size, and its accuracy as a function of sequence diversity.
+
+\subsubsection*{Generation of synthetic data}
+
+We simulated populations of size $N=100$ and sequence length $L=50\,000$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
+In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (respective rates $0.01$ and $0.05$ per generation), and horizontal sequence transfer that occurs with tunable rate $h$ per generation per genome (see SI Section I for details).
 The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
 The simulation framework is distributed within the PanGraph command line toolsuite for external use.
+
+\subsubsection*{Performances as a function of dataset size}
+
+The algorithmic complexity was measured empirically by simulating populations of increasing size and average sequence length, and generating pangenome graphs for the resulting sequences.
+The mean and standard deviation of results obtained from 50 iterations are shown in Fig. \ref{fig:toy-performance}.
+Importantly, PanGraph's computational complexity grows \emph{linearly} with the number of input genomes.
+We note that the PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of \textit{minimap2} \cite{li2018minimap2}.
+Benchamrks of CPU and memory usage are reported in SI Section IIA.
+
+\begin{figure}[htb]
+    \includegraphics[width=.4\textwidth]{figs/benchmark.pdf}
+    \caption{{\bf Algorithm performance}.
+        PanGraph scales linearly with the number of input genomes.
+        This is a direct result of the guide tree simplification.
+        The solid line and ribbons display the mean and standard deviation over 50 runs.
+        All runs were performed utilizing 8 cores.
+    }
+    \label{fig:toy-performance}
+\end{figure}
+
+\subsubsection*{Accuracy as a function of sequence divergence}
+
+We quantified the accuracy of PanGraph algorithm by computing the displacement of the inferred breakpoints relative to their known locus stored from the evolutionary simulation (cf.
+SI Section IIB for details).
+We evaluate this displacement by generating datasets with different rates of mutation $\mu$ and HGT $h$.
+For each $(h,\mu)$ pair we perform 25 different simulations, and build pangenome graphs using three different options for the alignment kernel, namely \textit{minimap2} \cite{li2018minimap2} with \textit{asm10} and \textit{asm20} option and \textit{mmseqs2} \cite{steinegger2017mmseqs2}.
+Critically, we found that the accuracy was \emph{independent} of the rate of horizontal gene transfer $h$ and thus underlying graph complexity.
+The predominant determinant of accuracy was the sample diversity, controlled by the mutation rate $\mu$ in our simulations (cf.
+SI Fig.~2).
+For low-diversity samples breakpoints are inferred with accuracy of few basepairs, while for highly diverged isolates most breakpoints are displaced by several hundreds of basepairs (cf.
+SI Fig.~3).
+The choice of alignment kernel influences the threshold diversity at which accuracy is lost.
+
+%  TODO: talk aboit the figure
+For approximately clonal populations, $\sim 80\%$ of breakpoints were estimated to be within $10$bp away from the ``true" location.
+Conversely, for samples with average pairwise diversity above $2\%$, $\sim 80\%$ of breakpoints are estimated only within $100$ bp.
+This limitation is primarily driven by lack of sensitivity in the pairwise alignment phase performed using \emph{minimap2} \cite{li2018minimap2}.
 
@@
 
-We quantified the accuracy of PanGraph algorithm by computing the distance, $\delta$, of the inferred breakpoints relative to their known locus stored from the evolutionary simulation.
-Critically, we found that the accuracy was \emph{independent} of the rate of horizontal gene transfer $r$ and thus underlying graph complexity.
-The predominant determinant of accuracy was the sample diversity, shown in Fig.
+% \subsection{Validation on synthetic data}
+
+% In order to assess the accuracy of the pangenome graph inferred by PanGraph, and to quantify its performance characteristics as a function of input size, we generated synthetic data.
+% Specifically, we simulated populations of size $N=100$ and sequence length $L=50000$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
+% In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (each with independent rates set to $10^{-2}$ per generation), and horizontal sequence transfer that occurs with tunable rate $r$ per generation per genome.
+% The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
+% The simulation framework is distributed within the PanGraph command line toolsuite for external use.
+
+% \begin{figure}[htb]
+%     \includegraphics[width=.4\textwidth]{figs/misplaced_fraction_vs_divergence.pdf}
+%     \caption{{\bf Accuracy against synthetic data}.
+%         We generate artificial data with varying degree of sequence divergence, and compare the real underlying pangenome graph with the one reconstructed by \textit{PanGraph}, for three different alignment kernels :\textit{minimap2} with \textit{asm10} and \textit{asm20} option, and \textit{mmseqs2}.
+In each comparison we evaluate the misplacement of breakpoints that we can pair on the two graphs within 1kbp.
+The plot displays the fraction of breakpoints that have misplacement greater than the standard \textit{PanGraph} precision threshold of 100bp, as a function of average pairwise sequence divergence.
+Line and shaded area represents mean and standard deviation over 25 repetitions.
+%         Graph reconstruction with \textit{minimap2} and \textit{asm20} is accurate up to around 5\% average sequence divergence, while for \textit{mmseqs2} accuracy is maintained until $\sim$12\% divergence.
+%     }
+%     \label{fig:toy-accuracy}
+% \end{figure}
+
+% We quantified the accuracy of PanGraph algorithm by computing the distance, $\delta$, of the inferred breakpoints relative to their known locus stored from the evolutionary simulation.
+% Critically, we found that the accuracy was \emph{independent} of the rate of horizontal gene transfer $r$ and thus underlying graph complexity.
+% The predominant determinant of accuracy was the sample diversity, shown in Fig.
 \ref{fig:toy-accuracy} which displays the distributions of $\delta$ for $100$ independent simulations ran for each mutation rate $\mu$.
-For approximately clonal populations, $\sim 80\%$ of breakpoints were estimated to be within $10$bp away from the ``true" location.
-Conversely, for samples with average pairwise diversity above $2\%$, $\sim 80\%$ of breakpoints are estimated only within $100$ bp.
-This limitation is primarily driven by lack of sensitivity in the pairwise alignment phase performed using \emph{minimap2} \cite{li2018minimap2}.
+% For approximately clonal populations, $\sim 80\%$ of breakpoints were estimated to be within $10$bp away from the ``true" location.
+% Conversely, for samples with average pairwise diversity above $2\%$, $\sim 80\%$ of breakpoints are estimated only within $100$ bp.
+% This limitation is primarily driven by lack of sensitivity in the pairwise alignment phase performed using \emph{minimap2} \cite{li2018minimap2}.
 
-\begin{figure}[htb]
-    \includegraphics[width=.4\textwidth]{figs/benchmark.pdf}
-    \caption{{\bf Algorithm performance}.
-        PanGraph scales linearly with the number of input genomes.
-        This is a direct result of the guide tree simplification.
-        The solid line and ribbons display the mean and standard deviation respectively.
-        All runs were performed on an AMD Ryzen 7 2700X utilizing 8 cores.
-    }
-    \label{fig:toy-performance}
-\end{figure}
+% \begin{figure}[htb]
+%     \includegraphics[width=.4\textwidth]{figs/benchmark.pdf}
+%     \caption{{\bf Algorithm performance}.
+%         PanGraph scales linearly with the number of input genomes.
+%         This is a direct result of the guide tree simplification.
+%         The solid line and ribbons display the mean and standard deviation respectively.
+%         All runs were performed on an AMD Ryzen 7 2700X utilizing 8 cores.
+%     }
+%     \label{fig:toy-performance}
+% \end{figure}
 
-The algorithmic complexity was measured empirically by simulating populations of increasing size and average sequence length.
-The mean and standard deviation of results obtained from 50 iterations are shown in Fig. \ref{fig:toy-performance}.
-Importantly, PanGraph's computational complexity grows \emph{linearly} with the number of input genomes.
-We note that the PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of Minimap2 \cite{li2018minimap2}.
+% The algorithmic complexity was measured empirically by simulating populations of increasing size and average sequence length.
+% The mean and standard deviation of results obtained from 50 iterations are shown in Fig. \ref{fig:toy-performance}.
+% Importantly, PanGraph's computational complexity grows \emph{linearly} with the number of input genomes.
+% We note that the PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of Minimap2 \cite{li2018minimap2}.
 
@@
 \begin{figure*}[htb]
-    \includegraphics[width=.8\textwidth]{figs/proj_single_escherichia_coli.pdf}
+    \includegraphics[width=\textwidth]{figs/marginalize_test.pdf}
     \caption{{\bf Test of graph marginalization}.}

### 2022-09-29 main.tex
Update on Overleaf.

@@
 SI Fig.~3).
-The choice of alignment kernel influences the threshold diversity at which accuracy is lost.
-
-%  TODO: talk aboit the figure
-For approximately clonal populations, $\sim 80\%$ of breakpoints were estimated to be within $10$bp away from the ``true" location.
-Conversely, for samples with average pairwise diversity above $2\%$, $\sim 80\%$ of breakpoints are estimated only within $100$ bp.
-This limitation is primarily driven by lack of sensitivity in the pairwise alignment phase performed using \emph{minimap2} \cite{li2018minimap2}.
+The choice of alignment kernel influences the threshold diversity at which accuracy is lost.\\
+For each alignment kernel and value of average sequence divergence in the population we evaluated the fraction of breakpoints that have displacement greater than the default precision threshold for Pangraph of 100 bp, displayed in Fig.~\ref{fig:toy-accuracy}.
+On the sequences generated by our simulations, \textit{minimap2} with option \textit{asm20} show a loss of accuracy at a threshold pairwise divergence of around 5\%, while \textit{mmseqs2} is accurate up to average sequence divergence of around 12\%, at the cost of higher computational time.
 
@@
 Line and shaded area represents mean and standard deviation over 25 repetitions.
-        Graph reconstruction with \textit{minimap2} and \textit{asm20} is accurate up to around 5\% average sequence divergence, while for \textit{mmseqs2} accuracy is maintained until $\sim$12\% divergence.
+\textit{mmseqs2} maintains accuracy at higher divergence, at the cost of more computational time.
     }
@@
 
-\begin{figure*}[htb]
-    \includegraphics[width=.9\textwidth]{figs/incr_size.pdf}
-    \caption{{\bf Pangenome graph properties vs. dataset size}.
-        We build pangenome graphs with increasing number of isolates from the E.~Coli dataset and measure the scaling of different properties of the graphs.
-Lines and shaded areas represent mean and standard deviation over 10 different repetitions on random subsets of isolates, except for the final point indicating the full graph (307 isolates).
-        The number and size of pancontigs scales sub-linearly with the number of isolates, with most of the pangenome being included in ~10\% of the pancontigs, and core pancontigs having higher-than-average size.
-\textit{PanGraph} is able to compress hundreds of genomes in a pangenome graph having only a fraction of the size.
-        \textbf{A}: n.
-of pancontigs in the graph.
-We count the total number of pacontigs (blue), the number of core pancontigs (orange) and the minimum number of pancontigs that contain more than 50\% of the pangenome (L50, black).
-        \textbf{B}: average size of pancontigs (blue), of only core pancontings (orange), and size of the smallest panconting in the minimal set that spans 50\% of the pangenome (N50, black).
-        \textbf{C}: cumulative size of all genomes in the pangenome graph (gray), total pangenome size (blue) and size of core pangenome (orange).
-    }
-    \label{fig:panx-size}
-\end{figure*}
+\subsubsection*{Dataset characterization}
+
+\subsubsection*{Benchmark}
 
@@
 
-The computational time and maximum memory required to build each alignment is summarized in Table \ref{table:panx-performance}.
+\subsubsection*{Scaling with dataset size}
 
-\begin{table}[hb]
-    \caption{{\bf Real-world Performance}.
-        Elapsed time and maximum RAM usage needed to compute the pangraph for the following collections of RefSeq genomes.
-        All runs were performed on an AMD Ryzen 7 2700X utilizing 8 cores.
-    }
-    \begin{tabular}{l r r r }
-        \hline\hline
-        Species         & Sample size & Wall time (min) & Max memory (GB) \\
-        \hline
-        E.~coli         & 307         & 426             & 11.7            \\
-        H.~pylori       & 85          & 102             & 1.2             \\
-        K.~pneumoniae   & 109         & 132             & 2.1             \\
-        M.~tuberculosis & 51          & 91              & 0.4             \\
-        P.~marinus      & 10          & 12              & 0.1             \\
-        \hline
-    \end{tabular}
-    \label{table:panx-performance}
-\end{table}
+\section{Graph marginalization}
 
@@
 \end{figure*}
+
+% \subsection{Validation on real data}
+
+% We additionally validated PanGraph on natural populations sampled from RefSeq \cite{o2016reference}.
+% Obviously the ``true" breakpoints are not known \emph{a priori} and thus the direct analog of Fig. \ref{fig:toy-accuracy} is not possible.
+% However, we postulate that most \emph{bona fide} breakpoints generated from evolutionary events will happen at or near gene boundaries; there is preliminary evidence for this assumption \cite{oliveira2017chromosomal}.
+% We compared the empirical distribution of distances to gene boundaries to a null distribution generated from the data, in which breakpoints are sampled uniformly from each genome, constrained to produce equivalent number of partitions as was measured by PanGraph.
+% For all tested cases, we find our boundaries cluster closer to genes than would be expected by chance.
+% The median distance from a Pangraph boundary to annotated gene ranges between 50-100 bp, depending on the species, while the random breakpoint is, on average, 600 bp off.
+
+% \begin{figure*}[htb]
+%     \includegraphics[width=.9\textwidth]{figs/incr_size.pdf}
+%     \caption{{\bf Pangenome graph properties vs. dataset size}.
+%         We build pangenome graphs with increasing number of isolates from the E.~Coli dataset and measure the scaling of different properties of the graphs.
+Lines and shaded areas represent mean and standard deviation over 10 different repetitions on random subsets of isolates, except for the final point indicating the full graph (307 isolates).
+%         The number and size of pancontigs scales sub-linearly with the number of isolates, with most of the pangenome being included in ~10\% of the pancontigs, and core pancontigs having higher-than-average size.
+\textit{PanGraph} is able to compress hundreds of genomes in a pangenome graph having only a fraction of the size.
+%         \textbf{A}: n.
+of pancontigs in the graph.
+We count the total number of pacontigs (blue), the number of core pancontigs (orange) and the minimum number of pancontigs that contain more than 50\% of the pangenome (L50, black).
+%         \textbf{B}: average size of pancontigs (blue), of only core pancontings (orange), and size of the smallest panconting in the minimal set that spans 50\% of the pangenome (N50, black).
+%         \textbf{C}: cumulative size of all genomes in the pangenome graph (gray), total pangenome size (blue) and size of core pangenome (orange).
+%     }
+%     \label{fig:panx-size}
+% \end{figure*}
+
+% % TODO: talk more extensively about the meaning of the results?
+% \begin{figure*}[htb]
+%     \includegraphics[width=\textwidth]{figs/panx_benchmark.pdf}
+%     \caption{{\bf Benchmark on real data}.
+%         We build pangenome graphs from fully-assembled chromosomes from 5 different bacterial species.
+For each species we build graphs with three different alignment kernel options (\textit{minimap2} with \textit{asm10} or \textit{asm20} option and \textit{mmseqs2}) and two different settings for the pseudo-energy parameters (standard or null values).
+%         \textbf{A}: \textit{PanGraph} wall-time when run in parallel on 8 cores.
+%         \textbf{B}: fraction of core pangenome in the pangenome graph.
+%         \textbf{C}: sequence compression, measured as the ratio between the pangenome graph size and the cumulative size of all the sequences contained in the graph.
+%     }
+%     \label{fig:panx-benchmark}
+% \end{figure*}
+
+% The computational time and maximum memory required to build each alignment is summarized in Table \ref{table:panx-performance}.
+
+% \begin{table}[hb]
+%     \caption{{\bf Real-world Performance}.
+%         Elapsed time and maximum RAM usage needed to compute the pangraph for the following collections of RefSeq genomes.
+%         All runs were performed on an AMD Ryzen 7 2700X utilizing 8 cores.
+%     }
+%     \begin{tabular}{l r r r }
+%         \hline\hline
+%         Species         & Sample size & Wall time (min) & Max memory (GB) \\
+%         \hline
+%         E.~coli         & 307         & 426             & 11.7            \\
+%         H.~pylori       & 85          & 102             & 1.2             \\
+%         K.~pneumoniae   & 109         & 132             & 2.1             \\
+%         M.~tuberculosis & 51          & 91              & 0.4             \\
+%         P.~marinus      & 10          & 12              & 0.1             \\
+%         \hline
+%     \end{tabular}
+%     \label{table:panx-performance}
+% \end{table}
+
+% \begin{figure*}[htb]
+%     \includegraphics[width=\textwidth]{figs/marginalize_test.pdf}
+%     \caption{{\bf Test of graph marginalization}.}
+%     \label{fig:marginalization}
+% \end{figure*}
 

### 2022-09-29 main.tex
started panx section in main

--- 2022-09-13 main.tex
@@
 
-\graphicspath{{../figs/}{./figs}}
+\newcommand{\avg}[1]{\langle #1 \rangle}
+\newcommand{\Lcore}{L_\text{core}}
+\newcommand{\Lpang}{L_\text{pang}}
+\newcommand{\Lgen}{L_\text{gen}}
+\newcommand{\dcore}{\langle d_\text{core} \rangle}
 
@@
 \maketitle
+
+% TODO: check consistency of numberings of figures and sections between SI and main text
+% TODO: real data
+% TODO: - dataset
+% TODO:     - table
+% TODO: - size
+% TODO:     - figure
+% TODO: - benchmark
+% TODO:     - figure
+% TODO: - marginalization
+% TODO: 
+% TODO: Table: leave all the columns? Improve description? Depending on what is useful afterwards
+% TODO: SI Section IIA: superfluous?
+% TODO: minimap2 asm10 superfluous?
 
@@
 
-In order to assess the accuracy of the pangenome graph inferred by PanGraph, and to quantify its performance characteristics as a function of input size, we generated synthetic data.
-Specifically, we simulated populations of size $N=100$ and sequence length $L=50000$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
-In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (each with independent rates set to $10^{-2}$ per generation), and horizontal sequence transfer that occurs with tunable rate $r$ per generation per genome.
+As a first validation step we use generated synthetic data to quantify the performance characteristics of \textit{PanGraph} as a function of input size, and its accuracy as a function of sequence diversity.
+
+\paragraph*{Generation of synthetic data}
+
+We simulated populations of size $N=100$ and sequence length $L=50\,000$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
+In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (respective rates $0.01$ and $0.05$ per generation), and horizontal sequence transfer that occurs with tunable rate $h$ per generation per genome (see SI Section I for details).
 The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
 The simulation framework is distributed within the PanGraph command line toolsuite for external use.
+
+\paragraph*{Performances as a function of dataset size}
+
+The algorithmic complexity was measured empirically by simulating populations of increasing size and average sequence length, and generating pangenome graphs for the resulting sequences.
+The mean and standard deviation of results obtained from 50 iterations are shown in Fig. \ref{fig:toy-performance}.
+Importantly, PanGraph's computational complexity grows \emph{linearly} with the number of input genomes.
+We note that the PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of \textit{minimap2} \cite{li2018minimap2}.
+Benchamrks of CPU and memory usage are reported in SI Section IIA.
+
+\begin{figure}[htb]
+    \includegraphics[width=.4\textwidth]{figs/benchmark.pdf}
+    \caption{{\bf Algorithm performance}.
+        PanGraph scales linearly with the number of input genomes.
+        This is a direct result of the guide tree simplification.
+        The solid line and ribbons display the mean and standard deviation over 50 runs.
+        All runs were performed utilizing 8 cores.
+    }
+    \label{fig:toy-performance}
+\end{figure}
+
+\paragraph*{Accuracy as a function of sequence divergence}
+
+We quantified the accuracy of PanGraph algorithm by computing the displacement of the inferred breakpoints relative to their known locus stored from the evolutionary simulation (cf.
+SI Section IIB for details).
+We evaluate this displacement by generating datasets with different rates of mutation $\mu$ and HGT $h$.
+For each $(h,\mu)$ pair we perform 25 different simulations, and build pangenome graphs using three different options for the alignment kernel, namely \textit{minimap2} \cite{li2018minimap2} with \textit{asm10} and \textit{asm20} option and \textit{mmseqs2} \cite{steinegger2017mmseqs2}.
+Critically, we found that the accuracy was \emph{independent} of the rate of horizontal gene transfer $h$ and thus underlying graph complexity.
+The predominant determinant of accuracy was the sample diversity, controlled by the mutation rate $\mu$ in our simulations (cf.
+SI Fig.~2).
+For low-diversity samples breakpoints are inferred with accuracy of few basepairs, while for highly diverged isolates most breakpoints are displaced by several hundreds of basepairs (cf.
+SI Fig.~3).
+The choice of alignment kernel influences the threshold diversity at which accuracy is lost.\\
+For each alignment kernel and value of average sequence divergence in the population we evaluated the fraction of breakpoints that have displacement greater than the default precision threshold for Pangraph of 100 bp, displayed in Fig.~\ref{fig:toy-accuracy}.
+On the sequences generated by our simulations, \textit{minimap2} with option \textit{asm20} show a loss of accuracy at a threshold pairwise divergence of around 5\%, while \textit{mmseqs2} is accurate up to average sequence divergence of around 12\%, at the cost of higher computational time.
 
@@
 
-We quantified the accuracy of PanGraph algorithm by computing the distance, $\delta$, of the inferred breakpoints relative to their known locus stored from the evolutionary simulation.
-Critically, we found that the accuracy was \emph{independent} of the rate of horizontal gene transfer $r$ and thus underlying graph complexity.
-The predominant determinant of accuracy was the sample diversity, shown in Fig.
-\ref{fig:toy-accuracy} which displays the distributions of $\delta$ for $100$ independent simulations ran for each mutation rate $\mu$.
-For approximately clonal populations, $\sim 80\%$ of breakpoints were estimated to be within $10$bp away from the ``true" location.
-Conversely, for samples with average pairwise diversity above $2\%$, $\sim 80\%$ of breakpoints are estimated only within $100$ bp.
-This limitation is primarily driven by lack of sensitivity in the pairwise alignment phase performed using \emph{minimap2} \cite{li2018minimap2}.
-
-\begin{figure}[htb]
-    \includegraphics[width=.4\textwidth]{figs/benchmark.pdf}
-    \caption{{\bf Algorithm performance}.
-        PanGraph scales linearly with the number of input genomes.
-        This is a direct result of the guide tree simplification.
-        The solid line and ribbons display the mean and standard deviation respectively.
-        All runs were performed on an AMD Ryzen 7 2700X utilizing 8 cores.
-    }
-    \label{fig:toy-performance}
-\end{figure}
-
-The algorithmic complexity was measured empirically by simulating populations of increasing size and average sequence length.
-The mean and standard deviation of results obtained from 50 iterations are shown in Fig. \ref{fig:toy-performance}.
-Importantly, PanGraph's computational complexity grows \emph{linearly} with the number of input genomes.
-We note that the PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of Minimap2 \cite{li2018minimap2}.
-
 \subsection{Validation on real data}
 
-We additionally validated PanGraph on natural populations sampled from RefSeq \cite{o2016reference}.
-Obviously the ``true" breakpoints are not known \emph{a priori} and thus the direct analog of Fig. \ref{fig:toy-accuracy} is not possible.
-However, we postulate that most \emph{bona fide} breakpoints generated from evolutionary events will happen at or near gene boundaries; there is preliminary evidence for this assumption \cite{oliveira2017chromosomal}.
-We compared the empirical distribution of distances to gene boundaries to a null distribution generated from the data, in which breakpoints are sampled uniformly from each genome, constrained to produce equivalent number of partitions as was measured by PanGraph.
-For all tested cases, we find our boundaries cluster closer to genes than would be expected by chance.
-The median distance from a Pangraph boundary to annotated gene ranges between 50-100 bp, depending on the species, while the random breakpoint is, on average, 600 bp off.
+We additionally validated PanGraph on natural populations sampled from RefSeq \cite{o2016reference}, focusing on the properties of the resulting pangenome graphs for datasets of different size and diversity.
+
+\paragraph*{Dataset characterization}
+
+We downloaded from RefSeq \cite{o2016reference} completely assembled chromosomes from 5 different bacterial species: \textit{Escherichia Coli} (EC), \textit{Klebsiella Pneumoniae} (KP), \textit{Helicobacter Pylori} (HP), \textit{Mycobacterium Tuberculosis} (MT) and \textit{Prochlorococcus Marinus} (PM).
+These data had been previously analyzed using PanX \cite{ding2018panx}, which allowed us to estimate the size of the pangenome and core genome, and the average pairwise divergence on core genes (cf.
+Table~\ref{table:panx-dataset} and SI Section~IIIA).
+% The \textit{E.~Coli} (EC) dataset is the one with the highest number of isolates (307 chromosomes) and the largest pangenome.
+The species with the highest core genome divergence are \textit{H.~Pylori} (HP) and \textit{P.
+Marinus} (PM).
+For the former, the divergce sits at the limit of what the \textit{minimap2} alignment kernel is able to accurately merge (cf.
+Fig.~\ref{fig:toy-accuracy}), while for the latter the diversity is beyond even the limits of \textit{mmseqs2}.
+
+\begin{table}[h]
+    \setlength{\tabcolsep}{6pt}
+    \begin{tabular}{l c c c c c }
+        \hline\hline
+        Species         & N.  & $\avg{\Lgen}$ & $\Lpang$ & $\Lcore$ & $\dcore$ \\
+        \hline
+        E.~Coli         & 307 & 5.0           & 17.7     & 0.7      & 1.6\%    \\
+        K.~Pneumoniae   & 109 & 5.3           & 13.0     & 2.3      & 0.8\%    \\
+        H.~Pylori       & 85  & 1.6           & 1.9      & 0.6      & 4.2\%    \\
+        M.~Tuberculosis & 51  & 4.4           & 4.1      & 2.4      & 0.03\%   \\
+        P.~Marinus      & 10  & 1.8           & 3.3      & 0.7      & 26.9\%   \\
+        \hline
+    \end{tabular}
+    \caption{{\bf Dataset properties}. Columns represent:
+        (1) number of isolates,
+        (2) average chromosome lenght in Mbps,
+        (3) total pangenome legnth in Mbp,
+        (4) total core genome length in Mbp,
+        (5) average pairwise divergence of core genes.
+    }
+    \label{table:panx-dataset}
+\end{table}
+
+\paragraph*{Benchmark}
+
+Using \textit{PanGraph}, we build multiple pangenome graphs for each species in the dataset.
+We test all the different alignment kernels (\textit{minimap2} with \textit{asm10} or \textit{asm20} option, or \textit{mmseqs2}).
+Moreover we also vary the value of the pseudo-energy parameters $\alpha,\beta$ from eq.~(\ref{eq:pseudo-energy}), using either the standard values ($\alpha=100$, $\beta=10$) or null values ($\alpha = \beta = 0$).\\
+\textit{PanGraph} can build pangenome graphs comprising hundreds of isolates in a few hours (5-6h for 307 EC isolates with \textit{minimap2} kernel, cf.
+Fig~\ref{fig:panx-benchmark}A).
+Use of \textit{mmseqs2} consistently requires longer time, but provides higher accuracy when merging highly diverged sequences.
+This is the case for the HP dataset, whose core genome divergence sits at the limit of what the \textit{minimap2} kernel can accurately align.
+In this case the use of \textit{mmseqs2} improves both sequence compression and core pangenome fraction (cf.
+Fig~\ref{fig:panx-benchmark}B-C).\\
+% beta poses a limit to the divergence. Removing it..
+% but increases fragmentation (cite supplementary)
+% talk about PM
+% average compression
+% core genome fraction compatible?
+
+% TODO: talk more extensively about the meaning of the results?
+\begin{figure*}[htb]
+    \includegraphics[width=\textwidth]{figs/panx_benchmark.pdf}
+    \caption{{\bf Benchmark on real data}.
+        We build pangenome graphs from fully-assembled chromosomes from 5 different bacterial species.
+For each species we build graphs with three different alignment kernel options (\textit{minimap2} with \textit{asm10} or \textit{asm20} option and \textit{mmseqs2}) and two different settings for the pseudo-energy parameters (standard or null values).
+        \textbf{A}: \textit{PanGraph} wall-time when run in parallel on 8 cores.
+        \textbf{B}: fraction of core pangenome in the pangenome graph.
+        \textbf{C}: sequence compression, measured as the ratio between the pangenome graph size and the cumulative size of all the sequences contained in the graph.
+    }
+    \label{fig:panx-benchmark}
+\end{figure*}
+
+\paragraph*{Scaling with dataset size}
 
@@
 
-% TODO: talk more extensively about the meaning of the results?
-\begin{figure*}[htb]
-    \includegraphics[width=\textwidth]{figs/panx_benchmark.pdf}
-    \caption{{\bf Benchmark on real data}.
-        We build pangenome graphs from fully-assembled chromosomes from 5 different bacterial species.
-For each species we build graphs with three different alignment kernel options (\textit{minimap2} with \textit{asm10} or \textit{asm20} option and \textit{mmseqs2}) and two different settings for the pseudo-energy parameters (standard or null values).
-        \textbf{A}: \textit{PanGraph} wall-time when run in parallel on 8 cores.
-        \textbf{B}: fraction of core pangenome in the pangenome graph.
-        \textbf{C}: sequence compression, measured as the ratio between the pangenome graph size and the cumulative size of all the sequences contained in the graph.
-    }
-    \label{fig:panx-benchmark}
-\end{figure*}
-
-The computational time and maximum memory required to build each alignment is summarized in Table \ref{table:panx-performance}.
-
-\begin{table}[hb]
-    \caption{{\bf Real-world Performance}.
-        Elapsed time and maximum RAM usage needed to compute the pangraph for the following collections of RefSeq genomes.
-        All runs were performed on an AMD Ryzen 7 2700X utilizing 8 cores.
-    }
-    \begin{tabular}{l r r r }
-        \hline\hline
-        Species         & Sample size & Wall time (min) & Max memory (GB) \\
-        \hline
-        E.~coli         & 307         & 426             & 11.7            \\
-        H.~pylori       & 85          & 102             & 1.2             \\
-        K.~pneumoniae   & 109         & 132             & 2.1             \\
-        M.~tuberculosis & 51          & 91              & 0.4             \\
-        P.~marinus      & 10          & 12              & 0.1             \\
-        \hline
-    \end{tabular}
-    \label{table:panx-performance}
-\end{table}
+\section{Graph marginalization}
 
 \begin{figure*}[htb]
-    \includegraphics[width=.8\textwidth]{figs/proj_single_escherichia_coli.pdf}
+    \includegraphics[width=\textwidth]{figs/marginalize_test.pdf}
     \caption{{\bf Test of graph marginalization}.}
@@
 \end{figure*}
+
+% % TODO: talk more extensively about the meaning of the results?
+% \begin{figure*}[htb]
+%     \includegraphics[width=\textwidth]{figs/panx_benchmark.pdf}
+%     \caption{{\bf Benchmark on real data}.
+%         We build pangenome graphs from fully-assembled chromosomes from 5 different bacterial species.
+For each species we build graphs with three different alignment kernel options (\textit{minimap2} with \textit{asm10} or \textit{asm20} option and \textit{mmseqs2}) and two different settings for the pseudo-energy parameters (standard or null values).
+%         \textbf{A}: \textit{PanGraph} wall-time when run in parallel on 8 cores.
+%         \textbf{B}: fraction of core pangenome in the pangenome graph.
+%         \textbf{C}: sequence compression, measured as the ratio between the pangenome graph size and the cumulative size of all the sequences contained in the graph.
+%     }
+%     \label{fig:panx-benchmark}
+% \end{figure*}
+
+% The computational time and maximum memory required to build each alignment is summarized in Table \ref{table:panx-performance}.
+
+% \begin{figure*}[htb]
+%     \includegraphics[width=\textwidth]{figs/marginalize_test.pdf}
+%     \caption{{\bf Test of graph marginalization}.}
+%     \label{fig:marginalization}
+% \end{figure*}
 

### 2022-09-29 main.tex
Update on Overleaf.

@@
 
-\paragraph*{Benchmark}
+\paragraph*{Benchmark on real data}
 
-Using \textit{PanGraph}, we build multiple pangenome graphs for each species in the dataset.
-We test all the different alignment kernels (\textit{minimap2} with \textit{asm10} or \textit{asm20} option, or \textit{mmseqs2}).
-Moreover we also vary the value of the pseudo-energy parameters $\alpha,\beta$ from eq.~(\ref{eq:pseudo-energy}), using either the standard values ($\alpha=100$, $\beta=10$) or null values ($\alpha = \beta = 0$).\\
+Using \textit{PanGraph}, we build multiple pangenome graphs for each species in the dataset (cf.
+SI Section~IIIB and IIIC).
+These differ by the alignment kernel used (\textit{minimap2} with \textit{asm10} or \textit{asm20} option, and \textit{mmseqs2}) or by the value of the pseudo-energy parameters $\alpha,\beta$ from eq.~(\ref{eq:pseudo-energy}).
+Different alignment kernels are expected to reach different accuracy on datasets with different diversities (cf.
+Fig.~\ref{fig:toy-accuracy}).
+Moreover the use of null energy parameters ($\alpha = \beta = 0$) is expected to remove the threshold divergence of 10\% associated with the standard values of parameters ($\alpha=100$, $\beta=10$) at the cost of more fragmented pangenome graphs.\\
 \textit{PanGraph} can build pangenome graphs comprising hundreds of isolates in a few hours (5-6h for 307 EC isolates with \textit{minimap2} kernel, cf.
 Fig~\ref{fig:panx-benchmark}A).
-Use of \textit{mmseqs2} consistently requires longer time, but provides higher accuracy when merging highly diverged sequences.
-This is the case for the HP dataset, whose core genome divergence sits at the limit of what the \textit{minimap2} kernel can accurately align.
-In this case the use of \textit{mmseqs2} improves both sequence compression and core pangenome fraction (cf.
-Fig~\ref{fig:panx-benchmark}B-C).\\
-% beta poses a limit to the divergence. Removing it..
-% but increases fragmentation (cite supplementary)
-% talk about PM
-% average compression
-% core genome fraction compatible?
+Use of \textit{mmseqs2} consistently requires longer computational time, but provides higher accuracy when merging highly diverged sequences.
+This is the case for the HP dataset, whose core genome divergence sits at the limit of what can be accurately merged with the \textit{minimap2} kernel.
+In this case the use of \textit{mmseqs2} and null energy parameters increases core pangenome fraction, decreases the total pangenome size and and does not compromise the fragmentation of the graph (cf.
+Fig~\ref{fig:panx-benchmark}B-C and SI Fig.~4).
+The divergence of the PM datasets sits instead beyond the capabilities of \textit{Pangraph}.
+In this case none of the alignment kernel reaches satisfactory sequence compression, and only \textit{mmseqs2} combined with null energy parameters is able to retrieve few core pancontigs.\\
+For the other species, ... 
+% TODO: finish
+For these species the use of null energy parameters slightly increases merging and sequence compression, at the cost of slightly more fragmented graphs (cf. SI Section IIIC and SI Fig.~4). 
+Overall this benchmark showcases the capabilities and limits of \textit{PanGraph}, and demonstrates the value of adapting the pseudo-energy parameter and choice of alignment kernel to the expected diversity of the dataset considered.
 

### 2022-09-30 main.tex
finished draft of benchmark

@@
 In this case none of the alignment kernel reaches satisfactory sequence compression, and only \textit{mmseqs2} combined with null energy parameters is able to retrieve few core pancontigs.\\
-For the other species, ... 
-% TODO: finish
+For the other species sequence compression is of the order of few percents, graphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in blocks spanning several kbps.
 For these species the use of null energy parameters slightly increases merging and sequence compression, at the cost of slightly more fragmented graphs (cf. SI Section IIIC and SI Fig.~4).
@@
         We build pangenome graphs from fully-assembled chromosomes from 5 different bacterial species.
-For each species we build graphs with three different alignment kernel options (\textit{minimap2} with \textit{asm10} or \textit{asm20} option and \textit{mmseqs2}) and two different settings for the pseudo-energy parameters (standard or null values).
+For each species we build graphs with three different alignment kernel options (\textit{minimap2} with \textit{asm10} or \textit{asm20} option and \textit{mmseqs2}) and two different settings for the pseudo-energy parameters $\alpha, \beta$ (standard or null values).
         \textbf{A}: \textit{PanGraph} wall-time when run in parallel on 8 cores.
         \textbf{B}: fraction of core pangenome in the pangenome graph.
-        \textbf{C}: sequence compression, measured as the ratio between the pangenome graph size and the cumulative size of all the sequences contained in the graph.
+        \textbf{C}: sequence compression, defined as the ratio between the pangenome graph size and the cumulative size of all the sequences contained in the graph.
     }
@@
 
+We then explore how the properties of the pangenome graph scale with the dataset size.
+To do so we build pangenome graphs from random sets of isolates of increasing size from the EC dataset.
+Graphs are built with standard values of the parameter and using the \textit{minimap2} kernel with \textit{asm20} option.
+For each size 10 different random graphs are constructed.
+Results are reported in Fig.~\ref{fig:panx-size}.\\
+The number and size of pancontigs scales sub-linearly with the number of isolates, with most of the pangenome being included in ~10\% of the pancontigs, and core pancontigs having higher-than-average size.
+Adding more strains does not generate excessive fragmentation of these pancontigs, and their size remains of several kbps even as hundreds of strains are added.
+The size of the core genome does not decrease significantly with the addition of new strains, while the total pangenome size remains orders of magnitude smaller than the total size of all genomes included in the graph.
+
 \begin{figure*}[htb]
@@
 Lines and shaded areas represent mean and standard deviation over 10 different repetitions on random subsets of isolates, except for the final point indicating the full graph (307 isolates).
-        The number and size of pancontigs scales sub-linearly with the number of isolates, with most of the pangenome being included in ~10\% of the pancontigs, and core pancontigs having higher-than-average size.
-\textit{PanGraph} is able to compress hundreds of genomes in a pangenome graph having only a fraction of the size.
         \textbf{A}: n.
@@
 \end{figure*}
-
-% % TODO: talk more extensively about the meaning of the results?
-% \begin{figure*}[htb]
-%     \includegraphics[width=\textwidth]{figs/panx_benchmark.pdf}
-%     \caption{{\bf Benchmark on real data}.
-%         We build pangenome graphs from fully-assembled chromosomes from 5 different bacterial species.
-For each species we build graphs with three different alignment kernel options (\textit{minimap2} with \textit{asm10} or \textit{asm20} option and \textit{mmseqs2}) and two different settings for the pseudo-energy parameters (standard or null values).
-%         \textbf{A}: \textit{PanGraph} wall-time when run in parallel on 8 cores.
-%         \textbf{B}: fraction of core pangenome in the pangenome graph.
-%         \textbf{C}: sequence compression, measured as the ratio between the pangenome graph size and the cumulative size of all the sequences contained in the graph.
-%     }
-%     \label{fig:panx-benchmark}
-% \end{figure*}
-
-% The computational time and maximum memory required to build each alignment is summarized in Table \ref{table:panx-performance}.
-
-% \begin{figure*}[htb]
-%     \includegraphics[width=\textwidth]{figs/marginalize_test.pdf}
-%     \caption{{\bf Test of graph marginalization}.}
-%     \label{fig:marginalization}
-% \end{figure*}
 

### 2022-10-02 main.tex
continued marginalize section

@@
 
+Interpretation of a pangenome graph containing hundreds of strains can be challenging.
+To this end it is often informative to inspect simpler sub-graphs containing only a small subset of isolates.
+However building multiple such sub-graphs can be computationally intensive.
+To facilitate this task \textit{PanGraph} provides the \verb|marginalize| command, that can be used to marginalize a large pangenome graph on a small subset of strains, removing all the other paths and merging transitive pancontigs.
+This operation is computationally much cheaper than building a new graph for the subset of strains considered.\\
+To verify that marginalized graphs are compatible with newly-built graphs, we built a pangenome graph for 50 randomly selected strains from the \textit{Klebsiella Pneumoniae} dataset.
+We then picked 50 random pairs of strains from the same set and built two graphs for each pair: the graph obtained by marginalizing the complete graph on the pair, and the graph built directly from the pair (cf.
+Fig.~\ref{fig:marginalization}A, top).
+We then compared these two graphs by considering the partition they generate on the genome of the isolates they include.
+Each genome is partitioned in pancontigs that are either shared on the pair, or private to one isolate
+(cf.
+Fig.~\ref{fig:marginalization}A, bottom).
+We can classify the segments in the intersection of the two partitions in two categories: either the two partitions agree or they do not on the segment kind.
+And agreement can be further divided in agreement on the segments being shared or agreement on them being private.
+% TODO: Compatibility between the two graphs would required segments on which the two graphs disagree to be short and few.\\
+
+% TODO: cite supplementary
+
 \begin{figure*}[htb]

### 2022-10-03 main.tex
completed draft of marginalize paragraph

@@
 
-Interpretation of a pangenome graph containing hundreds of strains can be challenging.
-To this end it is often informative to inspect simpler sub-graphs containing only a small subset of isolates.
-However building multiple such sub-graphs can be computationally intensive.
+The interpretation of a pangenome graph containing hundreds of strains can be challenging.
+To this end it is often informative to inspect simpler sub-graphs comprising only a small subset of isolates.
+However building multiple such sub-graphs is computationally intensive.
 To facilitate this task \textit{PanGraph} provides the \verb|marginalize| command, that can be used to marginalize a large pangenome graph on a small subset of strains, removing all the other paths and merging transitive pancontigs.
@@
 Fig.~\ref{fig:marginalization}A, top).
-We then compared these two graphs by considering the partition they generate on the genome of the isolates they include.
-Each genome is partitioned in pancontigs that are either shared on the pair, or private to one isolate
-(cf.
+We compare these two graphs by considering the partition they generate on the genome of the isolates they include.
+Each genome is partitioned in pancontigs that are either shared on the pair, or private to one isolate (cf.
 Fig.~\ref{fig:marginalization}A, bottom).
-We can classify the segments in the intersection of the two partitions in two categories: either the two partitions agree or they do not on the segment kind.
-And agreement can be further divided in agreement on the segments being shared or agreement on them being private.
-% TODO: Compatibility between the two graphs would required segments on which the two graphs disagree to be short and few.\\
-
-% TODO: cite supplementary
+We can classify the segments in the intersection of the two partitions in different categories, depending on whether the two partitions agree on whether the segment is shared or private.
+Moreover segments on which the two partitions agree are further sub-divided depending on whether they are private or shared on both partitions.\\
+Compatibility between the two graphs requires segments on which the two partitions disagree to be few and short.
+Indeed, we verified that over the pairs we picked these segments cover a very small fraction of the genome ($<1\%$), and have average size compatible with the default 100 bp precision threshold of \textit{PanGraph} (cf.
+Fig.~\ref{fig:marginalization}B-C).
+This is contrary to agreement segments, that have average size of several kbps.
+The prevalence of agreement segment cannot be explained only by the fact that most of the genome is shared, since segments that are shared on both graphs cover on average only 85\% of the sequence.
+We confirmed this results using the fraction of shared k-mers, using an approach inspired by PopPUNK \cite{lees2019fast}.
+Namely, we approximate the fraction of homologous sequence between any two pairs using the fraction of shared k-mers (in our case we take $k=21$).
+This fraction is however an underestimate of the fraction of homologous sequence because of mutations.
+We therefore correct it with the factor $(1-d)^{-k}$, where $d$ is the average pairwise divergence on core genes for the pair considered.
+The resulting distribution of shared sequence is indeed in very good agreement with the fraction of segments that are shared on both graphs (cf.
+Fig.~\ref{fig:marginalization}B), suggesting that homologous sequences are correctly merged on the complete, marginalized and pairwise graphs.\\
+We performed the same test on species from the other datasets, obtaining similar results on all species whose divergence is compatible with \textit{PanGraph} capabilities (cf.
+SI Section IV and SI Fig.~5).
 
@@
     \includegraphics[width=\textwidth]{figs/marginalize_test.pdf}
-    \caption{{\bf Test of graph marginalization}.}
+    \caption{{\bf Test of graph marginalization}.
+        \textbf{A}: we built a pangenome graph from 50 randomly chosen strains from the \textit{K.
+Pneumoniae} dataset.
+We then randomly picked 50 pairs of strains.
+For each pair we compared the pangenome graph obtained by marginalizing the complete graph on the pair of strain, and the one obtained by building a new graph for the pair (top).
+The comparison is done by considering that each graph partitions a genome in shared and private segments.
+By combining the partitions generated by the marginalized and pairwise graphs we categorize segments in three categories, depending on whether the two partitions agree or not, and if they agree depending on whether segments are shared or private.
+All graphs were built using \textit{minimap2} alignment kernel with \textit{asm20} option and default value for the energy parameters.
+        \textbf{B}: distribution of the average fraction of the genome covered by segments of each category, over the 50 pairs considered (2 entries per pair).
+The last line represents the distribution of shared sequence, approximated using the fraction of shared k-mers corrected using sequence divergence as described in the main text.
+        \textbf{C}: distribution of average segment lengths for each category over the 50 pairs considered (2 entries per pair).
+    }
     \label{fig:marginalization}

### 2022-10-03 main.tex
modified conclusions

@@
 \section{Discussion}
-While single nucleotide differences in the core genome are straightforward to analyze with existing tools, these analysis miss the great majority of genetic diversity.
-The ability to rapidly align large sets of closely related, fully assembled genomes is a critical tool for the investigation of the processes governing microbial evolution.
-We hope PanGraph will spur novel insight microbial diversity and the processes by which bacteria adapt and change.
 
-Currently, PanGraph is limited to collections of genomes that are closely related with typical sequence divergence smaller that than 3\%.
-This limitation originates from the core tool used for pairwise alignment, \texttt{minimap2}, that was determined to lose sensitivity if sequences are sufficiently diverged under the chosen parameters.
-Homologous sequence that is more diverged than this cut-off will not be reliably merged by pangraph.
-This effectively limits the TMRCA of sequences within a block.
-More diverged sequence, for example originating from a cross-species transfer event, will thus be represented as alternative path.
-In addition to the intrinsic limit of \texttt{minimap2}, this cut-off can be tuned with the parameter $\alpha$ and can for example be used to restrict merging to very similar sequences and thus capture all recent horizontal events as divergent paths in the graph.
+While single nucleotide differences in the core genome are straightforward to analyze with existing tools, these analyses miss the great majority of genetic diversity.
+The ability to rapidly align large sets of closely related, fully assembled genomes is crucial for the investigation of the processes governing microbial evolution.
+We developed \textit{PanGraph} as a tool that can extract information on homology, diversity and syntheny, both on core and accessory genome, in a scalable way.
 
-We emphasize that PanGraph was designed as a modular tool.
-In the future, we anticipate utilizing other pairwise ``alignment" kernels in addition to minimap2 that will allow the user to trade-off speed for sensitivity.
+In our analysis we demonstrated the capabilities and limits of this tool, both on synthetic and real sequences.
+The efficient implementation of \textit{PanGraph} allows it to create pangenome graphs containing hundreds of isolates in a few hours on a 8-cores machine.
+The size of pancontigs and the fraction of core sequece have good scaling properties with the number of isolates in the graph, indicating that \textit{PanGraph} is able to successfully capture pangenome properties.\\
+One of the main limitations is the diversity of the input sequences.
+The default \textit{minimap2} kernel is able to correctly merge sequences up to around 5\% divergence.
+Sequences with higher divergence (up to 10-15\%) can be merged using the \textit{mmseqs2} alignment kernel, and tuning the $\alpha, \beta$ energy parameters (cf.
+eq.~(\ref{eq:pseudo-energy})), trading off speed for sensitivity.\\
+We also provide the ability to quickly marginalize big graps on a subset of strains, obtaining simpler graphs that can more easily be explored.
+Graphs can also be exported in different formats for further analysis and visualization.
+
+We hope that these characteristics will make \textit{PanGraph} a valuable tool, with the potential to spur novel insight on microbial diversity and the processes by which bacteria adapt and change.
 

### 2022-10-04 main.tex
Pangraph 2.0

@@
 
-% TODO: check consistency of numberings of figures and sections between SI and main text
-% TODO: real data
-% TODO: - dataset
-% TODO:     - table
-% TODO: - size
-% TODO:     - figure
-% TODO: - benchmark
-% TODO:     - figure
-% TODO: - marginalization
-% TODO: 
-% TODO: Table: leave all the columns? Improve description? Depending on what is useful afterwards
 % TODO: SI Section IIA: superfluous?
@@
 In particular we decided to include the option to use \emph{mmseqs2} \cite{steinegger2017mmseqs2} as an alternative alignment kernel, because of its sensitivity on highly diverged sequences at the cost of higher computational time.
-In general, the kernel interface expects two lists of \emph{pancontigs} to align as input and will output a list of putative overlapping alignments between both input lists.
+% In general, the kernel interface expects two lists of \emph{pancontigs} to align as input and will output a list of putative overlapping alignments between both input lists.
 
@@
 Additionally, $\alpha$ and $\beta$ are hyperparameters of the algorithm, respectively controlling the tradeoff between fragmentation of the graph and the maximum diversity within each block.
-Only alignments with negative pseudo-energy are performed.
+Only mergers whose alignment has negative pseudo-energy are performed.
 
@@
 All data and scripts used to validate PanGraph are available within the same repository.
-% TODO: add data?
 
@@
 We note that the PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of \textit{minimap2} \cite{li2018minimap2}.
-Benchamrks of CPU and memory usage are reported in SI Section IIA.
+Benchmarks of CPU and memory usage are reported in SI Section IIA.
 
@@
         The solid line and ribbons display the mean and standard deviation over 50 runs.
-        All runs were performed utilizing 8 cores.
+        All runs were performed utilizing 8 cores, and with the default \textit{minimap2} alignment kernel and \textit{asm20} option.
     }
@@
 SI Section IIB for details).
-We evaluate this displacement by generating datasets with different rates of mutation $\mu$ and HGT $h$.
+We evaluate this displacement by generating datasets with different rates of mutation $\mu$ and horizontal sequence transfer $h$.
 For each $(h,\mu)$ pair we perform 25 different simulations, and build pangenome graphs using three different options for the alignment kernel, namely \textit{minimap2} \cite{li2018minimap2} with \textit{asm10} and \textit{asm20} option and \textit{mmseqs2} \cite{steinegger2017mmseqs2}.
-Critically, we found that the accuracy was \emph{independent} of the rate of horizontal gene transfer $h$ and thus underlying graph complexity.
+Critically, we found that the accuracy was \emph{independent} of the rate of horizontal transfer $h$ and thus underlying graph complexity.
 The predominant determinant of accuracy was the sample diversity, controlled by the mutation rate $\mu$ in our simulations (cf.
@@
 The choice of alignment kernel influences the threshold diversity at which accuracy is lost.\\
-For each alignment kernel and value of average sequence divergence in the population we evaluated the fraction of breakpoints that have displacement greater than the default precision threshold for Pangraph of 100 bp, displayed in Fig.~\ref{fig:toy-accuracy}.
+For each alignment kernel and value of average sequence divergence in the population we evaluated the fraction of breakpoints that have displacement greater than the default precision threshold for Pangraph of 100 bp (cf.
+Fig.~\ref{fig:toy-accuracy}).
 On the sequences generated by our simulations, \textit{minimap2} with option \textit{asm20} show a loss of accuracy at a threshold pairwise divergence of around 5\%, while \textit{mmseqs2} is accurate up to average sequence divergence of around 12\%, at the cost of higher computational time.
@@
     \caption{{\bf Accuracy against synthetic data}.
-        We generate artificial data with varying degree of sequence divergence, and compare the real underlying pangenome graph with the one reconstructed by \textit{PanGraph}, for three different alignment kernels :\textit{minimap2} with \textit{asm10} and \textit{asm20} option, and \textit{mmseqs2}.
+        We generate artificial data with varying degree of sequence divergence, and compare the real underlying pangenome graph with the one reconstructed by \textit{PanGraph}, for three different alignment kernels :\textit{minimap2} with \textit{asm10} or \textit{asm20} option, and \textit{mmseqs2}.
 In each comparison we evaluate the misplacement of breakpoints that we can pair on the two graphs within 1kbp.
@@
 Line and shaded area represents mean and standard deviation over 25 repetitions.
-\textit{mmseqs2} maintains accuracy at higher divergence, at the cost of more computational time.
+\textit{mmseqs2} maintains accuracy at higher divergence, at the cost of higher computational time.
     }
@@
 
-We additionally validated PanGraph on natural populations sampled from RefSeq \cite{o2016reference}, focusing on the properties of the resulting pangenome graphs for datasets of different size and diversity.
+We additionally validated PanGraph on natural populations sampled from RefSeq \cite{o2016reference}, focusing on the properties of the resulting pangenome graphs as a function of dataset size and diversity.
 
@@
     \caption{{\bf Dataset properties}. Columns represent:
-        (1) number of isolates,
-        (2) average chromosome lenght in Mbps,
-        (3) total pangenome legnth in Mbp,
-        (4) total core genome length in Mbp,
-        (5) average pairwise divergence of core genes.
+        number of isolates (N.),
+        average chromosome length in Mbps ($\avg{\Lgen}$),
+        total pangenome length in Mbps ($\Lpang$),
+        total core genome length in Mbps ($\Lcore$),
+        average pairwise divergence of core genes ($\dcore$).
     }
@@
 Moreover the use of null energy parameters ($\alpha = \beta = 0$) is expected to remove the threshold divergence of 10\% associated with the standard values of parameters ($\alpha=100$, $\beta=10$) at the cost of more fragmented pangenome graphs.\\
-\textit{PanGraph} can build pangenome graphs comprising hundreds of isolates in a few hours (5-6h for 307 EC isolates with \textit{minimap2} kernel, cf.
+\textit{PanGraph} can build pangenome graphs comprising hundreds of isolates in a few hours (5h for 307 EC isolates with \textit{minimap2} kernel, cf.
 Fig~\ref{fig:panx-benchmark}A).
@@
 The divergence of the PM datasets sits instead beyond the capabilities of \textit{Pangraph}.
-In this case none of the alignment kernel reaches satisfactory sequence compression, and only \textit{mmseqs2} combined with null energy parameters is able to retrieve few core pancontigs.\\
+In this case no alignment kernel can reach satisfactory sequence compression, and only \textit{mmseqs2} combined with null energy parameters is able to retrieve few core pancontigs.\\
 For the other species sequence compression is of the order of few percents, graphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in blocks spanning several kbps.
@@
 
-% TODO: talk more extensively about the meaning of the results?
 \begin{figure*}[htb]
@@
 To do so we build pangenome graphs from random sets of isolates of increasing size from the EC dataset.
-Graphs are built with standard values of the parameter and using the \textit{minimap2} kernel with \textit{asm20} option.
-For each size 10 different random graphs are constructed.
 Results are reported in Fig.~\ref{fig:panx-size}.\\
-The number and size of pancontigs scales sub-linearly with the number of isolates, with most of the pangenome being included in ~10\% of the pancontigs, and core pancontigs having higher-than-average size.
+The number and size of pancontigs scale sub-linearly with the number of isolates, with more than 50\% of the pangenome being included in the 10\% longest pancontigs, and core pancontigs having higher-than-average size.
 Adding more strains does not generate excessive fragmentation of these pancontigs, and their size remains of several kbps even as hundreds of strains are added.
@@
         We build pangenome graphs with increasing number of isolates from the E.~Coli dataset and measure the scaling of different properties of the graphs.
+Graphs were built using \textit{minimap2} alignment kernel with \textit{asm20} option.
 Lines and shaded areas represent mean and standard deviation over 10 different repetitions on random subsets of isolates, except for the final point indicating the full graph (307 isolates).
@@
 To verify that marginalized graphs are compatible with newly-built graphs, we built a pangenome graph for 50 randomly selected strains from the \textit{Klebsiella Pneumoniae} dataset.
-We then picked 50 random pairs of strains from the same set and built two graphs for each pair: the graph obtained by marginalizing the complete graph on the pair, and the graph built directly from the pair (cf.
-Fig.~\ref{fig:marginalization}A, top).
+We then picked 50 random pairs of strains from the same set and built two graphs for each pair: the graph obtained by marginalizing the complete graph on the pair (i.e.
+pairwise graph), and the graph built directly from the pair (i.e.
+marginalized graph, cf.
+Fig.~\ref{fig:marginalization}A top).
 We compare these two graphs by considering the partition they generate on the genome of the isolates they include.
 Each genome is partitioned in pancontigs that are either shared on the pair, or private to one isolate (cf.
-Fig.~\ref{fig:marginalization}A, bottom).
+Fig.~\ref{fig:marginalization}A bottom).
 We can classify the segments in the intersection of the two partitions in different categories, depending on whether the two partitions agree on whether the segment is shared or private.
@@
 Fig.~\ref{fig:marginalization}B-C).
-This is contrary to agreement segments, that have average size of several kbps.
+Conversely, agreement segments have average size of several kbps.
 The prevalence of agreement segment cannot be explained only by the fact that most of the genome is shared, since segments that are shared on both graphs cover on average only 85\% of the sequence.
 We confirmed this results using the fraction of shared k-mers, using an approach inspired by PopPUNK \cite{lees2019fast}.
-Namely, we approximate the fraction of homologous sequence between any two pairs using the fraction of shared k-mers (in our case we take $k=21$).
-This fraction is however an underestimate of the fraction of homologous sequence because of mutations.
-We therefore correct it with the factor $(1-d)^{-k}$, where $d$ is the average pairwise divergence on core genes for the pair considered.
+Namely, we approximate the fraction of homologous sequence between any two pairs using the fraction of shared k-mers ($k=21$) divided by $(1-d)^{k}$, where $d$ is the average pairwise divergence on core genes for the pair considered.
+This divisor corrects for k-mers on homologous sequence that are not shared due to mutations.
 The resulting distribution of shared sequence is indeed in very good agreement with the fraction of segments that are shared on both graphs (cf.
@@
 The ability to rapidly align large sets of closely related, fully assembled genomes is crucial for the investigation of the processes governing microbial evolution.
-We developed \textit{PanGraph} as a tool that can extract information on homology, diversity and syntheny, both on core and accessory genome, in a scalable way.
+We developed \textit{PanGraph} as a tool that can adress this need, by being able to capture the structural and nucleotide diversity in both the core and accessory genome in a scalable way.
 

### 2022-10-04 supplementary.tex
feat: modified main
Pangraph 2.0

@@
 Each individual in the population has a circular genome of initial size $L$.
-At each generation single-nucleotide mutations occur at a rate $\mu$ per position per generation.\\
+At each generation single-nucleotide mutations occur at a rate $\mu$ per position per generation.
 Deletions can occur at a rate $d$ per genome per generation.
@@
 \begin{table}[hb]
-    \begin{tabular}{c l l}
+    \setlength{\tabcolsep}{9pt}
+    \begin{tabular}{r l l}
         \hline\hline
@@
         $T$       & n. of simulated generations & 50                              \\
-        $L$       & average genome size         & $5 \cdot 10^5$ [bp]             \\
+        $L$       & average genome size         & $50\, 000$ [bp]                 \\
         $\sigma$  & s.t.d. of genome size       & $L/10$ [bp]                     \\
@@
 Results are displayed in Fig.~\ref{fig:snps-suppl}.\\
-While the rate of HGT does not influence the average pairwise divergence of pancontigs, the choice of alignment kernel controls their average divergence.
-With the \textit{asm20} option, \textit{Minimap2} is able to correctly merge pangenome graphs with up to populations with $\avg{d} \sim 7\%$ in our simulations, while \textit{mmseqs2} reaches $\avg{d} \sim 10\%$ at the expense of higher processing time.\\
+For high values of $\mu$, the average pairwise divergence of pancontigs is not influenced by the rate of HGT $h$, but depends strongly on the choice of alignment kernel.
+With the \textit{asm20} option, \textit{minimap2} is able to correctly merge genomes with average pairwsise divergence $\avg{d} \sim 7\%$ in our simulations, while \textit{mmseqs2} reaches $\avg{d} \sim 10\%$ at the expense of higher processing time.\\
 By performing a linear fit on the datapoints with $\mu \leq 0.002$ we are able to recover the conversion factor between the mutation rate and average pairwise divergence of pancontigs in our simulations: $\avg{d} \sim 20.8 \, \mu$.
@@
     \avg{\Lgen}$, the total pangenome length $\Lpang$ and core genome length $\Lcore$. These quantities are reported in Table~\ref{table:panx-dataset}.\\
-EC is the dataset with the highest number of isolates (307) and is also the one with the biggest pangenome genome.
+EC is the dataset with the highest number of isolates (307) and is also the one with the biggest pangenome.
 MT is the species with the smallest sequence divergence ($\dcore \sim 0.03\%$) and the biggest core genome.
@@
 Fig.~\ref{fig:benchmark-accuracy-suppl}), but is merged correctly by the \textit{mmseqs2} kernel.
-The sequence divergence of PM is instead too high and cannot be handled correctly by any of the kernels.
+The sequence divergence of PM sits instead beyond the capabilities of all alignment kernels.
 
@@
 Aligning sequences with \textit{mmseqs2} consistently requires more time than \textit{minimap2}.
-The time required to build a pangenome graph for the 307 chromosomes of EC is only 5-6 h for \textit{minimap2}, and around 24h for \textit{mmseqs2}.
+The time required to build a pangenome graph for the 307 chromosomes of EC is only 5h for \textit{minimap2}, and around 24h for \textit{mmseqs2}.
 The latter also has a higher minimum memory requirement (panel C, around 8 Gb when running on 8 cores) but this is comparable to \textit{minimap2} when aligning several tens of isolates.\\
-We quantify the properties of pancontings by looking at three main statistics: their total number (panel D) the size of the minimal set of pancontigs that includes at least 50\% of all the graph sequence (L50 statistics, panel E) and the minimum length of pancontigs in this minimal set (N50 statistics, panel F).
+We quantify the properties of pancontigs by looking at three main statistics: their total number (panel D) the size of the minimal set of pancontigs that includes at least 50\% of all the graph sequence (L50 statistics, panel E) and the length of the shortest pancontig in this minimal set (N50 statistics, panel F).
 As expected, the use of null energy parameters results on average in a more fragmented graph, having more and shorter pancontigs.
@@
 When setting the energy parameters to zero more diverged sequences are merged into the same pancontigs and the total pangenome graph size is consistently smaller.
-The total size of the pangenome graph is in general much smaller than the cumulative size of the genomes it contains (panel I), with compression up to 1-2\% of the total size for graphs with a high number of isolates (EC) or with small accessory genome (MT).
-The fraction of core pancontigs (panel H) is consistent with the species expectation, with MT having the highest core fraction.\\
+The total graph size is in general much smaller than the cumulative size of the genomes it contains (panel I), with compression up to 1-2\% of the total size for graphs with a high number of isolates (EC) or with small accessory genome (MT).
+The fraction of core sequence (panel H) is consistent with the species expectation, with MT having the highest core fraction.\\
 
@@
 Fig.~\ref{fig:benchmark-accuracy-suppl}).
-The HP datasets represents an example in which relaxing the default divergence threshold of pangraph and using a more accurate alignment kernel results in a more complete merging of homologous sequence.\\
+The HP datasets represents an example in which relaxing the default divergence threshold of pangraph and using a more accurate alignment kernel results in a more consistent pangenome graph.\\
 Conversely, the PM dataset contains highly diverged isolates, whose divergence go beyond the capabilities of both alignment kernels ($\dcore \sim 27 \%$, cf.
 Table~\ref{table:panx-dataset}).
-Only \textit{mmseqs2} manages to partially merge some of the least diverged core sequences, but is still unable to merge the majority of the core genes.
+Only \textit{mmseqs2} manages to partially merge some of the least diverged core sequences.
 This results in a small core genome fraction and poor sequence compression.
@@
         \textbf{A}: number of isolates for each species.
-        \textbf{B}: algoritm wall-time.
+        \textbf{B}: algorithm wall-time.
         \textbf{C}: maximum memory usage.
         \textbf{D}: total number of pancontigs in the pangenome graph.
-        \textbf{E}: minimum number of pancontigs that can add up to 50\% of the total pagenome graph length.
-        \textbf{F}: threshold length such that if all pancontigs longer than this threshold are added they cover more than 50\% of the full pangenome graph.
+        \textbf{E}: minimum number of pancontigs that can add up to 50\% of the total pangenome graph length.
+        \textbf{F}: threshold length such that the cumulative size of all pancontigs longer than this threshold is more than 50\% of the full pangenome graph.
         \textbf{G}: total length of the pangenome graph (sum of all pancontig lengths).
         \textbf{H}: fraction of the pangenome graph that is composed of core pancontigs, i.e. pancontigs that occur exactly once per isolate.
-        \textbf{I}: ratio between the length of the pangenome graph and the total length of all chromosomes included in the graph.
+        \textbf{I}: ratio between the length of the pangenome graph and the total length of all genomes included in the graph.
     }
@@
 For each species we selected 50 different isolates (10 for PM) and built pangenome graphs using standard parameter values and the \textit{minimap2} kernel with \textit{asm20} option.
-For each species we then randomly picked 50 different pairs of isolates (all 45 for PM).\footnote{The list of selected accession numbers and selected pairs for each species is available on the repository: \url{https://github.com/neherlab/pangraph/blob/feat/finalize-paper/script/config/projection_strains.json}} For each pair we then built pairwise pangenome graphs using the same parameters, and compared them to the corresponding marginalized graph, obtained by projecting the full pangenome graph for the 50 isolates on the selected pair (cf.
+For each species we then randomly picked 50 different pairs of isolates (all 45 for PM).\footnote{The list of selected accession numbers and selected pairs for each species is available on the repository: \url{https://github.com/neherlab/pangraph/blob/feat/finalize-paper/script/config/projection_strains.json}} For each pair we then built pairwise graphs using the same parameters, and compared them to the corresponding marginalized graph, obtained by projecting the full pangenome graph for the 50 isolates on the selected pair (cf.
 Fig.~6A, main text).\\
@@
 Fig.~6A, main text).
-For each isolate we consider the two partitions defined by the marginalized and pairwise graph, and generate their intersection.
-Each block in this intersection can belong to two different categories: either the two original partitions agree or they disagree on whether the block is shared.
+For each isolate we consider the two partitions defined by the marginalized and pairwise graphs, and generate their intersection.
+Each block in this intersection can belong to two different categories, depending on whether the two partitions agree or disagree on whether the block is shared.
 Moreover, we can separate blocks on which we find agreement in two further categories: blocks that are shared and blocks that are private on both graphs.\\
 For each pair of genomes and each type of block we evaluate the fraction of the genome that they cover and their average size.
-In Fig.~\ref{fig:marginalize-suppl} we display the distribution of these quantities across all pairs for the different species considered.
-We consider the four categories described above: blocks on which we find agreement (\textit{agree}) or not (\textit{disagree}), and we separate agreement blocks in \textit{shared} and \textit{private} blocks.\\
-In all species except HP and PM we find very good agreement between pairwise and marginalized graphs: segments on which the graphs \textit{disagree} represent only a very minor fraction of the genome (much smaller than the fraction of shared or private blocks), and have size compatible with the default block length threhsold of \textit{PanGraph} (100 bp).
+In Fig.~\ref{fig:marginalize-suppl} we display the distribution of these quantities across all pairs for the different species considered and the block types described above.\\
+In all species except HP and PM we find very good agreement between pairwise and marginalized graphs: segments on which the graphs \textit{disagree} represent only a very minor fraction of the genome (much smaller than the fraction of shared or private blocks), and have size compatible with the default block length threshold of \textit{PanGraph} (100 bp).
 As described in the previous section, the divergence of the HP dataset sit on the limit of what the \textit{minimap2} alignment kernel can successfully handle.
 In this case the results of merging different graphs together can depend on the order of merging, and pairwise and marginalized graphs can potentially show some disagreement.
-In this case the disagreement remains minor and only extends to few percents of the genome size, with disagreement blocks having size only slightly bigger than the 100 bp threshold of PanGraph.
+In this case the disagreement remains minor and only extends to few percents of the genome size, with disagreement blocks having size only slightly bigger than the 100 bp threshold of \textit{PanGraph}.
 PM represents instead an example of a dataset with divergence much beyond the limit of the alignment kernel ($\dcore \sim 27 \%$).
@@
 The figure displays the distribution of the average size of these segments (right column) and the average fraction of genome that they cover (left column) for the 50 different pairs considered (45 for PM).
-        For all species with sequence divergence compatible with the limit of the alignment kernel used (KP, MT, EC), we find very good agreement between marginalized and pairwise graphs, with disagagreement segment only covering a very small fraction of the genome and having size compatible with \textit{PanGraph} threshold pancontig length (100 bp).
+        For all species with sequence divergence compatible with the limit of the alignment kernel used (KP, MT, EC), we find very good agreement between marginalized and pairwise graphs, with disagreement segment only covering a very small fraction of the genome and having size compatible with \textit{PanGraph} threshold pancontig length (100 bp).
     }

### 2022-10-10 main.tex
Update on Overleaf.

@@
 \author{Nicholas Noll}
+\thanks{These two authors contributed equally}
 \affiliation{Kavli Institute for Theoretical Physics, University of California, Santa Barbara \looseness=-5}
 \author{Marco Molari}
+\thanks{These two authors contributed equally}
 \affiliation{Swiss Institute of Bioinformatics, Basel, Switzerland \looseness=-5}
 \affiliation{Biozentrum, University of Basel, Basel, Switzerland \looseness=-5}
+\author{Liam Shaw}
+\affiliation{Department of Biology, University of Oxford, Oxford, UK\looseness=-5}
 \author{Richard A.~Neher}
@@
     The genomic diversity of microbes is commonly parameterized as population genetic polymorphisms relative to a reference genome of a well-characterized, but arbitrary, isolate.
-    Reference genomes contain a fraction of the microbial \emph{pangenome}, the set of genes observed within \emph{all} isolates of a given species, and are thus blind to both the dynamics of the accessory genome, as well as variation within gene order and copy number.
+    However, any reference genome contains only a fraction of the microbial \emph{pangenome}, the \textit{total} set of genes observed in a given species.
+    Reference-based approaches are thus blind to the dynamics of the accessory genome, as well as variation within gene order and copy number.
     With the wide-spread usage of long-read sequencing, the number of high-quality, complete genome assemblies has increased dramatically.
     Traditional computational approaches towards whole-genome analysis either scale poorly, or treat genomes as dissociated bags of genes, and thus are not suited for this new era.
-    Here, we present \emph{PanGraph}, a Julia based library and command line interface for aligning whole genomes into a graph, wherein each genome is represented as an undirected path along vertices, which in turn, encapsulate homologous multiple sequence alignments.
+    Here, we present \emph{PanGraph}, a Julia-based library and command line interface for aligning whole genomes into a graph. 
+    Each genome is represented as an undirected path along vertices, which in turn, encapsulate homologous multiple sequence alignments.
     The resultant data structure succinctly summarizes population-level nucleotide and structural polymorphisms and can be exported into a several common formats for either downstream analysis or immediate visualization.
@@
 %\section{Introduction}
-During evolution, microbial genomes change by both local modifications and large-scale alterations \cite{arnold2021horizontal}.
+During evolution, microbial genomes change by both local mutations and large-scale alterations \cite{arnold2021horizontal}.
 Local mutations conserve sequence architecture and only change a few nucleotides by substitution, insertion or deletion.
-Conversely, large-scale mutations reorganize the sequence, and involve either the homologous recombination of large segments or wholesale gene loss or gain from the environment.
-The accumulation of such polymorphisms over time complicates comparative genomic analyses of present day isolates; microbial genomes are mosaics of homologous sequence, locally related by an asexual phylogenetic tree \cite{sakoparnig2021whole}.
+Conversely, large-scale alterations reorganize the sequence, and involve either the homologous recombination of large segments or wholesale gene loss or gain.
+The accumulation of such changes over time complicates comparative genomic analyses of present day isolates; microbial genomes are mosaics of homologous sequence, locally related by an asexual phylogenetic tree \cite{sakoparnig2021whole}.
+% comment: rewritten to use consistent terminology (I think 'mutation' and 'polymorphism' for most people make them think of local changes only)
 %Scalable inference techniques of this complicated relatedness structure would, in part, spur the advance of critical epidemiological surveillance tools, analogous to NextStrain, for bacterial pathogens.
 
-Recent advances in long-read sequencing have enabled the low-cost assembly of isolated genomes at the quality of reference databases \cite{whibley2021changing}.
-The accumulation of many complete genomes promises to rapidly improve our ability to quantify the evolutionary dynamics that drive microbial diversity in natural populations.
-Unfortunately, comparative genomic analyses have not kept pace with the technological advance, opting either for reference-based approaches that only partially capture microbial diversity \cite{tettelin2008comparative}  or to construct a \emph{pangenome} that accurately captures nucleotide polymorphisms within genes but approximates structural polymorphisms as simple gene presence-absence relationships \cite{page2015roary,ding2018panx} .
+Recent advances in long-read sequencing have enabled the low-cost assembly of complete genomes at the quality of reference databases \cite{whibley2021changing}.
+The accumulation of so many complete genomes promises to rapidly improve our ability to quantify the evolutionary dynamics that drive microbial diversity in natural populations.
+Unfortunately, comparative genomic analyses have not kept pace with these technological advances.
+Since reference-based approaches only partially capture microbial diversity \cite{tettelin2008comparative}, the concept of the \emph{pangenome} has motivated the development of methods that account for substantial variation in gene content.
+However, though such approaches can accurately capture nucleotide polymorphisms within genes, they approximate structural polymorphisms as gene presence-absence relationships \cite{page2015roary,ding2018panx}.
 This new era of pangenomics demands novel data structures to encapsulate the \emph{complete} diversity of a given genomic sample set.
@@
 While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging.
-Colored generalizations of the de Bruijin graph-based assemblers have been successively used to build graphs from large sequence sets, however the underlying efficiency derives from a fixed kmer size which prevents modelling long-range homology \cite{iqbal2012novo,muggli2017succinct}.
+Colored generalizations of the de Bruijin graph-based assemblers have been successively used to build graphs from large sequence sets, although the underlying efficiency derives from a fixed k-mer size which prevents modelling long-range homology \cite{iqbal2012novo,muggli2017succinct}.
 An orthogonal approach has been to formulate the inference of the pangenome graph as a multiple genome alignment.
@@
 Here we present PanGraph, a Julia \cite{bezanson2017julia} library and command line interface, designed to efficiently align large sets of closely related genomes into a pangenome graph on personal computers.
-The resulting graph both compresses the input sequence set and succinctly captures the population diversity at multiple scales ranging from nucleotide mutations and indels to structural polymorphisms driven by inversions, rearrangements, and gene gain/loss.
+The resulting graph both compresses the input sequence set and succinctly captures the population diversity at multiple scales: from nucleotide mutations and indels to structural polymorphisms driven by inversions, rearrangements, and gene gain/loss.
 The underlying graph data structure can be exported into numerous formats for downstream analysis and visualization in software such as Bandage \cite{wick2015bandage}.
@@
 The final guide tree is constructed as the balanced binary tree constrained to reproduce the topological ordering of leaves found initially.
-See Fig \ref{fig:visualization}A for a graphical depiction of the guide tree.
+See Fig. \ref{fig:visualization}A for a graphical depiction of the guide tree.
 
@@
 The identified homologous intervals of the pancontig are then merged, thereby creating shorter contigs that represent homologous sections of multiple input genomes, see Fig.
-\ref{fig:visualization}B for a graphical deption.
+\ref{fig:visualization}B for a graphical depiction.
 These steps of pairwise graph alignment and merging of homologous intervals are repeated until the root of the guide tree is reached.
@@
 This allows for visualization in software such as Bandage \cite{wick2015bandage}.
-Lastly, we provide functionality to export as a conventional pangenome, however wherein \emph{pancontigs} supplant putative gene clusters, that can be visualized directly by the PanX toolkit \cite{ding2018panx}.
+Lastly, we provide functionality to export as a conventional presence/absence pangenome {--} albeit one with \emph{pancontigs} taking the place of putative gene clusters {--} that can be visualized directly by the PanX toolkit \cite{ding2018panx}.
 % TODO: add GFA version?
@@
 The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
-The simulation framework is distributed within the PanGraph command line toolsuite for external use.
+The simulation framework is distributed within the PanGraph command line tools for external use.
 
@@
 Importantly, PanGraph's computational complexity grows \emph{linearly} with the number of input genomes.
-We note that the PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of \textit{minimap2} \cite{li2018minimap2}.
+We note that PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of \textit{minimap2} \cite{li2018minimap2}.
 Benchmarks of CPU and memory usage are reported in SI Section IIA.
@@
 
-We quantified the accuracy of PanGraph algorithm by computing the displacement of the inferred breakpoints relative to their known locus stored from the evolutionary simulation (cf.
+We quantified the accuracy of PanGraph by computing the displacement of the inferred breakpoints relative to their known locus stored from the evolutionary simulation (cf.
 SI Section IIB for details).
 We evaluate this displacement by generating datasets with different rates of mutation $\mu$ and horizontal sequence transfer $h$.
-For each $(h,\mu)$ pair we perform 25 different simulations, and build pangenome graphs using three different options for the alignment kernel, namely \textit{minimap2} \cite{li2018minimap2} with \textit{asm10} and \textit{asm20} option and \textit{mmseqs2} \cite{steinegger2017mmseqs2}.
+For each $(h,\mu)$ pair we perform 25 different simulations, and build pangenome graphs using three different options for the alignment kernel: \textit{minimap2} \cite{li2018minimap2} with \textit{asm10} and \textit{asm20} options and \textit{mmseqs2} \cite{steinegger2017mmseqs2}.
 Critically, we found that the accuracy was \emph{independent} of the rate of horizontal transfer $h$ and thus underlying graph complexity.
@@
 SI Fig.~2).
-For low-diversity samples breakpoints are inferred with accuracy of few basepairs, while for highly diverged isolates most breakpoints are displaced by several hundreds of basepairs (cf.
+For low-diversity datasets, breakpoints are inferred with accuracy of few basepairs, while for highly diverged isolates most breakpoints are displaced by several hundreds of basepairs (cf.
 SI Fig.~3).
@@
 Fig.~\ref{fig:toy-accuracy}).
-On the sequences generated by our simulations, \textit{minimap2} with option \textit{asm20} show a loss of accuracy at a threshold pairwise divergence of around 5\%, while \textit{mmseqs2} is accurate up to average sequence divergence of around 12\%, at the cost of higher computational time.
+On the sequences generated by our simulations, \textit{minimap2} with option \textit{asm20} show a loss of accuracy at an average pairwise sequence divergence of around 5\%, while \textit{mmseqs2} is accurate up to around 12\%, at the cost of higher computational time.
 
@@
 \paragraph*{Dataset characterization}
-
-We downloaded from RefSeq \cite{o2016reference} completely assembled chromosomes from 5 different bacterial species: \textit{Escherichia Coli} (EC), \textit{Klebsiella Pneumoniae} (KP), \textit{Helicobacter Pylori} (HP), \textit{Mycobacterium Tuberculosis} (MT) and \textit{Prochlorococcus Marinus} (PM).
+We downloaded from RefSeq \cite{o2016reference} completely assembled chromosomes from 5 different bacterial species: \textit{Escherichia coli} (EC), \textit{Klebsiella pneumoniae} (KP), \textit{Helicobacter pylori} (HP), \textit{Mycobacterium tuberculosis} (MT) and \textit{Prochlorococcus ,arinus} (PM).
 These data had been previously analyzed using PanX \cite{ding2018panx}, which allowed us to estimate the size of the pangenome and core genome, and the average pairwise divergence on core genes (cf.
@@
         Species         & N.  & $\avg{\Lgen}$ & $\Lpang$ & $\Lcore$ & $\dcore$ \\
+                        &     & [Mbp]         & [Mbp]    & [Mbp]    &          \\
         \hline
-        E.~Coli         & 307 & 5.0           & 17.7     & 0.7      & 1.6\%    \\
-        K.~Pneumoniae   & 109 & 5.3           & 13.0     & 2.3      & 0.8\%    \\
-        H.~Pylori       & 85  & 1.6           & 1.9      & 0.6      & 4.2\%    \\
-        M.~Tuberculosis & 51  & 4.4           & 4.1      & 2.4      & 0.03\%   \\
-        P.~Marinus      & 10  & 1.8           & 3.3      & 0.7      & 26.9\%   \\
+        \textit{E.~coli}         & 307 & 5.0           & 17.7     & 0.7      & 1.6\%    \\
+        \textit{K.~pneumoniae}   & 109 & 5.3           & 13.0     & 2.3      & 0.8\%    \\
+        \textit{H.~pylori}       & 85  & 1.6           & 1.9      & 0.6      & 4.2\%    \\
+        \textit{M.~tuberculosis} & 51  & 4.4           & 4.1      & 2.4      & 0.03\%   \\
+        \textit{P.~marinus}      & 10  & 1.8           & 3.3      & 0.7      & 26.9\%   \\
         \hline
@@
         number of isolates (N.),
-        average chromosome length in Mbps ($\avg{\Lgen}$),
-        total pangenome length in Mbps ($\Lpang$),
-        total core genome length in Mbps ($\Lcore$),
+        average chromosome length ($\avg{\Lgen}$),
+        total pangenome length ($\Lpang$),
+        total core genome length ($\Lcore$),
         average pairwise divergence of core genes ($\dcore$).
@@
 \paragraph*{Benchmark on real data}
-
-Using \textit{PanGraph}, we build multiple pangenome graphs for each species in the dataset (cf.
+Using \textit{PanGraph}, we built multiple pangenome graphs for each species in the dataset (cf.
 SI Section~IIIB and IIIC).
@@
 \textit{PanGraph} can build pangenome graphs comprising hundreds of isolates in a few hours (5h for 307 EC isolates with \textit{minimap2} kernel, cf.
-Fig~\ref{fig:panx-benchmark}A).
+Fig.~\ref{fig:panx-benchmark}A).
 Use of \textit{mmseqs2} consistently requires longer computational time, but provides higher accuracy when merging highly diverged sequences.
@@
 In this case the use of \textit{mmseqs2} and null energy parameters increases core pangenome fraction, decreases the total pangenome size and and does not compromise the fragmentation of the graph (cf.
-Fig~\ref{fig:panx-benchmark}B-C and SI Fig.~4).
+Fig.~\ref{fig:panx-benchmark}B-C and SI Fig.~4).
 The divergence of the PM datasets sits instead beyond the capabilities of \textit{Pangraph}.
-In this case no alignment kernel can reach satisfactory sequence compression, and only \textit{mmseqs2} combined with null energy parameters is able to retrieve few core pancontigs.\\
-For the other species sequence compression is of the order of few percents, graphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in blocks spanning several kbps.
+In this case no alignment kernel can reach satisfactory sequence compression, and only \textit{mmseqs2} combined with null energy parameters is able to retrieve a few core pancontigs.\\
+For the other species the sequence compression (i.e.
+the ratio of the length of the sequence in the final pangenome graph to the total length of all input genomes) is around 1-4\%.
+Graphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in blocks spanning several kbps.
 For these species the use of null energy parameters slightly increases merging and sequence compression, at the cost of slightly more fragmented graphs (cf. SI Section IIIC and SI Fig.~4).
@@
 This operation is computationally much cheaper than building a new graph for the subset of strains considered.\\
-To verify that marginalized graphs are compatible with newly-built graphs, we built a pangenome graph for 50 randomly selected strains from the \textit{Klebsiella Pneumoniae} dataset.
+To verify that marginalized graphs are compatible with newly-built graphs, we built a pangenome graph for 50 randomly selected strains from the \textit{K.
+pneumoniae} dataset.
 We then picked 50 random pairs of strains from the same set and built two graphs for each pair: the graph obtained by marginalizing the complete graph on the pair (i.e.
@@
 Compatibility between the two graphs requires segments on which the two partitions disagree to be few and short.
-Indeed, we verified that over the pairs we picked these segments cover a very small fraction of the genome ($<1\%$), and have average size compatible with the default 100 bp precision threshold of \textit{PanGraph} (cf.
+Indeed, we verified that over the pairs we picked, these segments cover a very small fraction of the genome ($<1\%$) and have average size compatible with the default 100 bp precision threshold of \textit{PanGraph} (cf.
 Fig.~\ref{fig:marginalization}B-C).
@@
 The prevalence of agreement segment cannot be explained only by the fact that most of the genome is shared, since segments that are shared on both graphs cover on average only 85\% of the sequence.
-We confirmed this results using the fraction of shared k-mers, using an approach inspired by PopPUNK \cite{lees2019fast}.
+We confirmed these results using the fraction of shared k-mers, using an approach inspired by PopPUNK \cite{lees2019fast}.
 Namely, we approximate the fraction of homologous sequence between any two pairs using the fraction of shared k-mers ($k=21$) divided by $(1-d)^{k}$, where $d$ is the average pairwise divergence on core genes for the pair considered.
@@
         \textbf{A}: we built a pangenome graph from 50 randomly chosen strains from the \textit{K.
-Pneumoniae} dataset.
+pneumoniae} dataset.
 We then randomly picked 50 pairs of strains.
@@
 While single nucleotide differences in the core genome are straightforward to analyze with existing tools, these analyses miss the great majority of genetic diversity.
-The ability to rapidly align large sets of closely related, fully assembled genomes is crucial for the investigation of the processes governing microbial evolution.
+The ability to rapidly align large sets of complete genomes of a bacterial species is crucial for the investigation of the processes governing microbial evolution.
 We developed \textit{PanGraph} as a tool that can adress this need, by being able to capture the structural and nucleotide diversity in both the core and accessory genome in a scalable way.
@@
 In our analysis we demonstrated the capabilities and limits of this tool, both on synthetic and real sequences.
-The efficient implementation of \textit{PanGraph} allows it to create pangenome graphs containing hundreds of isolates in a few hours on a 8-cores machine.
+The efficient implementation of \textit{PanGraph} allows it to create pangenome graphs containing hundreds of isolates in a few hours on a 8-core machine.
 The size of pancontigs and the fraction of core sequece have good scaling properties with the number of isolates in the graph, indicating that \textit{PanGraph} is able to successfully capture pangenome properties.\\
@@
 eq.~(\ref{eq:pseudo-energy})), trading off speed for sensitivity.\\
-We also provide the ability to quickly marginalize big graps on a subset of strains, obtaining simpler graphs that can more easily be explored.
+We also provide the ability to quickly marginalize big graphs on a subset of strains, obtaining simpler graphs that can more easily be explored.
 Graphs can also be exported in different formats for further analysis and visualization.
 
-We hope that these characteristics will make \textit{PanGraph} a valuable tool, with the potential to spur novel insight on microbial diversity and the processes by which bacteria adapt and change.
+We hope that these features will make \textit{PanGraph} a valuable tool, with the potential to spur new insights into microbial diversity and the processes by which bacteria adapt and change.
 

### 2022-10-10 supplementary.tex
Update on Overleaf.

@@
 \begin{itemize}
-    \itemsep1pt
+    \itemsep0pt
     \item \textit{minimap2} with option \textit{asm10},
@@
 For high values of $\mu$, the average pairwise divergence of pancontigs is not influenced by the rate of HGT $h$, but depends strongly on the choice of alignment kernel.
-With the \textit{asm20} option, \textit{minimap2} is able to correctly merge genomes with average pairwsise divergence $\avg{d} \sim 7\%$ in our simulations, while \textit{mmseqs2} reaches $\avg{d} \sim 10\%$ at the expense of higher processing time.\\
+With the \textit{asm20} option, \textit{minimap2} is able to correctly merge genomes with average pairwsise divergence $\avg{d} \sim 7\%$ in our simulations, while \textit{mmseqs2} reaches $\avg{d} \sim 10\%$ at the expense of higher processing time.
+Notice that these values are lower than the threshold sensitivity of these aligners, which are expected to find matches up to respectively around 10\%, 20\% and 30\% sequence divergence.
+This is due to the fact that homologous sequences in our simulations are generated by an evolutionary process, they can have a wide distribution of diversities and hierarchical population structure.
+Inability to merge sequences above a threshold divergence in a single panconting is expected to result in multiple pancontigs containing sub-clades with higher similarity, and average diversity that is appreciably lower than the aligner threshold.\\
 By performing a linear fit on the datapoints with $\mu \leq 0.002$ we are able to recover the conversion factor between the mutation rate and average pairwise divergence of pancontigs in our simulations: $\avg{d} \sim 20.8 \, \mu$.
@@
 We formulate this problem as an assignment problem, and solve it numerically using the Hungarian method \cite{kuhn1955hungarian}.
-For each isolate we call $b_i$ for $i=1,\ldots,I$ the panconting breakpoints on the real pangenome graph, and $b_j$ for $j=1,\ldots,J$ the ones on the reconstructed pangenome graph.\footnote{To avoid artifacts generated by the default threshold distance $d=100$ bp of
-    \textit{PanGraph}, if any pair of breakpoints on the same graph sits at a distance smaller than this threshold, we remove one of them.} We define a cost matrix $D_{ij} = d(b_i,b_j)$ whose elements are distances on the genome of pairs of breakpoints from the two graphs.
+For each isolate we call $b_i$ for $i=1,\ldots,I$ the panconting breakpoints on the real pangenome graph, and $b_j$ for $j=1,\ldots,J$ the ones on the reconstructed pangenome graph.\footnote{To avoid artifacts generated by the default threshold distance $d=100$ bp of \textit{PanGraph}, if any pair of breakpoints on the same graph sits at a distance smaller than this threshold, we remove one of them.} We define a cost matrix $D_{ij} = d(b_i,b_j)$ whose elements are distances on the genome of pairs of breakpoints from the two graphs.
 We then numerically solve the assignment problem on this matrix, and obtain a set of $K = \min\{ I,J \}$ pairings between breakpoints such that the distance between the pairs is minimal and all the breakpoints of the graph with minimum number of breakpoints have been paired.
@@
 
-We downloaded sets of annotated and complete chromosomes from RefSeq \cite{o2016reference} for 5 different bacterial species: \textit{Klebsiella Pneumoniae} (KP), \textit{Helicobacter Pylori} (HP), \textit{Prochlorococcus Marinus} (PM), \textit{Mycobacterium Tuberculosis} (MT) and \textit{Escherichia Coli} (EC).
-In Table~\ref{table:panx-dataset} we report some summary statistics for these datasets.
+We downloaded sets of annotated and complete chromosomes from RefSeq \cite{o2016reference} for 5 different bacterial species: \textit{Klebsiella pneumoniae} (KP), \textit{Helicobacter pylori} (HP), \textit{Prochlorococcus marinus} (PM), \textit{Mycobacterium tuberculosis} (MT) and \textit{Escherichia coli} (EC).
+In Table~I in the main text we report some summary statistics for these datasets.
 The number of isolates per species ranges from several hundreds for EC to 10 for PM.
@@
 This allowed us to identify core genes, and evaluate their average pairwise sequence divergence $\avg{d_\text{core}}$.
-We also evaluated the average genome length $
-    \avg{\Lgen}$, the total pangenome length $\Lpang$ and core genome length $\Lcore$. These quantities are reported in Table~\ref{table:panx-dataset}.\\
+We also evaluated the average genome length $\avg{\Lgen}$, the total pangenome length $\Lpang$ and core genome length $\Lcore$ (see Table~I in the main text).\\
 EC is the dataset with the highest number of isolates (307) and is also the one with the biggest pangenome.
-MT is the species with the smallest sequence divergence ($\dcore \sim 0.03\%$) and the biggest core genome.
+MT is the species with the smallest sequence divergence ($\dcore \sim 0.03\%$) and the biggest core genome fraction.
 The two datasets with highest sequence divergence are HP ($\dcore \sim 4\%$) and PM ($\dcore \sim 24\%$).
@@
 The sequence divergence of PM sits instead beyond the capabilities of all alignment kernels.
-
-\begin{table}[hb]
-    \setlength{\tabcolsep}{9pt}
-    \begin{tabular}{lccccc}
-        \hline\hline
-        Species         & N. isolates & $\avg{\Lgen}$ [Mbp] & $\Lpang$ [Mbp] & $\Lcore$ [Mbp] & $\dcore$ \\
-        \hline
-        E.~Coli         & 307         & 5.0                 & 17.7           & 0.7            & 1.6\%    \\
-        K.~Pneumoniae   & 109         & 5.3                 & 13.0           & 2.3            & 0.8\%    \\
-        H.~Pylori       & 85          & 1.6                 & 1.9            & 0.6            & 4.2\%    \\
-        M.~Tuberculosis & 51          & 4.4                 & 4.1            & 2.4            & 0.03\%   \\
-        P.~Marinus      & 10          & 1.8                 & 3.3            & 0.7            & 26.9\%   \\
-        \hline
-    \end{tabular}
-    \caption{{\bf Dataset overview}.
-We downloaded complete chromosome sequences from RefSeq for 5 different bacterial species.
-For each species we report the number of isolates downloaded, along with the average chromosome length $\avg{\Lgen}$, total size of pangenome  $\Lpang$ and core genome $\Lcore$, and average pairwise divergence on core genes $\dcore$.}
-    \label{table:panx-dataset}
-\end{table}
 

### 2022-11-18 main.tex
Update on Overleaf.

@@
 \usepackage{enumerate}
+\usepackage{color}
 \usepackage[final]{hyperref}
@@
 \newcommand{\dcore}{\langle d_\text{core} \rangle}
+\newcommand{\Richard}[1]{{\color{red}Richard: #1}}
 
@@
 \affiliation{Biozentrum, University of Basel, Basel, Switzerland \looseness=-5}
-\author{Liam Shaw}
+\author{Liam P. Shaw}
 \affiliation{Department of Biology, University of Oxford, Oxford, UK\looseness=-5}
@@
 \begin{abstract}
-    The genomic diversity of microbes is commonly parameterized as population genetic polymorphisms relative to a reference genome of a well-characterized, but arbitrary, isolate.
+    The genomic diversity of microbes is commonly parameterized as single nucleotide polymorphisms relative to a reference genome of a well-characterized, but arbitrary, isolate.
     However, any reference genome contains only a fraction of the microbial \emph{pangenome}, the \textit{total} set of genes observed in a given species.
@@
     With the wide-spread usage of long-read sequencing, the number of high-quality, complete genome assemblies has increased dramatically.
-    Traditional computational approaches towards whole-genome analysis either scale poorly, or treat genomes as dissociated bags of genes, and thus are not suited for this new era.
+    Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.
     Here, we present \emph{PanGraph}, a Julia-based library and command line interface for aligning whole genomes into a graph. 
@@
 During evolution, microbial genomes change by both local mutations and large-scale alterations \cite{arnold2021horizontal}.
-Local mutations conserve sequence architecture and only change a few nucleotides by substitution, insertion or deletion.
-Conversely, large-scale alterations reorganize the sequence, and involve either the homologous recombination of large segments or wholesale gene loss or gain.
-The accumulation of such changes over time complicates comparative genomic analyses of present day isolates; microbial genomes are mosaics of homologous sequence, locally related by an asexual phylogenetic tree \cite{sakoparnig2021whole}.
+Local mutations only change a few nucleotides by substitution, insertion or deletion.
+Conversely, large-scale alterations reorganize the sequence, and involve either the homologous recombination of large segments, gene loss or gain, inversions, or mobilisation of genetic elements. 
+The accumulation of such changes over time complicates comparative genomic analyses of present day isolates. 
+Homologous recombination is rapid enough that most genes in many bacterial core genomes have distinct phylogenies \cite{sakoparnig2021whole} and even closely related genomes differ dramatically in gene content.
 % comment: rewritten to use consistent terminology (I think 'mutation' and 'polymorphism' for most people make them think of local changes only)
@@
 The accumulation of so many complete genomes promises to rapidly improve our ability to quantify the evolutionary dynamics that drive microbial diversity in natural populations.
-Unfortunately, comparative genomic analyses have not kept pace with these technological advances.
 Since reference-based approaches only partially capture microbial diversity \cite{tettelin2008comparative}, the concept of the \emph{pangenome} has motivated the development of methods that account for substantial variation in gene content.
-However, though such approaches can accurately capture nucleotide polymorphisms within genes, they approximate structural polymorphisms as gene presence-absence relationships \cite{page2015roary,ding2018panx}.
+However, though such approaches can accurately capture nucleotide polymorphisms within genes, they approximate structural polymorphisms as gene presence-absence relationships \cite{page2015roary,ding2018panx} irrespective of gene order or orientation.
 This new era of pangenomics demands novel data structures to encapsulate the \emph{complete} diversity of a given genomic sample set.
@@
 In recent years, efforts have focused on generalizing the \emph{pangenome} framework of microbial diversity to graphical models \cite{eizenga2020pangenome}.
-At a high level, pangenome graphs generalize the reference sequence coordinate system conventionally used; microbial diversity is, instead, encapsulated by a graph-based data structure, with vertices or edges labelled by homologous DNA sequences, in which a walk recapitulates a subset of mosaic genomes.
+At a high level, pangenome graphs generalize the reference sequence coordinate system conventionally used and encode genomes as directed paths through the graph consisting of node that represent sequence.
+In multicellular eukaryotes, pangenome graphs usually have an overall linear structure in which structural diversity can be encoded as short range  excursions. 
+In microbial genomics, in contrast, large scale rearrangements and inversions give rise to complex graph topologies. 
+Here, we focus on pangenome graphs of closely related bacterial genomes.
+\Richard{I feel the following is more confusing than helping.}
 As such, pangenome graphs can be thought of as an analogous construct to the well-known assembly graph, in which whole genomes are individual ``reads" of the underlying microbial structure \cite{myers2005fragment}.
@@
 \emph{Pancontigs} are connected by an edge if they are syntenic on at least one input sequence; individual sequences are then recapitulated by contiguous \emph{paths} through the graph.
+Pancontigs are directed, meaning their orientation in a path can be either 5' to 3' or vice versa.
 
-\emph{PanGraph's} overarching strategy is to approximate multiple-genome alignment by iterative pairwise alignment, in the spirit of progressive alignment tools \cite{darling2010progressivemauve,armstrong2020progressive}.
-A guide tree is utilized to linearize the problem's complexity by approximating multiple-sequence alignment as a quenched order of successive pairwise alignments \cite{feng1987progressive}.
-Pairwise graph alignment is performed by an all-to-all alignment of the \emph{pancontigs} between both graphs.
+To construct a pangraph, the algorithm needs to find homologous sequence within and among all input genomes. 
+\emph{PanGraph's} overarching strategy is to approximate multiple-genome alignment by iterative pairwise alignment of graphs of subsets of sequences, in the spirit of progressive alignment tools \cite{feng1987progressive,darling2010progressivemauve,armstrong2020progressive}.
+Pairwise graph alignment is performed by an all-to-all alignment of the \emph{pancontigs} between both graphs and the order of pairwise alignments is determined by a guide tree.
 
 \subsection{Guide tree construction}
-The alignment guide tree is constructed subject to three design constraints: (i) sequences are aligned sequentially based upon similarity, (ii) the similarity computation scales sub-quadratically with the number of input sequences, and (iii) the resultant tree is balanced.
+The alignment guide tree is constructed subject to three design constraints: (i) similar sequences are aligned first, (ii) the similarity computation scales sub-quadratically with the number of input sequences, and (iii) the resultant tree is balanced to maximize parallelism.
 To this end, we formulate the algorithm as a two step process.
@@
 Computationally, each sequence can be sketched into its set of minimizers in linear time while the cardinality of all pairwise intersections can be computed by sorting the list of all minimizers to efficiently count overlaps.
-Hence, the pairwise distance matrix is estimated in log-linear time.
+Hence, the pairwise distance matrix is estimated in a time that increases log-linearly in the total length of sequence that is analyzed. 
 The final guide tree is constructed as the balanced binary tree constrained to reproduce the topological ordering of leaves found initially.
+This balancing maximizes the number of independent pairwise alignments and thereby allows efficient parallelism. 
 See Fig. \ref{fig:visualization}A for a graphical depiction of the guide tree.
@@
 \subsection{Iterative graph alignment}
-The iterative pangraph construction is illustrated in Fig.~\ref{fig:visualization}.
-The pangraph is initialized as one subgraph per input genome, each representing its respective genome as a single \emph{pancontig}.
-Pairs of subgraphs are aligned in a postorder traversal of the constructed guide tree.
+The full pangraph representing all genomes is constructed by aligning/merging graphs that represent subsets of the genomes in an iterative manner illustrated in Fig.~\ref{fig:visualization}.
+The iteration starts with one subgraph per input genome, each representing its respective genome as a single \emph{pancontig}.
+Pairs of subgraphs are aligned in a postorder traversal of the guide tree.
 The identified homologous intervals of the pancontig are then merged, thereby creating shorter contigs that represent homologous sections of multiple input genomes, see Fig.
@@
 These steps of pairwise graph alignment and merging of homologous intervals are repeated until the root of the guide tree is reached.
+\emph{Pancontigs} encapsulate linear multiple-sequence alignments which are modelled internally by a star phylogeny, i.e.
+are assumed to be well-described by a reference sequence augmented by independent SNPs and indels for each contained isolate.
 
-\paragraph{Pairwise alignment.}
-Full genome alignment between two closely related isolates is a well-studied problem with many high-quality tools available \cite{li2018minimap2,marccais2018mummer4}.
+\paragraph*{Pairwise alignment.}
+To align two graphs, the consensus sequences of all pancontigs in both graphs are searched for homologies and aligned. 
+Full genome alignment between two closely related isolates is a well-studied problem with many sensitive and efficient tools available \cite{li2018minimap2,marccais2018mummer4}.
 We chose to use \emph{minimap2} as the core pairwise genome aligner for its proven speed, sensitivity, and easy-to-use exposed library API \cite{li2018minimap2}.
@@
 
-\paragraph{Merging of homologous sequence.}
-\emph{Pancontigs} encapsulate linear multiple-sequence alignments which are modelled internally by a star phylogeny, i.e.
-are assumed to be well-described by a reference sequence augmented by independent SNPs and indels for each contained isolate.
-Importantly, during the all-to-all alignment phase, \emph{pancontigs} are aligned based \emph{solely} upon their consensus sequence.
-All putative homologous alignments found by \emph{pancontig} alignment are ranked according to the pseudo-energy
+\paragraph*{Merging of homologous sequence.}
+If the above alignment step detected homologous stretches between two pancontigs, these pancontigs, or parts of them, can be merged. 
+It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order. 
+We rank each alignment between two pancontigs according to the pseudo-energy
 \begin{equation}\label{eq:pseudo-energy}
@@
 See Fig \ref{fig:visualization}B for a graphical definition for each term.
-Additionally, $\alpha$ and $\beta$ are hyperparameters of the algorithm, respectively controlling the tradeoff between fragmentation of the graph and the maximum diversity within each block.
+Additionally, $\alpha$ and $\beta$ are hyperparameters of the algorithm. 
 Only mergers whose alignment has negative pseudo-energy are performed.
+The parameter $\beta$ controls the maximal sequence divergence between the two merger-candidates and no mergers will be performed if the Hamming distance exceeds $1/\beta$. 
+Importantly, this divergence threshold is applied to the comparison of the consensus sequences of the merger candidates.
+The parameter $\alpha$ controls the fragmentation of the graph and imposes a minimal length on the resulting pan-contigs resulting from the merger.
+The merger is always performed if the merger doesn't increase the number of contigs, but mergers that introduce 1, 2, 3 or 4 cuts are only performed for sufficiently long homologous stretches as parameterized by $\alpha$.
+In addition, there is parameter that controls the minimal size of contigs produced by the algorithm.
 
@@
 Once both sets of sequences are placed onto a common coordinate system, the resultant consensus sequence, and thus polymorphisms, are recomputed.
-This procedure can be viewed as an online multiple sequence alignment algorithm.
+This procedure can be viewed as an approximate multiple sequence alignment algorithm that aligns homologous parts without further investigating insertions in the two sub-alignments.
 
 The above procedure is repeated until no alignments with negative energy remain.
-Upon completion, transitive edges within the graph are removed by merging adjacent \emph{pancontigs}.
+Upon completion, transitive edges within the graph, i.e., edges along which all path are co-linear, are removed by merging adjacent \emph{pancontigs}.
 
 \subsection{Parallelism}
-\emph{PanGraph} is designed with a message-passing architecture to enable scalable parallelism, as shown in Fig \ref{fig:visualization}A for a cartoon example.
+\emph{PanGraph} guide trees are balanced binary trees that break the task into many independent sub-problems and thereby enable scalable parallelism, as shown in Fig \ref{fig:visualization}A for a cartoon example.
 Each internal node of the guide tree represents a job that performs a single pairwise graph alignment between its two children.
@@
 Additionally, pre-built binaries are available as GitHub releases.
-All data and scripts used to validate PanGraph are available within the same repository.
+All data and scripts used to validate PanGraph are available within the same repository.\footnote{Instructions on how to run the analysis scripts are available at \url{https://github.com/neherlab/pangraph/blob/master/script/README.md}}
 
@@
 
-\paragraph*{Generation of synthetic data}
+\paragraph*{Generation of synthetic data.}
 
-We simulated populations of size $N=100$ and sequence length $L=50\,000$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
+We simulated populations of size $N=100$ and varying sequence length up to $L=500\,000$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
 In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (respective rates $0.01$ and $0.05$ per generation), and horizontal sequence transfer that occurs with tunable rate $h$ per generation per genome (see SI Section I for details).
@@
 
-\paragraph*{Performances as a function of dataset size}
-
-The algorithmic complexity was measured empirically by simulating populations of increasing size and average sequence length, and generating pangenome graphs for the resulting sequences.
-The mean and standard deviation of results obtained from 50 iterations are shown in Fig. \ref{fig:toy-performance}.
-Importantly, PanGraph's computational complexity grows \emph{linearly} with the number of input genomes.
+\paragraph*{Performances as a function of dataset size.}
+The algorithmic complexity was measured empirically by constructing pangenome graphs from data generated by simulating populations of increasing size and average sequence length.
+The mean and standard deviation of the run-time obtained from 50 iterations are shown in Fig. \ref{fig:toy-performance}.
+Importantly, PanGraph's computational complexity grows \emph{linearly} with the number of input genomes once the number of genomes exceeds a certain threshold and parallelism can be exploited.
 We note that PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of \textit{minimap2} \cite{li2018minimap2}.
@@
 
-\begin{figure}[htb]
+\begin{figure}[tb]
     \includegraphics[width=.4\textwidth]{figs/benchmark.pdf}
@@
 \paragraph*{Accuracy as a function of sequence divergence}
-
-We quantified the accuracy of PanGraph by computing the displacement of the inferred breakpoints relative to their known locus stored from the evolutionary simulation (cf.
+To be useful and informative beyond data compression of microbial genomes, the contigs and break-points of the reconstructed graphs should correspond, at least approximately, to evolutionary events in the history of the sample.
+We quantified PanGraph's ability to accurately reconstruct the true graph by computing the displacement of the inferred breakpoints relative to their known locus stored from the evolutionary simulation (cf.
 SI Section IIB for details).
@@
 SI Fig.~3).
-The choice of alignment kernel influences the threshold diversity at which accuracy is lost.\\
-For each alignment kernel and value of average sequence divergence in the population we evaluated the fraction of breakpoints that have displacement greater than the default precision threshold for Pangraph of 100 bp (cf.
+The choice of alignment kernel influences the threshold diversity at which accuracy is lost, suggesting that inability to detect homology between diverged sequences is the reason.\\
+For each alignment kernel and value of average sequence divergence in the population we evaluated the fraction of breakpoints that have displacement greater than the default minimal block size for Pangraph of 100 bp (cf.
 Fig.~\ref{fig:toy-accuracy}).
 On the sequences generated by our simulations, \textit{minimap2} with option \textit{asm20} show a loss of accuracy at an average pairwise sequence divergence of around 5\%, while \textit{mmseqs2} is accurate up to around 12\%, at the cost of higher computational time.
+These threshold are lower than the nominal sensitivity of the aligners, which are 10\% and 20\% for minimap2 with settings \textit{asm10} and \textit{asm20}, respectively, and around 30\% for \textit{mmseqs2}.
+This discrepancy is due to the fact that for accurate graph constructions, the largest divergence, rather than the average divergence, is the relevant quantity.
 
-\begin{figure}[htb]
+\begin{figure}[tb]
     \includegraphics[width=.4\textwidth]{figs/misplaced_fraction_vs_divergence.pdf}
     \caption{{\bf Accuracy against synthetic data}.
-        We generate artificial data with varying degree of sequence divergence, and compare the real underlying pangenome graph with the one reconstructed by \textit{PanGraph}, for three different alignment kernels :\textit{minimap2} with \textit{asm10} or \textit{asm20} option, and \textit{mmseqs2}.
+        We generate artificial data with varying degree of sequence divergence, and compare the real underlying pangenome graph with the one reconstructed by \textit{PanGraph}, for three different alignment kernels \textit{minimap2} with \textit{asm10} or \textit{asm20} option, and \textit{mmseqs2}.
 In each comparison we evaluate the misplacement of breakpoints that we can pair on the two graphs within 1kbp.
@@
 
-We additionally validated PanGraph on natural populations sampled from RefSeq \cite{o2016reference}, focusing on the properties of the resulting pangenome graphs as a function of dataset size and diversity.
+We additionally validated PanGraph on genomes from natural populations sampled from RefSeq \cite{o2016reference}, focusing on the properties of the resulting pangenome graphs as a function of dataset size and diversity.
 
-\paragraph*{Dataset characterization}
-We downloaded from RefSeq \cite{o2016reference} completely assembled chromosomes from 5 different bacterial species: \textit{Escherichia coli} (EC), \textit{Klebsiella pneumoniae} (KP), \textit{Helicobacter pylori} (HP), \textit{Mycobacterium tuberculosis} (MT) and \textit{Prochlorococcus ,arinus} (PM).
+\paragraph*{Dataset characterization.}
+We downloaded from RefSeq \cite{o2016reference} completely assembled chromosomes from 5 different bacterial species: \textit{Escherichia coli} (EC), \textit{Klebsiella pneumoniae} (KP), \textit{Helicobacter pylori} (HP), \textit{Mycobacterium tuberculosis} (MT) and \textit{Prochlorococcus marinus} (PM).
 These data had been previously analyzed using PanX \cite{ding2018panx}, which allowed us to estimate the size of the pangenome and core genome, and the average pairwise divergence on core genes (cf.
 Table~\ref{table:panx-dataset} and SI Section~IIIA).
+
+Among these 5 data sets, the \textit{E.~coli}, \textit{K.~pneumoniae}, \textit{M.~tuberculosis} genomes have core genome diversity below the thresholds of the minimap2 aligner, while \textit{H.~pylori} is diverse enough that we don't expect minimap2 to find all relevant matches, while mmseq2 should process this set without problems.
+\textit{P.~marinus}, on the other hand, is very diverse with an average core genome diversity of 26\% and is beyond what we expect PanGraph to handle.
+
 % The \textit{E.~Coli} (EC) dataset is the one with the highest number of isolates (307 chromosomes) and the largest pangenome.
@@
 
-\paragraph*{Benchmark on real data}
+\paragraph*{Benchmark on real data.}
 Using \textit{PanGraph}, we built multiple pangenome graphs for each species in the dataset (cf.
@@
 Moreover the use of null energy parameters ($\alpha = \beta = 0$) is expected to remove the threshold divergence of 10\% associated with the standard values of parameters ($\alpha=100$, $\beta=10$) at the cost of more fragmented pangenome graphs.\\
-\textit{PanGraph} can build pangenome graphs comprising hundreds of isolates in a few hours (5h for 307 EC isolates with \textit{minimap2} kernel, cf.
-Fig.~\ref{fig:panx-benchmark}A).
-Use of \textit{mmseqs2} consistently requires longer computational time, but provides higher accuracy when merging highly diverged sequences.
-This is the case for the HP dataset, whose core genome divergence sits at the limit of what can be accurately merged with the \textit{minimap2} kernel.
+\textit{PanGraph} can build pangenome graphs comprising hundreds of isolates in a few hours (5h for 307 E.~Coli isolates with \textit{minimap2} kernel, cf. Fig.~\ref{fig:panx-benchmark}A). 
+Use of \textit{mmseqs2} consistently requires longer computational time, but provides higher sensitivity when merging highly diverged sequences.
+
+Fig.~\ref{fig:panx-benchmark}B-C and SI Fig.~4 quantify the size of the core genome, i.e.~the sum of all pancontigs present once in every genome, identified by PanGraph and the compression of the dataset, i.e.~the ratio for the sum of the lengths of all pancontigs and the sum of the lengths of all input genomes, for different dataset and the different parameters.
+As expected, the very homogenous \textit{M.~tuberculosis} dataset has a large core genome with and a compression ratio that is almost equal to the inverse number of input genomes, independent of the aligner or parameters used.
+\textit{E.~coli} and \textit{K.~pneumoniae} compress similarly well, but their core genome is a much smaller fraction of the pangenome. 
+Using the more sensitive aligner mmseqs2 and relaxing the merger parameters leads to some additional compression, presumably due to merging of diverged paralogs.
+
+The \textit{H.~pylori}  dataset sits at the limit of what can be accurately merged with the \textit{minimap2} kernel. 
 In this case the use of \textit{mmseqs2} and null energy parameters increases core pangenome fraction, decreases the total pangenome size and and does not compromise the fragmentation of the graph (cf.
 Fig.~\ref{fig:panx-benchmark}B-C and SI Fig.~4).
-The divergence of the PM datasets sits instead beyond the capabilities of \textit{Pangraph}.
+The divergence of the \textit{P.~marinus} dataset sits instead beyond the capabilities of \textit{Pangraph}.
 In this case no alignment kernel can reach satisfactory sequence compression, and only \textit{mmseqs2} combined with null energy parameters is able to retrieve a few core pancontigs.\\
-For the other species the sequence compression (i.e.
-the ratio of the length of the sequence in the final pangenome graph to the total length of all input genomes) is around 1-4\%.
-Graphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in blocks spanning several kbps.
+
+For species other than \textit{P.~marinus}, the pangraphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in blocks spanning several kbps (see SI Fig.~4).
 For these species the use of null energy parameters slightly increases merging and sequence compression, at the cost of slightly more fragmented graphs (cf. SI Section IIIC and SI Fig.~4).
@@
 
-\paragraph*{Scaling with dataset size}
+\paragraph*{Scaling with dataset size.}
 
-We then explore how the properties of the pangenome graph scale with the dataset size.
-To do so we build pangenome graphs from random sets of isolates of increasing size from the EC dataset.
+We then explored how the properties of the pangraph scale with the dataset size.
+To do so, we build pangenome graphs from randomized subsets of isolates of increasing size from the \textit{E.~coli} dataset.
 Results are reported in Fig.~\ref{fig:panx-size}.\\
@@
 The interpretation of a pangenome graph containing hundreds of strains can be challenging.
-To this end it is often informative to inspect simpler sub-graphs comprising only a small subset of isolates.
-However building multiple such sub-graphs is computationally intensive.
-To facilitate this task \textit{PanGraph} provides the \verb|marginalize| command, that can be used to marginalize a large pangenome graph on a small subset of strains, removing all the other paths and merging transitive pancontigs.
-This operation is computationally much cheaper than building a new graph for the subset of strains considered.\\
-To verify that marginalized graphs are compatible with newly-built graphs, we built a pangenome graph for 50 randomly selected strains from the \textit{K.
-pneumoniae} dataset.
-We then picked 50 random pairs of strains from the same set and built two graphs for each pair: the graph obtained by marginalizing the complete graph on the pair (i.e.
-pairwise graph), and the graph built directly from the pair (i.e.
-marginalized graph, cf.
-Fig.~\ref{fig:marginalization}A top).
-We compare these two graphs by considering the partition they generate on the genome of the isolates they include.
+To this end, it is often informative to inspect simpler sub-graphs comprising only a subset of isolates.
+However building many such sub-graphs is computationally intensive.
+To facilitate this task \textit{PanGraph} provides the \verb|marginalize| command, that can be used to project a large pangenome graph on a small subset of strains, removing all the other paths and merging transitive pancontigs.
+This operation is computationally much cheaper than building a new graph for the subset of strains considered.
+
+In addition to being a useful operation, marginalization allows us to quantify the consistency and robustness of pangraph construction on real data, where the ground truth is unknown.
+To verify that marginalized graphs are compatible with newly-built graphs, we built a pangenome graph for 50 randomly selected strains from the \textit{K.~pneumoniae} dataset.
+We then picked 50 random pairs of strains from the same set and built two graphs for each pair: the graph obtained by marginalizing the complete graph on the pair ( (i.e.
+marginalized graph in Fig.~\ref{fig:marginalization}A top), and the graph built directly from the pair (i.e.
+pairwise graph in Fig.~\ref{fig:marginalization}A top).
+In order to compare both graphs, we compute the partition they generated on the genome of the isolates they include.
 Each genome is partitioned in pancontigs that are either shared on the pair, or private to one isolate (cf.
@@
 Moreover segments on which the two partitions agree are further sub-divided depending on whether they are private or shared on both partitions.\\
-Compatibility between the two graphs requires segments on which the two partitions disagree to be few and short.
+The compatibility between the two graphs requires segments on which the two partitions disagree to be few and short.
 Indeed, we verified that over the pairs we picked, these segments cover a very small fraction of the genome ($<1\%$) and have average size compatible with the default 100 bp precision threshold of \textit{PanGraph} (cf.
 Fig.~\ref{fig:marginalization}B-C).
-Conversely, agreement segments have average size of several kbps.
-The prevalence of agreement segment cannot be explained only by the fact that most of the genome is shared, since segments that are shared on both graphs cover on average only 85\% of the sequence.
+Conversely, segments on which the partitions agree have average size of several kbps.
+The fact that the two partitions almost completely agree cannot be explained simply by the fact that most of the genome is shared, since segments that are shared on both graphs cover on average only 85\% of the sequence.
 We confirmed these results using the fraction of shared k-mers, using an approach inspired by PopPUNK \cite{lees2019fast}.
@@
 This divisor corrects for k-mers on homologous sequence that are not shared due to mutations.
-The resulting distribution of shared sequence is indeed in very good agreement with the fraction of segments that are shared on both graphs (cf.
+The resulting distribution of shared sequence is in very good agreement with the fraction of segments that are shared on both graphs (cf.
 Fig.~\ref{fig:marginalization}B), suggesting that homologous sequences are correctly merged on the complete, marginalized and pairwise graphs.\\
@@
 The ability to rapidly align large sets of complete genomes of a bacterial species is crucial for the investigation of the processes governing microbial evolution.
-We developed \textit{PanGraph} as a tool that can adress this need, by being able to capture the structural and nucleotide diversity in both the core and accessory genome in a scalable way.
+We developed \textit{PanGraph} as a tool that can address this need, by being able to capture the structural and nucleotide diversity in both the core and accessory genome in a scalable way.
 
@@
 The efficient implementation of \textit{PanGraph} allows it to create pangenome graphs containing hundreds of isolates in a few hours on a 8-core machine.
-The size of pancontigs and the fraction of core sequece have good scaling properties with the number of isolates in the graph, indicating that \textit{PanGraph} is able to successfully capture pangenome properties.\\
+The size of pancontigs and the fraction of core sequence have good scaling properties with the number of isolates in the graph, indicating that \textit{PanGraph} is able to successfully capture pangenome properties.
+By construction \textit{PanGraph} operates in a gene-agnostic way, being only based on sequence homology, and is thus robust to gene annotation errors.\\
 One of the main limitations is the diversity of the input sequences.
-The default \textit{minimap2} kernel is able to correctly merge  sequences up to around 5\% divergence.
+With the default \textit{minimap2} alignment kernel, \textit{PanGraph} is able to correctly merge genomes with up to 5\% average divergence.
 Sequences with higher divergence (up to 10-15\%) can be merged using the \textit{mmseqs2} alignment kernel, and tuning the $\alpha, \beta$ energy parameters (cf.
-eq.~(\ref{eq:pseudo-energy})), trading off speed for sensitivity.\\
+eq.~(\ref{eq:pseudo-energy})), trading off speed for sensitivity.
+For more diverged data sets, homology detection tools that use protein sequences would likely be necessary.
+
 We also provide the ability to quickly marginalize big graphs on a subset of strains, obtaining simpler graphs that can more easily be explored.
@@
 
-We hope that these features will make \textit{PanGraph} a valuable tool, with the potential to spur new insights into microbial diversity and the processes by which bacteria adapt and change.
+The applications of \textit{PanGraph} are not limited to whole bacterial chromosomes.
+Plasmids and other Mobile Genetic Elements (MGEs) play a fundamental role in microbial evolution \cite{haudiquet2022selfish,frost2005mobile}, including in the spread of antibiotic resistance \cite{van2018spread}.
+MGEs often display a wide variety of sequence and structural diversity.
+Being able to capture and represent this diversity is key to understanding their evolution, particularly on epidemiologically relevant timescales where very few SNPs accumulate but large-scale structural changes are frequent \cite{Sheppard2016, Noll2018}.
+In this sense, the pangenome graph is a naturally powerful representation, and \textit{PanGraph} can be used to explore this diversity \cite{liam_preparation}.
+Some example applications on plasmids are available in \textit{PanGraph}'s documentation\footnote{See the tutorial section in \url{https://neherlab.github.io/pangraph}}.\\
+
+We hope that \textit{PanGraph} will prove a valuable tool, with the potential to spur new insights into microbial diversity and the processes by which bacteria adapt and change.
 
@@
 This study was funded by the University of Basel and the NSF.
+\Richard{Liam: do you need to add a funding statement?}
 

### 2022-11-18 supplementary.tex
Update on Overleaf.

@@
 
-\graphicspath{{../figs_suppl/}{./figs_suppl}}
-
 \begin{document}
@@
     \includegraphics[width=.9\textwidth]{figs_suppl/benchmark_suppl.pdf}
-    \caption{{\bf Computational performance of PanGraph algorightm on synthetic data, as a function of dataset size}.
+    \caption{{\bf Computational performance of PanGraph algorithm on synthetic data, as a function of dataset size}.
 We run PanGraph on artificially generated datasets consisting of a variable number of isolates with genomes of variable length (colorscale).
@@
 To link the mutation rate parameter $\mu$ with the average pairwise SNPs distance $\avg{d}$ between homologous segments in the population we evaluated the average pairwise distance for every pancontig in the pangenome graph with depth greater than one.
-For each graph we then perform the weighted average of these divergences, using as weight the length of the panconting.
+For each graph we then perform the weighted average of these divergences, using as weight the length of the pancontig.
 Finally, for every pair of parameters $(\mu,h)$ we average these numbers over the 25 different simulations.
@@
 For high values of $\mu$, the average pairwise divergence of pancontigs is not influenced by the rate of HGT $h$, but depends strongly on the choice of alignment kernel.
-With the \textit{asm20} option, \textit{minimap2} is able to correctly merge genomes with average pairwsise divergence $\avg{d} \sim 7\%$ in our simulations, while \textit{mmseqs2} reaches $\avg{d} \sim 10\%$ at the expense of higher processing time.
+With the \textit{asm20} option, \textit{minimap2} is able to correctly merge genomes with average pairwise divergence $\avg{d} \sim 7\%$ in our simulations, while \textit{mmseqs2} reaches $\avg{d} \sim 10\%$ at the expense of higher processing time.
 Notice that these values are lower than the threshold sensitivity of these aligners, which are expected to find matches up to respectively around 10\%, 20\% and 30\% sequence divergence.
@@
 We formulate this problem as an assignment problem, and solve it numerically using the Hungarian method \cite{kuhn1955hungarian}.
-For each isolate we call $b_i$ for $i=1,\ldots,I$ the panconting breakpoints on the real pangenome graph, and $b_j$ for $j=1,\ldots,J$ the ones on the reconstructed pangenome graph.\footnote{To avoid artifacts generated by the default threshold distance $d=100$ bp of \textit{PanGraph}, if any pair of breakpoints on the same graph sits at a distance smaller than this threshold, we remove one of them.} We define a cost matrix $D_{ij} = d(b_i,b_j)$ whose elements are distances on the genome of pairs of breakpoints from the two graphs.
+For each isolate we call $b_i$ for $i=1,\ldots,I$ the pancontig breakpoints on the real pangenome graph, and $b_j$ for $j=1,\ldots,J$ the ones on the reconstructed pangenome graph.\footnote{To avoid artifacts generated by the default threshold distance $d=100$ bp of \textit{PanGraph}, if any pair of breakpoints on the same graph sits at a distance smaller than this threshold, we remove one of them.} We define a cost matrix $D_{ij} = d(b_i,b_j)$ whose elements are distances on the genome of pairs of breakpoints from the two graphs.
 We then numerically solve the assignment problem on this matrix, and obtain a set of $K = \min\{ I,J \}$ pairings between breakpoints such that the distance between the pairs is minimal and all the breakpoints of the graph with minimum number of breakpoints have been paired.
@@
 At low divergence the position of most breakpoints is correctly estimated withing a few bps, while at high divergence most breakpoints can be misplaced by multiple hundreds of bps.
-The transition value for the divergence depends on the alignemnt kernel used, with \textit{mmseqs2} having the highest accuracy at the cost of higher computational time.}
+The transition value for the divergence depends on the alignment kernel used, with \textit{mmseqs2} having the highest accuracy at the cost of higher computational time.}
     \label{fig:benchmark-accuracy-suppl}
@@
 For each species we selected 50 different isolates (10 for PM) and built pangenome graphs using standard parameter values and the \textit{minimap2} kernel with \textit{asm20} option.
-For each species we then randomly picked 50 different pairs of isolates (all 45 for PM).\footnote{The list of selected accession numbers and selected pairs for each species is available on the repository: \url{https://github.com/neherlab/pangraph/blob/feat/finalize-paper/script/config/projection_strains.json}} For each pair we then built pairwise graphs using the same parameters, and compared them to the corresponding marginalized graph, obtained by projecting the full pangenome graph for the 50 isolates on the selected pair (cf.
+For each species we then randomly picked 50 different pairs of isolates (all 45 for PM).\footnote{The list of selected accession numbers and selected pairs for each species is available on the repository: \url{https://github.com/neherlab/pangraph/blob/master/script/config/projection_strains.json}} For each pair we then built pairwise graphs using the same parameters, and compared them to the corresponding marginalized graph, obtained by projecting the full pangenome graph for the 50 isolates on the selected pair (cf.
 Fig.~6A, main text).\\

### 2022-11-23 main.tex
fix: algorithm figure
Update on Overleaf.

@@
 Here, we focus on pangenome graphs of closely related bacterial genomes.
-\Richard{I feel the following is more confusing than helping.}
-As such, pangenome graphs can be thought of as an analogous construct to the well-known assembly graph, in which whole genomes are individual ``reads" of the underlying microbial structure \cite{myers2005fragment}.
 
 While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging.
-Colored generalizations of the de Bruijin graph-based assemblers have been successively used to build graphs from large sequence sets, although the underlying efficiency derives from a fixed k-mer size which prevents modelling long-range homology \cite{iqbal2012novo,muggli2017succinct}.
+Colored generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets, although the underlying efficiency derives from a fixed k-mer size which prevents modelling long-range homology \cite{iqbal2012novo,muggli2017succinct}.
 An orthogonal approach has been to formulate the inference of the pangenome graph as a multiple genome alignment.
@@
 
-Here we present PanGraph, a Julia \cite{bezanson2017julia} library and command line interface, designed to efficiently align large sets of closely related genomes into a pangenome graph on personal computers.
+Here we present \emph{PanGraph}, a Julia \cite{bezanson2017julia} library and command line interface, designed to efficiently align large sets of closely related genomes into a pangenome graph on personal computers.
 The resulting graph both compresses the input sequence set and succinctly captures the population diversity at multiple scales: from nucleotide mutations and indels to structural polymorphisms driven by inversions, rearrangements, and gene gain/loss.
@@
 \begin{figure}[htb]
-    \includegraphics[width=.45\textwidth]{figs/algorithm.png}
+    \includegraphics[width=.45\textwidth]{figs/algorithm.pdf}
     \caption{{\bf Overview of PanGraph algorithm}.
@@
 To verify that marginalized graphs are compatible with newly-built graphs, we built a pangenome graph for 50 randomly selected strains from the \textit{K.~pneumoniae} dataset.
-We then picked 50 random pairs of strains from the same set and built two graphs for each pair: the graph obtained by marginalizing the complete graph on the pair ( (i.e.
+We then picked 50 random pairs of strains from the same set and built two graphs for each pair: the graph obtained by marginalizing the complete graph on the pair (i.e.
 marginalized graph in Fig.~\ref{fig:marginalization}A top), and the graph built directly from the pair (i.e.
@@
 This study was funded by the University of Basel and the NSF.
-\Richard{Liam: do you need to add a funding statement?}
+L.P.S. is a Sir Henry Wellcome Postdoctoral Fellow funded by Wellcome (Grant 220422/Z/20/Z).
 

### 2023-02-14 main.tex
submitted-version

@@
-\documentclass[aps,rmp,reprint,superscriptaddress,notitlepage,10pt]{revtex4-1}
+\documentclass[aps,rmp,preprint,superscriptaddress,10pt,linenumbers]{revtex4-1}
+%\documentclass[aps,rmp,reprint,superscriptaddress,10pt]{revtex4-1}
 \usepackage[utf8x]{inputenc}
@@
 \author{Richard A.~Neher}
+\email{corresponding author: richard.neher@unibas.ch}
 \affiliation{Swiss Institute of Bioinformatics, Basel, Switzerland \looseness=-5}
@@
 
+\section*{Conflict of Interest}
+The authors declare that they are not aware of relevant conflicts of interests.
+
 \bibliography{cite}{}

### 2023-02-14 supplementary.tex
submitted-version

@@
 Each individual in the population has a circular genome of initial size $L$.
-At each generation single-nucleotide mutations occur at a rate $\mu$ per position per generation.
+At each generation single-nucleotide mutations occur at a rate $\mu$ per position.
 Deletions can occur at a rate $d$ per genome per generation.
@@
 Notice that these values are lower than the threshold sensitivity of these aligners, which are expected to find matches up to respectively around 10\%, 20\% and 30\% sequence divergence.
-This is due to the fact that homologous sequences in our simulations are generated by an evolutionary process, they can have a wide distribution of diversities and hierarchical population structure.
+This is due to the fact that homologous sequences in our simulations are generated by an evolutionary process, and they can have a wide distribution of diversities and hierarchical population structure.
 Inability to merge sequences above a threshold divergence in a single panconting is expected to result in multiple pancontigs containing sub-clades with higher similarity, and average diversity that is appreciably lower than the aligner threshold.\\
@@
 Conversely, the PM dataset contains highly diverged isolates, whose divergence go beyond the capabilities of both alignment kernels ($\dcore \sim 27 \%$, cf.
-Table~\ref{table:panx-dataset}).
+Table~I in the main text).
 Only \textit{mmseqs2} manages to partially merge some of the least diverged core sequences.
 This results in a small core genome fraction and poor sequence compression.
-In its current state \textit{PanGraph} cannot meaningfully merge genome with such high divergence into a pangenome graph.
+In its current state \textit{PanGraph} cannot meaningfully merge genomes with such high divergence into a pangenome graph.
 
@@
 In all species except HP and PM we find very good agreement between pairwise and marginalized graphs: segments on which the graphs \textit{disagree} represent only a very minor fraction of the genome (much smaller than the fraction of shared or private blocks), and have size compatible with the default block length threshold of \textit{PanGraph} (100 bp).
-As described in the previous section, the divergence of the HP dataset sit on the limit of what the \textit{minimap2} alignment kernel can successfully handle.
+As described in the previous section, the divergence of the HP dataset sits on the limit of what the \textit{minimap2} alignment kernel can successfully handle.
 In this case the results of merging different graphs together can depend on the order of merging, and pairwise and marginalized graphs can potentially show some disagreement.

### 2023-02-14 reviewer_response.tex
feat: add reviewer response scaffold

\documentclass{article}

\usepackage{hyperref}

\newcommand{\review}[1]{\textbf{#1}}
\newcommand{\response}[1]{\textbf{Response:} \textit{#1}}

\title{Response to Reviewers}
\begin{document}
\maketitle

\section*{Reviewer 1}

\review{R1} Bacterial evolution is driven both by local (vertical) mutation and larger scale (horizontal) genome alternations resulting in gain, loss, inversion or reassortment of genetic material. A single linear reference cannot capture this larger scale diversity, and existing graph reference structures either depend on coloured de Bruijn graphs (losing long range homology) or simplify the complex topologies by partitioning the pangenome and considering genes/loci in isolation. PanGraph provides a library and data structure to create a pangenome reference graph for closely related bacteria based on a progressive multiple sequence alignment approach.

It is not easy to handle this sort of diversity efficiently and I was excited to read about the neat tricks employed to allow for a progressive alignment approach within reasonable resources, including parallelisation.

This paper was clearly organised and a real pleasure to read. The niche was clearly described and demonstrated. The tests were well thought through. A lot of care had been taken to demonstrate the effects of different levels of bacterial diversity on the resulting graphs. The deliberate inclusion of P. marinus provided a clear informative demonstration of the limit of the software capabilities.

The software was easy to install and the documentation available online is impressive.

There was no comparison with other tools - would be appropriate if e.g. downstream mapping performance were evaluated but that is outside the scope here.\\

On reading I had a few queries, which could perhaps be clarified further in the text.

\begin{enumerate}
    \item A pancontig is defined as a multiple sequence (sub)alignment. In line 54 you say "Pairwise graph alignment is performed by an all-to-all alignment of the pancontigs" which I interpreted to mean all sequences in the alignment vs all sequences in another alignment. You do later add that "To align two graphs, the consensus sequences of all pancontigs in both graphs are searched for homologies and aligned". Adding the word consensus in the earlier sentence might avoid confusion.

    \item Figure 4C: The main text talks about how to interpret this, as ~1/N for TB and a bit less compression for E. coli and K. pneumoniae where the core is a smaller fraction, but this is not so easy to interpret from the figure. Would it be possible to rescale this plot to orders of magnitude of N? Or in another way aid this reading?

    \item Figure 5 has 2 typos: pacontigs and panconting.

    \item Throughout the text E. Coli and K. Pneumoniae etc capitalization.

    \item Line 213: Core pancontigs have higher than average size - is this what you would expect given that if they are core genes then they will be present in more/all of the samples so there could be more diversity due to being more sampled?

    \item Figure 6B: could you explore another way to project this data so it is not squished at the ends? I can't read off the fraction agree (shared+private) or disagree which I would like to (although at a glance this is exactly the result you want to see, so happy for it to stand if it is the best projection).

    \item Have you tried running PanGraph on more than 300 references? I'd love to see another order of magnitude (not necessary for the paper, just curiosity)

    \item Discussion: Do you know of/have you tried any mapping approaches which could take this reference as an input. Could you touch on the downstream applications of this graph structure? I would like to see discussion of how the graph can be used downstream - e.g. touching on what software can be used to map/align to such a data structure? Or how it can be used to analysis genome structure.

    \item Supplementary Information: simulation - could you comment on how/if homologous recombination is included in this model e.g. within a gene.

    \item Supplementary Information C: You comment that the use of null energy parameters results on average in "more and shorter pancontigs" but also that they result in more diverged sequences being merged into the same pancontig. Why is a more diverse pancontig more likely to become fragmented?
\end{enumerate}

\section*{Reviewer 2}

\review{R2} In this paper, Noll et al set out to make a novel contribution to the pangenomics field. As it stands prior to this work, pangenomics splits into 3 different approaches

\begin{enumerate}
    \item People working in human genomics are working to develop a graph human reference. The challenge there is one of scale - large genomes with comparatively small pan-genomes (in the classical/bacterial sense) except for some devilish mess at the centromeres. This is essentially a subfield of its own focussed on the needs of human and clinical genetics.

    \item People working in bacterial genomics who do reference-based SNP calling plus gene-presense/absence, or more recently building gene order graphs (panaroo) and even gene calling /detection on graphs (ggcaller).

    \item People breaking the pan-genome up into a collection of orthologous blocks and then building a collection of pan-genome graphs (pandora), at the cost of losing order information.
\end{enumerate}

Noll et al set out to build a representation which combines a global pan-genome with local MSAs, and to do so in a way which scales well, and which provides easy to use visualisations. It is an ambitious aim, and in my opinion this is great work. They have developed something noone else has (see enumeration above), and the paper does a good job of describing and then evaluating their work - they are to be congratulated. This tool is sure to be extremely useful, and very valuable to the community. I have no serious reservations about the quality or calibre of the work, i admire it greatly.\\

Nevertheless, there are a number of clarifications which need to be addressed before the paper is suitable for publication.\\

\subsection*{Major Issues}

\begin{enumerate}
    \item Line 46: when you say "PanGraph transforms an arbitrary set of genomes into a graph that simultaneously compresses the collection of sequences and exhaustively summarizes both the structural and nucleotide-level polymorphisms." - I am a bit unclear on how the nucleotide level variation is summarised. The paper is a bit hand-wavey about this. When i dug around on github , i found this helpful page \url{https://neherlab.github.io/pangraph/tutorials/tutorial_2/} Which answers and raises questions. So, my question: for each pancontig (which i think is the same as a block), i think what you do is infer a consensus (what is this? majority base at each position but somehow also choose an allele when there is an indel?), and then you describe the other alleles in the MSA as edits from this? You surely do not do all pairwise descriptions of SNPS, right? On that same webpage you have a section on "block consensus to node alignment" - how do you choose the order of alleles here? This is all valuable information and it would be great to have it in the supplementary or main methods somewhere.

    \item Line 61: I agree and understand how each sequence can be sketched in linear time. I don't follow how you calculate all pairwise intersections through sorting the list of minimizers, nor do i see how this ends up being log-linear. For each genome, you can sort it's minimizers in nlog(n) (ie log linear), where n ~genome length. How do you get all pairwise distances from that without another multiplier of n?

    \item The section on iterative graph alignment (lines 69-77) is in one sense quite clear and intuitive, but there is a lot of detail missing which leaves the reader a bit confused when they try to think about details. I would appreciate some detail being added, perhaps in the supplementary methods. For example:
          \begin{itemize}
              \item at the start you just have two genomes, and you scan them for orthologous regions, and you decide which to merge. As i understand it, you merge them in order of pseudoenergy ranking, which makes sense (what do you do for precise duplicates, where a region on genome 1 is a precise match for 2 regions on genome 2? random choice?)
              \item as you move on, you are aligning graphs, and you really skim over the issue of how you do that. How does minimap or mmseqs allow you to look for homology between two graphs? The only thing i can imagine you are doing is only looking for homology between the genomes in one graph, and the genomes in the second graph. I would really appreciate it if you could clarify this and even ideally add some example figures in the supplement showing with some toy genomes how you align two graphs.
          \end{itemize}

    \item Line 92-94, same section. You give two statements:
          "Only mergers whose alignment has negative pseudo-energy are performed"
          and
          "no mergers will be performed if the Hamming distance exceeds $1/\beta$."
          These are separate criteria that prevent mergers, right? So any potential merger with negative pseudoenergy that does not fail these criteria will happen, with the caveat that you perform mergers in order?

          One interesting thing here is that you are effectively merging things in reverse evolutionary order, as you have lower energy for longer more similar things which presumably diverged more recently. Not sure if that tells us anything.

    \item I very much appreciated that you measure the accuracy of pangraph via this displacement idea. I started writing a kindly worded criticism that I thought your method was flawed - surely the displacement depends not only on the order of doing merges, and decisions made about what to do when there are two equally good places to merge with. I guess i was saying that slightly different evolutionary events could give the same genomes, and thereby justify the errors. However i very much like the approach you detail in the methods, numerically solving the assignment problem. So, this started as a major issue, but in the end is a compliment.

    \item Line 183 - I really do not understand what is going on with your alpha,beta being set to zero. Are you saying if you stop penalising merges for having too many SNPs, then you will be able to merge more, and this will stop you from havign a 10\% divergence limit? It will also mean you make more false merges, right?

    \item Why does the L50 peak go up for null-scores for mmseqs for Mtb, Pm, Kpn and Ec (Figure 4E, purple with lines) but down for helicobacter?

    \item Just to make sure i understand - is it possible for sequence to be "left out" in between pancontigs?

    \item About data accessibilityW
          \begin{enumerate}
              \item i don't see how to get the list of accessions for section IIB? Saying they were used in the Ding et al paper is not quite enough for me to be sure which ones you used.
              \item for the graph marginalisation section, could the accessions of the strains used be in a supplementary file as well as on github? Github repos are great, but they can be deleted.
          \end{enumerate}
\end{enumerate}

\subsection*{Minor Issues}

\begin{enumerate}
    \item line 13 - it would be good to give a reference for the statement that evem closelt related genomes can differ widely in gene content.
          I don't know what the first paper to show this was, the earliest i am aware of is from Eduardo Rocha's team, which annpoyingly i can't locate right now, but Figure 2 here shows it \url{https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1008866} - you can be very close in patristic distance and have very different gene repertoire.
          Feel free to replace with a better reference, this si just meant to be helpful.

    \item Why on earth is the "get an MSA of a block" option in pangraph called "polish"?

    \item line 48: "Pancontigs are connected by an edge if they are syntenic on at least one input sequence;" - i think this is a bit of an abuse of modern language. The word syntenic is generally used to mean two genes/whatever have order conserved on multiple chromosomes (although some googling tells me that in classical genetics it meant things were on the same chromosome). Anyway, my understanding is pancontigs are connected by an edge if one of the input genomes is spelled out by going from one to the other. So it's about adjacency rather than synteny.

    \item Line 131: was N really=100? on the plot in figure 2, you have n.isolates going up to 1000

    \item Figure 5 legend, lower case C in "coli" of "E. coli". Also italics.
\end{enumerate}

It is a bit depressing and unhelpful to structure a review only around problems and quibbles, so i wanted also to raise a number of
Things I liked:

\begin{enumerate}
    \item The paper is concise and well written, i really enjoyed reading it
    \item I liked the structure of the algorithms section, again i congratulate the authors, this is so much clearer than most papers. eg i was particularly happy to see the Guide Tree section start with your design constraints/aims.
    \item Fantastic choice of species in your empirical dataset - P. marinus diversity is insane.
    \item Line 215 - this is a remarkable achievement: "Adding more strains does not generate excessive fragmentation of these pancontigs, and their size remains of several kbps even as hundreds of strains are added"
    \item Congratulations
\end{enumerate}

\section*{Reviewer 3}

\review{R3} The article "PanGraph: scalable bacterial pan-genome graph construction" reports a new method for constructing a Pan-Genome Graph (PGG) in an efficient manner with an open and extractable grah structure. Unfortunately, the authors appear to either be unaware of or downplay the existence of previous methods for the same purpose.\textbf{ The authors need to survey the existing literature and compare and contrast with their method to show the utility of their approach.}\\

For instance the authors state: "Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated "bags of genes", and thus are not suited for this new era.". There are many pan-genome analysis tools which treat the pan-genome as a graph and generate a PGG. There are two primary differences in these existing tools: those which generate a graph at the nucleotide level with nodes representing basepairs (sometimes nonbranching runs of nodes are collapsed) and those where nodes represent larger features such as genes.\\
PanGraph appears to wed these approaches to some extent in that the alignments are done at the nucleotide level and the features are extracted as a characteristic of the alignments.\\

An obvious lack in PanGraph is an attempt to map existing annotations back onto the generated PGG which is a necessity for usefulness to biologists. The mechanism for this is obvious but providing a useful user interface might be tedious.\\

This statement: "While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging. Colored generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets, although the underlying efficiency derives from a fixed k-mer size which prevents modelling long-range homology (Iqbal et al., 2012; Muggli et al., 2017)." appears to be incomplete at best. De Bruijn graph-based assemblers routinely collapse nonbranching portions of the graph into longer sequences than the k-mer size used for construction.\\
One early approach that does not explicitly generate a pan-genome graph but deals with some of the rearrangement issues cited by the authors is:
\begin{itemize}
    \item Angiuoli SV, Dunning Hotopp JC, Salzberg SL, Tettelin H. Improving pan-genome annotation using whole genome multiple alignment. BMC Bioinformatics. 2011 Jun 30;12:272. doi: 10.1186/1471-2105-12-272. PMID: 21718539; PMCID: PMC3142524.
    \item Angiuoli SV, Salzberg SL. Mugsy: fast multiple alignment of closely related whole genomes. Bioinformatics. 2011 Feb 1;27(3):334-42. doi: 10.1093/bioinformatics/btq665. Epub 2010 Dec 9. PMID: 21148543; PMCID: PMC3031037.
\end{itemize}
While the underlying approach is quite different, a very similar PGG to PanGraph is produced by a different method with similar nodes represented by multiple sequence alignments and edges with the same meaning:
\begin{itemize}
    \item Sutton G, Fogel GB, Abramson B, Brinkac L, Michael T, Liu ES, Thomas S. A pan-genome method to determine core regions of the Bacillus subtilis and Escherichia coli genomes. F1000Res. 2021 Apr 13;10:286. doi: 10.12688/f1000research.51873.2. PMID: 34113437; PMCID: PMC8156514.
    \item Chan AP, Sutton G, DePew J, Krishnakumar R, Choi Y, Huang XZ, Beck E, Harkins DM, Kim M, Lesho EP, Nikolich MP, Fouts DE. A novel method of consensus pan-chromosome assembly and large-scale comparative analysis reveal the highly flexible pan-genome of Acinetobacter baumannii. Genome Biol. 2015 Jul 21;16(1):143. doi: 10.1186/s13059-015-0701-6. PMID: 26195261; PMCID: PMC4507327.
\end{itemize}

I believe this statement: "It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order." is the essential step in the process and perhaps deserves more focus. It captures two important issues: \textbf{aligning orthologs not paralogs and keeping the fracturing of the PGG nodes to a minimum}. The pseudo-energy captures the alignment length (longer is better for orthologs versus paralogs) but does not appear to capture the surrounding context of the iterative PGGs. For initial alignments of single genomes to single genomes this doesn't matter for the most part because there is a single pancontig for each each genome and PGG context is mostly captured in the alignment. I am not so sure this is true as the PGG nodes fracture and become much smaller. When choosing between competing mergers of paralogs/orthologs the context of the PGG and other possible mergers is important if the pancontigs have been reduced so that they contain only repetitive sequence without unique flaning sequence. I think this is a very important point as distinguishing orthologs from paralogs is essential for correct PGG construction and the authors need to explicitly address it. This could be measured as part of the simulated data where identical paralog sequences of various lengths and copy number could be modeled.\\

In the pseudo-energy discussion a further parameter is mentioned: "In addition, there is parameter that controls the minimal size of contigs produced by the algorithm.". The value chosen for this parameter should be explicitly stated along with a rationale for the choice. It is also not clear if any of the splitting pancontigs are less than this value if the entire merger is dropped or just massaged to not have the too small new pancontig.\\

How are pancontig consensus sequences determined and updated? I could not find any details about this. This is not a trivial operation. My reading of the article seems to imply that a complete multiple sequence alignment is not done for each merger. Does the consensus sequence degenerate iteratively with mergers? For the simulated data how accurate are the final pancontig consensus sequences?\\

The statement: "Upon completion, transitive 108 edges within the graph, i.e., edges along which all path are co-linear, are removed by merging adjacent pancontigs." needs clarification. Why would these pancontigs have been split in the first place? Is there any necessary special issues with producing the merged multiple sequence alignment?\\

It is not clear to me what Table 1 shows. Is this published data from PanX or data from PanGraph? Either way the value of 0.7 Mbp for the E. coli core genome length seems implausibly low for high quality complete genomes. The authors should justify this. The authors should also include analysis on the calculated core genome length versus actual core genome length for the simulated data. The authors might want to consider some form of post processing of the PGG to collapse what have been called bubbles in the graph if alignment of the pancontigs on different edges of the bubble warrant it.\\

Overall the paper is well written. I like the basic methodology and the attention to using robust efficient existing algorithms and combining them to keep efficiency in mind. I am very concerned about the potential fracturing of the PGG where unfortunately the devil is in the details and am not convinced that the fracturing will not prevent some limit on the number of genomes which can be used for a pangenome.\\

An issue I have struggled with in my own work is how to represent an orthologous gene cluster (node/vertex) within the PGG when the gene sequence in some genomes has an insertion event such as an IS element within the gene. I assume for PanGraph that this would necessitate splitting the node into multiple pancontigs. How is this sustainable for species with large numbers of active IS elements? This could also be investigated for PanGrah using suitable simulated data.

\end{document}

### 2023-02-14 reviewer_response.tex
feat: review comments

@@
 \usepackage{hyperref}
+\usepackage{xcolor}
 
@@
 \newcommand{\response}[1]{\textbf{Response:} \textit{#1}}
+\newcommand{\Marco}[1]{{\color{gray}Marco: #1}}
 
@@
 Adding the word consensus in the earlier sentence might avoid confusion.
+          \Marco{changed to ``Pairwise graph alignment is performed by an all-to-all alignment of the \emph{pancontigs} consensus sequences between both graphs, and the order of pairwise alignments is determined by a guide tree." Too compressed?
+I could also break in two sentences and expand a bit more.}
 
@@
 Would it be possible to rescale this plot to orders of magnitude of N?
-Or in another way aid this reading?
+Or in another way aid this reading?\\
+          \Marco{maybe try visualizing it as deviation from "one strain" compression (1/N)?
+But the "maximal" compression depends not only on the sample size N but also on the gene frequency distribution.
+Another possible comparison might be to pangraph's estimation of the pangenome size, but again we might have different effects coming into play.}
 
-    \item Figure 5 has 2 typos: pacontigs and panconting.
+    \item Figure 5 has 2 typos: pacontigs and panconting.\\
+          \Marco{fixed.}
 
-    \item Throughout the text E. Coli and K. Pneumoniae etc capitalization.
+    \item Throughout the text E. Coli and K. Pneumoniae etc capitalization.\\
+          \Marco{fixed.}
 
-    \item Line 213: Core pancontigs have higher than average size - is this what you would expect given that if they are core genes then they will be present in more/all of the samples so there could be more diversity due to being more sampled?
+    \item Line 213: Core pancontigs have higher than average size - is this what you would expect given that if they are core genes then they will be present in more/all of the samples so there could be more diversity due to being more sampled?\\
+          \Marco{here the pancontig size is the consensus sequence length.
+I would say that the fact that they have higher-than-average size is mostly due to synteny.
+Long blocks (several kbps) contain core genes that always appear in the same order, without any accessory sequence ever occurring in these regions.
+As opposed to accessory regions with a lot of structural diversity, that generates short fragments.
+But here the main message is that core blocks are well-behaved (close to N50 line in the figure), as opposed to fine-grained structural diversity, part of which contains artifacts generated by the block size limit (many short 100bp blocks).
+Should we show block frequency distributions in the supplementary?}
 
@@
 I can't read off the fraction agree (shared+private) or disagree which I would like to (although at a glance this is exactly the result you want to see, so happy for it to stand if it is the best projection).
+          \Marco{maybe add mean and standard deviation as numbers on the side?}
 
     \item Have you tried running PanGraph on more than 300 references? I'd love to see another order of magnitude (not necessary for the paper, just curiosity)
+          \Marco{not really, I never tried on ~500 or ~1000 strains. I could try but would not consider this a priority.}
 
@@
 touching on what software can be used to map/align to such a data structure?
-Or how it can be used to analysis genome structure.
+Or how it can be used to analysis genome structure.\\
+          \Marco{other than explicitly exporting block sequences and mapping to those, we do not really using the pangenome graph as a reference against which to map new strains...
+But we could expand the discussion with more ideas on applications to study pangenome organization and evolution.
+For software for further analysis and exploration I have \href{https://github.com/mmolari/pypangraph}{this package} to explore and interact with graphs in python, but I don't think it's really relevant.}
 
-    \item Supplementary Information: simulation - could you comment on how/if homologous recombination is included in this model e.g. within a gene.
+    \item Supplementary Information: simulation - could you comment on how/if homologous recombination is included in this model e.g. within a gene.\\
+          \Marco{homologous recombination is not implemented directly in the model. Only sequence transfer between strains. Genes are not explicitly modeled in the simulation.}
 
     \item Supplementary Information C: You comment that the use of null energy parameters results on average in "more and shorter pancontigs" but also that they result in more diverged sequences being merged into the same pancontig.
-Why is a more diverse pancontig more likely to become fragmented?
+Why is a more diverse pancontig more likely to become fragmented?\\
+          \Marco{I think that basically the reason is that merging more distant paralogs/orthologs can fragment more the graph. Maybe we could explain this more explicitly in the text.}
 \end{enumerate}

### 2023-02-14 main.tex
feat: review comments

@@
 \emph{PanGraph's} overarching strategy is to approximate multiple-genome alignment by iterative pairwise alignment of graphs of subsets of sequences, in the spirit of progressive alignment tools \cite{feng1987progressive,darling2010progressivemauve,armstrong2020progressive}.
-Pairwise graph alignment is performed by an all-to-all alignment of the \emph{pancontigs} between both graphs and the order of pairwise alignments is determined by a guide tree.
+Pairwise graph alignment is performed by an all-to-all alignment of the \emph{pancontigs} consensus sequences between both graphs, and the order of pairwise alignments is determined by a guide tree.
 
@@
 
-% The \textit{E.~Coli} (EC) dataset is the one with the highest number of isolates (307 chromosomes) and the largest pangenome.
-The species with the highest core genome divergence are \textit{H.~Pylori} (HP) and \textit{P.
-Marinus} (PM).
+% The \textit{E.~coli} (EC) dataset is the one with the highest number of isolates (307 chromosomes) and the largest pangenome.
+The species with the highest core genome divergence are \textit{H.~pylori} (HP) and \textit{P.~marinus} (PM).
 For the former, the divergce sits at the limit of what the \textit{minimap2} alignment kernel is able to accurately merge (cf.
@@
 Moreover the use of null energy parameters ($\alpha = \beta = 0$) is expected to remove the threshold divergence of 10\% associated with the standard values of parameters ($\alpha=100$, $\beta=10$) at the cost of more fragmented pangenome graphs.\\
-\textit{PanGraph} can build pangenome graphs comprising hundreds of isolates in a few hours (5h for 307 E.~Coli isolates with \textit{minimap2} kernel, cf. Fig.~\ref{fig:panx-benchmark}A).
+\textit{PanGraph} can build pangenome graphs comprising hundreds of isolates in a few hours (5h for 307 \textit{E.~coli} isolates with \textit{minimap2} kernel, cf. Fig.~\ref{fig:panx-benchmark}A).
 Use of \textit{mmseqs2} consistently requires longer computational time, but provides higher sensitivity when merging highly diverged sequences.
@@
     \caption{{\bf Pangenome graph properties vs. dataset size}.
-        We build pangenome graphs with increasing number of isolates from the E.~Coli dataset and measure the scaling of different properties of the graphs.
+        We build pangenome graphs with increasing number of isolates from the \textit{E.~coli} dataset and measure the scaling of different properties of the graphs.
 Graphs were built using \textit{minimap2} alignment kernel with \textit{asm20} option.
@@
 of pancontigs in the graph.
-We count the total number of pacontigs (blue), the number of core pancontigs (orange) and the minimum number of pancontigs that contain more than 50\% of the pangenome (L50, black).
-        \textbf{B}: average size of pancontigs (blue), of only core pancontings (orange), and size of the smallest panconting in the minimal set that spans 50\% of the pangenome (N50, black).
+We count the total number of pancontigs (blue), the number of core pancontigs (orange) and the minimum number of pancontigs that contain more than 50\% of the pangenome (L50, black).
+        \textbf{B}: average size of pancontigs (blue), of only core pancontigs (orange), and size of the smallest pancontig in the minimal set that spans 50\% of the pangenome (N50, black).
         \textbf{C}: cumulative size of all genomes in the pangenome graph (gray), total pangenome size (blue) and size of core pangenome (orange).
@@
     \caption{{\bf Test of graph marginalization}.
-        \textbf{A}: we built a pangenome graph from 50 randomly chosen strains from the \textit{K.
-pneumoniae} dataset.
+        \textbf{A}: we built a pangenome graph from 50 randomly chosen strains from the \textit{K.~pneumoniae} dataset.
 We then randomly picked 50 pairs of strains.

### 2023-02-15 reviewer_response.tex
feat: addedd comments for rev 2

@@
-\documentclass{article}
-
+\documentclass[aps,rmp,onecolumn]{revtex4-1}
+\usepackage{geometry}
 \usepackage{hyperref}
@@
 \newcommand{\review}[1]{\textbf{#1}}
-\newcommand{\response}[1]{\textbf{Response:} \textit{#1}}
 \newcommand{\Marco}[1]{{\color{gray}Marco: #1}}
+
+% \definecolor{response}{rgb}{0.1, 0.1, 0.1}
+% \newcommand{\reviewer}[2]{\textbf{#1:} #2\vskip 5mm}
+% \newcommand{\response}[1]{{\it {\color{response}\textbf{Response:} #1}}\vskip 5mm}
+% \newcommand{\responsedraft}[1]{{\it {\color{purple}\textbf{ResponseDraft:} #1}}\vskip 5mm}
 
@@
 
-\section*{Reviewer 1}
+\section*{Response to reviewers}
+
+\subsection*{Reviewer 1}
 
@@
 But we could expand the discussion with more ideas on applications to study pangenome organization and evolution.
-For software for further analysis and exploration I have \href{https://github.com/mmolari/pypangraph}{this package} to explore and interact with graphs in python, but I don't think it's really relevant.}
+For software for further analysis and exploration I have \href{https://github.com/mmolari/pypangraph}{this package} to explore and interact with graphs in python, but not sure it's worth mentioning.}
 
@@
 
-\section*{Reviewer 2}
+\subsection*{Reviewer 2}
 
@@
 
-\subsection*{Major Issues}
+\subsubsection*{Major Issues}
 
@@
 On that same webpage you have a section on "block consensus to node alignment" - how do you choose the order of alleles here?
-This is all valuable information and it would be great to have it in the supplementary or main methods somewhere.
+This is all valuable information and it would be great to have it in the supplementary or main methods somewhere.\\
+            \Marco{this point was also raised by R3.
+Mutations are always represented as variation on the consensus.
+The way consensus is propagated during graph merging might require more explanation.
+Maybe we could add a section in the supplementary with a somewhat more detailed description of the algorithmic steps we perform when merging two graphs?
+In the code \href{https://github.com/neherlab/pangraph/blob/e1c11012a3d19975d4b1dc61b1af3eb3949f7cd2/src/block.jl\#L1274}{this part} takes care of merging alignments between two blocks with only small (i.e.
+below cutoff) insertions and deletions.
+In short what we do is stitch the two alignments together using the cigar string as a guide, and re-compute the consensus on the newly-obtained alignment.
+This ensures that the consensus agrees with the majority of sequences in the alignment, but does not solve the full multiple sequence alignment problem.
+However we also provide the ``polish'' step for this, that can be executed after graph creation.
+This will take care of refining the block alignments.}
 
@@
 For each genome, you can sort it's minimizers in nlog(n) (ie log linear), where n ~genome length.
-How do you get all pairwise distances from that without another multiplier of n?
+How do you get all pairwise distances from that without another multiplier of n?\\
+            \Marco{in the text we write that ``the pairwise distance matrix is estimated in a time that increases log-linearly in the \textbf{total length} of sequence that is analyzed.'' This might generate some confusion since the scaling depends both on the number of genomes $N$ and their length $l$.
+The algorithmic time for all the pairwise distances should scale as $\sim N^2 l \log l$.
+We should be in a regime where $N$ is not a limiting factor, it's just a multiplicative constant, and thus I think our statement is fundamentally correct.
+But we can make it more precise.
+This scaling is at any rate not a limiting factor in pangraph.}
 
@@
 As i understand it, you merge them in order of pseudoenergy ranking, which makes sense (what do you do for precise duplicates, where a region on genome 1 is a precise match for 2 regions on genome 2?
-random choice?)
+random choice?)\\
+                        \Marco{the ranking would be the same and we perform only one of the two merges in the first phase, when comparing the two graphs.
+After this we align the graph on itself and the second merge should be performed at this point.
+This comment also suggests we could add a section in the supplementary with a more detailed description of the algorithm?}
                   \item as you move on, you are aligning graphs, and you really skim over the issue of how you do that.
@@
 The only thing i can imagine you are doing is only looking for homology between the genomes in one graph, and the genomes in the second graph.
-I would really appreciate it if you could clarify this and even ideally add some example figures in the supplement showing with some toy genomes how you align two graphs.
+I would really appreciate it if you could clarify this and even ideally add some example figures in the supplement showing with some toy genomes how you align two graphs.\\
+                        \Marco{another comment pointing at the fact that we could add a more detailed explanation about graph merging on the supplementary.
+What we do is using the aligner using as reference and queries the consensus sequences of each block in the two different graphs, and create a first version of the merged graph.
+We then repeat this procedure in an all-vs-all comparison of the consensus sequences of the blocks, and proceed with further merges.
+This is necessary because when two alignments overlap only one of the two is performed in the first alignment round, but the second will be performed in further rounds.}
             \end{itemize}
@@
             These are separate criteria that prevent mergers, right?
-So any potential merger with negative pseudoenergy that does not fail these criteria will happen, with the caveat that you perform mergers in order?
-
+So any potential merger with negative pseudoenergy that does not fail these criteria will happen, with the caveat that you perform mergers in order?\\
+            \Marco{the second is actually a consequence of the first, but except for that it is correct.}\\
             One interesting thing here is that you are effectively merging things in reverse evolutionary order, as you have lower energy for longer more similar things which presumably diverged more recently.
-Not sure if that tells us anything.
+Not sure if that tells us anything.\\
+            \Marco{indeed, the parameter $\beta$ could control how far in the past one wants to accept merges. Should we mention this explicilty in the text?}
 
@@
 Are you saying if you stop penalising merges for having too many SNPs, then you will be able to merge more, and this will stop you from havign a 10\% divergence limit?
-It will also mean you make more false merges, right?
+It will also mean you make more false merges, right?\\
+            \Marco{setting $\beta$ to zero will indeed remove the 10\% divergence limit set by the pseudo-energy.
+There is another limit that depends on the aligner sensitivity (20 to 30\%, depending on the aligner).
+This limit remains, and in this sense no ``false merges'' are performed?
+Not sure I interpret this correctly.}
 
-    \item Why does the L50 peak go up for null-scores for mmseqs for Mtb, Pm, Kpn and Ec (Figure 4E, purple with lines) but down for helicobacter?
+      \item Why does the L50 peak go up for null-scores for mmseqs for Mtb, Pm, Kpn and Ec (Figure 4E, purple with lines) but down for helicobacter?\\
+            \Marco{the effect of merging more blocks on L50 is not trivial.
+It depends on whether the merging occurs between long or short blocks, and how much fragmentation this causes.
+HP is a special case in that its diverge is on the threshold of what minimap2 can handle.
+Using mmseqs2 results in a pangenome that is half the size compared to standard parameters (fig.
+4G).
+This is an impressive amount of additional compression.
+At the same time the fraction of core pangenome more than doubles (4H).
+The total number of blocks decreases to roughly half (4D) while N50 increases (4F).
+For me this points to the fact that in the case of HP merging will split long blocks and create more fragmentation.
+For the reviewer I could produce the block frequency and lenght distribution for the two cases and comparing them would give an answer.
+I could do the same for EC or KP for comparison.}
 
-    \item Just to make sure i understand - is it possible for sequence to be "left out" in between pancontigs?
+      \item Just to make sure i understand - is it possible for sequence to be "left out" in between pancontigs?\\
+            \Marco{no, every part of the genome must be contained in a pancontig. It can be left out from the consensus if it is considered an insertion.}
 
-    \item About data accessibilityW
+      \item About data accessibility
             \begin{enumerate}
                   \item i don't see how to get the list of accessions for section IIB? Saying they were used in the Ding et al paper is not quite enough for me to be sure which ones you used.
-              \item for the graph marginalisation section, could the accessions of the strains used be in a supplementary file as well as on github? Github repos are great, but they can be deleted.
+                  \item for the graph marginalisation section, could the accessions of the strains used be in a supplementary file as well as on github?
+Github repos are great, but they can be deleted.
             \end{enumerate}
+            \Marco{on the repo we have \href{https://github.com/neherlab/pangraph/blob/master/script/config/accnums.json}{this file} for the list of all accession numbers we use, and \href{https://github.com/neherlab/pangraph/blob/master/script/config/projection_strains.json}{this file} for the marginalization pairs.
+I could create some excel spreadsheet with accession numbers that we could add to the submission?}
 \end{enumerate}
 
-\subsection*{Minor Issues}
+\subsubsection*{Minor Issues}
 
 \begin{enumerate}
-    \item line 13 - it would be good to give a reference for the statement that evem closelt related genomes can differ widely in gene content.
-          I don't know what the first paper to show this was, the earliest i am aware of is from Eduardo Rocha's team, which annpoyingly i can't locate right now, but Figure 2 here shows it \url{https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1008866} - you can be very close in patristic distance and have very different gene repertoire.
-          Feel free to replace with a better reference, this si just meant to be helpful.
+      \item line 13 - it would be good to give a reference for the statement that even closelt related genomes can differ widely in gene content.
+            I don't know what the first paper to show this was, the earliest i am aware of is from Eduardo Rocha's team, which annpoyingly i can't locate right now, but Figure 2 here shows it \url{https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1008866} \cite{touchon2020phylogenetic} - you can be very close in patristic distance and have very different gene repertoire.
+            Feel free to replace with a better reference, this si just meant to be helpful.\\
+            \Marco{I think this is a good reference, fig.
+2 is familiar to us.
+And also fig.2 in \cite{doolittle2009origin} which is a bit older.
+But not sure which other Rocha paper he is referring to.}
 
-    \item Why on earth is the "get an MSA of a block" option in pangraph called "polish"?
+      \item Why on earth is the "get an MSA of a block" option in pangraph called "polish"?\\
+            \Marco{i don't think this comes from the paper (I used grep on the tex file and we never mention polish).
+I think it comes from the documentation but not sure where.
+We explain polish \href{https://neherlab.github.io/pangraph/tutorials/tutorial_3/\#Polishing-the-alignments}{in this part} of the tutorial and I don't think it's ambiguous.
+If in the paper we explain how we propagate the consensus sequence and related problems we will also mention the polish command in the same part and make explicit that it can be used to polish alignments, this should solve the ambiguity.}
 
@@
 Anyway, my understanding is pancontigs are connected by an edge if one of the input genomes is spelled out by going from one to the other.
-So it's about adjacency rather than synteny.
+So it's about adjacency rather than synteny.\\
+            \Marco{I agree, we could use ``if they are ajacent/contiguous on at least one input sequence''?}
 
-    \item Line 131: was N really=100? on the plot in figure 2, you have n.isolates going up to 1000
+      \item Line 131: was N really=100? on the plot in figure 2, you have n.isolates going up to 1000\\
+            \Marco{$N=100$ was the standard value for the rest of the simulations (e.g.
+later for accuracy).
+I did a tentative correction: removed from the main the specification on population size and mentioned to check the SI for details on standard parameter values.}
 
-    \item Figure 5 legend, lower case C in "coli" of "E. coli". Also italics.
+      \item Figure 5 legend, lower case C in "coli" of "E. coli". Also italics.\\
+            \Marco{fixed.}
 \end{enumerate}
@@
 
-\section*{Reviewer 3}
+\subsection*{Reviewer 3}
 
@@
 
+\bibliography{cite}{}
+
 \end{document}

### 2023-02-15 main.tex
feat: addedd comments for rev 2

@@
 
-We simulated populations of size $N=100$ and varying sequence length up to $L=500\,000$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
-In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (respective rates $0.01$ and $0.05$ per generation), and horizontal sequence transfer that occurs with tunable rate $h$ per generation per genome (see SI Section I for details).
+We simulated populations of size $N$ and sequence length $L$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
+In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (respective rates $0.01$ and $0.05$ per generation), and horizontal sequence transfer that occurs with tunable rate $h$ per generation per genome (see SI Section I for details on the simulation and standard parameter values).
 The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
@@
 \paragraph*{Performances as a function of dataset size.}
-The algorithmic complexity was measured empirically by constructing pangenome graphs from data generated by simulating populations of increasing size and average sequence length.
+The algorithmic complexity was measured empirically by constructing pangenome graphs from data generated by simulating populations of varying size and sequence length.
 The mean and standard deviation of the run-time obtained from 50 iterations are shown in Fig. \ref{fig:toy-performance}.

### 2023-02-15 reviewer_response.tex
Update on Overleaf.

@@
 \newcommand{\Marco}[1]{{\color{gray}Marco: #1}}
+\newcommand{\Liam}[1]{{\color{darkgray}Liam: #1}}
 
@@
 You do later add that "To align two graphs, the consensus sequences of all pancontigs in both graphs are searched for homologies and aligned".
-Adding the word consensus in the earlier sentence might avoid confusion.
+Adding the word consensus in the earlier sentence might avoid confusion.\\
             \Marco{changed to ``Pairwise graph alignment is performed by an all-to-all alignment of the \emph{pancontigs} consensus sequences between both graphs, and the order of pairwise alignments is determined by a guide tree." Too compressed?
-I could also break in two sentences and expand a bit more.}
+I could also break in two sentences and expand a bit more.}\\
+            \Liam{that's clear to me.}
 
@@
 But the "maximal" compression depends not only on the sample size N but also on the gene frequency distribution.
-Another possible comparison might be to pangraph's estimation of the pangenome size, but again we might have different effects coming into play.}
+Another possible comparison might be to pangraph's estimation of the pangenome size, but again we might have different effects coming into play.}\\
+            \Liam{yes, I think just adding a dashed line for 1/N for each species to the plot would aid that reading and satisfy this point.}
 
@@
 But here the main message is that core blocks are well-behaved (close to N50 line in the figure), as opposed to fine-grained structural diversity, part of which contains artifacts generated by the block size limit (many short 100bp blocks).
-Should we show block frequency distributions in the supplementary?}
+Should we show block frequency distributions in the supplementary?}\\
+            \Liam{I don't understand their point on diversity - maybe they misunderstood the meaning of size.
+Maybe pancontig `length' is clearer?
+I agree with you about synteny and think this is a nice point to add here to explain the observation.}
 
       \item Figure 6B: could you explore another way to project this data so it is not squished at the ends?
-I can't read off the fraction agree (shared+private) or disagree which I would like to (although at a glance this is exactly the result you want to see, so happy for it to stand if it is the best projection).
-            \Marco{maybe add mean and standard deviation as numbers on the side?}
+I can't read off the fraction agree (shared+private) or disagree which I would like to (although at a glance this is exactly the result you want to see, so happy for it to stand if it is the best projection).\\
+            \Marco{maybe add mean and standard deviation as numbers on the side?}\\
+            \Liam{that seems ideal.}
 
-      \item Have you tried running PanGraph on more than 300 references? I'd love to see another order of magnitude (not necessary for the paper, just curiosity)
-            \Marco{not really, I never tried on ~500 or ~1000 strains. I could try but would not consider this a priority.}
+      \item Have you tried running PanGraph on more than 300 references? I'd love to see another order of magnitude (not necessary for the paper, just curiosity)\\
+            \Marco{not really, I never tried on ~500 or ~1000 strains. I could try but would not consider this a priority.}\\
+            \Liam{agree that it's not necessary.}
 
@@
 But we could expand the discussion with more ideas on applications to study pangenome organization and evolution.
-For software for further analysis and exploration I have \href{https://github.com/mmolari/pypangraph}{this package} to explore and interact with graphs in python, but not sure it's worth mentioning.}
+For software for further analysis and exploration I have \href{https://github.com/mmolari/pypangraph}{this package} to explore and interact with graphs in python, but not sure it's worth mentioning.}\\
+            \Liam{Some ideas that could be mentioned.
+(1) I think one obvious (bad!) use would be to take the core blocks and map reads to those as a `core reference'.
+(2) I know pandora has \href{https://github.com/iqbal-lab-org/make_prg/}{\texttt{make\_prg}} which will construct a `pangenome reference graph' from a set of aligned sequences - I didn't try it on pangraph output.
+(3) For genome structure, exploring `hotspots' in the graph as \href{https://www.nature.com/articles/s41467-017-00808-w}{this paper} does could be done using in/out degree of pancontigs.}
 
       \item Supplementary Information: simulation - could you comment on how/if homologous recombination is included in this model e.g. within a gene.\\
-            \Marco{homologous recombination is not implemented directly in the model. Only sequence transfer between strains. Genes are not explicitly modeled in the simulation.}
+            \Marco{homologous recombination is not implemented directly in the model. Only sequence transfer between strains. Genes are not explicitly modeled in the simulation.}\\
+            \Liam{think just adding a comment in the SI saying this is required.}
 
@@
 Why is a more diverse pancontig more likely to become fragmented?\\
-            \Marco{I think that basically the reason is that merging more distant paralogs/orthologs can fragment more the graph. Maybe we could explain this more explicitly in the text.}
+            \Marco{I think that basically the reason is that merging more distant paralogs/orthologs can fragment more the graph. Maybe we could explain this more explicitly in the text.}\\
+            \Liam{makes sense and explaining this reasoning sounds good.}
 \end{enumerate}
@@
 However we also provide the ``polish'' step for this, that can be executed after graph creation.
-This will take care of refining the block alignments.}
+This will take care of refining the block alignments.}\\
+            \Liam{I think a `narrative section' in the supplementary describing this in brief sounds great.}
 
@@
 But we can make it more precise.
-This scaling is at any rate not a limiting factor in pangraph.}
+This scaling is at any rate not a limiting factor in pangraph.}\\
+            \Liam{clarifying $N$ and $l$ makes sense.}
 
@@
 We then repeat this procedure in an all-vs-all comparison of the consensus sequences of the blocks, and proceed with further merges.
-This is necessary because when two alignments overlap only one of the two is performed in the first alignment round, but the second will be performed in further rounds.}
+This is necessary because when two alignments overlap only one of the two is performed in the first alignment round, but the second will be performed in further rounds.}\\
+                        \Liam{agree, sounds like a section in supplementary is a good idea to address these comments.}
             \end{itemize}
@@
             \Marco{the second is actually a consequence of the first, but except for that it is correct.}\\
+            \Liam{maybe say `$\beta$ controls the maximal sequence divergence between the two merger-candidates.
+The negative pseudo-energy requirement means that no merger will be performed if the Hamming distance exceeds $1/\beta$.'}\\
             One interesting thing here is that you are effectively merging things in reverse evolutionary order, as you have lower energy for longer more similar things which presumably diverged more recently.
 Not sure if that tells us anything.\\
-            \Marco{indeed, the parameter $\beta$ could control how far in the past one wants to accept merges. Should we mention this explicilty in the text?}
+            \Marco{indeed, the parameter $\beta$ could control how far in the past one wants to accept merges. Should we mention this explicilty in the text?}\\
+            \Liam{that's a nice observation.
+could be mentioned as an `intuitively, this should correspond to merging of homologous stretches of sequence in roughly reverse evolutionary order'.
+Risk is suggesting that this is the reason for doing it that way?}
 
@@
 This limit remains, and in this sense no ``false merges'' are performed?
-Not sure I interpret this correctly.}
+Not sure I interpret this correctly.}\\
+            \Liam{I think your interpretation is right.
+Important to clarify that there are those two limits and that $\beta$ only controls the divergence limit \textit{within} what one can actually align.}
 
@@
 For the reviewer I could produce the block frequency and lenght distribution for the two cases and comparing them would give an answer.
-I could do the same for EC or KP for comparison.}
+I could do the same for EC or KP for comparison.}\\
+            \Liam{makes sense.}
 
       \item Just to make sure i understand - is it possible for sequence to be "left out" in between pancontigs?\\
-            \Marco{no, every part of the genome must be contained in a pancontig. It can be left out from the consensus if it is considered an insertion.}
+            \Marco{no, every part of the genome must be contained in a pancontig. It can be left out from the consensus if it is considered an insertion.}\\
+            \Liam{if they possibly misunderstood, suggests that some statement that all sequence ends up in a pancontig should be added to manuscript.
+I suspect they are thinking of case of short ($<$100bp) stretches `in between' pancontigs, which (if I understand correctly) will be incorporated as insertions at the end of the consensus alignment of the adjacent pancontig.}
 
@@
             \Marco{on the repo we have \href{https://github.com/neherlab/pangraph/blob/master/script/config/accnums.json}{this file} for the list of all accession numbers we use, and \href{https://github.com/neherlab/pangraph/blob/master/script/config/projection_strains.json}{this file} for the marginalization pairs.
-I could create some excel spreadsheet with accession numbers that we could add to the submission?}
+I could create some excel spreadsheet with accession numbers that we could add to the submission?}\\
+            \Liam{for sure that's all they want (most biologists don't like json!)}
 \end{enumerate}
@@
 And also fig.2 in \cite{doolittle2009origin} which is a bit older.
-But not sure which other Rocha paper he is referring to.}
+But not sure which other Rocha paper he is referring to.}\\
+            \Liam{I think they mean Figure 5 of \cite{touchon2009} (\href{https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1000344}{here}).
+Citing Touchon 2009 and 2020 and the Doolittle paper seems fine to me.}
 
@@
 We explain polish \href{https://neherlab.github.io/pangraph/tutorials/tutorial_3/\#Polishing-the-alignments}{in this part} of the tutorial and I don't think it's ambiguous.
-If in the paper we explain how we propagate the consensus sequence and related problems we will also mention the polish command in the same part and make explicit that it can be used to polish alignments, this should solve the ambiguity.}
+If in the paper we explain how we propagate the consensus sequence and related problems we will also mention the polish command in the same part and make explicit that it can be used to polish alignments, this should solve the ambiguity.}\\
+            \Liam{this sounds good.
+I think they misunderstood that polish is not just `extract the MSA' but actually produces a pangraph with the same structure but redone alignments.
+I think what is in the documentation there is great and could almost go straight in the paper, something like: `At each stage pancontigs are compared and aligned using their consensus sequences.
+This shortens considerably the time needed, but might introduce minor inconsistencies and artifacts in the alignments.
+These can be removed at the end of the process by reconstructing the sequences in the block and performing a multiple sequence alignment (see Polish command in documentation).')}
 
@@
 So it's about adjacency rather than synteny.\\
-            \Marco{I agree, we could use ``if they are ajacent/contiguous on at least one input sequence''?}
+            \Marco{I agree, we could use ``if they are ajacent/contiguous on at least one input sequence''?}\\
+            \Liam{adjacent sounds good.}
 
@@
 The mechanism for this is obvious but providing a useful user interface might be tedious.\\
+\Liam{I think perhaps producing a helper python script that converts a genbank annotation file for a genome within a pangraph to an output gff/csv with the block locations and coordinates would please this reviewer.}\\
 
@@
 When choosing between competing mergers of paralogs/orthologs the context of the PGG and other possible mergers is important if the pancontigs have been reduced so that they contain only repetitive sequence without unique flaning sequence.
-I think this is a very important point as distinguishing orthologs from paralogs is essential for correct PGG construction and the authors need to explicitly address it.
+I think this is a very important point as \textbf{distinguishing orthologs from paralogs is essential for correct PGG construction and the authors need to explicitly address it}.
 This could be measured as part of the simulated data where identical paralog sequences of various lengths and copy number could be modeled.\\
+\Liam{I am not sure I agree that constructing a pangenome graph means one must distinguish orthologs from paralogs.
+The structure is still useful even if this isn't done.
+Maybe there is a way of doing something of what they propose with simulations, but I suspect this isn't necessary for the editor to accept the paper\ldots}\\
 
@@
 It is also not clear if any of the splitting pancontigs are less than this value if the entire merger is dropped or just massaged to not have the too small new pancontig.\\
+\Liam{this relates to R2's point above about all sequence ending up in pancontigs so I think is just clarification.}\\
 
@@
 For the simulated data how accurate are the final pancontig consensus sequences?\\
+\Liam{expanding discussion of doing an MSA at end as for R2's comments would partially address.
+Quantifying how the consensus sequence degenerates iteratively is an interesting point that seems possible?}\\
 
@@
 Is there any necessary special issues with producing the merged multiple sequence alignment?\\
+\Liam{don't understand their last sentence. Maybe should quantify how common the transitive edges typically are?}\\
 
@@
 The authors might want to consider some form of post processing of the PGG to collapse what have been called bubbles in the graph if alignment of the pancontigs on different edges of the bubble warrant it.\\
+\Liam{0.7Mbp does seem lower than I would expect - I guess that they expect >2 Mbp - but it shows that obsessing about core is a bit unwise.
+With slightly relaxed definitions (95\% of genomes) does the core get sensibly bigger?
+For comparison e.g.
+ggCaller gets ~2,500 core genes in 100\% of n=162 \textit{E.
+coli} (Table 1 of \href{https://www.biorxiv.org/content/10.1101/2023.01.24.524926v1.full.pdf}{here}).}\\
 
@@
 How is this sustainable for species with large numbers of active IS elements?
-This could also be investigated for PanGrah using suitable simulated data.
+This could also be investigated for PanGrah using suitable simulated data.\\
+\Liam{I think nothing to say here apart from `yes, active IS elements split pancontigs up' (and can themselves be split up due to similarity with other IS\ldots).
+For the future one might want to have pangenome methods where you can explicitly pass a list of predefined sequences that should not be split.
+But I think this is a general problem and not something to investigate with simulations or fix here.}\\
 

### 2023-02-15 reviewer_response.tex
chore: temporary push

--- 2023-02-15 reviewer_response.tex
@@
 \newcommand{\Marco}[1]{{\color{gray}Marco: #1}}
+\newcommand{\Marcotmp}[1]{{\color{purple}Marco: #1}}
 
@@
 You do later add that "To align two graphs, the consensus sequences of all pancontigs in both graphs are searched for homologies and aligned".
-Adding the word consensus in the earlier sentence might avoid confusion.
+Adding the word consensus in the earlier sentence might avoid confusion.\\
             \Marco{changed to ``Pairwise graph alignment is performed by an all-to-all alignment of the \emph{pancontigs} consensus sequences between both graphs, and the order of pairwise alignments is determined by a guide tree." Too compressed?
@@
 PanGraph appears to wed these approaches to some extent in that the alignments are done at the nucleotide level and the features are extracted as a characteristic of the alignments.\\
+\Marcotmp{in line 24 we cite \cite{eizenga2020pangenome}.}\\
 
@@
 The mechanism for this is obvious but providing a useful user interface might be tedious.\\
+\Marco{good point.
+I don't think this is something we should tackle for the resubmission, since deciding how to harmonize annotations might be non-trivial and might require some validation, but it is definitely something we want to address in the future.}\\
 
@@
 De Bruijn graph-based assemblers routinely collapse nonbranching portions of the graph into longer sequences than the k-mer size used for construction.\\
+\Marcotmp{from what I understand the structure in de Bruijn graphs is very fine-grained.
+Every SNP can generate a branching event.
+While in our case we have this intermediate corse-graining where homologous parts are collapsed together and the remaining structural variation represents large-scale structural variation in the genome.
+In this sense using k-mers limits long-range homology, because  }\\
 One early approach that does not explicitly generate a pan-genome graph but deals with some of the rearrangement issues cited by the authors is:
 \begin{itemize}
-      \item Angiuoli SV, Dunning Hotopp JC, Salzberg SL, Tettelin H.
-Improving pan-genome annotation using whole genome multiple alignment.
-BMC Bioinformatics.
-2011 Jun 30;12:272.
-doi: 10.1186/1471-2105-12-272.
-PMID: 21718539; PMCID: PMC3142524.
-      \item Angiuoli SV, Salzberg SL.
-Mugsy: fast multiple alignment of closely related whole genomes.
-Bioinformatics.
-2011 Feb 1;27(3):334-42.
-doi: 10.1093/bioinformatics/btq665.
-Epub 2010 Dec 9.
-PMID: 21148543; PMCID: PMC3031037.
+      \item \cite{angiuoli2011improving}
+      \item \cite{angiuoli2011mugsy}
 \end{itemize}
@@
 \begin{itemize}
-      \item Sutton G, Fogel GB, Abramson B, Brinkac L, Michael T, Liu ES, Thomas S.
-A pan-genome method to determine core regions of the Bacillus subtilis and Escherichia coli genomes.
-F1000Res.
-2021 Apr 13;10:286.
-doi: 10.12688/f1000research.51873.2.
-PMID: 34113437; PMCID: PMC8156514.
-      \item Chan AP, Sutton G, DePew J, Krishnakumar R, Choi Y, Huang XZ, Beck E, Harkins DM, Kim M, Lesho EP, Nikolich MP, Fouts DE.
-A novel method of consensus pan-chromosome assembly and large-scale comparative analysis reveal the highly flexible pan-genome of Acinetobacter baumannii.
-Genome Biol.
-2015 Jul 21;16(1):143.
-doi: 10.1186/s13059-015-0701-6.
-PMID: 26195261; PMCID: PMC4507327.
+      \item \cite{sutton2021pan}
+      \item \cite{chan2015novel}
 \end{itemize}
@@
 This could be measured as part of the simulated data where identical paralog sequences of various lengths and copy number could be modeled.\\
+\Marcotmp{}
 
@@
 It is also not clear if any of the splitting pancontigs are less than this value if the entire merger is dropped or just massaged to not have the too small new pancontig.\\
+\Marco{we set a minimal cutoff of 100bp.
+This is stated later in line 106.
+Maybe we could expand the sentence in line 99 and give the value and few more words on the meaning/effect.
+This controls in some sense the ``corse graining'' of the graph: what is the size of an event that we want to encode as structural variation in the graph, rather than as an indel in the alignment.
+In the algorithm this minimal cutoff has different effects depending on the context.
+When aligning, if the match lenght is less than this threshold the merge is refused.
+For long alignments, if the CIGAR string contains indels that are longer than this threshold, these are saved as separate blocks and not as indels in the block.
+The same goes for the flanking sequences, if merging two long blocks leaves overhangs that are less than this threshold size then these overhangs are saved as insertions on the side of the block (but I should double-check this).
+This might be explained in an extra SI section on the merging?}\\
 
@@
 For the simulated data how accurate are the final pancontig consensus sequences?\\
+\Marco{this could be addressed by adding a more detailed algorithm explanation of the merging algorithm in the SI.
+The short answer is that at each merging the full alignment is constructed using the CIGAR of the alignment of the consensus of the two blocks.
+Then consensus of the full alignment is re-evaluated.
+This does not solve the multiple alignment problem during constructions and there can be degeneration on where indels are mapped, but the final graph can be eventually polished.}\\
 
@@
 Is there any necessary special issues with producing the merged multiple sequence alignment?\\
+\Marco{maybe this is a subtlety and could be removed from the main text?
+I can think of one edge case which is due to periodic boundary conditions.
+Initially each circular genomes is saved as a single long block.
+If during alignment this unique blocks is split somewhere in the middle, the resulting edge blocks might be always colinear, since they are actually adjacent parts of the sequence that were artificially split to linearize the circular genome.
+One wants to collaps these two parts in a single block.
+This should happen often at the beginning of the merging of circular genomes.
+But might be unnecessary information when explaining the algorithm the first time, we might want to move it to the SI and spend some more words?}\\
 
@@
 The authors should justify this.
-The authors should also include analysis on the calculated core genome length versus actual core genome length for the simulated data.
+The authors should also include analysis on the calculated core genome length versus actual core genome length for the simulated data.\\
+\Marcotmp{}\\
 The authors might want to consider some form of post processing of the PGG to collapse what have been called bubbles in the graph if alignment of the pancontigs on different edges of the bubble warrant it.\\
+\Marco{if I understand correctly this is hinting at some form of post-processing step in which artifactual small blocks - generated at the edges of long block merges - could be absorbed in the flanking block alignments and removed?}\\
 
@@
 I am very concerned about the potential fracturing of the PGG where unfortunately the devil is in the details and am not convinced that the fracturing will not prevent some limit on the number of genomes which can be used for a pangenome.\\
+\Marco{if the fracturing derives from a real structural property of pangenomes (i.e. ISs that really move around and split long blocks) this might be an intrinsic real obstacle.}\\
 
@@
 How is this sustainable for species with large numbers of active IS elements?
-This could also be investigated for PanGrah using suitable simulated data.
+This could also be investigated for PanGrah using suitable simulated data.\\
+\Marcotmp{}
 

### 2023-02-15 reviewer_response.tex

--- 2023-02-15 reviewer_response.tex
@@
 But not sure which other Rocha paper he is referring to.}\\
-            \Liam{I think they mean Figure 5 of \cite{touchon2009} (\href{https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1000344}{here}).
+            \Liam{I think they mean Figure 5 of \cite{touchon2009organised} (\href{https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1000344}{here}).
 Citing Touchon 2009 and 2020 and the Doolittle paper seems fine to me.}
@@
 \Liam{I think perhaps producing a helper python script that converts a genbank annotation file for a genome within a pangraph to an output gff/csv with the block locations and coordinates would please this reviewer.}\\
+\Marco{Yeah good idea. Maybe something that just map each annotation into a block, without solving the problem of harmonizing annotations which might be some more work...}\\
 
@@
 De Bruijn graph-based assemblers routinely collapse nonbranching portions of the graph into longer sequences than the k-mer size used for construction.\\
+\Marco{from what I understand the structure in de Bruijn graphs is very fine-grained.
+Every SNP can generate a branching event.
+While in our case we have this intermediate corse-graining where homologous parts are collapsed together and the remaining structural variation represents large-scale structural variation in the genome.
+In this sense using k-mers limits long-range homology, because  }\\
 One early approach that does not explicitly generate a pan-genome graph but deals with some of the rearrangement issues cited by the authors is:
 \begin{itemize}
-      \item Angiuoli SV, Dunning Hotopp JC, Salzberg SL, Tettelin H.
-Improving pan-genome annotation using whole genome multiple alignment.
-BMC Bioinformatics.
-2011 Jun 30;12:272.
-doi: 10.1186/1471-2105-12-272.
-PMID: 21718539; PMCID: PMC3142524.
-      \item Angiuoli SV, Salzberg SL.
-Mugsy: fast multiple alignment of closely related whole genomes.
-Bioinformatics.
-2011 Feb 1;27(3):334-42.
-doi: 10.1093/bioinformatics/btq665.
-Epub 2010 Dec 9.
-PMID: 21148543; PMCID: PMC3031037.
+      \item \cite{angiuoli2011improving}
+      \item \cite{angiuoli2011mugsy}
 \end{itemize}
@@
 Maybe there is a way of doing something of what they propose with simulations, but I suspect this isn't necessary for the editor to accept the paper\ldots}\\
+\Marco{I agree, I think it's a different design choice.
+We could also mention that the marginalize command has an option to project out paralogs, but should be a post-processing step rather than something that we want to do while building the graph.}
 
@@
 \Liam{this relates to R2's point above about all sequence ending up in pancontigs so I think is just clarification.}\\
+\Marco{I agree.
+We set a minimal cutoff of 100bp.
+This is stated later in line 106.
+Maybe we could expand the sentence in line 99 and give the value and few more words on the meaning/effect.
+This controls in some sense the ``corse graining'' of the graph: what is the size of an event that we want to encode as structural variation in the graph, rather than as an indel in the alignment.
+In the algorithm this minimal cutoff has different effects depending on the context.
+When aligning, if the match lenght is less than this threshold the merge is refused.
+For long alignments, if the CIGAR string contains indels that are longer than this threshold, these are saved as separate blocks and not as indels in the block.
+The same goes for the flanking sequences, if merging two long blocks leaves overhangs that are less than this threshold size then these overhangs are saved as insertions on the side of the block (but I should double-check this).
+This might be explained in an extra SI section on the merging?}\\
 
@@
 Quantifying how the consensus sequence degenerates iteratively is an interesting point that seems possible?}\\
+\Marco{I agree, this could be addressed by adding a more detailed algorithm explanation of the merging algorithm in the SI.
+The short answer is that at each merging the full alignment is constructed using the CIGAR of the alignment of the consensus of the two blocks.
+Then consensus of the full alignment is re-evaluated.
+This does not solve the multiple alignment problem during constructions and there can be degeneration on where indels are mapped, but the final graph can be eventually polished.
+I don't think we ever quantified how much the consensus sequence degenerates during the iterations.
+Maybe one could devise a proper test set and test.}\\
 
@@
 \Liam{don't understand their last sentence. Maybe should quantify how common the transitive edges typically are?}\\
+\Marco{I can think of one edge case which is due to periodic boundary conditions.
+Initially each circular genomes is saved as a single long block.
+If during alignment this unique blocks is split somewhere in the middle, the resulting edge blocks might be always colinear, since they are actually adjacent parts of the sequence that were artificially split to linearize the circular genome.
+One wants to collaps these two parts in a single block.
+This should happen often at the beginning of the merging of circular genomes.
+But might be unnecessary information when explaining the algorithm the first time, we might want to move it to the SI section where we explain merging in more detail?}\\
 
@@
 coli} (Table 1 of \href{https://www.biorxiv.org/content/10.1101/2023.01.24.524926v1.full.pdf}{here}).}\\
+\Marco{if I understand correctly this is hinting at some form of post-processing step in which artifactual small blocks - generated at the edges of long block merges - could be absorbed in the flanking block alignments and removed?}\\
 
@@
 But I think this is a general problem and not something to investigate with simulations or fix here.}\\
+\Marco{agree. I think this is a structural property that we want to capture, the fact that these pancontigs are split.}
 

### 2023-02-16 reviewer_response.tex
Update on Overleaf.

@@
 \newcommand{\Marco}[1]{{\color{gray}Marco: #1}}
-\newcommand{\Liam}[1]{{\color{darkgray}Liam: #1}}
+\newcommand{\Liam}[1]{{\color{teal}Liam: #1}}
 
@@
             \Marco{indeed, the parameter $\beta$ could control how far in the past one wants to accept merges. Should we mention this explicilty in the text?}\\
-            \Liam{that's a nice observation.
-could be mentioned as an `intuitively, this should correspond to merging of homologous stretches of sequence in roughly reverse evolutionary order'.
-Risk is suggesting that this is the reason for doing it that way?}
+            \Liam{could be mentioned as an `intuitively, this should correspond to merging of homologous stretches of sequence in roughly reverse evolutionary order'.}
 
@@
 Every SNP can generate a branching event.
-While in our case we have this intermediate corse-graining where homologous parts are collapsed together and the remaining structural variation represents large-scale structural variation in the genome.
-In this sense using k-mers limits long-range homology, because  }\\
+While in our case we have this intermediate corse-graining where homologous parts are collapsed together and the remaining structure of the graph represents large-scale structural variation in the genomes.
+I don't think this is solved by collapsing nonbranching portions in dB graphs (but I might not know enough about this...)}\\
 One early approach that does not explicitly generate a pan-genome graph but deals with some of the rearrangement issues cited by the authors is:
@@
 \Marco{I agree, I think it's a different design choice.
-We could also mention that the marginalize command has an option to project out paralogs, but should be a post-processing step rather than something that we want to do while building the graph.}
+We could also mention that the marginalize command has an option to project out paralogs, but should be a post-processing step rather than something that we want to do while building the graph.
+One can a posteriori decide whether one wants to split paralogs using structural information (flanking blocks) or one can even prevent the merging using diversity and the $\beta$ parameter.}
 
@@
 For long alignments, if the CIGAR string contains indels that are longer than this threshold, these are saved as separate blocks and not as indels in the block.
-The same goes for the flanking sequences, if merging two long blocks leaves overhangs that are less than this threshold size then these overhangs are saved as insertions on the side of the block (but I should double-check this).
-This might be explained in an extra SI section on the merging?}\\
+The same goes for the flanking sequences, if merging two long blocks leaves overhangs that are less than this threshold size then these overhangs are saved as part of the merged block and not as separate small blocks.
+This might be explained in the SI?}\\
 
@@
 I don't think we ever quantified how much the consensus sequence degenerates during the iterations.
-Maybe one could devise a proper test set and test.}\\
+Maybe one could devise a proper test set and see how it degenerates over the iterations.}\\
 
@@
 This should happen often at the beginning of the merging of circular genomes.
-But might be unnecessary information when explaining the algorithm the first time, we might want to move it to the SI section where we explain merging in more detail?}\\
+But might be unnecessary detail when explaining the algorithm the first time, we might want to move it to the SI section where we explain merging in more detail?}\\
 
@@
 The authors should justify this.
-The authors should also include analysis on the calculated core genome length versus actual core genome length for the simulated data.
-The authors might want to consider some form of post processing of the PGG to collapse what have been called bubbles in the graph if alignment of the pancontigs on different edges of the bubble warrant it.\\
-\Liam{0.7Mbp does seem lower than I would expect - I guess that they expect >2 Mbp - but it shows that obsessing about core is a bit unwise.
+The authors should also include analysis on the calculated core genome length versus actual core genome length for the simulated data.\\
+\Liam{0.7Mbp does seem lower than I would expect - I guess that they expect $>$2 Mbp - but it shows that obsessing about core is a bit unwise.
 With slightly relaxed definitions (95\% of genomes) does the core get sensibly bigger?
@@
 coli} (Table 1 of \href{https://www.biorxiv.org/content/10.1101/2023.01.24.524926v1.full.pdf}{here}).}\\
+\Marco{I checked and soft core (gene in 95\% of strains) is 2Mbp in our dataset.
+Some of our problems might come from weird strains.
+E.g.
+the second strain with the least amount of soft-core genes is \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_020518.1?report=genbank}{NC\_020518}, an artificial genome that was removed from RefSeq.
+The strain that misses the most of the 1000 most frequent genes \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_017663.1?report=genbank}{NC\_017663} was also removed from RefSeq.
+The rest might be due to misannotations?
+At least one of the most common 1000 genes are missed by 87/307 strains.
+Interestingly in \cite{sutton2021pan} they report $\sim$700 core orthologous gene clusters for an E.coli dataset of $\sim$1000 genomes, if they do not harmonize annotations.
+Those increase to $\sim$1500 after refining their annotations.\\
+The easiest thing to do might be to just add the size of soft-core for all species? Maybe these values might also be added as dashed lines in fig. 4G for comparison with pangraph's estimation?}\\
+The authors might want to consider some form of post processing of the PGG to collapse what have been called bubbles in the graph if alignment of the pancontigs on different edges of the bubble warrant it.\\
 \Marco{if I understand correctly this is hinting at some form of post-processing step in which artifactual small blocks - generated at the edges of long block merges - could be absorbed in the flanking block alignments and removed?}\\
@@
 I am very concerned about the potential fracturing of the PGG where unfortunately the devil is in the details and am not convinced that the fracturing will not prevent some limit on the number of genomes which can be used for a pangenome.\\
-\Marco{if the fracturing derives from a real structural property of pangenomes (i.e. ISs that really move around and split long blocks) this might be an intrinsic real obstacle.}\\
+\Marco{if the fracturing derives from a real structural property of pangenomes (i.e.
+ISs that really move around and split long blocks) this might be an intrinsic limit that derives from the pangenome, I wouldn't consider it a problem with PanGraph.}\\
 
@@
 \end{document}
+
+% meeting notes:
+% SIAS -> (new) Supplementary Information Algorithm Section
+% make changes in the text traceable in the reviewer answers.
+% Reviewer 1:
+% - for blocks, mention that consensus is considered as the centroid of a star-shaped phylogeny?
+% - 2) add line to fig. on compression
+% - 5) the effect comes from the fact that core genes are in long syntenic stretches, compared to accessory that are often in messy islands (cite Rocha?). Just add half sentence to the paper.
+% - 6) annotate the figure with numbers
+% - 8) reference: graph alighner, ggCaller? Many take GFA as input.\
+% - 10) this is something that could be explained in the SIAS: one starts with long and shallow blocks and as graphs are merged and propagated down the tree the blocks become shorter and deeper.
+Removing energy threshold just causes more merging and deeper but shorter blocks.
+(figure?)
+% Reviewer 2:
+% - 1) do not mention blocks, only pancontigs
+% - specify better the meaning of null energy parameters and why more merging.
+% Reviewer 3:
+% - paralog merging is partially controlled by beta. As future direction one should build trees for every blok and split paralogs in a PanX fashion. 
+% - more nuanced treatment of the literature.

### 2023-02-17 reviewer_response.tex
Update on Overleaf.

@@
 
-\newcommand{\review}[1]{\textbf{#1}}
 \newcommand{\Marco}[1]{{\color{gray}Marco: #1}}
@@
 
-% \definecolor{response}{rgb}{0.1, 0.1, 0.1}
-% \newcommand{\reviewer}[2]{\textbf{#1:} #2\vskip 5mm}
-% \newcommand{\response}[1]{{\it {\color{response}\textbf{Response:} #1}}\vskip 5mm}
+\definecolor{response}{rgb}{0.1, 0.1, 0.1}
+\newcommand{\reviewer}[2]{\textbf{#1:} #2\vskip 5mm}
+\newcommand{\response}[1]{{\it {\color{response}\textbf{Response:} #1}}\vskip 5mm}
 % \newcommand{\responsedraft}[1]{{\it {\color{purple}\textbf{ResponseDraft:} #1}}\vskip 5mm}
@@
 
-\review{R1} Bacterial evolution is driven both by local (vertical) mutation and larger scale (horizontal) genome alternations resulting in gain, loss, inversion or reassortment of genetic material.
+\reviewer{R1}{Bacterial evolution is driven both by local (vertical) mutation and larger scale (horizontal) genome alternations resulting in gain, loss, inversion or reassortment of genetic material.
 A single linear reference cannot capture this larger scale diversity, and existing graph reference structures either depend on coloured de Bruijn graphs (losing long range homology) or simplify the complex topologies by partitioning the pangenome and considering genes/loci in isolation.
@@
 
-On reading I had a few queries, which could perhaps be clarified further in the text.
+On reading I had a few queries, which could perhaps be clarified further in the text.}
 
-\begin{enumerate}
-      \item A pancontig is defined as a multiple sequence (sub)alignment.
+\reviewer{R1.1}{A pancontig is defined as a multiple sequence (sub)alignment.
 In line 54 you say "Pairwise graph alignment is performed by an all-to-all alignment of the pancontigs" which I interpreted to mean all sequences in the alignment vs all sequences in another alignment.
 You do later add that "To align two graphs, the consensus sequences of all pancontigs in both graphs are searched for homologies and aligned".
-Adding the word consensus in the earlier sentence might avoid confusion.\\
+Adding the word consensus in the earlier sentence might avoid confusion.}
     \Marco{changed to ``Pairwise graph alignment is performed by an all-to-all alignment of the \emph{pancontigs} consensus sequences between both graphs, and the order of pairwise alignments is determined by a guide tree." Too compressed?
@@
 
-      \item Figure 4C: The main text talks about how to interpret this, as ~1/N for TB and a bit less compression for E.
+\reviewer{R1.2}{Figure 4C: The main text talks about how to interpret this, as ~1/N for TB and a bit less compression for E.
 coli and K.
@@
 Would it be possible to rescale this plot to orders of magnitude of N?
-Or in another way aid this reading?\\
+Or in another way aid this reading?}
     \Marco{maybe try visualizing it as deviation from "one strain" compression (1/N)?
@@
 
-      \item Figure 5 has 2 typos: pacontigs and panconting.\\
+\response{Test}
+
+\reviewer{R1.3}{Figure 5 has 2 typos: pacontigs and panconting.}
     \Marco{fixed.}
 
-      \item Throughout the text E. Coli and K. Pneumoniae etc capitalization.\\
+\reviewer{R1.4}{Throughout the text E. Coli and K. Pneumoniae etc capitalization.}
     \Marco{fixed.}
 
-      \item Line 213: Core pancontigs have higher than average size - is this what you would expect given that if they are core genes then they will be present in more/all of the samples so there could be more diversity due to being more sampled?\\
+\reviewer{R1.5}{Line 213: Core pancontigs have higher than average size - is this what you would expect given that if they are core genes then they will be present in more/all of the samples so there could be more diversity due to being more sampled?}
     \Marco{here the pancontig size is the consensus sequence length.
@@
 
-      \item Figure 6B: could you explore another way to project this data so it is not squished at the ends?
-I can't read off the fraction agree (shared+private) or disagree which I would like to (although at a glance this is exactly the result you want to see, so happy for it to stand if it is the best projection).\\
+\reviewer{R1.6}{Figure 6B: could you explore another way to project this data so it is not squished at the ends?
+I can't read off the fraction agree (shared+private) or disagree which I would like to (although at a glance this is exactly the result you want to see, so happy for it to stand if it is the best projection).}
     \Marco{maybe add mean and standard deviation as numbers on the side?}\\
@@
 
-      \item Have you tried running PanGraph on more than 300 references? I'd love to see another order of magnitude (not necessary for the paper, just curiosity)\\
+7. Have you tried running PanGraph on more than 300 references? I'd love to see another order of magnitude (not necessary for the paper, just curiosity)\\
     \Marco{not really, I never tried on ~500 or ~1000 strains. I could try but would not consider this a priority.}\\
@@
 
-      \item Discussion: Do you know of/have you tried any mapping approaches which could take this reference as an input.
+8.
+Discussion: Do you know of/have you tried any mapping approaches which could take this reference as an input.
 Could you touch on the downstream applications of this graph structure?
@@
 
-      \item Supplementary Information: simulation - could you comment on how/if homologous recombination is included in this model e.g. within a gene.\\
+9. Supplementary Information: simulation - could you comment on how/if homologous recombination is included in this model e.g. within a gene.\\
     \Marco{homologous recombination is not implemented directly in the model. Only sequence transfer between strains. Genes are not explicitly modeled in the simulation.}\\
@@
 
-      \item Supplementary Information C: You comment that the use of null energy parameters results on average in "more and shorter pancontigs" but also that they result in more diverged sequences being merged into the same pancontig.
+10.
+Supplementary Information C: You comment that the use of null energy parameters results on average in "more and shorter pancontigs" but also that they result in more diverged sequences being merged into the same pancontig.
 Why is a more diverse pancontig more likely to become fragmented?\\
@@
     \Liam{makes sense and explaining this reasoning sounds good.}
-\end{enumerate}
 
@@
 
-\review{R2} In this paper, Noll et al set out to make a novel contribution to the pangenomics field. As it stands prior to this work, pangenomics splits into 3 different approaches
+In this paper, Noll et al set out to make a novel contribution to the pangenomics field. As it stands prior to this work, pangenomics splits into 3 different approaches
 
@@
 
-\review{R3} The article "PanGraph: scalable bacterial pan-genome graph construction" reports a new method for constructing a Pan-Genome Graph (PGG) in an efficient manner with an open and extractable grah structure.
+The article "PanGraph: scalable bacterial pan-genome graph construction" reports a new method for constructing a Pan-Genome Graph (PGG) in an efficient manner with an open and extractable grah structure.
 Unfortunately, the authors appear to either be unaware of or downplay the existence of previous methods for the same purpose.\textbf{ The authors need to survey the existing literature and compare and contrast with their method to show the utility of their approach.}\\

### 2023-02-17 reviewer_response.tex
chore: cosmetic changes

@@
 
-In this paper, Noll et al set out to make a novel contribution to the pangenomics field. As it stands prior to this work, pangenomics splits into 3 different approaches
+\reviewer{R2}{In this paper, Noll et al set out to make a novel contribution to the pangenomics field. As it stands prior to this work, pangenomics splits into 3 different approaches
 
@@
 
-Nevertheless, there are a number of clarifications which need to be addressed before the paper is suitable for publication.\\
+      Nevertheless, there are a number of clarifications which need to be addressed before the paper is suitable for publication.}
 
@@
 
-\begin{enumerate}
-      \item Line 46: when you say "PanGraph transforms an arbitrary set of genomes into a graph that simultaneously compresses the collection of sequences and exhaustively summarizes both the structural and nucleotide-level polymorphisms." - I am a bit unclear on how the nucleotide level variation is summarised.
+\reviewer{R2.1}{Line 46: when you say "PanGraph transforms an arbitrary set of genomes into a graph that simultaneously compresses the collection of sequences and exhaustively summarizes both the structural and nucleotide-level polymorphisms." - I am a bit unclear on how the nucleotide level variation is summarised.
 The paper is a bit hand-wavey about this.
@@
 On that same webpage you have a section on "block consensus to node alignment" - how do you choose the order of alleles here?
-This is all valuable information and it would be great to have it in the supplementary or main methods somewhere.\\
+This is all valuable information and it would be great to have it in the supplementary or main methods somewhere.}
 \Marco{this point was also raised by R3.
@@
 
-      \item Line 61: I agree and understand how each sequence can be sketched in linear time.
+\reviewer{R2.2}{Line 61: I agree and understand how each sequence can be sketched in linear time.
 I don't follow how you calculate all pairwise intersections through sorting the list of minimizers, nor do i see how this ends up being log-linear.
 For each genome, you can sort it's minimizers in nlog(n) (ie log linear), where n ~genome length.
-How do you get all pairwise distances from that without another multiplier of n?\\
+How do you get all pairwise distances from that without another multiplier of n?}
 \Marco{in the text we write that ``the pairwise distance matrix is estimated in a time that increases log-linearly in the \textbf{total length} of sequence that is analyzed.'' This might generate some confusion since the scaling depends both on the number of genomes $N$ and their length $l$.
@@
 
-      \item The section on iterative graph alignment (lines 69-77) is in one sense quite clear and intuitive, but there is a lot of detail missing which leaves the reader a bit confused when they try to think about details.
+\reviewer{R2.3}{The section on iterative graph alignment (lines 69-77) is in one sense quite clear and intuitive, but there is a lot of detail missing which leaves the reader a bit confused when they try to think about details.
 I would appreciate some detail being added, perhaps in the supplementary methods.
@@
                   \Liam{agree, sounds like a section in supplementary is a good idea to address these comments.}
-            \end{itemize}
+      \end{itemize}}
 
-      \item Line 92-94, same section. You give two statements:
+\reviewer{R2.4}{Line 92-94, same section. You give two statements:
       "Only mergers whose alignment has negative pseudo-energy are performed"
@@
       These are separate criteria that prevent mergers, right?
-So any potential merger with negative pseudoenergy that does not fail these criteria will happen, with the caveat that you perform mergers in order?\\
+So any potential merger with negative pseudoenergy that does not fail these criteria will happen, with the caveat that you perform mergers in order?
       \Marco{the second is actually a consequence of the first, but except for that it is correct.}\\
@@
       One interesting thing here is that you are effectively merging things in reverse evolutionary order, as you have lower energy for longer more similar things which presumably diverged more recently.
-Not sure if that tells us anything.\\
+Not sure if that tells us anything.}
 \Marco{indeed, the parameter $\beta$ could control how far in the past one wants to accept merges. Should we mention this explicilty in the text?}\\
@@
 
-      \item I very much appreciated that you measure the accuracy of pangraph via this displacement idea.
+\reviewer{R2.5}{I very much appreciated that you measure the accuracy of pangraph via this displacement idea.
 I started writing a kindly worded criticism that I thought your method was flawed - surely the displacement depends not only on the order of doing merges, and decisions made about what to do when there are two equally good places to merge with.
@@
 However i very much like the approach you detail in the methods, numerically solving the assignment problem.
-So, this started as a major issue, but in the end is a compliment.
+So, this started as a major issue, but in the end is a compliment.}
 
-      \item Line 183 - I really do not understand what is going on with your alpha,beta being set to zero.
+\reviewer{R2.6}{Line 183 - I really do not understand what is going on with your alpha,beta being set to zero.
 Are you saying if you stop penalising merges for having too many SNPs, then you will be able to merge more, and this will stop you from havign a 10\% divergence limit?
-It will also mean you make more false merges, right?\\
+It will also mean you make more false merges, right?}
 \Marco{setting $\beta$ to zero will indeed remove the 10\% divergence limit set by the pseudo-energy.
@@
 Not sure I interpret this correctly.}\\
-            \Liam{I think your interpretation is right.
-Important to clarify that there are those two limits and that $\beta$ only controls the divergence limit \textit{within} what one can actually align.}
+\Liam{I think your interpretation is right. Important to clarify that there are those two limits and that $\beta$ only controls the divergence limit \textit{within} what one can actually align.}
 
-      \item Why does the L50 peak go up for null-scores for mmseqs for Mtb, Pm, Kpn and Ec (Figure 4E, purple with lines) but down for helicobacter?\\
+\reviewer{R2.7}{Why does the L50 peak go up for null-scores for mmseqs for Mtb, Pm, Kpn and Ec (Figure 4E, purple with lines) but down for helicobacter?}
 \Marco{the effect of merging more blocks on L50 is not trivial.
@@
 
-      \item Just to make sure i understand - is it possible for sequence to be "left out" in between pancontigs?\\
+\reviewer{R2.8}{Just to make sure i understand - is it possible for sequence to be "left out" in between pancontigs?}
 \Marco{no, every part of the genome must be contained in a pancontig. It can be left out from the consensus if it is considered an insertion.}\\
@@
 
-      \item About data accessibility
+\reviewer{R2.9}{About data accessibility
       \begin{enumerate}
             \item i don't see how to get the list of accessions for section IIB? Saying they were used in the Ding et al paper is not quite enough for me to be sure which ones you used.
-                  \item for the graph marginalisation section, could the accessions of the strains used be in a supplementary file as well as on github?
-Github repos are great, but they can be deleted.
-            \end{enumerate}
+            \item for the graph marginalisation section, could the accessions of the strains used be in a supplementary file as well as on github? Github repos are great, but they can be deleted.
+      \end{enumerate}}
 \Marco{on the repo we have \href{https://github.com/neherlab/pangraph/blob/master/script/config/accnums.json}{this file} for the list of all accession numbers we use, and \href{https://github.com/neherlab/pangraph/blob/master/script/config/projection_strains.json}{this file} for the marginalization pairs.
@@
 \Liam{for sure that's all they want (most biologists don't like json!)}
-\end{enumerate}
 
@@
 
-\begin{enumerate}
-      \item line 13 - it would be good to give a reference for the statement that even closelt related genomes can differ widely in gene content.
+\reviewer{R2.1}{line 13 - it would be good to give a reference for the statement that even closelt related genomes can differ widely in gene content.
       I don't know what the first paper to show this was, the earliest i am aware of is from Eduardo Rocha's team, which annpoyingly i can't locate right now, but Figure 2 here shows it \url{https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1008866} \cite{touchon2020phylogenetic} - you can be very close in patristic distance and have very different gene repertoire.
-            Feel free to replace with a better reference, this si just meant to be helpful.\\
-            \Marco{I think this is a good reference, fig.
-2 is familiar to us.
-And also fig.2 in \cite{doolittle2009origin} which is a bit older.
-But not sure which other Rocha paper he is referring to.}\\
+      Feel free to replace with a better reference, this si just meant to be helpful.}
+\Marco{I think this is a good reference, fig. 2 is familiar to us. And also fig.2 in \cite{doolittle2009origin} which is a bit older. But not sure which other Rocha paper he is referring to.}\\
 \Liam{I think they mean Figure 5 of \cite{touchon2009organised} (\href{https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1000344}{here}).
@@
 
-      \item Why on earth is the "get an MSA of a block" option in pangraph called "polish"?\\
+\reviewer{R2.2}{Why on earth is the "get an MSA of a block" option in pangraph called "polish"?}
 \Marco{i don't think this comes from the paper (I used grep on the tex file and we never mention polish).
@@
 
-      \item line 48: "Pancontigs are connected by an edge if they are syntenic on at least one input sequence;" - i think this is a bit of an abuse of modern language.
+\reviewer{R2.3}{line 48: "Pancontigs are connected by an edge if they are syntenic on at least one input sequence;" - i think this is a bit of an abuse of modern language.
 The word syntenic is generally used to mean two genes/whatever have order conserved on multiple chromosomes (although some googling tells me that in classical genetics it meant things were on the same chromosome).
 Anyway, my understanding is pancontigs are connected by an edge if one of the input genomes is spelled out by going from one to the other.
-So it's about adjacency rather than synteny.\\
+So it's about adjacency rather than synteny.}
 \Marco{I agree, we could use ``if they are ajacent/contiguous on at least one input sequence''?}\\
@@
 
-      \item Line 131: was N really=100? on the plot in figure 2, you have n.isolates going up to 1000\\
+\reviewer{R2.4}{Line 131: was N really=100? on the plot in figure 2, you have n.isolates going up to 1000}
 \Marco{$N=100$ was the standard value for the rest of the simulations (e.g.
@@
 
-      \item Figure 5 legend, lower case C in "coli" of "E. coli". Also italics.\\
+\reviewer{R2.5}{Figure 5 legend, lower case C in "coli" of "E. coli". Also italics.}
 \Marco{fixed.}
-\end{enumerate}
 
-It is a bit depressing and unhelpful to structure a review only around problems and quibbles, so i wanted also to raise a number of
+\reviewer{R2}{It is a bit depressing and unhelpful to structure a review only around problems and quibbles, so i wanted also to raise a number of
       Things I liked:
@@
       \end{enumerate}
+}
 
@@
 
-The article "PanGraph: scalable bacterial pan-genome graph construction" reports a new method for constructing a Pan-Genome Graph (PGG) in an efficient manner with an open and extractable grah structure.
-Unfortunately, the authors appear to either be unaware of or downplay the existence of previous methods for the same purpose.\textbf{ The authors need to survey the existing literature and compare and contrast with their method to show the utility of their approach.}\\
+\reviewer{R3}{The article "PanGraph: scalable bacterial pan-genome graph construction" reports a new method for constructing a Pan-Genome Graph (PGG) in an efficient manner with an open and extractable grah structure.
+Unfortunately, the authors appear to either be unaware of or downplay the existence of previous methods for the same purpose.\textbf{ The authors need to survey the existing literature and compare and contrast with their method to show the utility of their approach.}}
 
-For instance the authors state: "Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated "bags of genes", and thus are not suited for this new era.".
+\reviewer{R3}{For instance the authors state: "Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated "bags of genes", and thus are not suited for this new era.".
 There are many pan-genome analysis tools which treat the pan-genome as a graph and generate a PGG.
 There are two primary differences in these existing tools: those which generate a graph at the nucleotide level with nodes representing basepairs (sometimes nonbranching runs of nodes are collapsed) and those where nodes represent larger features such as genes.\\
-PanGraph appears to wed these approaches to some extent in that the alignments are done at the nucleotide level and the features are extracted as a characteristic of the alignments.\\
+      PanGraph appears to wed these approaches to some extent in that the alignments are done at the nucleotide level and the features are extracted as a characteristic of the alignments.}
 
-An obvious lack in PanGraph is an attempt to map existing annotations back onto the generated PGG which is a necessity for usefulness to biologists.
-The mechanism for this is obvious but providing a useful user interface might be tedious.\\
+\reviewer{R3}{An obvious lack in PanGraph is an attempt to map existing annotations back onto the generated PGG which is a necessity for usefulness to biologists.
+The mechanism for this is obvious but providing a useful user interface might be tedious.}
 \Liam{I think perhaps producing a helper python script that converts a genbank annotation file for a genome within a pangraph to an output gff/csv with the block locations and coordinates would please this reviewer.}\\
@@
 
-This statement: "While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging.
+\reviewer{R3}{This statement: "While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging.
 Colored generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets, although the underlying efficiency derives from a fixed k-mer size which prevents modelling long-range homology (Iqbal et al., 2012; Muggli et al., 2017)." appears to be incomplete at best.
-De Bruijn graph-based assemblers routinely collapse nonbranching portions of the graph into longer sequences than the k-mer size used for construction.\\
-\Marco{from what I understand the structure in de Bruijn graphs is very fine-grained.
+De Bruijn graph-based assemblers routinely collapse nonbranching portions of the graph into longer sequences than the k-mer size used for construction.
+      % \Marco{from what I understand the structure in de Bruijn graphs is very fine-grained.
 Every SNP can generate a branching event.
@@
             \item \cite{chan2015novel}
-\end{itemize}
+      \end{itemize}}
 
-I believe this statement: "It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order." is the essential step in the process and perhaps deserves more focus.
+\reviewer{R3}{I believe this statement: "It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order." is the essential step in the process and perhaps deserves more focus.
 It captures two important issues: \textbf{aligning orthologs not paralogs and keeping the fracturing of the PGG nodes to a minimum}.
@@
 I think this is a very important point as \textbf{distinguishing orthologs from paralogs is essential for correct PGG construction and the authors need to explicitly address it}.
-This could be measured as part of the simulated data where identical paralog sequences of various lengths and copy number could be modeled.\\
+This could be measured as part of the simulated data where identical paralog sequences of various lengths and copy number could be modeled.}
 \Liam{I am not sure I agree that constructing a pangenome graph means one must distinguish orthologs from paralogs.
@@
 
-In the pseudo-energy discussion a further parameter is mentioned: "In addition, there is parameter that controls the minimal size of contigs produced by the algorithm.".
+\reviewer{R3}{In the pseudo-energy discussion a further parameter is mentioned: "In addition, there is parameter that controls the minimal size of contigs produced by the algorithm.".
 The value chosen for this parameter should be explicitly stated along with a rationale for the choice.
-It is also not clear if any of the splitting pancontigs are less than this value if the entire merger is dropped or just massaged to not have the too small new pancontig.\\
+It is also not clear if any of the splitting pancontigs are less than this value if the entire merger is dropped or just massaged to not have the too small new pancontig.}
 \Liam{this relates to R2's point above about all sequence ending up in pancontigs so I think is just clarification.}\\
@@
 
-How are pancontig consensus sequences determined and updated?
+\reviewer{R3}{How are pancontig consensus sequences determined and updated?
 I could not find any details about this.
@@
 Does the consensus sequence degenerate iteratively with mergers?
-For the simulated data how accurate are the final pancontig consensus sequences?\\
+For the simulated data how accurate are the final pancontig consensus sequences?}
 \Liam{expanding discussion of doing an MSA at end as for R2's comments would partially address.
@@
 I don't think we ever quantified how much the consensus sequence degenerates during the iterations.
-Maybe one could devise a proper test set and see how it degenerates over the iterations.}\\
+Maybe one could devise a proper test set and see how it degenerates over the iterations.}
 
-The statement: "Upon completion, transitive 108 edges within the graph, i.e., edges along which all path are co-linear, are removed by merging adjacent pancontigs." needs clarification.
+\reviewer{R3}{The statement: "Upon completion, transitive 108 edges within the graph, i.e., edges along which all path are co-linear, are removed by merging adjacent pancontigs." needs clarification.
 Why would these pancontigs have been split in the first place?
-Is there any necessary special issues with producing the merged multiple sequence alignment?\\
+Is there any necessary special issues with producing the merged multiple sequence alignment?}
 \Liam{don't understand their last sentence. Maybe should quantify how common the transitive edges typically are?}\\
@@
 
-It is not clear to me what Table 1 shows.
+\reviewer{R3}{It is not clear to me what Table 1 shows.
 Is this published data from PanX or data from PanGraph?
@@
 The authors should justify this.
-The authors should also include analysis on the calculated core genome length versus actual core genome length for the simulated data.\\
+The authors should also include analysis on the calculated core genome length versus actual core genome length for the simulated data.}
 \Liam{0.7Mbp does seem lower than I would expect - I guess that they expect $>$2 Mbp - but it shows that obsessing about core is a bit unwise.
@@
 Those increase to $\sim$1500 after refining their annotations.\\
-The easiest thing to do might be to just add the size of soft-core for all species? Maybe these values might also be added as dashed lines in fig. 4G for comparison with pangraph's estimation?}\\
-The authors might want to consider some form of post processing of the PGG to collapse what have been called bubbles in the graph if alignment of the pancontigs on different edges of the bubble warrant it.\\
+      The easiest thing to do might be to just add the size of soft-core for all species?
+Maybe these values might also be added as dashed lines in fig.
+4G for comparison with pangraph's estimation?}\\
+\reviewer{R3}{The authors might want to consider some form of post processing of the PGG to collapse what have been called bubbles in the graph if alignment of the pancontigs on different edges of the bubble warrant it.}
 \Marco{if I understand correctly this is hinting at some form of post-processing step in which artifactual small blocks - generated at the edges of long block merges - could be absorbed in the flanking block alignments and removed?}\\
 
-Overall the paper is well written.
+\reviewer{R3}{Overall the paper is well written.
 I like the basic methodology and the attention to using robust efficient existing algorithms and combining them to keep efficiency in mind.
-I am very concerned about the potential fracturing of the PGG where unfortunately the devil is in the details and am not convinced that the fracturing will not prevent some limit on the number of genomes which can be used for a pangenome.\\
+I am very concerned about the potential fracturing of the PGG where unfortunately the devil is in the details and am not convinced that the fracturing will not prevent some limit on the number of genomes which can be used for a pangenome.}
 \Marco{if the fracturing derives from a real structural property of pangenomes (i.e.
@@
 
-An issue I have struggled with in my own work is how to represent an orthologous gene cluster (node/vertex) within the PGG when the gene sequence in some genomes has an insertion event such as an IS element within the gene.
+\reviewer{R3}{An issue I have struggled with in my own work is how to represent an orthologous gene cluster (node/vertex) within the PGG when the gene sequence in some genomes has an insertion event such as an IS element within the gene.
 I assume for PanGraph that this would necessitate splitting the node into multiple pancontigs.
 How is this sustainable for species with large numbers of active IS elements?
-This could also be investigated for PanGrah using suitable simulated data.\\
+This could also be investigated for PanGrah using suitable simulated data.}
 \Liam{I think nothing to say here apart from `yes, active IS elements split pancontigs up' (and can themselves be split up due to similarity with other IS\ldots).

### 2023-02-17 reviewer_response.tex
Update on Overleaf.

@@
 
-\definecolor{response}{rgb}{0.1, 0.1, 0.1}
+\definecolor{response}{rgb}{0.2, 0.2, 0.2}
 \newcommand{\reviewer}[2]{\textbf{#1:} #2\vskip 5mm}
@@
 Or in another way aid this reading?}
-\Marco{maybe try visualizing it as deviation from "one strain" compression (1/N)?
+\response{we agree with the reviewer that it is important to be able visually compare the compression with the number $N$ of isolates for each species, and we thank him for the suggestion.
+We modified fig.~4C in the main text and fig.~4I in SI marking the value of $1/N$ for each species with horizontal lines, and updated the captions accordingly.}
+% \Marco{maybe try visualizing it as deviation from "one strain" compression (1/N)?
 But the "maximal" compression depends not only on the sample size N but also on the gene frequency distribution.
 Another possible comparison might be to pangraph's estimation of the pangenome size, but again we might have different effects coming into play.}\\
-\Liam{yes, I think just adding a dashed line for 1/N for each species to the plot would aid that reading and satisfy this point.}
-
-\response{Test}
+% \Liam{yes, I think just adding a dashed line for 1/N for each species to the plot would aid that reading and satisfy this point.}
 
@@
       These are separate criteria that prevent mergers, right?
-So any potential merger with negative pseudoenergy that does not fail these criteria will happen, with the caveat that you perform mergers in order?
+So any potential merger with negative pseudoenergy that does not fail these criteria will happen, with the caveat that you perform mergers in order?\\
+      One interesting thing here is that you are effectively merging things in reverse evolutionary order, as you have lower energy for longer more similar things which presumably diverged more recently.
+Not sure if that tells us anything.}
 \Marco{the second is actually a consequence of the first, but except for that it is correct.}\\
@@
 The negative pseudo-energy requirement means that no merger will be performed if the Hamming distance exceeds $1/\beta$.'}\\
-      One interesting thing here is that you are effectively merging things in reverse evolutionary order, as you have lower energy for longer more similar things which presumably diverged more recently.
-Not sure if that tells us anything.}
 \Marco{indeed, the parameter $\beta$ could control how far in the past one wants to accept merges. Should we mention this explicilty in the text?}\\
@@
 So it's about adjacency rather than synteny.}
-\Marco{I agree, we could use ``if they are ajacent/contiguous on at least one input sequence''?}\\
-\Liam{adjacent sounds good.}
+% \Marco{I agree, we could use ``if they are ajacent/contiguous on at least one input sequence''?}\\
+% \Liam{adjacent sounds good.}
+\response{we agree with the reviewer observation. We modified ``syntenic on'' to ``adjacent in'' in line 48.}
 
 \reviewer{R2.4}{Line 131: was N really=100? on the plot in figure 2, you have n.isolates going up to 1000}
-\Marco{$N=100$ was the standard value for the rest of the simulations (e.g.
+% \Marco{$N=100$ was the standard value for the rest of the simulations (e.g.
 later for accuracy).
 I did a tentative correction: removed from the main the specification on population size and mentioned to check the SI for details on standard parameter values.}
+\response{The value $N=100$ is the standard value used in the rest of the simulation, but as the reviewer pointed out this value is varied for the results showed in fig.~2 and the statement was confusing.
+We modified line 131 removing the value of $N$ and referring to SI section 1 for the standard value of simulation parameters.}
 
 \reviewer{R2.5}{Figure 5 legend, lower case C in "coli" of "E. coli". Also italics.}
-\Marco{fixed.}
+% \Marco{fixed.}
+\response{The legend has been corrected.}
 
@@
 (figure?)
+% - test pangraph with 500 E.coli.
 % Reviewer 2:

### 2023-02-17 main.tex
Update on Overleaf.

@@
 The graph is composed of \emph{pancontigs}, which represent linear multiple-sequence alignments of homologous sequence found within one or more input genomes.
-\emph{Pancontigs} are connected by an edge if they are syntenic on at least one input sequence; individual sequences are then recapitulated by contiguous \emph{paths} through the graph.
+\emph{Pancontigs} are connected by an edge if they are adjacent in at least one input sequence; individual sequences are then recapitulated by contiguous \emph{paths} through the graph.
 Pancontigs are directed, meaning their orientation in a path can be either 5' to 3' or vice versa.
@@
         \textbf{C}: sequence compression, defined as the ratio between the pangenome graph size and the cumulative size of all the sequences contained in the graph.
+For reference, the inverse of this number is marked.
     }

### 2023-02-18 reviewer_response.tex
chore: partial update supplementary

@@
 Adding the word consensus in the earlier sentence might avoid confusion.}
-\Marco{changed to ``Pairwise graph alignment is performed by an all-to-all alignment of the \emph{pancontigs} consensus sequences between both graphs, and the order of pairwise alignments is determined by a guide tree." Too compressed?
+% \Marco{changed to ``Pairwise graph alignment is performed by an all-to-all alignment of the \emph{pancontigs} consensus sequences between both graphs, and the order of pairwise alignments is determined by a guide tree." Too compressed?
 I could also break in two sentences and expand a bit more.}\\
-\Liam{that's clear to me.}
+% \Liam{that's clear to me.}
+\response{We agree with the reviewer suggestion. The wording in Line 54 has been updated to ``all-to-all alignment of the \textit{pancontigs} consensus sequences''.}
 
@@
 Or in another way aid this reading?}
-\response{we agree with the reviewer that it is important to be able visually compare the compression with the number $N$ of isolates for each species, and we thank him for the suggestion.
+\response{We agree with the reviewer that it is important to be able visually compare the compression with the number $N$ of isolates for each species, and we thank him for the suggestion.
 We modified fig.~4C in the main text and fig.~4I in SI marking the value of $1/N$ for each species with horizontal lines, and updated the captions accordingly.}
@@
 \reviewer{R1.3}{Figure 5 has 2 typos: pacontigs and panconting.}
-\Marco{fixed.}
+% \Marco{fixed.}
+\response{The typos have been corrected.}
 
 \reviewer{R1.4}{Throughout the text E. Coli and K. Pneumoniae etc capitalization.}
-\Marco{fixed.}
+% \Marco{fixed.}
+\response{The capitalization has been corrected.}
 
@@
 % - more nuanced treatment of the literature.
+% - update marginalize figure legend
+% - update SI section numbering in the main text

### 2023-02-18 supplementary.tex
feat: review comments
Update on Overleaf.
chore: partial update supplementary

@@
 \newcommand{\dcore}{\langle d_\text{core} \rangle}
+\newcommand{\Lthr}{L_{\min}}
 
@@
 \maketitle
+
+\section{Graph merging algorithm}
+\label{supp:graph_merging}
+
+In this section we explain in more detail the procedure used to merge two pangenome graphs together. This procedure is at the core of PanGraphs's algorithm.
+
+As explained in the main text, the full pangenome graph is built by placing genomes on terminal leaves of a guide tree.
+These genomes are initially considered as elementary graphs composed of a single \textit{pancontig}.
+These graphs are then propagated along the tree towards the root, and at each intersection a merge operation is performed.
+The full pangenome graph is then collected from the root.
+
+\subsection*{pangenome graph representation}
+
+As a preliminary step in the explanation we spend few words on the pangenome graph representation of pangraph.
+A pangenome graph is essentially composed of two elements: pancontigs, which encode for alignment of homologous sequences, and paths, that correspond to representation for genomes as a sequence of pancontigs.
+
+Each \textbf{pancontig} contains the full information on the alignment of homologous sequences it contains.
+In PanGraph each pancontig is econded with a consensus sequence, and each line of the alignment is saved as variations over the consensus (mutations, insertions and deletions).
+More information on the precise encoding used can be found in PanGraph documentation.\footnote{See \url{https://neherlab.github.io/pangraph/tutorials/tutorial_2/\#From-block-consensus-to-node-alignment}}
+
+Each \textbf{path} is an oriented collection of pancontigs that represents a genome.
+For each item in the sequence along with the pacontig id PanGraph also stores the pancontig orientation (whether homology with the consensus is on the forward the reverse strain of the genome) and which variations on the pancontig consesus (i.e.
+which line in the pancontig alignment) correspond to this precise location in the genome.
+
+From knowledge of all the pancontigs and pahts, every single genome can be exactly reconstructed.
+Every part of the sequence is always encoded in a pancontig.
+``Small-scale'' difference between genomes (i.e.
+SNPs, small insertions and deletions) are encoded in pancontigs as variations over the consensus.
+``Large-scale'' differences, such as presence or absence of long stretch of sequences, are instead encoded in paths as presence or absence of pancontigs.
+This scale separation can be tuned using PanGraph parameters, as will be explained in the next section.
+
+\subsection*{graph merging}
+
+At each merge operation two pangenome graphs need to be compressed into a single one. Three of pangraph's parameters play a role in the outcome of this procedure:
+\begin{itemize}
+    \item the minimal block length $\Lthr$.
+    \item pseudo-energy parameters $\alpha$ and $\beta$ (cf. eq.~1 in the main text).
+    \item the choice of alignment kernel (minimap2 or mmseqs) and aligner precision (asm5, asm10 or asm20 for minimap2).
+\end{itemize}
+In the merging, these parameters control how much of the diversity between sequence must be stored as large-scare variation (difference in paths) and how much as fine-grained variation (differences in alignments, stored in pancontigs).
+This is done for example by setting a divergence limit for sequences to be stored in the same alignment.\\
+An implicit assumption of our procedure is that the phylogeny of sequences stored inside the same panconting is roughly star-like, with each sequence containing independent variations from the consensus.
+In such cases the consensus is a good ``average'' representation for the set of sequences in the alignment.
+In the next section we briefly discussion how PanGraph behaves when this assumption is not satisfied, such as in the case of merging ortholog and paralog sequences in the same pancontig.\\
+% move this paragraph or change?
+
+The merging operation is performed in the following steps:
+\begin{enumerate}
+    \item Using the chosen aligner kernel, we scan for homology between pancontigs of the two different graphs. This is done by using only the consensus sequence of each pancontig.
+    \item The aligner will return a set of matches.
+Each match is characterized by an alignment length and a measure of diversity.
+Using this information we rank matches by pseudo energy, so that longer and less diverged matches will be processed first.
+We refuse all matches with negative pseudo-energu and all matches shorter than the threshold length $\Lthr$.
+    \item For each of the remaining matches we attempt to merge pancontigs following these steps:
+          \begin{enumerate}
+              \item We check that no prior merges have been performed on the same two pancontigs.
+Sometimes multiple matches are found in the same pancontig, and in this case we prioritize the ones with higher pseudo-energy.
+Any following match will be discarded in this step.
+The same merge might however be performed later on in the self-map phase of the algorithm (see step 4).
+              \item Flanking regions of the consensus sequences that are not part of the matches are cut and not merged and stored in separate pancontigs.
+One exceptions is when one of the flanking regions is shorted than the threshold length $\Lthr$.
+In this case instead of creating a very short pancontig this region is stored as a small flanking \textbf{insertion (check)} in the merged pancontig.
+              \item The CIGAR string of the match is then parsed and subdivided in sub-regions.
+Regions can be either shared or private to query or reference.
+Shared regions contain Each insertion or deletion longer than $\Lthr$ will cause a split in the merge, and will be stored as a separate panconting.
+\textbf{(what happens with small matches in longer flanking indels?)}
+              \item
+          \end{enumerate}
+    \item Once the list of all matches has been exhausted and all merges have been performed, the two graphs have been turned into
+\end{enumerate}
+
+% post-processing step: transitive edges. Explain.
+
+% blocks become progressively deeper but shorter. The energy parameters and the threshold lengt determine when to stop.
+% implicit assumption of star-like topology.
+
+\subsection*{PanGraph parameters effect}
+% scale separation effect. Maybe move somewhere else?
+% also discussion on orthologs and paralogs.
 
@@
         \textbf{I}: ratio between the length of the pangenome graph and the total length of all genomes included in the graph.
+Since maximal compression also depends on the number of isolates $N$, we mark with horizontal lines the value of $1/N$ for each species.
     }

### 2023-02-18 supplementary.tex
chore: partial results

@@
 In this case instead of creating a very short pancontig this region is stored as a small flanking \textbf{insertion (check)} in the merged pancontig.
-              \item The CIGAR string of the match is then parsed and subdivided in sub-regions.
-Regions can be either shared or private to query or reference.
-Shared regions contain Each insertion or deletion longer than $\Lthr$ will cause a split in the merge, and will be stored as a separate panconting.
+              \item The remaining homolous region is then a candidate for the merging.
+However the merging might result in more than a single panconting, depending on the underlying alignment.
+If the alignment contains insertions or deletions longer thant $\Lthr$, these need to appear as separate matches.
+To this end the CIGAR string is parsed, and the long match is potentially split in multiple sub-matches, with interposed long insertions or deletions that will be saved as separate pancontigs.
 \textbf{(what happens with small matches in longer flanking indels?)}
-              \item
+              \item At this point every matching sub-region is processed to produce a single merged pancontig.
+This is done by expanding the alignments of the two original sub-contigs, and stitching them together following the instructions contained in the CIGAR string of the sub-match.
+Once the full alignment is constructed the new consensus sequence is calculated using the frequency of nucleotides at each position.
+Each sequence in the alignment is then encoded as variations over this consensus.
+This terminates the creation of a new pancontigs.
+\textbf{(check)}.
+              \item The merging is repeated for every matching sub-region of the alignment.
+Each merged pancontig is shorter in size than the startin pancontigs, since it only represents a sub part of the sequence, but its allignment has depth equal to the sum of the depth of the original pancontigs.
+The original pancontigs are then removed from the pangenome graph, and no new merges can be performed on them at this step of the algorithm.
+The newly-created pancontigs are added to the graph and the corresponding paths are updated.
           \end{enumerate}
-    \item Once the list of all matches has been exhausted and all merges have been performed, the two graphs have been turned into
+    \item Once the list of all matches has been exhausted the two graphs have been merged into one.
+However at the previous stage only one merge could be performed per pancontig, which could be insufficient to collapse all of the homology contained in the graph.
+To this end we repeat the procedure described in step 3, but instead of looking for matches between pancontigs of two different graphs, we search for matches between all pancontigs in the same graph.
+Merges are performed following the same rules, and this procedure is repeated until no new merges can be performed.
+    \item In the final step of the procedure we perform some final grooming of the graph.
+This includes for example removing transitive edges, i.e.
+pairs of pancontigs that are always traversed in the same way by all paths.
+These pairs are merged in a single pancontig.
+This can happen for example when dealing with circular sequences such as bacterial chromosomes.
+These sequences are transformed into a single linear block at the beginning of the algorithm, which introduces an artificial cut.
+When this elementary graph is propagated and the unique block is split, a transitive edge is generated across the artificial cut.
+We remove it in this step.
+\textbf{(check.)}
 \end{enumerate}
 
+% comment: a bit too verbose? The last part has been added to address the reviewer's comments but sounds superfluous even in the SI.
 % post-processing step: transitive edges. Explain.
@@
 % implicit assumption of star-like topology.
+% consensus updated but no solving of MSA
 
 \subsection*{PanGraph parameters effect}
+
+We conclude this section with some comments on the effect of PanGraph's parameters on the resulting pangenome graphs.
+
+The value of $\Lthr$ controls the minimal size of blocks.
+Indels shorter than this size are saved as variation in alignments, rather than as separate blocks.
+As such this parameter partially controls the fragmentation of the graph.
+It should be set
+The default value in PanGraph is set to 100 bp.
+A similar role is played by the parameter $\alpha$. This can be interpreted as the cost of...
+% discuss alpha with this?
+
+The value of $\beta$ and
 % scale separation effect. Maybe move somewhere else?
 % also discussion on orthologs and paralogs.
+% a summary of these considerations in the main text, with reference to this paragraph?
+% references to SI in the main text.
 

### 2023-02-20 reviewer_response.tex
feat: SI

@@
 
-7. Have you tried running PanGraph on more than 300 references? I'd love to see another order of magnitude (not necessary for the paper, just curiosity)\\
-\Marco{not really, I never tried on ~500 or ~1000 strains. I could try but would not consider this a priority.}\\
-\Liam{agree that it's not necessary.}
+\reviewer{R1.7}{Have you tried running PanGraph on more than 300 references? I'd love to see another order of magnitude (not necessary for the paper, just curiosity)}
+% \Marco{not really, I never tried on ~500 or ~1000 strains. I could try but would not consider this a priority.}\\
+% \Liam{agree that it's not necessary.}
+\response{The biggest dataset we tried PanGraph on is a collection of 500 E.coli isolates.
+We used minimap2 as alignment kernel, asm20 sensitivity and values of $\alpha = 100$, $\beta=20$.
+This took 18 hours and 12 minutes to complete and had maximum memory usage of 17Gb.}
 
-8.
-Discussion: Do you know of/have you tried any mapping approaches which could take this reference as an input.
+\reviewer{R1.8}{Discussion: Do you know of/have you tried any mapping approaches which could take this reference as an input.
 Could you touch on the downstream applications of this graph structure?
@@
 touching on what software can be used to map/align to such a data structure?
-Or how it can be used to analysis genome structure.\\
+Or how it can be used to analysis genome structure.}
 \Marco{other than explicitly exporting block sequences and mapping to those, we do not really using the pangenome graph as a reference against which to map new strains...
@@
 
-9. Supplementary Information: simulation - could you comment on how/if homologous recombination is included in this model e.g. within a gene.\\
+\reviewer{R1.9}{Supplementary Information: simulation - could you comment on how/if homologous recombination is included in this model e.g. within a gene.}
 \Marco{homologous recombination is not implemented directly in the model. Only sequence transfer between strains. Genes are not explicitly modeled in the simulation.}\\
@@
 
-10.
-Supplementary Information C: You comment that the use of null energy parameters results on average in "more and shorter pancontigs" but also that they result in more diverged sequences being merged into the same pancontig.
-Why is a more diverse pancontig more likely to become fragmented?\\
+\reviewer{R1.10}{Supplementary Information C: You comment that the use of null energy parameters results on average in "more and shorter pancontigs" but also that they result in more diverged sequences being merged into the same pancontig.
+Why is a more diverse pancontig more likely to become fragmented?}
 \Marco{I think that basically the reason is that merging more distant paralogs/orthologs can fragment more the graph. Maybe we could explain this more explicitly in the text.}\\

### 2023-02-20 supplementary.tex
feat: SI

@@
               \item We check that no prior merges have been performed on the same two pancontigs.
-Sometimes multiple matches are found in the same pancontig, and in this case we prioritize the ones with higher pseudo-energy.
-Any following match will be discarded in this step.
-The same merge might however be performed later on in the self-map phase of the algorithm (see step 4).
+Sometimes multiple matches are found on the same pancontig, and in this case we prioritize the ones with higher pseudo-energy.
+Any following match on the same pancontig will be discarded at this step.
+It will however be performed later on in the self-map phase of the algorithm (see step 4).
               \item Flanking regions of the consensus sequences that are not part of the matches are cut and not merged and stored in separate pancontigs.
-One exceptions is when one of the flanking regions is shorted than the threshold length $\Lthr$.
-In this case instead of creating a very short pancontig this region is stored as a small flanking \textbf{insertion (check)} in the merged pancontig.
+One exceptions is when any of these flanking regions is shorter than the threshold length $\Lthr$.
+In this case instead of creating a very short pancontig the region is stored as an indel in the merged pancontig alignment.
               \item The remaining homolous region is then a candidate for the merging.
 However the merging might result in more than a single panconting, depending on the underlying alignment.
-If the alignment contains insertions or deletions longer thant $\Lthr$, these need to appear as separate matches.
+If the alignment contains insertions or deletions longer thant $\Lthr$, these need to appear as separate pancontigs.
 To this end the CIGAR string is parsed, and the long match is potentially split in multiple sub-matches, with interposed long insertions or deletions that will be saved as separate pancontigs.
-\textbf{(what happens with small matches in longer flanking indels?)}
-              \item At this point every matching sub-region is processed to produce a single merged pancontig.
-This is done by expanding the alignments of the two original sub-contigs, and stitching them together following the instructions contained in the CIGAR string of the sub-match.
-Once the full alignment is constructed the new consensus sequence is calculated using the frequency of nucleotides at each position.
-Each sequence in the alignment is then encoded as variations over this consensus.
-This terminates the creation of a new pancontigs.
-\textbf{(check)}.
-              \item The merging is repeated for every matching sub-region of the alignment.
-Each merged pancontig is shorter in size than the startin pancontigs, since it only represents a sub part of the sequence, but its allignment has depth equal to the sum of the depth of the original pancontigs.
-The original pancontigs are then removed from the pangenome graph, and no new merges can be performed on them at this step of the algorithm.
+              \item At this point every (sub-)match is processed to produce a single merged pancontig.
+This is done by expanding the alignments of the two original pancontigs, and stitching them together following the instructions contained in the CIGAR string of the match.
+Once the full alignment is constructed the new consensus sequence is re-evaluated, and each sequence is then encoded as variations over this consensus.
+This completes the creation of a new pancontig.
+              \item The original starting pancontigs are then removed from the pangenome graph, and no new merges can be performed on them at this step of the algorithm.
 The newly-created pancontigs are added to the graph and the corresponding paths are updated.
@@
 We remove it in this step.
-\textbf{(check.)}
 \end{enumerate}
@@
 
-% blocks become progressively deeper but shorter. The energy parameters and the threshold lengt determine when to stop.
+% pancontigs become progressively deeper but shorter. The energy parameters and the threshold lengt determine when to stop.
 % implicit assumption of star-like topology.

### 2023-02-22 reviewer_response.tex
Update on Overleaf.

@@
 \newcommand{\response}[1]{{\it {\color{response}\textbf{Response:} #1}}\vskip 5mm}
-% \newcommand{\responsedraft}[1]{{\it {\color{purple}\textbf{ResponseDraft:} #1}}\vskip 5mm}
+\newcommand{\responsedraft}[1]{{\it {\color{purple}\textbf{ResponseDraft:} #1}}\vskip 5mm}
 
@@
 % \Marco{fixed.}
+
 \response{The typos have been corrected.}
@@
 \reviewer{R1.5}{Line 213: Core pancontigs have higher than average size - is this what you would expect given that if they are core genes then they will be present in more/all of the samples so there could be more diversity due to being more sampled?}
-\Marco{here the pancontig size is the consensus sequence length.
+% \Marco{here the pancontig size is the consensus sequence length.
 I would say that the fact that they have higher-than-average size is mostly due to synteny.
@@
 Should we show block frequency distributions in the supplementary?}\\
-\Liam{I don't understand their point on diversity - maybe they misunderstood the meaning of size.
+% \Liam{I don't understand their point on diversity - maybe they misunderstood the meaning of size.
 Maybe pancontig `length' is clearer?
 I agree with you about synteny and think this is a nice point to add here to explain the observation.}
+\response{The observation that core pancontigs are longer than average was in agreement with the expectation that core genes would be more syntenic and present less structural fragmentation compared to accessory genes.
+In contrast, we expected most short pancontigs to be located around recombination islands, where structural variation is high across different genomes.
+We modified the line to: ``Core pancontigs having higher-than-average length.
+This is expected given their strong synteny and the fact that they present less structural variation across genomes compared to accessory genes.''}
 
@@
 I can't read off the fraction agree (shared+private) or disagree which I would like to (although at a glance this is exactly the result you want to see, so happy for it to stand if it is the best projection).}
-\Marco{maybe add mean and standard deviation as numbers on the side?}\\
-\Liam{that seems ideal.}
+% \Marco{maybe add mean and standard deviation as numbers on the side?}\\
+% \Liam{that seems ideal.}
+\response{We considered different projections and in the end decided to maintain the linear projection for ease of interpretation, but we added next to each distribution its mean and standard deviation.
+In this way both deviations from 0 and 1 in the first two distributions can be quantified.}
 
@@
 \reviewer{R1.9}{Supplementary Information: simulation - could you comment on how/if homologous recombination is included in this model e.g. within a gene.}
-\Marco{homologous recombination is not implemented directly in the model. Only sequence transfer between strains. Genes are not explicitly modeled in the simulation.}\\
-\Liam{think just adding a comment in the SI saying this is required.}
+% \Marco{homologous recombination is not implemented directly in the model. Only sequence transfer between strains. Genes are not explicitly modeled in the simulation.}\\
+% \Liam{think just adding a comment in the SI saying this is required.}
+\response{Homologous recombination is not directly included in our simulated data generation procedure.
+We only simulate transfer of sequence between two genomes but this transfer is not based on homology.
+Moreover we simulate sequence evolution without a notion of genes.
+This is now specified in SI section II.
+To avoid confusion we also changed ``Horizontal Gene Transfer'' to ``Horizontal Sequence Transfer''.}
 
@@
 Why is a more diverse pancontig more likely to become fragmented?}
-\Marco{I think that basically the reason is that merging more distant paralogs/orthologs can fragment more the graph. Maybe we could explain this more explicitly in the text.}\\
-\Liam{makes sense and explaining this reasoning sounds good.}
+% \Marco{I think that basically the reason is that merging more distant paralogs/orthologs can fragment more the graph. Maybe we could explain this more explicitly in the text.}\\
+% \Liam{makes sense and explaining this reasoning sounds good.}
+\response{We use the pseudo-energy defined in eq.~(1) as an heuristic to decide whether a homologous match detected by our aligner should result in a merge between two pancontigs, depending on the match length and on sequence divergence.
+In particular any choice of $\beta$ would result in no merges between sequences with divergence higher than $1/\beta$.
+Setting $\beta=0$ instead does not impose a divergence limit on merges, other than the one intrinsically set by the aligner sensitivity.
+In this sense the choice of null energy parameter results in more diverged sequences being merged into the same pancontig.
+A consequence of more merging is also more fragmentation of the graph, since the merging of a homologous section of two pancontigs often requires splitting them in chunks.
+Therefore we expected that the choice of null energy parameters would lead to shorter pancontigs, and alignments containing more diverged sequences.
+To make this reasoning more explicit we added section I in the SI to better explain the merging procedure and the effect of tuning the pseudo-energy parameters.}
 
@@
 This is all valuable information and it would be great to have it in the supplementary or main methods somewhere.}
-\Marco{this point was also raised by R3.
+% \Marco{this point was also raised by R3.
 Mutations are always represented as variation on the consensus.
@@
 This will take care of refining the block alignments.}\\
-\Liam{I think a `narrative section' in the supplementary describing this in brief sounds great.}
+% \Liam{I think a `narrative section' in the supplementary describing this in brief sounds great.}
+\response{In our pangenome graph representation each pancontig encodes an alignment using a consensus sequence, and encoding each line as variations over the consensus.
+The consensus sequence is defined using the majority rule on the alignment and is updated at each pancontig merge.
+Each line in the alignment is then only described as variations over this consensus (mutations, insertions and deletions).\\
+To make this more clear we introduced section I in SI, in which we explain in more detail the merging procedure and how the consensus sequence is updated.
+We also refer to the documentation page mentioned by the reviewer.
+Moreover, to avoid ambiguity we also avoided referring to pancontigs as ``blocks'' in the main text.}
 
@@
 How do you get all pairwise distances from that without another multiplier of n?}
-\Marco{in the text we write that ``the pairwise distance matrix is estimated in a time that increases log-linearly in the \textbf{total length} of sequence that is analyzed.'' This might generate some confusion since the scaling depends both on the number of genomes $N$ and their length $l$.
+% \Marco{in the text we write that ``the pairwise distance matrix is estimated in a time that increases log-linearly in the \textbf{total length} of sequence that is analyzed.'' This might generate some confusion since the scaling depends both on the number of genomes $N$ and their length $l$.
 The algorithmic time for all the pairwise distances should scale as $\sim N^2 l \log l$.
@@
 This scaling is at any rate not a limiting factor in pangraph.}\\
-\Liam{clarifying $N$ and $l$ makes sense.}
+% \Liam{clarifying $N$ and $l$ makes sense.}
+\responsedraft{We agree with the reviewer that the formulation in the main text is too concise and not completely correct.
+There are two main variables at play: the length of a genomes $l$ and the number of isolates $n$.
+As the reviewer points out, the set of minimizers for each genome can be created and sorted in $l \log l$.
+Finding the set of shared minimizers would require comparing $\sim n^2$ sets of ordered lists of minimizers, which scales as $\sim n^2 l$.
+This would provide a scaling $O(n l \log l + n^2 l)$\\
+In practice in our implementation we achieve a better scaling by sketching and sorting all the genomes together, which can be done in time $L \log L$, where $L= n \times l$ is the total length of all genomes.
+We keep the information on each minimizer origin in the list.
+We then scroll the list and for each chunk of equal minimizers we update the corresponding entries of a $n^2$ matrix of minimizer hit counts.
+Depending on the amount of shared minimizers, this scales...\\
+% In practice we achieve a scaling that is between $L \log L$ and $n L \log L$.
+We stress however that the computational time to build the guide tree is usually negligible compared to the time required for graph merging.
+% missing: changes in main text. My plan: report the n^2 l log l scaling. 
+}
 

### 2023-02-22 main.tex
Update on Overleaf.

@@
 \newcommand{\Lcore}{L_\text{core}}
+\newcommand{\Lsoftcore}{L_\text{95\%}}
 \newcommand{\Lpang}{L_\text{pang}}
@@
 The choice of alignment kernel influences the threshold diversity at which accuracy is lost, suggesting that inability to detect homology between diverged sequences is the reason.\\
-For each alignment kernel and value of average sequence divergence in the population we evaluated the fraction of breakpoints that have displacement greater than the default minimal block size for Pangraph of 100 bp (cf.
+For each alignment kernel and value of average sequence divergence in the population we evaluated the fraction of breakpoints that have displacement greater than the default minimal pancontig size for Pangraph of 100 bp (cf.
 Fig.~\ref{fig:toy-accuracy}).
@@
     \setlength{\tabcolsep}{6pt}
-    \begin{tabular}{l c c c c c }
+    \begin{tabular}{l c c c c c c}
         \hline\hline
-        Species                  & N.  & $\avg{\Lgen}$ & $\Lpang$ & $\Lcore$ & $\dcore$ \\
-                                 &     & [Mbp]         & [Mbp]    & [Mbp]    &          \\
+        Species                  & N.  & $\avg{\Lgen}$ & $\Lpang$ & $\Lcore$ & $\Lsoftcore$ & $\dcore$ \\
+                                 &     & [Mbp]         & [Mbp]    & [Mbp]    & [Mbp]        & \\
         \hline
-        \textit{E.~coli}         & 307 & 5.0           & 17.7     & 0.7      & 1.6\%    \\
-        \textit{K.~pneumoniae}   & 109 & 5.3           & 13.0     & 2.3      & 0.8\%    \\
-        \textit{H.~pylori}       & 85  & 1.6           & 1.9      & 0.6      & 4.2\%    \\
-        \textit{M.~tuberculosis} & 51  & 4.4           & 4.1      & 2.4      & 0.03\%   \\
-        \textit{P.~marinus}      & 10  & 1.8           & 3.3      & 0.7      & 26.9\%   \\
+        \textit{E.~coli}         & 307 & 5.0           & 17.7     & 0.7      & 2.0          & 1.6\%    \\
+        \textit{K.~pneumoniae}   & 109 & 5.3           & 13.0     & 2.3      & 3.7          & 0.8\%    \\
+        \textit{H.~pylori}       & 85  & 1.6           & 1.9      & 0.6      & 1.0          & 4.2\%    \\
+        \textit{M.~tuberculosis} & 51  & 4.4           & 4.1      & 2.4      & 3.4          & 0.03\%   \\
+        \textit{P.~marinus}      & 10  & 1.8           & 3.3      & 0.7      & 0.7          & 26.9\%   \\
         \hline
@@
         total core genome length ($\Lcore$),
+        total soft-core genome length ($\Lsoftcore$, corresponding to genes shared by at least 95\% of the isolates), and
         average pairwise divergence of core genes ($\dcore$).
@@
 
-For species other than \textit{P.~marinus}, the pangraphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in blocks spanning several kbps (see SI Fig.~4).
+For species other than \textit{P.~marinus}, the pangraphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in pancontigs spanning several kbps (see SI Fig.~4).
 For these species the use of null energy parameters slightly increases merging and sequence compression, at the cost of slightly more fragmented graphs (cf. SI Section IIIC and SI Fig.~4).
@@
         \textbf{C}: sequence compression, defined as the ratio between the pangenome graph size and the cumulative size of all the sequences contained in the graph.
-For reference, the inverse of this number is marked.
+Since maximal compression depends on the number of isolates $n$ in the pangenome graph, we mark for reference the value of $1/n$ for each dataset.
     }
@@
 Results are reported in Fig.~\ref{fig:panx-size}.\\
-The number and size of pancontigs scale sub-linearly with the number of isolates, with more than 50\% of the pangenome being included in the 10\% longest pancontigs, and core pancontigs having higher-than-average size.
+The number and size of pancontigs scale sub-linearly with the number of isolates, with more than 50\% of the pangenome being included in the 10\% longest pancontigs.
+Core pancontigs having higher-than-average length.
+This is expected given their strong synteny and the fact that they present less structural variation across genomes compared to accessory genes.
 Adding more strains does not generate excessive fragmentation of these pancontigs, and their size remains of several kbps even as hundreds of strains are added.
@@
 The last line represents the distribution of shared sequence, approximated using the fraction of shared k-mers corrected using sequence divergence as described in the main text.
-        \textbf{C}: distribution of average segment lengths for each category over the 50 pairs considered (2 entries per pair).
+Next to each distribution we report its mean and standard deviation.
+        \textbf{C}: distribution of average segment lengths for each category over the 50 pairs considered (2 entries per pair). Mean and standard deviation are reported.
     }

### 2023-02-22 supplementary.tex
Update on Overleaf.

@@
 
-In this section we explain in more detail the procedure used to merge two pangenome graphs together. This procedure is at the core of PanGraphs's algorithm.
+In this section we describe in more detail the procedure used to merge two pangenome graphs together. This procedure is at the core of \textit{PanGraphs}'s algorithm.
 
@@
 
-As a preliminary step in the explanation we spend few words on the pangenome graph representation of pangraph.
+As a preliminary step in the explanation we spend few words on the pangenome graph representation of \textit{PanGraph}.
 A pangenome graph is essentially composed of two elements: pancontigs, which encode for alignment of homologous sequences, and paths, that correspond to representation for genomes as a sequence of pancontigs.
@@
 Each \textbf{pancontig} contains the full information on the alignment of homologous sequences it contains.
-In PanGraph each pancontig is econded with a consensus sequence, and each line of the alignment is saved as variations over the consensus (mutations, insertions and deletions).
+In PanGraph each pancontig is encoded with a consensus sequence, and each line of the alignment is saved as variations over this consensus (mutations, insertions and deletions).
 More information on the precise encoding used can be found in PanGraph documentation.\footnote{See \url{https://neherlab.github.io/pangraph/tutorials/tutorial_2/\#From-block-consensus-to-node-alignment}}
@@
 \begin{itemize}
-    \item the minimal block length $\Lthr$.
-    \item pseudo-energy parameters $\alpha$ and $\beta$ (cf. eq.~1 in the main text).
+    \item the minimal pancontig length $\Lthr$.
+    \item pseudo-energy parameters $\alpha$ and $\beta$ (cf. eq.~(1)in the main text).
     \item the choice of alignment kernel (minimap2 or mmseqs) and aligner precision (asm5, asm10 or asm20 for minimap2).
@@
 This can happen for example when dealing with circular sequences such as bacterial chromosomes.
-These sequences are transformed into a single linear block at the beginning of the algorithm, which introduces an artificial cut.
-When this elementary graph is propagated and the unique block is split, a transitive edge is generated across the artificial cut.
+These sequences are transformed into a single linear pancontig at the beginning of the algorithm, which introduces an artificial cut.
+When this elementary graph is propagated and the unique pancontig is split, a transitive edge is generated across the artificial cut.
 We remove it in this step.
 \end{enumerate}
+
+As graphs are propagated along the guide tree more and more merges are performed. This makes pancontigs progressively shorter but their alignments deeper.\\
+We also point out that in our merging procedure only the consensus sequence of any two pancontigs is used in their alignment.
+This makes this procedure fast and is sufficient to capture the large-scale structural variation of the pangenome, but does not solve the multiple sequence alignment problem for each pancontig alignment.
+To this end \textit{PanGraph} has the command \verb|polish|\footnote{See \url{https://neherlab.github.io/pangraph/tutorials/tutorial_3/}.}.
+This command ``polishes'' alignments by solving on each pancontig the multiple sequence alignment problem using MAFFT \cite{katoh2002mafft}.
+It has a simple interface: it takes as input a pangenome graph and produces another pangenome graph with the same structure but improved pancontig alignments.
+This efficiently separates the problem of finding large-scale structural variation, and nucleotide-level variation.
+When precise knowledge of the latter is needed, a graph can be \textit{polished} restricting the multiple sequence alignment problem only on the homologous sequences contained in each pancontig.
 
@@
 
-We conclude this section with some comments on the effect of PanGraph's parameters on the resulting pangenome graphs.
+After explaining the procedure we use for merging, we conclude this section with some comments on the effect of PanGraph's parameters on the resulting pangenome graphs.\\
 
-The value of $\Lthr$ controls the minimal size of blocks.
-Indels shorter than this size are saved as variation in alignments, rather than as separate blocks.
+Parameter $\Lthr$ controls the minimal size of pancontigs.
+Indels shorter than this size are saved as variation in alignments, rather than as separate pancontigs.
 As such this parameter partially controls the fragmentation of the graph.
-It should be set
-The default value in PanGraph is set to 100 bp.
-A similar role is played by the parameter $\alpha$. This can be interpreted as the cost of...
-% discuss alpha with this?
+The appropriate value depends on the scale at which one wants to study structural variations in genomes.
+The default value in PanGraph is set to 100 bp.\\
+A similar role is played by the parameter $\alpha$.
+This can be interpreted as the cost of introducing a cut in a pancontig.
+The length of homologous sequence to be merged must exceed this cost for a cut to be performed.\\
 
-The value of $\beta$ and
+The choice of alignment kernel and the value of $\beta$ set instead a limit on the divergence of sequences that are merged in the same pancontig.
+The first intrinsic limit is set by the choice of alignment kernel and its sensitivity: minimap2 is significantly faster than mmseqs2, but it is in general unable to find matches between homologous sequences with more than 20\% divergence (with asm20 sensitivity).\\
+On top of this intrinsic limit of the aligner, a more fine-tuned setting of the acceptable amount of divergence can be achieved by tuning the value of $\beta$.
+From the definition of the pseudo-energy (cf.
+eq.~(1) in the main text) it follows that matches between homologous sequence with divergence higher than $1/\beta$ will have negative pseudo-energy and will thus not result in a merge.
+If the molecular clock hypothesis holds, tuning this parameter is equivalent to setting an effective time limit beyond which sequences are not to be merged in the same alignment.
+This parameter can be tuned based on the diversity of the sample in study.
+It can for example be useful in resolving situations in which the hypothesis of a star-like phylogeny in a pancontig alignment is broken, such as to separate loosely related paralogs.
+
 % scale separation effect. Maybe move somewhere else?
@@
 % references to SI in the main text.
+% temporal horizon set with beta.
 
@@
 
-As a first test we benchmark the performances and accuracy of \textit{PanGraph} on synthetic data, obtained by simulating the evolution of a population in which individuals accumulate mutations, can loose part of the sequence from deletions, or gain new genetic material via Horizontal Gene Transfer (HGT) from other members of the population.\\
+As a first test we benchmark the performances and accuracy of \textit{PanGraph} on synthetic data, obtained by simulating the evolution of a population in which individuals accumulate mutations, can loose part of the sequence from deletions, or gain new genetic material via Horizontal Sequence Transfer (HST) from other members of the population.\\
 In the simulation a population of size $N$ is evolved for $T$ generations using a Wright-Fisher model \cite{hudson2002generating}.
@@
 This ensures that the length distribution of genomes in the population remains close to the desired length $L$, with $\sigma^2$ being a proxy for the variance.
-HGT occurs at a rate $h$ per genome per generation.
-Similarly to deletions, when a HGT event is suggested a new desired length $L'_n$ is extracted from the same Gaussian distribution, and this time the event is performed only if $\Delta = L'_n - L_n > 0$.
+HST occurs at a rate $h$ per genome per generation.
+Similarly to deletions, when a HST event is suggested a new desired length $L'_n$ is extracted from the same Gaussian distribution, and this time the event is performed only if $\Delta = L'_n - L_n > 0$.
 In this case a random chunk of sequence of length $\Delta$ is extracted from a random individual in the parent population (possibly the parent of isolate $n$ itself) and inserted in a random position of the genome of isolate $n$.
+All parts of the sequence have equal probability of being transferred, irrespective of homology.
 Finally, inversions occur at a rate $i$ per genome per generation.
-When an inversion is proposed a random length $\Delta$ is chosen in the same way, and a random chunk of length $|\Delta|$ is inverted.
+When an inversion is proposed a random length $\Delta$ is chosen in the same way, and a random chunk of sequence of this length is inverted.
+In the simulation we do not implement a notion of genes.
 Standard values of the simulation parameters are reported in Table~\ref{table:sim-params}.
@@
         $i$       & inversion rate              & 0.01 [1/generation]             \\
-        $h$       & HGT rate                    & 0.1 [1/generation]              \\
+        $h$       & HST rate                    & 0.1 [1/generation]              \\
         \hline
@@
 In Fig.~\ref{fig:marginalize-suppl} we display the distribution of these quantities across all pairs for the different species considered and the block types described above.\\
-In all species except HP and PM we find very good agreement between pairwise and marginalized graphs: segments on which the graphs \textit{disagree} represent only a very minor fraction of the genome (much smaller than the fraction of shared or private blocks), and have size compatible with the default block length threshold of \textit{PanGraph} (100 bp).
+In all species except HP and PM we find very good agreement between pairwise and marginalized graphs: segments on which the graphs \textit{disagree} represent only a very minor fraction of the genome (much smaller than the fraction of shared or private blocks), and have size compatible with the default pancontig length threshold of \textit{PanGraph} (100 bp).
 As described in the previous section, the divergence of the HP dataset sits on the limit of what the \textit{minimap2} alignment kernel can successfully handle.

### 2023-02-23 reviewer_response.tex
chore: update

@@
 You surely do not do all pairwise descriptions of SNPS, right?
-On that same webpage you have a section on "block consensus to node alignment" - how do you choose the order of alleles here?
+On that same webpage you have a section on ``block consensus to node alignment'' - how do you choose the order of alleles here?
 This is all valuable information and it would be great to have it in the supplementary or main methods somewhere.}
@@
 Finding the set of shared minimizers would require comparing $\sim n^2$ sets of ordered lists of minimizers, which scales as $\sim n^2 l$.
-This would provide a scaling $O(n l \log l + n^2 l)$\\
-In practice in our implementation we achieve a better scaling by sketching and sorting all the genomes together, which can be done in time $L \log L$, where $L= n \times l$ is the total length of all genomes.
+This would provide a scaling $O(n l \log l + n^2 l)$.\\
+      In our implementation we achieve a slightly better scaling as a function of $n$ by sketching and sorting all the genomes together, which can be done in time $L \log L$, where $L= n \times l$ is the total length of all genomes.
 We keep the information on each minimizer origin in the list.
 We then scroll the list and for each chunk of equal minimizers we update the corresponding entries of a $n^2$ matrix of minimizer hit counts.
-Depending on the amount of shared minimizers, this scales...\\
-% In practice we achieve a scaling that is between $L \log L$ and $n L \log L$.
-We stress however that the computational time to build the guide tree is usually negligible compared to the time required for graph merging.
+Depending on the amount of shared minimizers, this scales as $n^2 l$ in the worst case (all minimizers shared) and as $n l$ in the best case (no minimizer shared).
+Therefore depending on the diversity of the sample the scaling will be between $O(L \log L + n L)$ and $O(L \log L)$.\\
+      In practice for typical use cases of pangraph ($n \sim 100$ and $L \sim 10^6$) the computationally challenging part is the minimizer sketching, and not the distance matrix computation.
+This is partly due to some prefactors (e.g.
+the window size of 100 bp used for sketching makes so that the list of minimizer is significantly shorter than the genome, and the only operation performed $n^2$ times is a single increment of an integer in the minimizer hit matrix).
+We also stress that the computational time to build the guide tree is usually negligible compared to the time required for the graph merging procedure.\\
+      To correct this ambiguity we...
       % missing: changes in main text. My plan: report the n^2 l log l scaling. 
@@
 As i understand it, you merge them in order of pseudoenergy ranking, which makes sense (what do you do for precise duplicates, where a region on genome 1 is a precise match for 2 regions on genome 2?
-random choice?)\\
-                  \Marco{the ranking would be the same and we perform only one of the two merges in the first phase, when comparing the two graphs.
+random choice?)
+                  % \Marco{the ranking would be the same and we perform only one of the two merges in the first phase, when comparing the two graphs.
 After this we align the graph on itself and the second merge should be performed at this point.
@@
 The only thing i can imagine you are doing is only looking for homology between the genomes in one graph, and the genomes in the second graph.
-I would really appreciate it if you could clarify this and even ideally add some example figures in the supplement showing with some toy genomes how you align two graphs.\\
-                  \Marco{another comment pointing at the fact that we could add a more detailed explanation about graph merging on the supplementary.
+I would really appreciate it if you could clarify this and even ideally add some example figures in the supplement showing with some toy genomes how you align two graphs.
+                  % \Marco{another comment pointing at the fact that we could add a more detailed explanation about graph merging on the supplementary.
 What we do is using the aligner using as reference and queries the consensus sequences of each block in the two different graphs, and create a first version of the merged graph.
@@
 This is necessary because when two alignments overlap only one of the two is performed in the first alignment round, but the second will be performed in further rounds.}\\
-                  \Liam{agree, sounds like a section in supplementary is a good idea to address these comments.}
+                  % \Liam{agree, sounds like a section in supplementary is a good idea to address these comments.}
       \end{itemize}}
+\response{We followed the advice of the reviewer and added Section I in the SI, in which we expand on the explanation of how we merge two graphs.\\
+      In short, we perform the merging procedure in two steps.
+In the first we use the aligner (minimap2 or mmseqs2) to scan for homology matches between the consensus sequence of the pancontigs of one graph and of the other.
+These matches are ranked by pseudo-energy and performed in order.
+Each pancontig can be subject to only one merge at this stage.\\
+      To answer the first point, for precise duplicates only one of the two matches is merged at this stage, with pseudo-random ordering.
+However also the second region will be matched and merged before the merging of the two graphs is conlcuded.\\
+      After all of the matches have been exhausted, the two graphs have undergone a preliminary merging, with some new merged pancontigs being created.
+At this point we repeat the procedure looking for matches on the consensus sequence of all the pancontigs in an all-against-all comparison.
+This self-mapping on the graph will detect any homology that could not have been merged at the previous stage, and will perform the corresponding merging.
+The self-mapping is repeated iteratively until no new merges can be performed.
+}
 
 \reviewer{R2.4}{Line 92-94, same section. You give two statements:
-      "Only mergers whose alignment has negative pseudo-energy are performed"
+      ``Only mergers whose alignment has negative pseudo-energy are performed''
       and
-      "no mergers will be performed if the Hamming distance exceeds $1/\beta$."
+      ``no mergers will be performed if the Hamming distance exceeds $1/\beta$.''
       These are separate criteria that prevent mergers, right?
@@
 Not sure if that tells us anything.}
-\Marco{the second is actually a consequence of the first, but except for that it is correct.}\\
-\Liam{maybe say `$\beta$ controls the maximal sequence divergence between the two merger-candidates.
+% \Marco{the second is actually a consequence of the first, but except for that it is correct.}\\
+% \Liam{maybe say `$\beta$ controls the maximal sequence divergence between the two merger-candidates.
 The negative pseudo-energy requirement means that no merger will be performed if the Hamming distance exceeds $1/\beta$.'}\\
-\Marco{indeed, the parameter $\beta$ could control how far in the past one wants to accept merges. Should we mention this explicilty in the text?}\\
-\Liam{could be mentioned as an `intuitively, this should correspond to merging of homologous stretches of sequence in roughly reverse evolutionary order'.}
+% \Marco{indeed, the parameter $\beta$ could control how far in the past one wants to accept merges. Should we mention this explicilty in the text?}\\
+% \Liam{could be mentioned as an `intuitively, this should correspond to merging of homologous stretches of sequence in roughly reverse evolutionary order'.}\\
+\response{
+      In this case the second criterion is a consequence of the first.
+From the definition of the pseudo-energy (cf.
+eq.~(1) in the main text) it follows that for an alignment of pancontig consensus sequences to have negative pseudo-energy must be that the number of SNPs divided by the alignment length $N_m/l$ must be less than $1/\beta$.
+Moreover, as the reviewer pointed out, the pseudo-energy ranking of merges effectively results in merges being performed in order of similarity, effectively in reverse evolutionary order.\\
+      To make this more explicit we modified lines 93-95 in ``$\beta$ controls the maximal sequence divergence between the two merger-candidates.
+The negative pseudo-energy requirement means that no merger will be performed if the Hamming distance exceeds $1/\beta$...
+The pseudo-energy ranking also effectively results in merges being performed in reverse evolutionary order, with less diverged sequences being merged first.''
+      % QUESTION: is there ambiguity in the main text on the value of N_m? As SNPs in the alignment? Misleading figure?
+}
 
@@
 It will also mean you make more false merges, right?}
-\Marco{setting $\beta$ to zero will indeed remove the 10\% divergence limit set by the pseudo-energy.
+% \Marco{setting $\beta$ to zero will indeed remove the 10\% divergence limit set by the pseudo-energy.
 There is another limit that depends on the aligner sensitivity (20 to 30\%, depending on the aligner).
@@
 Not sure I interpret this correctly.}\\
-\Liam{I think your interpretation is right. Important to clarify that there are those two limits and that $\beta$ only controls the divergence limit \textit{within} what one can actually align.}
+% \Liam{I think your interpretation is right. Important to clarify that there are those two limits and that $\beta$ only controls the divergence limit \textit{within} what one can actually align.}
+\response{There are two factors limiting the divergence of merges that are performed.
+The first factor is the aligner used.
+Minimap2 is in general able to find matches up to roughly 20\% sequence divergence, while mmseqs2 can reach divergence up to roughly 30\%.
+Only homology matches of up to this divergence will be suggested by the aligners.
+On top of this, only matches with negative pseudo-energy are turned into merges.
+For any given value of $\beta$, this sets an additional limit to the divergence, which must be less than $1/\beta$.\\
+      This is now made more explicit in SI section I (PanGraph parameters effect).\\
+      When setting $\beta = 0$ only the intrinsic limit on divergence posed by the aligner sensitivity is left.
+In this sense more merges are allowed, but all mergess will have some level of homology (maximum 20\% or 30\% divergence, depending on the aligner).}
 
 \reviewer{R2.7}{Why does the L50 peak go up for null-scores for mmseqs for Mtb, Pm, Kpn and Ec (Figure 4E, purple with lines) but down for helicobacter?}
-\Marco{the effect of merging more blocks on L50 is not trivial.
+% \Marco{the effect of merging more blocks on L50 is not trivial.
 It depends on whether the merging occurs between long or short blocks, and how much fragmentation this causes.
@@
 I could do the same for EC or KP for comparison.}\\
-\Liam{makes sense.}
+% \Liam{makes sense.}
+\response{L50 measures the minimum number of pancontigs needed to cover at least 50\% of the pangenome graph sequence.
+This number is closely related to the total number of pancontigs.
+In general the use of null energy parameters result in more fragmented graphs, with a higher number of pancontigs (cf.
+fig.~4D in SI).
+Hp is an exception and has the opposite trend, both for the number of pancontig and for the value of L50.
+For this species the use of null energy parameters results in less fragmentation of the graph.
+This is compatible with the observation that the average divergence of this dataset is on the limit of what PanGraph can handle.
+Unless a highly sensitive aligner is used and the divergence barrier in pseudo-energy is removed, some highly-diverged homologous sequences remain split in different pancontigs.
+Conversely, the choice of these parameters generates a graph with less pancontigs, a lower L50, a larger fraction of core pangenome, an higher median length of pancontigs (N50).
+      % TODO: add mention in SI
+      % TODO: plot change for HP and KP/EC as comparison?
+      % QUESTION: this is more or less explicit in SI? Should I expand more?
+}
 
-\reviewer{R2.8}{Just to make sure i understand - is it possible for sequence to be "left out" in between pancontigs?}
+\reviewer{R2.8}{Just to make sure i understand - is it possible for sequence to be ``left out'' in between pancontigs?}
 \Marco{no, every part of the genome must be contained in a pancontig. It can be left out from the consensus if it is considered an insertion.}\\

### 2023-02-23 main.tex
chore: update

@@
 \usepackage{enumerate}
-\usepackage{color}
+\usepackage{xcolor}
+\usepackage[normalem]{ulem}
 \usepackage[final]{hyperref}
@@
 \newcommand{\Lgen}{L_\text{gen}}
+\newcommand{\Lthr}{L_{\min}}
 \newcommand{\dcore}{\langle d_\text{core} \rangle}
-\newcommand{\Richard}[1]{{\color{red}Richard: #1}}
+\newcommand{\modified}[1]{{\color{red}#1}}
+\newcommand{\oldnew}[2]{{\color{gray}\sout{#1}}{\color{red}#2}}
 
@@
 \paragraph*{Merging of homologous sequence.}
-If the above alignment step detected homologous stretches between two pancontigs, these pancontigs, or parts of them, can be merged.
+If the above alignment step detected homologous stretches between \oldnew{}{the consensus sequences of} two pancontigs, these pancontigs, or parts of them, can be merged.
 It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order.
@@
 Only mergers whose alignment has negative pseudo-energy are performed.
-The parameter $\beta$ controls the maximal sequence divergence between the two merger-candidates and no mergers will be performed if the Hamming distance exceeds $1/\beta$.
+The parameter $\beta$ controls the maximal sequence divergence between the two merger-candidates\oldnew{ and no mergers will be performed if the Hamming distance exceeds $1/\beta$.}{.
+The negative pseudo-energy requirement means that no merger will be performed if the Hamming distance exceeds $1/\beta$.}
 Importantly, this divergence threshold is applied to the comparison of the consensus sequences of the merger candidates.
-The parameter $\alpha$ controls the fragmentation of the graph and imposes a minimal length on the resulting pan-contigs resulting from the merger.
-The merger is always performed if the merger doesn't increase the number of contigs, but mergers that introduce 1, 2, 3 or 4 cuts are only performed for sufficiently long homologous stretches as parameterized by $\alpha$.
-In addition, there is parameter that controls the minimal size of contigs produced by the algorithm.
+\oldnew{}{The pseudo-energy ranking also effectively results in merges being performed in reverse evolutionary order, with less diverged sequences being merged first.}
+The parameter $\alpha$ controls the fragmentation of the graph and imposes a minimal length on the \oldnew{resulting}{} pan-contigs resulting from the merger.
+The merger is always performed if \oldnew{the merger}{it} doesn't increase the number of contigs, but mergers that introduce 1, 2, 3 or 4 cuts are only performed for sufficiently long homologous stretches as parameterized by $\alpha$.
+\oldnew{In addition, there is parameter that controls the minimal size of contigs produced by the algorithm.}{In addition to the fragmentation parameter $\alpha$, there is a further parameter $\Lthr$ that controls the minimal length of pancontigs.
+Structural variation at a scale smaller than this threshold (i.e.
+short indels) is stored in pancontig alignments, rather than in the graph structure, and no merges are performed if they have length shorter than this threshold.
+We set a default value of 100 bp for this parameter.}
 
@@
 Upon completion, transitive edges within the graph, i.e., edges along which all path are co-linear, are removed by merging adjacent \emph{pancontigs}.
+\oldnew{}{More details on the merging procedure and the effect of parameters can be found in SI section I.}
 
@@
 In each comparison we evaluate the misplacement of breakpoints that we can pair on the two graphs within 1kbp.
-The plot displays the fraction of breakpoints that have misplacement greater than the standard \textit{PanGraph} precision threshold of 100bp, as a function of average pairwise sequence divergence.
+The plot displays the fraction of breakpoints that have misplacement greater than the standard \textit{PanGraph} precision threshold of \oldnew{}{$\Lthr=$}100bp, as a function of average pairwise sequence divergence.
 Line and shaded area represents mean and standard deviation over 25 repetitions.
@@
         total core genome length ($\Lcore$),
-        total soft-core genome length ($\Lsoftcore$, corresponding to genes shared by at least 95\% of the isolates), and
-        average pairwise divergence of core genes ($\dcore$).
+        \oldnew{}{total soft-core genome length ($\Lsoftcore$, corresponding to genes shared by at least 95\% of the isolates),}
+        and average pairwise divergence of core genes ($\dcore$).
     }
@@
 The compatibility between the two graphs requires segments on which the two partitions disagree to be few and short.
-Indeed, we verified that over the pairs we picked, these segments cover a very small fraction of the genome ($<1\%$) and have average size compatible with the default 100 bp precision threshold of \textit{PanGraph} (cf.
+Indeed, we verified that over the pairs we picked, these segments cover a very small fraction of the genome ($<1\%$) and have average size compatible with the default 100 bp precision threshold of \textit{PanGraph} \oldnew{}{set by the value of $\Lthr$} (cf.
 Fig.~\ref{fig:marginalization}B-C).

### 2023-02-23 supplementary.tex
chore: update

@@
 \usepackage{xr} % reference to other external latex documents
+\usepackage[normalem]{ulem}
+\usepackage{xcolor}
 \usepackage[final]{hyperref}
@@
 \newcommand{\Lthr}{L_{\min}}
+\newcommand{\oldnew}[2]{{\color{gray}\sout{#1}}{\color{red}#2}}
 
@@
 Each match is characterized by an alignment length and a measure of diversity.
-Using this information we rank matches by pseudo energy, so that longer and less diverged matches will be processed first.
-We refuse all matches with negative pseudo-energu and all matches shorter than the threshold length $\Lthr$.
+Using this information we rank matches by pseudo-energy, so that longer and less diverged matches will be processed first.
+We refuse all matches with positive pseudo-energy and all matches shorter than the threshold length $\Lthr$.
     \item For each of the remaining matches we attempt to merge pancontigs following these steps:
@@
               \item We check that no prior merges have been performed on the same two pancontigs.
-Sometimes multiple matches are found on the same pancontig, and in this case we prioritize the ones with higher pseudo-energy.
+Sometimes multiple matches are found on the same pancontig, and in this case we prioritize the ones with more negative pseudo-energy.
 Any following match on the same pancontig will be discarded at this step.
@@
 In Fig.~\ref{fig:marginalize-suppl} we display the distribution of these quantities across all pairs for the different species considered and the block types described above.\\
-In all species except HP and PM we find very good agreement between pairwise and marginalized graphs: segments on which the graphs \textit{disagree} represent only a very minor fraction of the genome (much smaller than the fraction of shared or private blocks), and have size compatible with the default pancontig length threshold of \textit{PanGraph} (100 bp).
+In all species except HP and PM we find very good agreement between pairwise and marginalized graphs: segments on which the graphs \textit{disagree} represent only a very minor fraction of the genome (much smaller than the fraction of shared or private blocks), and have size compatible with the default pancontig length threshold of \textit{PanGraph} (\oldnew{}{$\Lthr =$}100 bp).
 As described in the previous section, the divergence of the HP dataset sits on the limit of what the \textit{minimap2} alignment kernel can successfully handle.

### 2023-02-23 reviewer_response.tex
update

@@
 \usepackage{geometry}
+\usepackage{graphicx}
 \usepackage{hyperref}
@@
 Unless a highly sensitive aligner is used and the divergence barrier in pseudo-energy is removed, some highly-diverged homologous sequences remain split in different pancontigs.
-Conversely, the choice of these parameters generates a graph with less pancontigs, a lower L50, a larger fraction of core pangenome, an higher median length of pancontigs (N50).
+Conversely, the choice of these parameters generates a graph with less pancontigs, a lower L50, a larger fraction of core pangenome, an higher median length of pancontigs (N50).\\
+      For a visual verification of this interpretation we display in fig.~\ref{fig:kp-vs-hp} how the distributions of pancontig lenght and frequency change with the use of different alignment kernels and pseudo-energy parameters.
+We compare the changes observed on the Hp dataset to the ones in Kp, used as a reference.
+In the Kp dataset increasing the aligner sensitivity results in shorter pancontigs (lower-right panel) and a small increase in core genome fraction.
+For Hp instead a more sensitive aligner results in the loss of small and intermediate size pancontigs, but not of long ones.
+It also results in a loss of low-frequency pancontigs, indicating that these had an homologous counterpart that could not be merged due to high-divergence, and the core pangenome fraction increases considerably.
       % TODO: add mention in SI
-      % TODO: plot change for HP and KP/EC as comparison?
       % QUESTION: this is more or less explicit in SI? Should I expand more?
 }
+
+\begin{figure}[htb]
+      \includegraphics[width=.7\textwidth]{figs_response/hp_vs_kp.pdf}
+      \caption{frequency and length distribution (respectively left and right panels) of pancontigs for the \textit{Helicobacter pylori} and \textit{Klebsiella pneumoniae} datasets (top and bottom).
+For each panel we report the results for pangenome graphs built using minimap2 aligner with standard parameter value and asm20 sensitivity (blue) and with mmseqs2 and null pseudo-energy parameters (orange).
+Distributions are weighed by pancontig length, so that the area under the curve represents the total pangenome graph size.}
+      \label{fig:kp-vs-hp}
+\end{figure}
 
@@
       \end{enumerate}}
-\Marco{on the repo we have \href{https://github.com/neherlab/pangraph/blob/master/script/config/accnums.json}{this file} for the list of all accession numbers we use, and \href{https://github.com/neherlab/pangraph/blob/master/script/config/projection_strains.json}{this file} for the marginalization pairs.
+% \Marco{on the repo we have \href{https://github.com/neherlab/pangraph/blob/master/script/config/accnums.json}{this file} for the list of all accession numbers we use, and \href{https://github.com/neherlab/pangraph/blob/master/script/config/projection_strains.json}{this file} for the marginalization pairs.
 I could create some excel spreadsheet with accession numbers that we could add to the submission?}\\
-\Liam{for sure that's all they want (most biologists don't like json!)}
+% \Liam{for sure that's all they want (most biologists don't like json!)}
+\responsedraft{We now included in the submission two excel files.
+``accession\_numbers.xlsx'' contains five sheets, one per species considered, each with a list of the accession numbers of isolates used in our datasets.
+``marginalize\_pairs.xlsx'' contains instead a list of all the pairs used in the graph marginalization section.
+      % TODO: archive repository state.
+}
 

### 2023-02-23 main.tex
update

@@
 The number and size of pancontigs scale sub-linearly with the number of isolates, with more than 50\% of the pangenome being included in the 10\% longest pancontigs.
-Core pancontigs having higher-than-average length.
-This is expected given their strong synteny and the fact that they present less structural variation across genomes compared to accessory genes.
-Adding more strains does not generate excessive fragmentation of these pancontigs, and their size remains of several kbps even as hundreds of strains are added.
+\oldnew{, and core pancontigs having higher-than-average size}{Core pancontigs have higher-than-average length.
+This is expected given their strong synteny and the fact that they present less structural variation across genomes compared to accessory genes.} Adding more strains does not generate excessive fragmentation of these pancontigs, and their size remains of several kbps even as hundreds of strains are added.
 The size of the core genome does not decrease significantly with the addition of new strains, while the total pangenome size remains orders of magnitude smaller than the total size of all genomes included in the graph.
@@
 The last line represents the distribution of shared sequence, approximated using the fraction of shared k-mers corrected using sequence divergence as described in the main text.
-Next to each distribution we report its mean and standard deviation.
+\oldnew{}{Next to each distribution we report its mean and standard deviation.}
         \textbf{C}: distribution of average segment lengths for each category over the 50 pairs considered (2 entries per pair). Mean and standard deviation are reported.

### 2023-02-23 supplementary.tex
update

@@
 
-As a first test we benchmark the performances and accuracy of \textit{PanGraph} on synthetic data, obtained by simulating the evolution of a population in which individuals accumulate mutations, can loose part of the sequence from deletions, or gain new genetic material via Horizontal Sequence Transfer (HST) from other members of the population.\\
+As a first test we benchmark the performances and accuracy of \textit{PanGraph} on synthetic data, obtained by simulating the evolution of a population in which individuals accumulate mutations, can loose part of the sequence from deletions, or gain new genetic material via Horizontal \oldnew{Gene}{Sequence} Transfer (HST) from other members of the population.\\
 In the simulation a population of size $N$ is evolved for $T$ generations using a Wright-Fisher model \cite{hudson2002generating}.
@@
 In this case a random chunk of sequence of length $\Delta$ is extracted from a random individual in the parent population (possibly the parent of isolate $n$ itself) and inserted in a random position of the genome of isolate $n$.
-All parts of the sequence have equal probability of being transferred, irrespective of homology.
+\oldnew{}{All parts of the sequence have equal probability of being transferred, irrespective of homology.}
 Finally, inversions occur at a rate $i$ per genome per generation.
 When an inversion is proposed a random length $\Delta$ is chosen in the same way, and a random chunk of sequence of this length is inverted.
-In the simulation we do not implement a notion of genes.
+\oldnew{}{In the simulation we do not implement a notion of genes}.
 Standard values of the simulation parameters are reported in Table~\ref{table:sim-params}.

### 2023-02-23 main.tex
chore: marked changes

@@
 The graph is composed of \emph{pancontigs}, which represent linear multiple-sequence alignments of homologous sequence found within one or more input genomes.
-\emph{Pancontigs} are connected by an edge if they are adjacent in at least one input sequence; individual sequences are then recapitulated by contiguous \emph{paths} through the graph.
+\emph{Pancontigs} are connected by an edge if they are \oldnew{syntenic on}{adjacent in} at least one input sequence; individual sequences are then recapitulated by contiguous \emph{paths} through the graph.
 Pancontigs are directed, meaning their orientation in a path can be either 5' to 3' or vice versa.
@@
 
-For species other than \textit{P.~marinus}, the pangraphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in pancontigs spanning several kbps (see SI Fig.~4).
+For species other than \textit{P.~marinus}, the pangraphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in \oldnew{blocks}{pancontigs} spanning several kbps (see SI Fig.~4).
 For these species the use of null energy parameters slightly increases merging and sequence compression, at the cost of slightly more fragmented graphs (cf. SI Section IIIC and SI Fig.~4).
@@
         \textbf{C}: sequence compression, defined as the ratio between the pangenome graph size and the cumulative size of all the sequences contained in the graph.
-Since maximal compression depends on the number of isolates $n$ in the pangenome graph, we mark for reference the value of $1/n$ for each dataset.
+\oldnew{}{Since maximal compression depends on the number of isolates $n$ in the pangenome graph, we mark for reference the value of $1/n$ for each dataset.}
     }
@@
 \oldnew{}{Next to each distribution we report its mean and standard deviation.}
-        \textbf{C}: distribution of average segment lengths for each category over the 50 pairs considered (2 entries per pair). Mean and standard deviation are reported.
+        \textbf{C}: distribution of average segment lengths for each category over the 50 pairs considered (2 entries per pair). \oldnew{}{Mean and standard deviation are reported.}
     }

### 2023-02-23 reviewer_response.tex
chore: marked changes
update

@@
 \usepackage{xcolor}
+
+\newcommand{\Lthr}{L_{\min}}
 
@@
 \reviewer{R2.8}{Just to make sure i understand - is it possible for sequence to be ``left out'' in between pancontigs?}
-\Marco{no, every part of the genome must be contained in a pancontig. It can be left out from the consensus if it is considered an insertion.}\\
-\Liam{if they possibly misunderstood, suggests that some statement that all sequence ends up in a pancontig should be added to manuscript.
+% \Marco{no, every part of the genome must be contained in a pancontig. It can be left out from the consensus if it is considered an insertion.}\\
+% \Liam{if they possibly misunderstood, suggests that some statement that all sequence ends up in a pancontig should be added to manuscript.
 I suspect they are thinking of case of short ($<$100bp) stretches `in between' pancontigs, which (if I understand correctly) will be incorporated as insertions at the end of the consensus alignment of the adjacent pancontig.}
+\response{No, every part of the sequence of each genome must be included in some pancontig.
+I might not necessarily appear in the pancontig consensus sequence if it is present in the alignment as a short ($< \Lthr$) insertion, but it will nonetheless be present in the alignment.
+In the particular case in which merging two pancontigs would leave overhangs shorter than $\Lthr$, these are included in the alignmen as flanking insertions.
+This is now explained in SI section I, and mentioned in the main text in section IB (line 103).}
 
@@
 
-\reviewer{R2.1}{line 13 - it would be good to give a reference for the statement that even closelt related genomes can differ widely in gene content.
+\reviewer{R2.1}{line 13 - it would be good to give a reference for the statement that even closely related genomes can differ widely in gene content.
       I don't know what the first paper to show this was, the earliest i am aware of is from Eduardo Rocha's team, which annpoyingly i can't locate right now, but Figure 2 here shows it \url{https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1008866} \cite{touchon2020phylogenetic} - you can be very close in patristic distance and have very different gene repertoire.
       Feel free to replace with a better reference, this si just meant to be helpful.}
-\Marco{I think this is a good reference, fig. 2 is familiar to us. And also fig.2 in \cite{doolittle2009origin} which is a bit older. But not sure which other Rocha paper he is referring to.}\\
-\Liam{I think they mean Figure 5 of \cite{touchon2009organised} (\href{https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1000344}{here}).
+% \Marco{I think this is a good reference, fig. 2 is familiar to us. And also fig.2 in \cite{doolittle2009origin} which is a bit older. But not sure which other Rocha paper he is referring to.}\\
+% \Liam{I think they mean Figure 5 of \cite{touchon2009organised} (\href{https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1000344}{here}).
 Citing Touchon 2009 and 2020 and the Doolittle paper seems fine to me.}
+\response{We thank the reviewer for the suggestion. We now cite the suggested article, along with \cite{touchon2009organised,doolittle2009origin}.}
 

### 2023-02-23 main.tex
update

@@
 The accumulation of such changes over time complicates comparative genomic analyses of present day isolates.
-Homologous recombination is rapid enough that most genes in many bacterial core genomes have distinct phylogenies \cite{sakoparnig2021whole} and even closely related genomes differ dramatically in gene content.
+Homologous recombination is rapid enough that most genes in many bacterial core genomes have distinct phylogenies \cite{sakoparnig2021whole} and even closely related genomes differ dramatically in gene content \oldnew{}{\cite{touchon2020phylogenetic,touchon2009organised,doolittle2009origin}}.
+% \cite{touchon2020phylogenetic,doolittle2009origin,touchon2009organised}.
 % comment: rewritten to use consistent terminology (I think 'mutation' and 'polymorphism' for most people make them think of local changes only)
@@
 short indels) is stored in pancontig alignments, rather than in the graph structure, and no merges are performed if they have length shorter than this threshold.
+For instance, when a merge would generate flanking contigs shorter than this threshold, the alignment is extended to include these regions.
 We set a default value of 100 bp for this parameter.}

### 2023-02-24 reviewer_response.tex
save progress

@@
 touching on what software can be used to map/align to such a data structure?
-Or how it can be used to analysis genome structure.}
+Or how it can be used to analyze genome structure.}
 \Marco{other than explicitly exporting block sequences and mapping to those, we do not really using the pangenome graph as a reference against which to map new strains...
@@
 \responsedraft{We agree with the reviewer that the formulation in the main text is too concise and not completely correct.
-There are two main variables at play: the length of a genomes $l$ and the number of isolates $n$.
+There are two main variables at play: the length of a genome $l$ and the number of isolates $n$.
 As the reviewer points out, the set of minimizers for each genome can be created and sorted in $l \log l$.
@@
 \reviewer{R2.2}{Why on earth is the "get an MSA of a block" option in pangraph called "polish"?}
-\Marco{i don't think this comes from the paper (I used grep on the tex file and we never mention polish).
+% \Marco{i don't think this comes from the paper (I used grep on the tex file and we never mention polish).
 I think it comes from the documentation but not sure where.
@@
 If in the paper we explain how we propagate the consensus sequence and related problems we will also mention the polish command in the same part and make explicit that it can be used to polish alignments, this should solve the ambiguity.}\\
-\Liam{this sounds good.
+% \Liam{this sounds good.
 I think they misunderstood that polish is not just `extract the MSA' but actually produces a pangraph with the same structure but redone alignments.
@@
 These can be removed at the end of the process by reconstructing the sequences in the block and performing a multiple sequence alignment (see Polish command in documentation).')}
+\response{In our method pancontig mergers are performed using only the consensus sequence as a guide.
+This makes the merging procedure very fast, but can introduce minor inconsistencies and artifacts with insertions in the alignment.
+To remove these inconsistencies, after building the graph the user can optionally decide to ``polish'' pancontig alignments with the omonymous command.
+This command takes a graph as input and produces a graph with the same topology, but with improved alignments.
+In practice for each block it collects all of the sequences and uses MAFFT \cite{katoh2002mafft} to perform a multiple sequence alignment, obtaining an improved consensus and more consistent insertion pattern.\\
+      This command was not mentioned in the paper. This has now been amended, and we briefly described in in section IB in the main text (lines ...) and at in section I in the SI (graph merging) }
 
@@
 We could also mention that the marginalize command has an option to project out paralogs, but should be a post-processing step rather than something that we want to do while building the graph.
-One can a posteriori decide whether one wants to split paralogs using structural information (flanking blocks) or one can even prevent the merging using diversity and the $\beta$ parameter.}
+One can a posteriori decide whether one wants to split paralogs using structural information (flanking blocks) or one can even prevent the merging using diversity and the $\beta$ parameter.}\\
 
@@
 I don't think we ever quantified how much the consensus sequence degenerates during the iterations.
-Maybe one could devise a proper test set and see how it degenerates over the iterations.}
+Maybe one could devise a proper test set and see how it degenerates over the iterations.}\\
 
@@
 Is there any necessary special issues with producing the merged multiple sequence alignment?}
-\Liam{don't understand their last sentence. Maybe should quantify how common the transitive edges typically are?}\\
-\Marco{I can think of one edge case which is due to periodic boundary conditions.
+% \Liam{don't understand their last sentence. Maybe should quantify how common the transitive edges typically are?}\\
+% \Marco{I can think of one edge case which is due to periodic boundary conditions.
 Initially each circular genomes is saved as a single long block.
@@
 But might be unnecessary detail when explaining the algorithm the first time, we might want to move it to the SI section where we explain merging in more detail?}\\
+\response{Transitive pancontigs are an artifact of periodic boundary conditions when merging circular genomes, such as bacterial chromosomes.
+At the beginning of the procedure the circular genome is saved as a single linear pancontig.
+This introduces an artificial cut.
+When during merging this unique pancontig is further split, the flanking pancontigs could be crossing the artificial cut in a co-linear way on all paths.
+This step of the procedure resolves the artificial cut buy merging these pancontigs together.\\
+      We think that the mention of this step of the algorithm in this part of the paper was a bit confusing and needed more clarification.
+We moved the statement from section IB of the main text to SI section I (point 5 of the merging algorithm explanation).}
 
@@
 4G for comparison with pangraph's estimation?}\\
+
 \reviewer{R3}{The authors might want to consider some form of post processing of the PGG to collapse what have been called bubbles in the graph if alignment of the pancontigs on different edges of the bubble warrant it.}

### 2023-02-24 main.tex
save progress

@@
 short indels) is stored in pancontig alignments, rather than in the graph structure, and no merges are performed if they have length shorter than this threshold.
-For instance, when a merge would generate flanking contigs shorter than this threshold, the alignment is extended to include these regions.
+For instance, when a merge would generate flanking contigs shorter than this threshold, the merging is performed but the alignment is extended to include these regions.
 We set a default value of 100 bp for this parameter.}
@@
 At the graph level, the merger of two \emph{pancontigs} defines a new \emph{pancontig}, connected on both sides by edges to the neighboring \emph{pancontigs} of both inputs, and thus locally collapses the two graphs under consideration.
-At the nucleotide level, the pairwise alignment of two \emph{pancontigs} maps the reference of one onto the other; the merger of two \emph{pancontigs} requires the application of the map onto the underlying multiple-sequence alignment.
+At the nucleotide level, the pairwise alignment of two \emph{pancontigs} maps the \oldnew{reference}{consensus} of one onto the other; the merger of two \emph{pancontigs} requires the application of the map onto the underlying multiple-sequence alignment.
 Once both sets of sequences are placed onto a common coordinate system, the resultant consensus sequence, and thus polymorphisms, are recomputed.
 This procedure can be viewed as an approximate multiple sequence alignment algorithm that aligns homologous parts without further investigating insertions in the two sub-alignments.
+\oldnew{}{This shortens considerably the time needed, but might introduce minor inconsistencies and artifacts in the alignments.
+After building the graph, these can be removed by performing a multiple sequence alignment on each pancontig  (see ``polish'' command in documentation).}
 
 The above procedure is repeated until no alignments with negative energy remain.
-Upon completion, transitive edges within the graph, i.e., edges along which all path are co-linear, are removed by merging adjacent \emph{pancontigs}.
-\oldnew{}{More details on the merging procedure and the effect of parameters can be found in SI section I.}
+\oldnew{Upon completion, transitive edges within the graph, i.e., edges along which all path are co-linear, are removed by merging adjacent \emph{pancontigs}.}{More details on the merging procedure and the effect of parameters can be found in SI section I.}
 

### 2023-02-24 supplementary.tex
save progress

@@
 We also point out that in our merging procedure only the consensus sequence of any two pancontigs is used in their alignment.
-This makes this procedure fast and is sufficient to capture the large-scale structural variation of the pangenome, but does not solve the multiple sequence alignment problem for each pancontig alignment.
+This makes this procedure fast and is sufficient to capture the large-scale structural variation of the pangenome, but might introduce minor inconsistencies and artifacts in the alignment.
 To this end \textit{PanGraph} has the command \verb|polish|\footnote{See \url{https://neherlab.github.io/pangraph/tutorials/tutorial_3/}.}.

### 2023-02-24 reviewer_response.tex
update

@@
 It is also not clear if any of the splitting pancontigs are less than this value if the entire merger is dropped or just massaged to not have the too small new pancontig.}
-\Liam{this relates to R2's point above about all sequence ending up in pancontigs so I think is just clarification.}\\
-\Marco{I agree.
+% \Liam{this relates to R2's point above about all sequence ending up in pancontigs so I think is just clarification.}\\
+% \Marco{I agree.
 We set a minimal cutoff of 100bp.
@@
 This might be explained in the SI?}\\
+\responsedraft{
+      The minimal size parameter $\Lthr$ is set to 100 bp in PanGraph.
+This parameter controls the minimal size of blocks.
+In general, this parameter sets a scale separation between structural variation that is encoded in the graph structure or in the pancontig alignment.
+For instance, an insertion could appear either as a separate pancontig, or as an insertion in the pancontig alignment, depending on its size.
+In practice this parameter is used in three ways:
+      \begin{itemize}
+            \item In addition to the pseudo-energy costraint, merges are refused if they have alignment length shorter than $\Lthr$.
+            \item During merging, if the alignment contains long ($> \Lthr$) indels, these will cause a split in the merged pancontigs and will be treated as separate pancontigs.
+            \item If a merger of two pancontigs would produce flanking overhangs shorter than $\Lthr$, these are not saved as separate pancontigs but are included in the alignment as flanking indels.
+      \end{itemize}
+      As pointed out by the reviewer this parameter was not properly discussed in the text.
+We now extended its definition in IB in the main text, and give a more extended discussion in SI section I.\\
+      % TODO: missing: motivation for this choice in the text.
+      % TODO: Homogenization in the main text with discussion on precision. Extend.
+}
 

### 2023-02-24 reviewer_response.tex
Update on Overleaf.

@@
 % \Liam{that's clear to me.}
-\response{We agree with the reviewer suggestion. The wording in Line 54 has been updated to ``all-to-all alignment of the \textit{pancontigs} consensus sequences''.}
+\response{We agree with the reviewer suggestion. The wording in Line 54 has been updated to ``all-to-all alignment of the consensus sequences of all  \textit{pancontigs}''.} TODO
 
@@
 Or in another way aid this reading?}
-\response{We agree with the reviewer that it is important to be able visually compare the compression with the number $N$ of isolates for each species, and we thank him for the suggestion.
+\response{We agree with the reviewer that it is important to be able visually compare the compression with the number $N$ of isolates for each species, and we thank them for the suggestion.
 We modified fig.~4C in the main text and fig.~4I in SI marking the value of $1/N$ for each species with horizontal lines, and updated the captions accordingly.}
@@
 I agree with you about synteny and think this is a nice point to add here to explain the observation.}
-\response{The observation that core pancontigs are longer than average was in agreement with the expectation that core genes would be more syntenic and present less structural fragmentation compared to accessory genes.
-In contrast, we expected most short pancontigs to be located around recombination islands, where structural variation is high across different genomes.
-      We modified the line to: ``Core pancontigs having higher-than-average length.
-This is expected given their strong synteny and the fact that they present less structural variation across genomes compared to accessory genes.''}
+\response{The observation that core pancontigs are longer than average is in agreement with our prior the expectation since multiple core genes are often syntenic even in diverse species, compared to accessory genes that are often found in different contexts.
+This effect seems to outweigh the possibility of increased fragmentation due to the large number of mergers that common (core) blocks undergo.
+We modified the line to: ``Core pancontigs having higher-than-average length. This is expected given that core genes tend to stay synteny over large evolutionary distances.''} TODO
 
@@
 (3) For genome structure, exploring `hotspots' in the graph as \href{https://www.nature.com/articles/s41467-017-00808-w}{this paper} does could be done using in/out degree of pancontigs.}
+\response{We have so far not explored the use of pangraph as input for graph based mapping approaches like Pandora or GramTools.
+But this is indeed an obvious downstream application of pangraph.
+We have included a brief discussion of these approaches, but we feel that testing and implementation of such approaches is beyond the scope of this article.} TODO
 
@@
 
-\reviewer{R1.10}{Supplementary Information C: You comment that the use of null energy parameters results on average in "more and shorter pancontigs" but also that they result in more diverged sequences being merged into the same pancontig.
+\reviewer{R1.10}{Supplementary Information C: You comment that the use of null energy parameters results on average in ``more and shorter pancontigs" but also that they result in more diverged sequences being merged into the same pancontig.
 Why is a more diverse pancontig more likely to become fragmented?}
@@
 \response{We use the pseudo-energy defined in eq.~(1) as an heuristic to decide whether a homologous match detected by our aligner should result in a merge between two pancontigs, depending on the match length and on sequence divergence.
-In particular any choice of $\beta$ would result in no merges between sequences with divergence higher than $1/\beta$.
-Setting $\beta=0$ instead does not impose a divergence limit on merges, other than the one intrinsically set by the aligner sensitivity.
+In particular any choice of $\beta$ would prevent mergers between sequences with divergence higher than $1/\beta$.
+Setting $\beta=0$ instead does not impose a divergence limit on mergers other than the one implicitly set by the aligner sensitivity.
 In this sense the choice of null energy parameter results in more diverged sequences being merged into the same pancontig.
@@
 
+\response{We thank the reviewer for this succinct summary of the field and the essence of our contribution.}
+
 \subsubsection*{Major Issues}
 
-\reviewer{R2.1}{Line 46: when you say "PanGraph transforms an arbitrary set of genomes into a graph that simultaneously compresses the collection of sequences and exhaustively summarizes both the structural and nucleotide-level polymorphisms." - I am a bit unclear on how the nucleotide level variation is summarised.
+\reviewer{R2.1}{Line 46: when you say ``PanGraph transforms an arbitrary set of genomes into a graph that simultaneously compresses the collection of sequences and exhaustively summarizes both the structural and nucleotide-level polymorphisms." - I am a bit unclear on how the nucleotide level variation is summarised.
 The paper is a bit hand-wavey about this.
@@
 % \Liam{I think a `narrative section' in the supplementary describing this in brief sounds great.}
-\response{In our pangenome graph representation each pancontig encodes an alignment using a consensus sequence, and encoding each line as variations over the consensus.
-The consensus sequence is defined using the majority rule on the alignment and is updated at each pancontig merge.
-Each line in the alignment is then only described as variations over this consensus (mutations, insertions and deletions).\\
+\response{This point was also raised by other reviewers and we appreciate the need to make this pointer clearer.
+In our pangenome graph representation each pancontig encodes an alignment using a consensus sequence and list of edits required to obtain each individual sequence from the consensus sequence.
+These edits consist of single nucleotide changes (mutations) as well as insertions and deletions. 
+The consensus sequence is defined using the majority rule on the alignment and is updated after each merger.
+
 To make this more clear we introduced section I in SI, in which we explain in more detail the merging procedure and how the consensus sequence is updated.
 We also refer to the documentation page mentioned by the reviewer.
-Moreover, to avoid ambiguity we also avoided referring to pancontigs as ``blocks'' in the main text.}
+Moreover, to avoid ambiguity we also avoided referring to pancontigs as ``blocks'' in the main text and use pancontigs throughout.}
 
@@
 % \Liam{clarifying $N$ and $l$ makes sense.}
-\responsedraft{We agree with the reviewer that the formulation in the main text is too concise and not completely correct.
+\responsedraft{We agree with the reviewer that the formulation in the main text is too concise and a bit inaccurate. 
 There are two main variables at play: the length of a genome $l$ and the number of isolates $n$.
@@
 We also stress that the computational time to build the guide tree is usually negligible compared to the time required for the graph merging procedure.\\
+We slightly elaborated this point in the main text and added a some of the above explanation to the supplement.
 To correct this ambiguity we...
       % missing: changes in main text. My plan: report the n^2 l log l scaling. 
-}
+} TODO
 
@@
       \end{itemize}}
-\response{We followed the advice of the reviewer and added Section I in the SI, in which we expand on the explanation of how we merge two graphs.\\
+\response{We thank the reviewer for point out that this aspects require more explanation (the point was also raised by other reviewers). 
+We followed the advice of the reviewer and added Section I in the SI, in which we expand on the explanation of how we merge two graphs.\\
 In short, we perform the merging procedure in two steps.
-In the first we use the aligner (minimap2 or mmseqs2) to scan for homology matches between the consensus sequence of the pancontigs of one graph and of the other.
-These matches are ranked by pseudo-energy and performed in order.
-Each pancontig can be subject to only one merge at this stage.\\
-      To answer the first point, for precise duplicates only one of the two matches is merged at this stage, with pseudo-random ordering.
-However also the second region will be matched and merged before the merging of the two graphs is conlcuded.\\
+In the first we use the aligner (minimap2 or mmseqs2) to scan for homology matches between the consensus sequences of all pancontigs of one graph and of the other.
+This is achieved by simply passing all all these consensi in a single fasta file into mmseqs2 or via the API into minimap2. 
+The resulting matches are ranked by pseudo-energy and mergers performed in order. 
+Each pancontig can be subject to only one merge at this stage.
+For precise ties in this ranking of of matches, the behavior is not explicitly defined but depends on the order in which the matches appear in the output of the homology search tool.
+
 After all of the matches have been exhausted, the two graphs have undergone a preliminary merging, with some new merged pancontigs being created.
@@
 The self-mapping is repeated iteratively until no new merges can be performed.
+Exact duplicate matches would thus be merged in successive iterations of this merge-process.
 }
@@
 \response{
+Apologies for these slighly confusion statements.  
 In this case the second criterion is a consequence of the first.
@@
 eq.~(1) in the main text) it follows that for an alignment of pancontig consensus sequences to have negative pseudo-energy must be that the number of SNPs divided by the alignment length $N_m/l$ must be less than $1/\beta$.
-Moreover, as the reviewer pointed out, the pseudo-energy ranking of merges effectively results in merges being performed in order of similarity, effectively in reverse evolutionary order.\\
+Moreover, as the reviewer pointed out, the pseudo-energy ranking of merges effectively results in merges being performed in order of similarity, effectively in reverse evolutionary order.
+\\
+
 To make this more explicit we modified lines 93-95 in ``$\beta$ controls the maximal sequence divergence between the two merger-candidates.
@@
 
-\reviewer{R2.6}{Line 183 - I really do not understand what is going on with your alpha,beta being set to zero.
+\response{Thank you!}
+
+\reviewer{R2.6}{Line 183 - I really do not understand what is going on with your alpha, beta being set to zero.
 Are you saying if you stop penalising merges for having too many SNPs, then you will be able to merge more, and this will stop you from havign a 10\% divergence limit?
@@
 % \Liam{I think your interpretation is right. Important to clarify that there are those two limits and that $\beta$ only controls the divergence limit \textit{within} what one can actually align.}
-\response{There are two factors limiting the divergence of merges that are performed.
+\response{There are two factors limiting the divergence of mergers that are performed.
 The first factor is the aligner used.
@@
 Only homology matches of up to this divergence will be suggested by the aligners.
-On top of this, only matches with negative pseudo-energy are turned into merges.
-For any given value of $\beta$, this sets an additional limit to the divergence, which must be less than $1/\beta$.\\
+On top of this, only matches with negative pseudo-energy are accepted as mergers. 
+The parameter $\beta$ thus allows to explicitly set a maximal divergence ($1/\beta$) allowed in mergers, rather than relying on the implicit cut-off that comes from the sensitiviy of the aligner. \\
 This is now made more explicit in SI section I (PanGraph parameters effect).\\
-      When setting $\beta = 0$ only the intrinsic limit on divergence posed by the aligner sensitivity is left.
-In this sense more merges are allowed, but all mergess will have some level of homology (maximum 20\% or 30\% divergence, depending on the aligner).}
+When setting $\beta = 0$ the maximal divergence is determined by the aligner sensitivity.
+In this sense more merges are allowed, but all mergers will have some implicit maximal divergence threshold (20\% or 30\%, depending on the aligner).}
 
@@
 fig.~4D in SI).
-Hp is an exception and has the opposite trend, both for the number of pancontig and for the value of L50.
-For this species the use of null energy parameters results in less fragmentation of the graph.
-This is compatible with the observation that the average divergence of this dataset is on the limit of what PanGraph can handle.
-Unless a highly sensitive aligner is used and the divergence barrier in pseudo-energy is removed, some highly-diverged homologous sequences remain split in different pancontigs.
-Conversely, the choice of these parameters generates a graph with less pancontigs, a lower L50, a larger fraction of core pangenome, an higher median length of pancontigs (N50).\\
-      For a visual verification of this interpretation we display in fig.~\ref{fig:kp-vs-hp} how the distributions of pancontig lenght and frequency change with the use of different alignment kernels and pseudo-energy parameters.
+Helicobacter is an exception with the opposite trend where the use of null energy parameters results in less fragmentation of the graph (both in the number of pancontigs and for the value of L50).
+This might be explained by that fact that the divergence of some parts of the core genome in this dataset is close to $1/\beta$ for $
+\beta=10$ used in the figure. 
+Hence some more diverse parts of the genome might remain split in different pancontigs and removing this threshold imposed by $\beta$ results in a larger longer consecutive merger and a less fragmented graph.
+For a visual verification of this interpretation we display in fig.~\ref{fig:kp-vs-hp} how the distributions of pancontig length and frequency change with the use of different alignment kernels and pseudo-energy parameters.
 We compare the changes observed on the Hp dataset to the ones in Kp, used as a reference.
@@
 \response{No, every part of the sequence of each genome must be included in some pancontig.
-I might not necessarily appear in the pancontig consensus sequence if it is present in the alignment as a short ($< \Lthr$) insertion, but it will nonetheless be present in the alignment.
-In the particular case in which merging two pancontigs would leave overhangs shorter than $\Lthr$, these are included in the alignmen as flanking insertions.
-This is now explained in SI section I, and mentioned in the main text in section IB (line 103).}
+It might not necessarily appear in the pancontig consensus sequence if it is present in the alignment as a short ($< \Lthr$) insertion, but it will nonetheless be present in the alignment.
+In the particular case in which merging two pancontigs would leave overhangs shorter than $\Lthr$, these are included in the alignment as flanking insertions.
+This is now explained in SI section I, and mentioned in the main text in section IB (line 103).
+In fact, we used the fact that genomes can be reconstructed exactly from the pangraph to test and debug the tool. }
 
@@
       % TODO: archive repository state.
-}
+} TODO -- make csv
 
@@
 
-\reviewer{R2.2}{Why on earth is the "get an MSA of a block" option in pangraph called "polish"?}
+\reviewer{R2.2}{Why on earth is the ``get an MSA of a block" option in pangraph called ``polish"?}
 % \Marco{i don't think this comes from the paper (I used grep on the tex file and we never mention polish).
@@
 \response{In our method pancontig mergers are performed using only the consensus sequence as a guide.
-This makes the merging procedure very fast, but can introduce minor inconsistencies and artifacts with insertions in the alignment.
-To remove these inconsistencies, after building the graph the user can optionally decide to ``polish'' pancontig alignments with the omonymous command.
+This makes the merging procedure very fast, but can introduce minor inconsistencies and artifacts with insertions in the alignment as insertions are not aligned to each other in the procedure.
+To remove these inconsistencies, after building the graph the user can optionally decide to ``polish'' pancontig alignments with the \texttt{polish} command.
 This command takes a graph as input and produces a graph with the same topology, but with improved alignments.
@@
 
-\reviewer{R2.3}{line 48: "Pancontigs are connected by an edge if they are syntenic on at least one input sequence;" - i think this is a bit of an abuse of modern language.
+\reviewer{R2.3}{line 48: ``Pancontigs are connected by an edge if they are syntenic on at least one input sequence;" - i think this is a bit of an abuse of modern language.
 The word syntenic is generally used to mean two genes/whatever have order conserved on multiple chromosomes (although some googling tells me that in classical genetics it meant things were on the same chromosome).
@@
 % \Liam{adjacent sounds good.}
-\response{we agree with the reviewer observation. We modified ``syntenic on'' to ``adjacent in'' in line 48.}
+\response{we agree with the reviewer's observation. We modified ``syntenic on'' to ``adjacent in'' in line 48.}
 
@@
 }
+\response{Thank you!}
 

### 2023-02-24 reviewer_response.tex
save progress

@@
 
+\responsedraft{We thank the reviewer for their appreciation and for their critical review of our work.}
+
 \reviewer{R1.1}{A pancontig is defined as a multiple sequence (sub)alignment.
@@
 % \Liam{that's clear to me.}
-\response{We agree with the reviewer suggestion. The wording in Line 54 has been updated to ``all-to-all alignment of the consensus sequences of all  \textit{pancontigs}''.} TODO
+\response{We agree with the reviewer's suggestion. The wording in Line 54 has been updated to ``all-to-all alignment of the consensus sequences of all \textit{pancontigs}''.}
 
@@
 This effect seems to outweigh the possibility of increased fragmentation due to the large number of mergers that common (core) blocks undergo.
-We modified the line to: ``Core pancontigs having higher-than-average length. This is expected given that core genes tend to stay synteny over large evolutionary distances.''} TODO
+      We modified the line to: ``Core pancontigs having higher-than-average length. This is expected given that core genes tend to stay synteny over large evolutionary distances.''}
 
@@
 Or how it can be used to analyze genome structure.}
-\Marco{other than explicitly exporting block sequences and mapping to those, we do not really using the pangenome graph as a reference against which to map new strains...
+% \Marco{other than explicitly exporting block sequences and mapping to those, we do not really using the pangenome graph as a reference against which to map new strains...
 But we could expand the discussion with more ideas on applications to study pangenome organization and evolution.
 For software for further analysis and exploration I have \href{https://github.com/mmolari/pypangraph}{this package} to explore and interact with graphs in python, but not sure it's worth mentioning.}\\
-\Liam{Some ideas that could be mentioned.
+% \Liam{Some ideas that could be mentioned.
 (1) I think one obvious (bad!) use would be to take the core blocks and map reads to those as a `core reference'.
@@
 (3) For genome structure, exploring `hotspots' in the graph as \href{https://www.nature.com/articles/s41467-017-00808-w}{this paper} does could be done using in/out degree of pancontigs.}
-\response{We have so far not explored the use of pangraph as input for graph based mapping approaches like Pandora or GramTools.
+\responsedraft{We have so far not explored the use of pangraph as input for graph based mapping approaches like Pandora or GramTools.
 But this is indeed an obvious downstream application of pangraph.
-We have included a brief discussion of these approaches, but we feel that testing and implementation of such approaches is beyond the scope of this article.} TODO
+We have included a brief discussion of these approaches, but we feel that testing and implementation of such approaches is beyond the scope of this article.}
 
@@
       We slightly elaborated this point in the main text and added a some of the above explanation to the supplement.
-To correct this ambiguity we...
-      % missing: changes in main text. My plan: report the n^2 l log l scaling. 
-} TODO
+}
 
@@
       On top of this, only matches with negative pseudo-energy are accepted as mergers.
-The parameter $\beta$ thus allows to explicitly set a maximal divergence ($1/\beta$) allowed in mergers, rather than relying on the implicit cut-off that comes from the sensitiviy of the aligner. \\
+      The parameter $\beta$ thus allows to explicitly set a maximal divergence ($1/\beta$) allowed in mergers, rather than relying on the implicit cut-off that comes from the sensitiviy of the aligner.
+\\
       This is now made more explicit in SI section I (PanGraph parameters effect).\\
@@
 % \Liam{for sure that's all they want (most biologists don't like json!)}
-\responsedraft{We now included in the submission two excel files.
-``accession\_numbers.xlsx'' contains five sheets, one per species considered, each with a list of the accession numbers of isolates used in our datasets.
-``marginalize\_pairs.xlsx'' contains instead a list of all the pairs used in the graph marginalization section.
+\response{We now included in the submission two csv files.
+\texttt{accession\_numbers.csv} contains the list of accession numbers of all isolates in our dataset for all species considered.
+\texttt{marginalize\_pairs.csv} contains instead a list of all the pairs used in the graph marginalization section.
       % TODO: archive repository state.
-} TODO -- make csv
+}
 
@@
 
-\reviewer{R2.2}{Why on earth is the ``get an MSA of a block" option in pangraph called ``polish"?}
+\reviewer{R2.2}{Why on earth is the ``get an MSA of a block'' option in pangraph called ``polish''?}
 % \Marco{i don't think this comes from the paper (I used grep on the tex file and we never mention polish).
@@
 This step of the procedure resolves the artificial cut buy merging these pancontigs together.\\
-      We think that the mention of this step of the algorithm in this part of the paper was a bit confusing and needed more clarification.
-We moved the statement from section IB of the main text to SI section I (point 5 of the merging algorithm explanation).}
+      We think that the mention of this step of the algorithm in this part of the paper might be a bit confusing and needs more clarification.
+We moved and expanded the statement from section IB of the main text to SI section I (point 5 of the merging algorithm explanation).}
 
@@
 % - update SI section numbering in the main text
+
+% meeting notes:
+% - no alignment "line"
+% - [x] csv file 
+% - [ ] supplementary: move graph structure to new section and add two figures from the docs.
+% - [x] R1.1
+% - [x] R1.5
+% - [ ] R1.8
+% - [ ] R2.2
+% - [x] R2.6
+% - [x] R2.9
+% - [x] remove \oldnew and use git diff
+% pangraph automatic testing -> sequence reconstruction
+
+%  TODO: before submission
+% - [ ] archive repository state and add in the response
+% - [ ] update SI section numbers and line references in the main text and in the response.

### 2023-02-24 main.tex
save progress

--- 2023-02-23 main.tex
@@
 \newcommand{\modified}[1]{{\color{red}#1}}
-\newcommand{\oldnew}[2]{{\color{gray}\sout{#1}}{\color{red}#2}}
+% \newcommand{\oldnew}[2]{{\color{gray}\sout{#1}}{\color{red}#2}}
 
@@
 The accumulation of such changes over time complicates comparative genomic analyses of present day isolates.
-Homologous recombination is rapid enough that most genes in many bacterial core genomes have distinct phylogenies \cite{sakoparnig2021whole} and even closely related genomes differ dramatically in gene content.
+Homologous recombination is rapid enough that most genes in many bacterial core genomes have distinct phylogenies \cite{sakoparnig2021whole} and even closely related genomes differ dramatically in gene content \cite{touchon2020phylogenetic,touchon2009organised,doolittle2009origin}.
+% \cite{touchon2020phylogenetic,doolittle2009origin,touchon2009organised}.
 % comment: rewritten to use consistent terminology (I think 'mutation' and 'polymorphism' for most people make them think of local changes only)
@@
 The graph is composed of \emph{pancontigs}, which represent linear multiple-sequence alignments of homologous sequence found within one or more input genomes.
-\emph{Pancontigs} are connected by an edge if they are \oldnew{syntenic on}{adjacent in} at least one input sequence; individual sequences are then recapitulated by contiguous \emph{paths} through the graph.
+\emph{Pancontigs} are connected by an edge if they are adjacent in at least one input sequence; individual sequences are then recapitulated by contiguous \emph{paths} through the graph.
 Pancontigs are directed, meaning their orientation in a path can be either 5' to 3' or vice versa.
@@
 \emph{PanGraph's} overarching strategy is to approximate multiple-genome alignment by iterative pairwise alignment of graphs of subsets of sequences, in the spirit of progressive alignment tools \cite{feng1987progressive,darling2010progressivemauve,armstrong2020progressive}.
-Pairwise graph alignment is performed by an all-to-all alignment of the \emph{pancontigs} consensus sequences between both graphs, and the order of pairwise alignments is determined by a guide tree.
+Pairwise graph alignment is performed by an all-to-all alignment of the consensus sequence of all \emph{pancontigs} between both graphs, and the order of pairwise alignments is determined by a guide tree.
 
@@
 \paragraph*{Merging of homologous sequence.}
-If the above alignment step detected homologous stretches between \oldnew{}{the consensus sequences of} two pancontigs, these pancontigs, or parts of them, can be merged.
+If the above alignment step detected homologous stretches between the consensus sequences of two pancontigs, these pancontigs, or parts of them, can be merged.
 It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order.
@@
 Only mergers whose alignment has negative pseudo-energy are performed.
-The parameter $\beta$ controls the maximal sequence divergence between the two merger-candidates\oldnew{ and no mergers will be performed if the Hamming distance exceeds $1/\beta$.}{.
-The negative pseudo-energy requirement means that no merger will be performed if the Hamming distance exceeds $1/\beta$.}
+The parameter $\beta$ controls the maximal sequence divergence between the two merger-candidates.
+The negative pseudo-energy requirement means that no merger will be performed if the Hamming distance exceeds $1/\beta$.
 Importantly, this divergence threshold is applied to the comparison of the consensus sequences of the merger candidates.
-\oldnew{}{The pseudo-energy ranking also effectively results in merges being performed in reverse evolutionary order, with less diverged sequences being merged first.}
-The parameter $\alpha$ controls the fragmentation of the graph and imposes a minimal length on the \oldnew{resulting}{} pan-contigs resulting from the merger.
-The merger is always performed if \oldnew{the merger}{it} doesn't increase the number of contigs, but mergers that introduce 1, 2, 3 or 4 cuts are only performed for sufficiently long homologous stretches as parameterized by $\alpha$.
-\oldnew{In addition, there is parameter that controls the minimal size of contigs produced by the algorithm.}{In addition to the fragmentation parameter $\alpha$, there is a further parameter $\Lthr$ that controls the minimal length of pancontigs.
+The pseudo-energy ranking also effectively results in merges being performed in reverse evolutionary order, with less diverged sequences being merged first.
+The parameter $\alpha$ controls the fragmentation of the graph and imposes a minimal length on the pan-contigs resulting from the merger.
+The merger is always performed if it doesn't increase the number of contigs, but mergers that introduce 1, 2, 3 or 4 cuts are only performed for sufficiently long homologous stretches as parameterized by $\alpha$.
+In addition to the fragmentation parameter $\alpha$, there is a further parameter $\Lthr$ that controls the minimal length of pancontigs.
 Structural variation at a scale smaller than this threshold (i.e.
 short indels) is stored in pancontig alignments, rather than in the graph structure, and no merges are performed if they have length shorter than this threshold.
-We set a default value of 100 bp for this parameter.}
+For instance, when a merge would generate flanking contigs shorter than this threshold, the merging is performed but the alignment is extended to include these regions.
+We set a default value of 100 bp for this parameter.
 
@@
 At the graph level, the merger of two \emph{pancontigs} defines a new \emph{pancontig}, connected on both sides by edges to the neighboring \emph{pancontigs} of both inputs, and thus locally collapses the two graphs under consideration.
-At the nucleotide level, the pairwise alignment of two \emph{pancontigs} maps the reference of one onto the other; the merger of two \emph{pancontigs} requires the application of the map onto the underlying multiple-sequence alignment.
+At the nucleotide level, the pairwise alignment of two \emph{pancontigs} maps the consensus of one onto the other; the merger of two \emph{pancontigs} requires the application of the map onto the underlying multiple-sequence alignment.
 Once both sets of sequences are placed onto a common coordinate system, the resultant consensus sequence, and thus polymorphisms, are recomputed.
 This procedure can be viewed as an approximate multiple sequence alignment algorithm that aligns homologous parts without further investigating insertions in the two sub-alignments.
+This shortens considerably the time needed, but might introduce minor inconsistencies and artifacts in the alignments.
+After building the graph, these can be removed by performing a multiple sequence alignment on each pancontig  (see ``polish'' command in documentation).
 
 The above procedure is repeated until no alignments with negative energy remain.
-Upon completion, transitive edges within the graph, i.e., edges along which all path are co-linear, are removed by merging adjacent \emph{pancontigs}.
-\oldnew{}{More details on the merging procedure and the effect of parameters can be found in SI section I.}
+More details on the merging procedure and the effect of parameters can be found in SI section I.
 
@@
 In each comparison we evaluate the misplacement of breakpoints that we can pair on the two graphs within 1kbp.
-The plot displays the fraction of breakpoints that have misplacement greater than the standard \textit{PanGraph} precision threshold of \oldnew{}{$\Lthr=$}100bp, as a function of average pairwise sequence divergence.
+The plot displays the fraction of breakpoints that have misplacement greater than the standard \textit{PanGraph} precision threshold of $\Lthr=$100bp, as a function of average pairwise sequence divergence.
 Line and shaded area represents mean and standard deviation over 25 repetitions.
@@
         total core genome length ($\Lcore$),
-        \oldnew{}{total soft-core genome length ($\Lsoftcore$, corresponding to genes shared by at least 95\% of the isolates),}
+        total soft-core genome length ($\Lsoftcore$, corresponding to genes shared by at least 95\% of the isolates),
         and average pairwise divergence of core genes ($\dcore$).
@@
 
-For species other than \textit{P.~marinus}, the pangraphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in \oldnew{blocks}{pancontigs} spanning several kbps (see SI Fig.~4).
+For species other than \textit{P.~marinus}, the pangraphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in pancontigs spanning several kbps (see SI Fig.~4).
 For these species the use of null energy parameters slightly increases merging and sequence compression, at the cost of slightly more fragmented graphs (cf. SI Section IIIC and SI Fig.~4).
@@
         \textbf{C}: sequence compression, defined as the ratio between the pangenome graph size and the cumulative size of all the sequences contained in the graph.
-\oldnew{}{Since maximal compression depends on the number of isolates $n$ in the pangenome graph, we mark for reference the value of $1/n$ for each dataset.}
+Since maximal compression depends on the number of isolates $n$ in the pangenome graph, we mark for reference the value of $1/n$ for each dataset.
     }
@@
 The number and size of pancontigs scale sub-linearly with the number of isolates, with more than 50\% of the pangenome being included in the 10\% longest pancontigs.
-\oldnew{, and core pancontigs having higher-than-average size}{Core pancontigs have higher-than-average length.
-This is expected given their strong synteny and the fact that they present less structural variation across genomes compared to accessory genes.} Adding more strains does not generate excessive fragmentation of these pancontigs, and their size remains of several kbps even as hundreds of strains are added.
+Core pancontigs having higher-than-average length.
+This is expected given that core genes tend to stay synteny over large evolutionary distances.
+Adding more strains does not generate excessive fragmentation of these pancontigs, and their size remains of several kbps even as hundreds of strains are added.
 The size of the core genome does not decrease significantly with the addition of new strains, while the total pangenome size remains orders of magnitude smaller than the total size of all genomes included in the graph.
@@
 The compatibility between the two graphs requires segments on which the two partitions disagree to be few and short.
-Indeed, we verified that over the pairs we picked, these segments cover a very small fraction of the genome ($<1\%$) and have average size compatible with the default 100 bp precision threshold of \textit{PanGraph} \oldnew{}{set by the value of $\Lthr$} (cf.
+Indeed, we verified that over the pairs we picked, these segments cover a very small fraction of the genome ($<1\%$) and have average size compatible with the default 100 bp precision threshold of \textit{PanGraph} set by the value of $\Lthr$ (cf.
 Fig.~\ref{fig:marginalization}B-C).
@@
 The last line represents the distribution of shared sequence, approximated using the fraction of shared k-mers corrected using sequence divergence as described in the main text.
-\oldnew{}{Next to each distribution we report its mean and standard deviation.}
-        \textbf{C}: distribution of average segment lengths for each category over the 50 pairs considered (2 entries per pair). \oldnew{}{Mean and standard deviation are reported.}
+Next to each distribution we report its mean and standard deviation.
+        \textbf{C}: distribution of average segment lengths for each category over the 50 pairs considered (2 entries per pair). Mean and standard deviation are reported.
     }

### 2023-02-24 supplementary.tex
save progress

--- 2023-02-23 supplementary.tex
@@
 \newcommand{\Lthr}{L_{\min}}
-\newcommand{\oldnew}[2]{{\color{gray}\sout{#1}}{\color{red}#2}}
+% \newcommand{\oldnew}[2]{{\color{gray}\sout{#1}}{\color{red}#2}}
 
@@
 In Fig.~\ref{fig:marginalize-suppl} we display the distribution of these quantities across all pairs for the different species considered and the block types described above.\\
-In all species except HP and PM we find very good agreement between pairwise and marginalized graphs: segments on which the graphs \textit{disagree} represent only a very minor fraction of the genome (much smaller than the fraction of shared or private blocks), and have size compatible with the default pancontig length threshold of \textit{PanGraph} (\oldnew{}{$\Lthr =$}100 bp).
+In all species except HP and PM we find very good agreement between pairwise and marginalized graphs: segments on which the graphs \textit{disagree} represent only a very minor fraction of the genome (much smaller than the fraction of shared or private blocks), and have size compatible with the default pancontig length threshold of \textit{PanGraph} ($\Lthr =$100 bp).
 As described in the previous section, the divergence of the HP dataset sits on the limit of what the \textit{minimap2} alignment kernel can successfully handle.

### 2023-02-27 main.tex
updated SI and response

@@
 \usepackage{xcolor}
+\usepackage{xr} % external references (reference supplementary)
 \usepackage[normalem]{ulem}
 \usepackage[final]{hyperref}
+
+\externaldocument[sp-]{supplementary}
 
@@
 \subsection{Guide tree construction}
-The alignment guide tree is constructed subject to three design constraints: (i) similar sequences are aligned first, (ii) the similarity computation scales sub-quadratically with the number of input sequences, and (iii) the resultant tree is balanced to maximize parallelism.
+The alignment guide tree is constructed subject to three design constraints: (i) similar sequences are aligned first, (ii) the similarity computation must have good scaling properties with the length and number of input sequences, and (iii) the resultant tree is balanced to maximize parallelism.
 To this end, we formulate the algorithm as a two step process.
@@
 Computationally, each sequence can be sketched into its set of minimizers in linear time while the cardinality of all pairwise intersections can be computed by sorting the list of all minimizers to efficiently count overlaps.
-Hence, the pairwise distance matrix is estimated in a time that increases log-linearly in the total length of sequence that is analyzed.
+Hence, the pairwise distance matrix is estimated in a time that increases log-linearly in the length of sequences, and in our implementation can scale sub-quadratically with the number of isolates (see SI sect.\ref{sp-sec:pangraph-algo} for details).
 The final guide tree is constructed as the balanced binary tree constrained to reproduce the topological ordering of leaves found initially.

### 2023-02-27 supplementary.tex
updated SI and response

@@
 \newcommand{\Lthr}{L_{\min}}
+\newcommand{\red}[1]{{\color{red}#1}}
 % \newcommand{\oldnew}[2]{{\color{gray}\sout{#1}}{\color{red}#2}}
@@
 
-\section{Graph merging algorithm}
-\label{supp:graph_merging}
+\section{Pangenome graph representation}
+\label{sec:graph-repr}
 
-In this section we describe in more detail the procedure used to merge two pangenome graphs together. This procedure is at the core of \textit{PanGraphs}'s algorithm.
+This section contains a short explanation of the pangenome graph representation used in \textit{PanGraph}.
+A pangenome graph is essentially composed of two elements: pancontigs, which encode for alignment of homologous sequences, and paths, that correspond to representation for genomes as a sequence of pancontigs (see fig.~\ref{fig:graph-representation}).\\
 
-As explained in the main text, the full pangenome graph is built by placing genomes on terminal leaves of a guide tree.
-These genomes are initially considered as elementary graphs composed of a single \textit{pancontig}.
-These graphs are then propagated along the tree towards the root, and at each intersection a merge operation is performed.
-The full pangenome graph is then collected from the root.
+\begin{figure}[htb]
+    \includegraphics[width=0.5\textwidth]{figs_suppl/graph_representation.pdf}
+    \caption{{\bf Pangenome graph representation}.
+Given a set of genomes \textit{PanGraph} will construct a pangenome graph by condensing all stretches of homologous sequences in \textit{pancontigs}, containing the alignment of these sequences.
+Genomes can then be represented as \textit{paths}, i.e.
+lists of pancontigs.
+        In our representation a pangenome graph is in essence a collection of \textit{pancontigs} and \textit{paths}.
+    }
+    \label{fig:graph-representation}
+\end{figure}
 
-\subsection*{pangenome graph representation}
+Each \textbf{pancontig} carries the full information on the alignment of homologous sequences it contains.
+It possesses a single consensus sequence, and each sequence in the alignment is encoded as variations over this consensus.
+These can be either mutations of single nucleotides, gaps, deletions or insertions.
+The alignment sequence can be reconstructed by performing these changes in order, as displayed in fig.~\ref{fig:alignment-reconstruction}.
+More information on the precise encoding used can be found in PanGraph documentation.\footnote{See \url{https://neherlab.github.io/pangraph/tutorials/tutorial_2}}\\
 
-As a preliminary step in the explanation we spend few words on the pangenome graph representation of \textit{PanGraph}.
-A pangenome graph is essentially composed of two elements: pancontigs, which encode for alignment of homologous sequences, and paths, that correspond to representation for genomes as a sequence of pancontigs.
-
-Each \textbf{pancontig} contains the full information on the alignment of homologous sequences it contains.
-In PanGraph each pancontig is encoded with a consensus sequence, and each line of the alignment is saved as variations over this consensus (mutations, insertions and deletions).
-More information on the precise encoding used can be found in PanGraph documentation.\footnote{See \url{https://neherlab.github.io/pangraph/tutorials/tutorial_2/\#From-block-consensus-to-node-alignment}}
+\begin{figure}[htb]
+    \includegraphics[width=0.5\textwidth]{figs_suppl/pancontig_alignment.pdf}
+    \caption{{\bf From pancontig consensus to alignment}.
+In our pancontig representation, each sequence in the alignment is encoded as variations over the pancontig consensus.
+In practice in the pangraph output \texttt{json} file each alignment sequence is saved as a list of gaps to be added to the consenus, sites to be mutated, insertions to be performed in gaps, and deletions.
+By performing these changes in order on the consensus each sequence in the alignment can be retrieved.
+In this example we display how each of these variations are encoded and performed.
+    }
+    \label{fig:alignment-reconstruction}
+\end{figure}
 
@@
 For each item in the sequence along with the pacontig id PanGraph also stores the pancontig orientation (whether homology with the consensus is on the forward the reverse strain of the genome) and which variations on the pancontig consesus (i.e.
-which line in the pancontig alignment) correspond to this precise location in the genome.
+which sequence in the pancontig alignment) correspond to this precise location in the genome.\\
 
@@
 ``Large-scale'' differences, such as presence or absence of long stretch of sequences, are instead encoded in paths as presence or absence of pancontigs.
-This scale separation can be tuned using PanGraph parameters, as will be explained in the next section.
+This scale separation can be tuned using PanGraph parameters, as will be explained in following sections.
+
+\section{PanGraph algorithm}
+\label{sec:pangraph-algo}
+
+In this section we describe in more detail the algorithmic procedure used to build a pangenome graph.
+In essence, the full pangenome graph is built by placing genomes on terminal leaves of a guide tree, built from a sequence minimizer distance matrix.
+These genomes are initially considered as elementary graphs composed of a single \textit{pancontig}.
+These graphs are then propagated along the tree towards the root, and at each intersection a merge operation is performed.
+The full pangenome graph is then collected from the root.
+
+We focus on three aspects:
+\begin{itemize}
+    \item The estimation of pairwise distances between genomes, used to build the guide tree.
+    \item The procedure used to merge two pangenome graphs together, which is at the core of \textit{PanGraphs}'s algorithm.
+    \item The effect of the main parameters of the algorithm on the resulting graph structure.
+\end{itemize}
+
+\subsection*{guide tree construction and distance matrix estimation}
+
+As stated in the main text, the guide tree is built by neighbor-joining, with the pairwise distance between sequences being approximated by the Jaccard distance between sequence minimizers.
+To maximize parallelization, the final tree is a balanced binary tree with the same ordering of leaves as the one found by neighbour-joining.
+Here we briefly discuss how the complexity of this distance matrix estimation scales with the number $n$ and length $l$ of input sequences.\\
+As a first step of the distance estimation, we sketch all input sequences together and build a unique sorted list of minimizers.
+This can be done in time $O(L \log L)$, where $L = n \times l$ is the total length of input sequences.
+With each minimizer in the list we also keep the information of the genome of origin.\\
+As a second step we count the number of minimizers shared by each pair of sequences.
+This can be done efficiently by scrolling the sorted list of minimizers (having size proportional to $L$) only once.
+For each chunk of equal minimizers in the list (having size between 1 and $n$, depending on how shared is the original k-mer) we increment by one the corresponding pairs in a minimizer count matrix.
+Depending on how similar the input sequences are, this operation has complexity between $O(n^2 l) = O(nL)$ (all minimizers shared) and $O(nl) = O(L)$ (no minimizer shared).\\
+In conclusion, depending on the similarity of the input sequences, evaluation of the pairwise distance matrix has complexity between $O(L \log L)$ and $O(L \log L + nL)$.\\
+
+In practice for typical use cases of pangraph ($n \sim 100$ and $L \sim 10^6$) the computationally challenging part is the minimizer sketching, and not the distance matrix computation.
+This is partly due to some prefactors (e.g.
+the window size of 100 bp used for sketching makes so that the list of minimizer is significantly shorter than the genome, and the only operation performed $n^2$ times is a single increment of an integer in the minimizer hit matrix).
+Moreover, the computational time to build the guide tree is usually negligible compared to the time required for the graph merging procedure.\\
 
@@
 The appropriate value depends on the scale at which one wants to study structural variations in genomes.
-The default value in PanGraph is set to 100 bp.\\
+The default value in PanGraph is set to 100 bp.
+This is done to capture structural variation at the gene scale, while ignoring small indels.
+The value can hoewever be tuned by the user depending on the specific use-case.\\
 A similar role is played by the parameter $\alpha$.
@@
 \section{Generation of synthetic data}
-\label{supp:data_generation}
+\label{sec:data_generation}
 
@@
 
-To test the accuracy of \textit{PanGraph} in reconstructing the pangenome graph of a set of isolates we generated artificial data using the procedure described in Section~\ref{supp:data_generation}.
+To test the accuracy of \textit{PanGraph} in reconstructing the pangenome graph of a set of isolates we generated artificial data using the procedure described in Section~\ref{sec:data_generation}.
 For each simulation we obtain both a set of genome sequences and the underlying real pangenome graph.

### 2023-02-27 reviewer_response.tex
updated SI and response
update

@@
 \usepackage{graphicx}
+\usepackage{xcolor}
 \usepackage{hyperref}
-\usepackage{xcolor}
 
@@
 \newcommand{\responsedraft}[1]{{\it {\color{purple}\textbf{ResponseDraft:} #1}}\vskip 5mm}
+\newcommand{\crossref}[1]{{\color{purple}\{check cross-reference\}#1}} % cross-reference, to be later checked
 
@@
 % \Liam{clarifying $N$ and $l$ makes sense.}
-\responsedraft{We agree with the reviewer that the formulation in the main text is too concise and a bit inaccurate.
+\response{We agree with the reviewer that the formulation in the main text is too concise and a bit inaccurate.
       There are two main variables at play: the length of a genome $l$ and the number of isolates $n$.
@@
             \item Fantastic choice of species in your empirical dataset - P. marinus diversity is insane.
-            \item Line 215 - this is a remarkable achievement: "Adding more strains does not generate excessive fragmentation of these pancontigs, and their size remains of several kbps even as hundreds of strains are added"
+            \item Line 215 - this is a remarkable achievement: ``Adding more strains does not generate excessive fragmentation of these pancontigs, and their size remains of several kbps even as hundreds of strains are added''
             \item Congratulations
@@
 
-\reviewer{R3}{The article "PanGraph: scalable bacterial pan-genome graph construction" reports a new method for constructing a Pan-Genome Graph (PGG) in an efficient manner with an open and extractable grah structure.
+\reviewer{R3}{The article ``PanGraph: scalable bacterial pan-genome graph construction'' reports a new method for constructing a Pan-Genome Graph (PGG) in an efficient manner with an open and extractable grah structure.
 Unfortunately, the authors appear to either be unaware of or downplay the existence of previous methods for the same purpose.\textbf{ The authors need to survey the existing literature and compare and contrast with their method to show the utility of their approach.}}
 
-\reviewer{R3}{For instance the authors state: "Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated "bags of genes", and thus are not suited for this new era.".
+\reviewer{R3}{For instance the authors state: ``Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.''.
 There are many pan-genome analysis tools which treat the pan-genome as a graph and generate a PGG.
@@
 
-\reviewer{R3}{This statement: "While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging.
-Colored generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets, although the underlying efficiency derives from a fixed k-mer size which prevents modelling long-range homology (Iqbal et al., 2012; Muggli et al., 2017)." appears to be incomplete at best.
+\reviewer{R3}{This statement: ``While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging.
+Colored generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets, although the underlying efficiency derives from a fixed k-mer size which prevents modelling long-range homology (Iqbal et al., 2012; Muggli et al., 2017).'' appears to be incomplete at best.
 De Bruijn graph-based assemblers routinely collapse nonbranching portions of the graph into longer sequences than the k-mer size used for construction.
@@
 
-\reviewer{R3}{I believe this statement: "It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order." is the essential step in the process and perhaps deserves more focus.
+\reviewer{R3}{I believe this statement: ``It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order.'' is the essential step in the process and perhaps deserves more focus.
 It captures two important issues: \textbf{aligning orthologs not paralogs and keeping the fracturing of the PGG nodes to a minimum}.
@@
 This might be explained in the SI?}\\
-\responsedraft{
-      The minimal size parameter $\Lthr$ is set to 100 bp in PanGraph.
+\response{The minimal size parameter $\Lthr$ is set to 100 bp in PanGraph.
 This parameter controls the minimal size of blocks.
@@
       \end{itemize}
-      As pointed out by the reviewer this parameter was not properly discussed in the text.
-We now extended its definition in IB in the main text, and give a more extended discussion in SI section I.\\
-      % TODO: missing: motivation for this choice in the text.
+      We choose a value of 100 bp in our analysis (and as standard value) to be able to detect variations at the gene scale, while encoding variations of up to few tens of bps in the alignments.
+This value can however be tuned by the user depending on the specific needs.\\
+      We appreciate the need to better explain the role of this parameter in the merging procedure and in the structure of the resulting graph.
+To this end we extended its definition in section IB in the main text, and give a more extended discussion in SI section~\crossref{II}.\\
       % TODO: Homogenization in the main text with discussion on precision. Extend.
@@
 % - [ ] update SI section numbers and line references in the main text and in the response.
+% - [ ] update SI figure numbers.

### 2023-02-27 main.tex
update

@@
 
-\externaldocument[sp-]{supplementary}
-
 \newcommand{\avg}[1]{\langle #1 \rangle}
@@
 \newcommand{\dcore}{\langle d_\text{core} \rangle}
+\newcommand{\crossref}[1]{{\color{purple}\{check cross-refs\}#1}} % cross-reference, to be later checked
 \newcommand{\modified}[1]{{\color{red}#1}}
@@
 Computationally, each sequence can be sketched into its set of minimizers in linear time while the cardinality of all pairwise intersections can be computed by sorting the list of all minimizers to efficiently count overlaps.
-Hence, the pairwise distance matrix is estimated in a time that increases log-linearly in the length of sequences, and in our implementation can scale sub-quadratically with the number of isolates (see SI sect.\ref{sp-sec:pangraph-algo} for details).
+Hence, the pairwise distance matrix is estimated in a time that increases log-linearly in the length of sequences, and in our implementation can scale sub-quadratically with the number of isolates (see SI sect.~\crossref{II} for details).
 The final guide tree is constructed as the balanced binary tree constrained to reproduce the topological ordering of leaves found initially.
@@
 The above procedure is repeated until no alignments with negative energy remain.
-More details on the merging procedure and the effect of parameters can be found in SI section I.
+More details on the merging procedure and the effect of parameters can be found in SI sect.~\crossref{II}.
 
@@
 We simulated populations of size $N$ and sequence length $L$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
-In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (respective rates $0.01$ and $0.05$ per generation), and horizontal sequence transfer that occurs with tunable rate $h$ per generation per genome (see SI Section I for details on the simulation and standard parameter values).
+In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (respective rates $0.01$ and $0.05$ per generation), and horizontal sequence transfer that occurs with tunable rate $h$ per generation per genome (see SI sect.~\crossref{III} for details on the simulation and standard parameter values).
 The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
@@
 We note that PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of \textit{minimap2} \cite{li2018minimap2}.
-Benchmarks of CPU and memory usage are reported in SI Section IIA.
+Benchmarks of CPU and memory usage are reported in SI sect~\crossref{IVA}.
 
@@
 The predominant determinant of accuracy was the sample diversity, controlled by the mutation rate $\mu$ in our simulations (cf.
-SI Fig.~2).
+SI Fig.~\crossref{4}).
 For low-diversity datasets, breakpoints are inferred with accuracy of few basepairs, while for highly diverged isolates most breakpoints are displaced by several hundreds of basepairs (cf.
-SI Fig.~3).
+SI Fig.~\crossref{5}).
 The choice of alignment kernel influences the threshold diversity at which accuracy is lost, suggesting that inability to detect homology between diverged sequences is the reason.\\
@@
 These data had been previously analyzed using PanX \cite{ding2018panx}, which allowed us to estimate the size of the pangenome and core genome, and the average pairwise divergence on core genes (cf.
-Table~\ref{table:panx-dataset} and SI Section~IIIA).
+Table~\ref{table:panx-dataset} and SI Section~\crossref{VA}).
 
@@
 Using \textit{PanGraph}, we built multiple pangenome graphs for each species in the dataset (cf.
-SI Section~IIIB and IIIC).
+SI section~\crossref{VB and VC}).
 These differ by the alignment kernel used (\textit{minimap2} with \textit{asm10} or \textit{asm20} option, and \textit{mmseqs2}) or by the value of the pseudo-energy parameters $\alpha,\beta$ from eq.~(\ref{eq:pseudo-energy}).
@@
 
-Fig.~\ref{fig:panx-benchmark}B-C and SI Fig.~4 quantify the size of the core genome, i.e.~the sum of all pancontigs present once in every genome, identified by PanGraph and the compression of the dataset, i.e.~the ratio for the sum of the lengths of all pancontigs and the sum of the lengths of all input genomes, for different dataset and the different parameters.
+Fig.~\ref{fig:panx-benchmark}B-C and SI Fig.~\crossref{6} quantify the size of the core genome, i.e.~the sum of all pancontigs present once in every genome, identified by PanGraph and the compression of the dataset, i.e.~the ratio for the sum of the lengths of all pancontigs and the sum of the lengths of all input genomes, for different dataset and the different parameters.
 As expected, the very homogenous \textit{M.~tuberculosis} dataset has a large core genome with and a compression ratio that is almost equal to the inverse number of input genomes, independent of the aligner or parameters used.
@@
 
-For species other than \textit{P.~marinus}, the pangraphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in pancontigs spanning several kbps (see SI Fig.~4).
-For these species the use of null energy parameters slightly increases merging and sequence compression, at the cost of slightly more fragmented graphs (cf. SI Section IIIC and SI Fig.~4).
+For species other than \textit{P.~marinus}, the pangraphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in pancontigs spanning several kbps (see SI Fig.~\crossref{6}).
+For these species the use of null energy parameters slightly increases merging and sequence compression, at the cost of slightly more fragmented graphs (cf.
+SI Section \crossref{VC} and SI Fig.~\crossref{6}).
 Overall this benchmark showcases the capabilities and limits of \textit{PanGraph}, and demonstrates the value of adapting the pseudo-energy parameter and choice of alignment kernel to the expected diversity of the dataset considered.
@@
 We performed the same test on species from the other datasets, obtaining similar results on all species whose divergence is compatible with \textit{PanGraph} capabilities (cf.
-SI Section IV and SI Fig.~5).
+\crossref{SI Section VI and SI Fig.~7}).
 

### 2023-02-28 reviewer_response.tex
update

@@
 \newcommand{\crossref}[1]{{\color{purple}\{check cross-reference\}#1}} % cross-reference, to be later checked
+
+\newcommand{\SIalgo}{2}
 
@@
 \response{We agree with the reviewer that it is important to be able visually compare the compression with the number $N$ of isolates for each species, and we thank them for the suggestion.
-We modified fig.~4C in the main text and fig.~4I in SI marking the value of $1/N$ for each species with horizontal lines, and updated the captions accordingly.}
+We modified \crossref{fig.~4C in the main text and fig.~6I in SI} marking the value of $1/N$ for each species with horizontal lines, and updated the captions accordingly.}
 % \Marco{maybe try visualizing it as deviation from "one strain" compression (1/N)?
@@
 \reviewer{R1.3}{Figure 5 has 2 typos: pacontigs and panconting.}
-% \Marco{fixed.}
 
@@
 \reviewer{R1.4}{Throughout the text E. Coli and K. Pneumoniae etc capitalization.}
-% \Marco{fixed.}
+
 \response{The capitalization has been corrected.}
@@
 We used minimap2 as alignment kernel, asm20 sensitivity and values of $\alpha = 100$, $\beta=20$.
-This took 18 hours and 12 minutes to complete and had maximum memory usage of 17Gb.}
+Running on 8 cores this took 18 hours and 12 minutes to complete and had maximum memory usage of 17Gb.}
 
@@
 (3) For genome structure, exploring `hotspots' in the graph as \href{https://www.nature.com/articles/s41467-017-00808-w}{this paper} does could be done using in/out degree of pancontigs.}
-\responsedraft{We have so far not explored the use of pangraph as input for graph based mapping approaches like Pandora or GramTools.
+\response{We have so far not explored the use of pangraph as input for graph based mapping approaches like Pandora or GramTools.
 But this is indeed an obvious downstream application of pangraph.
@@
 \reviewer{R3}{The article ``PanGraph: scalable bacterial pan-genome graph construction'' reports a new method for constructing a Pan-Genome Graph (PGG) in an efficient manner with an open and extractable grah structure.
-Unfortunately, the authors appear to either be unaware of or downplay the existence of previous methods for the same purpose.\textbf{ The authors need to survey the existing literature and compare and contrast with their method to show the utility of their approach.}}
-
-\reviewer{R3}{For instance the authors state: ``Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.''.
+Unfortunately, the authors appear to either be unaware of or downplay the existence of previous methods for the same purpose.\textbf{ The authors need to survey the existing literature and compare and contrast with their method to show the utility of their approach.}\\
+      For instance the authors state: ``Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.''.
 There are many pan-genome analysis tools which treat the pan-genome as a graph and generate a PGG.
@@
       PanGraph appears to wed these approaches to some extent in that the alignments are done at the nucleotide level and the features are extracted as a characteristic of the alignments.}
+
+\responsedraft{We appreciate the reviewer's remarks and apologize fot our summary discussion of the literature.
+To this end we modified the introduction with a more in-depth analysis of the existing methods, and how pangraph compares to them.\\
+      In summary...}
 
@@
 \Marco{Yeah good idea. Maybe something that just map each annotation into a block, without solving the problem of harmonizing annotations which might be some more work...}\\
+\responsedraft{}
 
@@
 One can a posteriori decide whether one wants to split paralogs using structural information (flanking blocks) or one can even prevent the merging using diversity and the $\beta$ parameter.}\\
+\responsedraft{
+      Different design choice.
+      Use of beta parameter for partial control.
+}
 
@@
 For the simulated data how accurate are the final pancontig consensus sequences?}
-\Liam{expanding discussion of doing an MSA at end as for R2's comments would partially address.
+% \Liam{expanding discussion of doing an MSA at end as for R2's comments would partially address.
 Quantifying how the consensus sequence degenerates iteratively is an interesting point that seems possible?}\\
-\Marco{I agree, this could be addressed by adding a more detailed algorithm explanation of the merging algorithm in the SI.
+% \Marco{I agree, this could be addressed by adding a more detailed algorithm explanation of the merging algorithm in the SI.
 The short answer is that at each merging the full alignment is constructed using the CIGAR of the alignment of the consensus of the two blocks.
@@
 Maybe one could devise a proper test set and see how it degenerates over the iterations.}\\
+\responsedraft{This point was raised by other reviewers, and we appreciate the need to make this part of the alorithm clearer.
+To this end we added section \SIalgo in the SI (see ``graph merging''), with a detailed explanation of how the merging procedure works.\\
+      In essence, during graph merging alignment is only performed between the consensus sequences of the merged pancontigs.
+We avoid performing a multiple sequence alignment during merging to maintain the algorithm scalable.
+However we do update the consensus of the pangontig at each merge using the majority rule on the merged alignment.\\
+      The idea behind this approach is that the consensus sequence is needed to find long-range homology, and resolving the fine-grained structure of the alignment is not needed at this step and would require an eccessive amount of resources.
+However after the graph has been built alignments can be refined with \textit{PanGraph}'s \texttt{polish} command.
+This command performs a multiple sequence alignment on all the sequences in each pancontig and updates the consensus sequence accordingly.
+This separates efficiently the problem of resolving the large-scale and nucleotide-level diversity between sequences.\\
+      We did not quantify explicitly the divergence of consensus sequences during graph building, but we observed that, even without ``polishing'', alignments of up to hundreds of sequences look consistent.
+      % TODO: quantify this in simulations if there is time?
+}
 
@@
 The authors should also include analysis on the calculated core genome length versus actual core genome length for the simulated data.}
-\Liam{0.7Mbp does seem lower than I would expect - I guess that they expect $>$2 Mbp - but it shows that obsessing about core is a bit unwise.
+% \Liam{0.7Mbp does seem lower than I would expect - I guess that they expect $>$2 Mbp - but it shows that obsessing about core is a bit unwise.
 With slightly relaxed definitions (95\% of genomes) does the core get sensibly bigger?
@@
 4G for comparison with pangraph's estimation?}\\
+\responsedraft{
+      We thank the reviewer for pointing out the inconsistency in the E.~coli dataset and we apologize for not having investigated this further.\\
+      The isolates analyzed in our papers were downloaded from RefSeq.
+The accession numbers are now explicitly included in the sumbissions (file \texttt{accession\_numbers.csv}).
+We rely on the PanX pipeline \cite{ding2018panx} for the clustering of homologous genes, and thus also for the detection of core genome.
+The value of 0.7 Mbp results from this analysis.
+This value is indeed considerably less than what usually considered core genome for E.~coli.
+      For instance in a recent preprint \cite{horsfield2023accurate} by using pangenome graphs to improve gene annotations Horsefield and colleagues find around 2500 core genes in a set of 162 E.~coli isolates (see table 1 in the publication).\\
+      Since core-genome is sensitive to gene annotation problems, we
+      A deeper analysis of our dataset highlighted that part of the problem might be due to some problematic strains that are have been removed from RefSeq. For example...
+
+      Soft Core...
+      Reasons why this is wrong: misannotations + artificial genomes... This is consistent with misannotations before refining...
+      Added soft genome.
+      Added to the main text in a footnote? Or in the SI?
+      % TODO: core genome for simulated data if there is time?
+}
 
@@
 % - update marginalize figure legend
-% - update SI section numbering in the main text
-
-% meeting notes:
-% - no alignment "line"
-% - [x] csv file 
-% - [ ] supplementary: move graph structure to new section and add two figures from the docs.
-% - [x] R1.1
-% - [x] R1.5
-% - [ ] R1.8
-% - [ ] R2.2
-% - [x] R2.6
-% - [x] R2.9
-% - [x] remove \oldnew and use git diff
-% pangraph automatic testing -> sequence reconstruction
 
@@
 % - [ ] archive repository state and add in the response
+% - [ ] add files with accession numbers and pairs to the submission
 % - [ ] update SI section numbers and line references in the main text and in the response.
 % - [ ] update SI figure numbers.
+% - [ ] add mention of ``polish'' in main text.
+% - [ ] for the future: add sequence reconstruction to pangraph automatic testing.

### 2023-02-28 main.tex
update

@@
 \newcommand{\modified}[1]{{\color{red}#1}}
-% \newcommand{\oldnew}[2]{{\color{gray}\sout{#1}}{\color{red}#2}}
+% \newcommand{\Marco}[1]{{\color{teal}[Marco: #1]}}
 
@@
 The ability to rapidly align large sets of complete genomes of a bacterial species is crucial for the investigation of the processes governing microbial evolution.
-We developed \textit{PanGraph} as a tool that can address this need, by being able to capture the structural and nucleotide diversity in both the core and accessory genome in a scalable way.
+We developed \textit{PanGraph} as a tool that can address this need, by being able to capture the structural and nucleotide diversity in both the core and accessory genome in a scalable way.\\
+Importantly, this diversity is encoded in different ways depending on the scale, with large-scale variation affecting the graph structure, while nucleotide-level variation being included in pancontig alignments.\\
 
@@
 Some example applications on plasmids are available in \textit{PanGraph}'s documentation\footnote{See the tutorial section in \url{https://neherlab.github.io/pangraph}}.\\
+Even if \textit{PanGraph} was not developed with this aim in mind, its outputs can be used as input for other tools that perform variant calling and genotyping using Pangenome Reference Graphs (PRGs), such as \textit{pandora} \cite{colquhoun2021pandora} and \textit{gramtools} \cite{letcher2021gramtools}.
+To this end the MSA of each block can be produced using the \texttt{export} command (see \textit{PanGraph}'s documentation), and can be transformed in a set of PRGs using \texttt{make\_prg} (see \url{https://github.com/iqbal-lab-org/make_prg}).
+We however did not directly test or benchmark these approeaches in the current paper.\\
 

### 2023-03-01 reviewer_response.tex
Update on Overleaf.

@@
 coli} (Table 1 of \href{https://www.biorxiv.org/content/10.1101/2023.01.24.524926v1.full.pdf}{here}).}\\
-\Marco{I checked and soft core (gene in 95\% of strains) is 2Mbp in our dataset.
+% \Marco{I checked and soft core (gene in 95\% of strains) is 2Mbp in our dataset.
 Some of our problems might come from weird strains.
@@
 Interestingly in \cite{sutton2021pan} they report $\sim$700 core orthologous gene clusters for an E.coli dataset of $\sim$1000 genomes, if they do not harmonize annotations.
-Those increase to $\sim$1500 after refining their annotations.\\
-      The easiest thing to do might be to just add the size of soft-core for all species?
-Maybe these values might also be added as dashed lines in fig.
-4G for comparison with pangraph's estimation?}\\
+Those increase to $\sim$1500 after refining their annotations.
+% The easiest thing to do might be to just add the size of soft-core for all species? Maybe these values might also be added as dashed lines in fig. 4G for comparison with pangraph's estimation?}\\
 \responsedraft{
-      We thank the reviewer for pointing out the inconsistency in the E.~coli dataset and we apologize for not having investigated this further.\\
+      We thank the reviewer for pointing out the inconsistency in the E.~coli dataset and we apologize for not having investigated this further before submission.\\
       The isolates analyzed in our papers were downloaded from RefSeq.
-The accession numbers are now explicitly included in the sumbissions (file \texttt{accession\_numbers.csv}).
-We rely on the PanX pipeline \cite{ding2018panx} for the clustering of homologous genes, and thus also for the detection of core genome.
+The accession numbers are now explicitly included in the sumbission (file \texttt{accession\_numbers.csv}).
+These data were previously analyzed in our group with the PanX pipeline \cite{ding2018panx}.
+We rely on this pipeline for the clustering of homologous genes, and thus also for the detection of core genome.
 The value of 0.7 Mbp results from this analysis.
-This value is indeed considerably less than what usually considered core genome for E.~coli.
-      For instance in a recent preprint \cite{horsfield2023accurate} by using pangenome graphs to improve gene annotations Horsefield and colleagues find around 2500 core genes in a set of 162 E.~coli isolates (see table 1 in the publication).\\
-      Since core-genome is sensitive to gene annotation problems, we
-      A deeper analysis of our dataset highlighted that part of the problem might be due to some problematic strains that are have been removed from RefSeq. For example...
+As noticed by the reviewer, this value is indeed smaller than what usually considered core genome for E.~coli.
+      For instance in a recent preprint \cite{horsfield2023accurate} by using pangenome graphs to improve gene annotations Horsefield and colleagues find around 2500 core genes in a set of 162 E.~coli isolates (see table 1 in their publication).\\
+      This discrepancy in size could have different origins: problematic assemblies in our datasets, misannotations or problems with gene clustering.\\
+      As a first step we checked the size of the soft-core genome, defined as the set of gene clusters that are shared by at least 95\% of the strains.
+For our E.~coli dataset this has size of 2 Mbp, which is closer to our expectation.\\
+      We then ranked strains by the amount of soft-core genes that they include, and found that indeed some problematic assemblies are present in the dataset.
+Some of the isolates that have the lowest fraction of soft-core genes in our dataset were subsequently removed from RefSeq, for example \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_020518.1?report=genbank}{NC\_020518}) (an artificial genome) and  \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_017663.1?report=genbank}{NC\_017663}.\\
+      These problematic strains however do not completely explain the small core-genome size that we find, since for example at least one of the most common 1000 genes are missed by 87/307 isolates.\\
+      The remaining discrepancy in size is to be attributed to misannotations.
+      This is a common problem of strictly-core genome in annotation pipelines, if no harmonization strategy is used.
+For example in \cite{sutton2021pan} the authors report finding $\sim$700 core orthologous gene clusters for an E.~coli dataset of $\sim$1000 genomes, prior to applying their harmonization strategy.
+Those increase to $\sim$1500 after refining their annotations.
+
+      We now mention the problematic size of the core genome in the discussion
 

### 2023-03-01 reviewer_response.tex
EC core genome size

@@
 
-\newcommand{\SIalgo}{2}
+\newcommand{\SIalgo}{II}
+\newcommand{\SIsynthdata}{III}
+\newcommand{\SIsimBench}{IV}
+\newcommand{\SIdataBenchmark}{V}
+\newcommand{\SImarginalize}{VI}
+\newcommand{\SIfigMutrate}{4}
+\newcommand{\SIfigKernels}{5}
+\newcommand{\SIfigBenchmark}{6}
+\newcommand{\SIfigMarg}{7}
 
@@
       We appreciate the need to better explain the role of this parameter in the merging procedure and in the structure of the resulting graph.
-To this end we extended its definition in section IB in the main text, and give a more extended discussion in SI section~\crossref{II}.\\
+To this end we extended its definition in section IB in the main text, and give a more extended discussion in SI section~\SIalgo.\\
       % TODO: Homogenization in the main text with discussion on precision. Extend.
@@
 These data were previously analyzed in our group with the PanX pipeline \cite{ding2018panx}.
-We rely on this pipeline for the clustering of homologous genes, and thus also for the detection of core genome.
-The value of 0.7 Mbp results from this analysis.
-As noticed by the reviewer, this value is indeed smaller than what usually considered core genome for E.~coli.
-      For instance in a recent preprint \cite{horsfield2023accurate} by using pangenome graphs to improve gene annotations Horsefield and colleagues find around 2500 core genes in a set of 162 E.~coli isolates (see table 1 in their publication).\\
-      This discrepancy in size could have different origins: problematic assemblies in our datasets, misannotations or problems with gene clustering.\\
+We rely on this pipeline for the clustering of homologous genes, and the values shown in table~1 results from this analysis.
+This is now made explicit in the caption.
+This table is only meant to provide some general characterization of the size and diversity of the various datasets, and PanX analysis results are not used in PanGraph.\\
+
+As noted by the reviewer, the 0.7 Mbp size for E.~coli core genome found by PanX is smaller than what usually reported in the literature.
+Estimates of E.~coli pangenome size are usually around 2000-2500 genes \cite{park2019large,horsfield2023accurate,sutton2021pan}.
+However these estimates often either use a soft-core definition for the core genome \cite{park2019large} or use some form of annotation refinement to mitigate misannotations \cite{horsfield2023accurate,sutton2021pan}.\\
+In our case since we are using a strict definition for core genome the discrepancy could have different origins: problematic assemblies in our datasets, misannotations or problems with gene clustering.\\
 As a first step we checked the size of the soft-core genome, defined as the set of gene clusters that are shared by at least 95\% of the strains.
 For our E.~coli dataset this has size of 2 Mbp, which is closer to our expectation.\\
-      We then ranked strains by the amount of soft-core genes that they include, and found that indeed some problematic assemblies are present in the dataset.
-Some of the isolates that have the lowest fraction of soft-core genes in our dataset were subsequently removed from RefSeq, for example \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_020518.1?report=genbank}{NC\_020518}) (an artificial genome) and  \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_017663.1?report=genbank}{NC\_017663}.\\
-      These problematic strains however do not completely explain the small core-genome size that we find, since for example at least one of the most common 1000 genes are missed by 87/307 isolates.\\
-      The remaining discrepancy in size is to be attributed to misannotations.
-      This is a common problem of strictly-core genome in annotation pipelines, if no harmonization strategy is used.
-For example in \cite{sutton2021pan} the authors report finding $\sim$700 core orthologous gene clusters for an E.~coli dataset of $\sim$1000 genomes, prior to applying their harmonization strategy.
-Those increase to $\sim$1500 after refining their annotations.
+We then ranked strains by the amount of soft-core genes that they include, and found some problematic assemblies.
+Some of the isolates with the lowest fraction of soft-core genes were subsequently removed from RefSeq, for example \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_020518.1?report=genbank}{NC\_020518} (an artificial genome) and  \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_017663.1?report=genbank}{NC\_017663}.\\
+These problematic strains however do not explain the small core-genome size that we find.
+For instance, over the 307 strains in our dataset, 87 are missing at least one of the most common 1000 genes.
+This discrepancy is however compatible with misannotations.
+For instance by using a strict core definition in \cite{sutton2021pan} the authors report finding $\sim$700 core orthologous gene clusters for an E.~coli dataset of $\sim$1000 genomes (see table 2 in their publication).
+This would be compatible with our estimate of 0.7 Mbp.
+After refining the annotations using structural infromation from a pangenome graph, the number of strict core gene clusters increases to $\sim$1500.
+In our PanX analysis we do not perform this harmonization and our esitmate remains biased by annotation errors.\\
 
-      We now mention the problematic size of the core genome in the discussion
-
-      Soft Core...
-      Reasons why this is wrong: misannotations + artificial genomes... This is consistent with misannotations before refining...
-      Added soft genome.
-      Added to the main text in a footnote? Or in the SI?
+To amend this we added the size of soft-core genome for each species in table~1 in the main text, and added part of this discussion in SI sect.~{\SIdataBenchmark}A.
 % TODO: core genome for simulated data if there is time?

### 2023-03-01 main.tex
EC core genome size

@@
 \newcommand{\dcore}{\langle d_\text{core} \rangle}
-\newcommand{\crossref}[1]{{\color{purple}\{check cross-refs\}#1}} % cross-reference, to be later checked
 \newcommand{\modified}[1]{{\color{red}#1}}
 % \newcommand{\Marco}[1]{{\color{teal}[Marco: #1]}}
+
+\newcommand{\SIalgo}{II}
+\newcommand{\SIsynthdata}{III}
+\newcommand{\SIsimBench}{IV}
+\newcommand{\SIdataBenchmark}{V}
+\newcommand{\SImarginalize}{VI}
+\newcommand{\SIfigMutrate}{4}
+\newcommand{\SIfigKernels}{5}
+\newcommand{\SIfigBenchmark}{6}
+\newcommand{\SIfigMarg}{7}
 
@@
 Computationally, each sequence can be sketched into its set of minimizers in linear time while the cardinality of all pairwise intersections can be computed by sorting the list of all minimizers to efficiently count overlaps.
-Hence, the pairwise distance matrix is estimated in a time that increases log-linearly in the length of sequences, and in our implementation can scale sub-quadratically with the number of isolates (see SI sect.~\crossref{II} for details).
+Hence, the pairwise distance matrix is estimated in a time that increases log-linearly in the length of sequences, and in our implementation can scale sub-quadratically with the number of isolates (see SI sect.~\SIalgo for details).
 The final guide tree is constructed as the balanced binary tree constrained to reproduce the topological ordering of leaves found initially.
@@
 The above procedure is repeated until no alignments with negative energy remain.
-More details on the merging procedure and the effect of parameters can be found in SI sect.~\crossref{II}.
+More details on the merging procedure and the effect of parameters can be found in SI sect.~\SIalgo.
 
@@
 We simulated populations of size $N$ and sequence length $L$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
-In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (respective rates $0.01$ and $0.05$ per generation), and horizontal sequence transfer that occurs with tunable rate $h$ per generation per genome (see SI sect.~\crossref{III} for details on the simulation and standard parameter values).
+In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (respective rates $0.01$ and $0.05$ per generation), and horizontal sequence transfer that occurs with tunable rate $h$ per generation per genome (see SI sect.~\SIsynthdata for details on the simulation and standard parameter values).
 The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
@@
 We note that PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of \textit{minimap2} \cite{li2018minimap2}.
-Benchmarks of CPU and memory usage are reported in SI sect~\crossref{IVA}.
+Benchmarks of CPU and memory usage are reported in SI sect~{\SIsimBench}A.
 
@@
 The predominant determinant of accuracy was the sample diversity, controlled by the mutation rate $\mu$ in our simulations (cf.
-SI Fig.~\crossref{4}).
+SI fig.~\SIfigMutrate).
 For low-diversity datasets, breakpoints are inferred with accuracy of few basepairs, while for highly diverged isolates most breakpoints are displaced by several hundreds of basepairs (cf.
-SI Fig.~\crossref{5}).
+SI fig.~\SIfigKernels).
 The choice of alignment kernel influences the threshold diversity at which accuracy is lost, suggesting that inability to detect homology between diverged sequences is the reason.\\
@@
 These data had been previously analyzed using PanX \cite{ding2018panx}, which allowed us to estimate the size of the pangenome and core genome, and the average pairwise divergence on core genes (cf.
-Table~\ref{table:panx-dataset} and SI Section~\crossref{VA}).
+Table~\ref{table:panx-dataset} and SI Section~{\SIdataBenchmark}A).
 
@@
     \end{tabular}
-    \caption{{\bf Dataset properties}. Columns represent:
+    \caption{{\bf Dataset properties}. We use the PanX pipeline to cluster homologous genes and characterize core genome size and diversity of our datasets. Columns represent:
         number of isolates (N.),
@@
         total pangenome length ($\Lpang$),
-        total core genome length ($\Lcore$),
+        total core genome length ($\Lcore$, corresponding to genes present in every isolate),
         total soft-core genome length ($\Lsoftcore$, corresponding to genes shared by at least 95\% of the isolates),
@@
 Using \textit{PanGraph}, we built multiple pangenome graphs for each species in the dataset (cf.
-SI section~\crossref{VB and VC}).
+SI section~{\SIdataBenchmark}B and {\SIdataBenchmark}C).
 These differ by the alignment kernel used (\textit{minimap2} with \textit{asm10} or \textit{asm20} option, and \textit{mmseqs2}) or by the value of the pseudo-energy parameters $\alpha,\beta$ from eq.~(\ref{eq:pseudo-energy}).
@@
 Moreover the use of null energy parameters ($\alpha = \beta = 0$) is expected to remove the threshold divergence of 10\% associated with the standard values of parameters ($\alpha=100$, $\beta=10$) at the cost of more fragmented pangenome graphs.\\
-\textit{PanGraph} can build pangenome graphs comprising hundreds of isolates in a few hours (5h for 307 \textit{E.~coli} isolates with \textit{minimap2} kernel, cf. Fig.~\ref{fig:panx-benchmark}A).
+\textit{PanGraph} can build pangenome graphs comprising hundreds of isolates in a few hours (5h for 307 \textit{E.~coli} isolates with \textit{minimap2} kernel, cf. fig.~\ref{fig:panx-benchmark}A).
 Use of \textit{mmseqs2} consistently requires longer computational time, but provides higher sensitivity when merging highly diverged sequences.
 
-Fig.~\ref{fig:panx-benchmark}B-C and SI Fig.~\crossref{6} quantify the size of the core genome, i.e.~the sum of all pancontigs present once in every genome, identified by PanGraph and the compression of the dataset, i.e.~the ratio for the sum of the lengths of all pancontigs and the sum of the lengths of all input genomes, for different dataset and the different parameters.
+Fig.~\ref{fig:panx-benchmark}B-C and SI fig.~\SIfigBenchmark quantify the size of the core genome, i.e.~the sum of all pancontigs present once in every genome, identified by PanGraph and the compression of the dataset, i.e.~the ratio for the sum of the lengths of all pancontigs and the sum of the lengths of all input genomes, for different dataset and the different parameters.
 As expected, the very homogenous \textit{M.~tuberculosis} dataset has a large core genome with and a compression ratio that is almost equal to the inverse number of input genomes, independent of the aligner or parameters used.
@@
 
-For species other than \textit{P.~marinus}, the pangraphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in pancontigs spanning several kbps (see SI Fig.~\crossref{6}).
+For species other than \textit{P.~marinus}, the pangraphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in pancontigs spanning several kbps (see SI Fig.~\SIfigBenchmark).
 For these species the use of null energy parameters slightly increases merging and sequence compression, at the cost of slightly more fragmented graphs (cf.
-SI Section \crossref{VC} and SI Fig.~\crossref{6}).
+SI Section {\SIdataBenchmark}C and SI fig.~\SIfigBenchmark).
 Overall this benchmark showcases the capabilities and limits of \textit{PanGraph}, and demonstrates the value of adapting the pseudo-energy parameter and choice of alignment kernel to the expected diversity of the dataset considered.
@@
 We performed the same test on species from the other datasets, obtaining similar results on all species whose divergence is compatible with \textit{PanGraph} capabilities (cf.
-\crossref{SI Section VI and SI Fig.~7}).
+SI Section \SImarginalize and SI Fig.~\SIfigMarg).
 

### 2023-03-01 supplementary.tex
EC core genome size

@@
 \newcommand{\red}[1]{{\color{red}#1}}
-% \newcommand{\oldnew}[2]{{\color{gray}\sout{#1}}{\color{red}#2}}
 
@@
 The pipeline takes as input a set of annotated genomes, and it clusters and aligns homologous genes.
-This allowed us to identify core genes, and evaluate their average pairwise sequence divergence $\avg{d_\text{core}}$.
-We also evaluated the average genome length $\avg{\Lgen}$, the total pangenome length $\Lpang$ and core genome length $\Lcore$ (see Table~I in the main text).\\
+This allowed us to identify core genes (i.e.
+genes present in all isolates), and evaluate their average pairwise sequence divergence $\avg{d_\text{core}}$.
+We also evaluated the average genome length $\avg{\Lgen}$, the total pangenome length $\Lpang$, the total core genome length $\Lcore$, and the total soft-core genome length $\Lsoftcore$, where soft-core genes are defined as genes being present in at least 95\% of the isolates.
+These quantities are reported in Table~I in the main text.\\
 EC is the dataset with the highest number of isolates (307) and is also the one with the biggest pangenome.
@@
 Fig.~\ref{fig:benchmark-accuracy-suppl}), but is merged correctly by the \textit{mmseqs2} kernel.
-The sequence divergence of PM sits instead beyond the capabilities of all alignment kernels.
+The sequence divergence of PM sits instead beyond the capabilities of all alignment kernels.\\
+We point out that the core genome size of the EC dataset found by PanX is significantly smaller than typical values seen in the literature, which are closer to 2 Mbp (see for example \cite{horsfield2023accurate}).\\
+We point out that the value of $\Lcore = 0.7$ Mbp for EC is likely to be biased by annotation errors, to which the PanX pipeline is sensitive.
+Values for the core genome size of E.~coli reported in the literature are usually between 1500 and 2500 genes \cite{sutton2021pan,horsfield2023accurate,park2019large}, depending on the core genome definition used and the method used to mitigate annotation errors.
+For example using a strict definition for core genes the authors in \cite{sutton2021pan} initially find a core genome size of $\sim 700$ genes on a dataset of $\sim 1000$ isolates.
+However correcting for annotation errors this estimate increases to $\sim 1500$ genes.
+In our case we do not correct for these errors, since we only use core gene alignment to estimate the diversity of our dataset and since PanGraph is not sensitive to annotation errors.
+But we point out that for the EC dataset the value of $\Lsoftcore$ is likely to provide a better estimate of the core genome size.
 

### 2023-03-01 main.tex
update

@@
 \usepackage{xcolor}
-\usepackage{xr} % external references (reference supplementary)
-\usepackage[normalem]{ulem}
 \usepackage[final]{hyperref}
@@
 
+\newcommand{\SIgraph}{I}
 \newcommand{\SIalgo}{II}
@@
 Pancontigs are directed, meaning their orientation in a path can be either 5' to 3' or vice versa.
+More details on the structure of our graph representation can be found in SI sect.~\SIgraph{}.
 
@@
 Computationally, each sequence can be sketched into its set of minimizers in linear time while the cardinality of all pairwise intersections can be computed by sorting the list of all minimizers to efficiently count overlaps.
-Hence, the pairwise distance matrix is estimated in a time that increases log-linearly in the length of sequences, and in our implementation can scale sub-quadratically with the number of isolates (see SI sect.~\SIalgo for details).
+Hence, the pairwise distance matrix is estimated in a time that increases log-linearly in the length of sequences, and in our implementation can scale sub-quadratically with the number of isolates (see SI sect.~\SIalgo{} for details).
 The final guide tree is constructed as the balanced binary tree constrained to reproduce the topological ordering of leaves found initially.

### 2023-03-02 reviewer_response.tex
update
chore

@@
       \caption{frequency and length distribution (respectively left and right panels) of pancontigs for the \textit{Helicobacter pylori} and \textit{Klebsiella pneumoniae} datasets (top and bottom).
-For each panel we report the results for pangenome graphs built using minimap2 aligner with standard parameter value and asm20 sensitivity (blue) and with mmseqs2 and null pseudo-energy parameters (orange).
+For each panel we report the results for pangenome graphs built using minimap2 aligner with standard parameter values and asm20 sensitivity (blue) and with mmseqs2 and null pseudo-energy parameters (orange).
 Distributions are weighed by pancontig length, so that the area under the curve represents the total pangenome graph size.}
@@
 The mechanism for this is obvious but providing a useful user interface might be tedious.}
-\Liam{I think perhaps producing a helper python script that converts a genbank annotation file for a genome within a pangraph to an output gff/csv with the block locations and coordinates would please this reviewer.}\\
-\Marco{Yeah good idea. Maybe something that just map each annotation into a block, without solving the problem of harmonizing annotations which might be some more work...}\\
+% \Liam{I think perhaps producing a helper python script that converts a genbank annotation file for a genome within a pangraph to an output gff/csv with the block locations and coordinates would please this reviewer.}\\
+% \Marco{Yeah good idea. Maybe something that just map each annotation into a block, without solving the problem of harmonizing annotations which might be some more work...}\\
 \responsedraft{}
@@
       \end{itemize}}
+\responsedraft{}
 
@@
 This could be measured as part of the simulated data where identical paralog sequences of various lengths and copy number could be modeled.}
-\Liam{I am not sure I agree that constructing a pangenome graph means one must distinguish orthologs from paralogs.
+% \Liam{I am not sure I agree that constructing a pangenome graph means one must distinguish orthologs from paralogs.
 The structure is still useful even if this isn't done.
 Maybe there is a way of doing something of what they propose with simulations, but I suspect this isn't necessary for the editor to accept the paper\ldots}\\
-\Marco{I agree, I think it's a different design choice.
+% \Marco{I agree, I think it's a different design choice.
 We could also mention that the marginalize command has an option to project out paralogs, but should be a post-processing step rather than something that we want to do while building the graph.
@@
 \responsedraft{
+      We agree with the reviewer's
       Different design choice.
       Use of beta parameter for partial control.
+      Some form of PanX postprocessing paralog split. Partially done.
 }
@@
 \reviewer{R3}{The authors might want to consider some form of post processing of the PGG to collapse what have been called bubbles in the graph if alignment of the pancontigs on different edges of the bubble warrant it.}
-\Marco{if I understand correctly this is hinting at some form of post-processing step in which artifactual small blocks - generated at the edges of long block merges - could be absorbed in the flanking block alignments and removed?}\\
+% \Marco{if I understand correctly this is hinting at some form of post-processing step in which artifactual small blocks - generated at the edges of long block merges - could be absorbed in the flanking block alignments and removed?}\\
+\responsedraft{
+      We appreciate the importance of the point raised by the reviewer.
+      In it's current version \textit{PanGraph} does not support a post-processing refinement step in which alignments of adjacent edges of pancontigs are harmonized, potentially also removing very small pancontigs when these could be appropriately absorbed in the alignment of larger ones.
+      We agree with the reviewer that this post-processing smoothing of the graph would be an important improvement, and we consider adding it in subsequent versions of \textit{PanGraph}.
+}
 
@@
 I am very concerned about the potential fracturing of the PGG where unfortunately the devil is in the details and am not convinced that the fracturing will not prevent some limit on the number of genomes which can be used for a pangenome.}
-\Marco{if the fracturing derives from a real structural property of pangenomes (i.e.
+% \Marco{if the fracturing derives from a real structural property of pangenomes (i.e.
 ISs that really move around and split long blocks) this might be an intrinsic limit that derives from the pangenome, I wouldn't consider it a problem with PanGraph.}\\
+\responsedraft{
+      We thank the reviewer for their appreciation of the paper and of our methodology.\\
+      We agree with their concern that the level of fragmentation of the graph constitutes a limitation in the number and diversity of genomes that can meaningfully be added in the pangenome graph.\\
+      An increase in the number or diversity of isolates included in the graph will cause an increase in structural breakpoints in the graph.
+If subsequent structural changes occurr at a scale smaller than $\Lthr$, these will not correctly be captured by the graph topology.
+An excessive number of pancontigs would also increase the amount of computational resources used in the alignment.\\
+      % Other than being introduced by the diversity of the dataset, fragmentation could also be introduced artificially in the merging procedure.
+We tested for excessive fragmentation in simulated data using partition entropy, comparing the real underlying structural partition to the one reconstructed by pangraph (not shown in the paper).
+We found that fragmentation was present only when the diversity of the sample was too high for the sensitivity of the aligner.\\
+      Irrespective of its origin, a possible way to detect excessive fragmentation and check for consistency in the graph structure is to visualize the distribution of block size in the graph (see for instance fig.~\ref{fig:kp-vs-hp} lower right panel).
+Excessive fragmentation in the graph would be detectable as most of the sequence being condensed in blocks having size close to $\Lthr$.\\
+      (As soon as they are done I could put the time and block size distribution for graphs with 500/1000 E.coli.)
+      % TODO: show distributions for pangenome graphs with 500/1k isolates? Mention times?
+      % TODO: add this discussion in the text?
+      % TODO: mention the partition entropy? This could be an argument against artificial fragmentation.
+      % TODO: maybe add in the SI the block frequency distribution for different species/kernels?
+}
 
@@
 This could also be investigated for PanGrah using suitable simulated data.}
-\Liam{I think nothing to say here apart from `yes, active IS elements split pancontigs up' (and can themselves be split up due to similarity with other IS\ldots).
+% \Liam{I think nothing to say here apart from `yes, active IS elements split pancontigs up' (and can themselves be split up due to similarity with other IS\ldots).
 For the future one might want to have pangenome methods where you can explicitly pass a list of predefined sequences that should not be split.
 But I think this is a general problem and not something to investigate with simulations or fix here.}\\
-\Marco{agree. I think this is a structural property that we want to capture, the fact that these pancontigs are split.}
+% \Marco{agree. I think this is a structural property that we want to capture, the fact that these pancontigs are split.}
+\responsedraft{
+      We agree with the reviewer's observation that events such as the presence of an IS in an orthologous gene cluster would split the cluster in multiple pancontigs.
+For datasets in which isolates have a high amount of IS (larger than $\Lthr$), this would indeed result in multiple splits in the graph structure.\\
+      This is in part an explicit design choice in \textit{PanGraph}.
+We plan to use this tool for the study of the evolution of bacterial pangenome, and we wanted this feature of the pangenome - namely the fragmentation due to presence of the same IS in multiple contexts - to be explcitly encoded in the graph.
+By condensing all occurrences of such an IS in a single pancontig we also condense the information on all of the different contexts in which the said IS appears, at the cost of fragmenting the graph.
+We are already using this property in a work in preparation \cite{shaw_preparation}, in we compare the similarity between different contexts to obtain insights in the evolution of such sequences.
+This is especially important given the fact that IS often carry limited phylogenetic information, due to their limited size.
+      % TODO: maybe a bit off the mark? Not necessary to talk about the work in preparation?
+}
 
@@
 % - [ ] update SI figure numbers.
-% - [ ] add mention of ``polish'' in main text.
+% - [x] add mention of ``polish'' in main text.
 % - [ ] for the future: add sequence reconstruction to pangraph automatic testing.
+% - [ ] check that all SI sections have been mentioned in the main text
+% - [ ] add in SI the frequency-size distribution of blocks for the different species? To discuss fragmentation?

### 2023-03-02 main.tex
update

@@
 Being able to capture and represent this diversity is key to understanding their evolution, particularly on epidemiologically relevant timescales where very few SNPs accumulate but large-scale structural changes are frequent \cite{Sheppard2016, Noll2018}.
-In this sense, the pangenome graph is a naturally powerful representation, and \textit{PanGraph} can be used to explore this diversity \cite{liam_preparation}.
+In this sense, the pangenome graph is a naturally powerful representation, and \textit{PanGraph} can be used to explore this diversity \cite{shaw_preparation}.
 Some example applications on plasmids are available in \textit{PanGraph}'s documentation\footnote{See the tutorial section in \url{https://neherlab.github.io/pangraph}}.\\

### 2023-03-02 reviewer_response.tex
update

@@
 \newcommand{\responsedraft}[1]{{\it {\color{purple}\textbf{ResponseDraft:} #1}}\vskip 5mm}
-\newcommand{\crossref}[1]{{\color{purple}\{check cross-reference\}#1}} % cross-reference, to be later checked
 
@@
 
-\responsedraft{We thank the reviewer for their appreciation and for their critical review of our work.}
+\responsedraft{We thank the reviewer for their appreciation of our contribution and for their critical review of our work.}
 
@@
 \response{We agree with the reviewer that it is important to be able visually compare the compression with the number $N$ of isolates for each species, and we thank them for the suggestion.
-We modified \crossref{fig.~4C in the main text and fig.~6I in SI} marking the value of $1/N$ for each species with horizontal lines, and updated the captions accordingly.}
+We modified fig.~4C in the main text and fig.~{\SIfigBenchmark}I in SI marking the value of $1/N$ for each species with horizontal lines, and updated the captions accordingly.}
 % \Marco{maybe try visualizing it as deviation from "one strain" compression (1/N)?
@@
 % \Marco{Yeah good idea. Maybe something that just map each annotation into a block, without solving the problem of harmonizing annotations which might be some more work...}\\
-\responsedraft{}
+\responsedraft{We thank the reviewer for this important suggestion.
+We appreciate the need to include annotation in our pangenome graph structure.
+We created a script to decorate each annotation in a \texttt{gff} file with information on the pancontig location.
+The script and instructions on its usage can be found at \url{https://github.com/liampshaw/pangraph\_annotations}.
+We plan on extending this in the future to use directly the pangenome graph as a reference and add annotations as features of pancontigs.
+      % TODO: mention this in the text.
+      % TODO: make the repository public.
+}
 
@@
 \responsedraft{
+      We thank the reviewer for raising an important point that necessitates of more discussion in the paper.
+      As stated by the reviewer, we perform merges prioritizing longer and less diverged alignments.
+However \textit{PanGraph} in its current state does not explicity distinguish between orthologs and paralogs.
+
+      Currently two different ways exist of performing some effective form of paralog splitting. The first one consist in tuning
+
+      In the future we plan on providing an option to perform a more robust form of paralag splitting.
+This could be done using both topological information on the neighborhood of a duplicated pancontig, as currently done in the \texttt{marginalize} command, and phylogenetic information, by building a phylogenetic tree from the alignment of each pancontig and splitting diverged branches.
+This functionality is not however present in the current version of \textit{PanGraph}.
+
       We agree with the reviewer's
@@
       Use of beta parameter for partial control.
-      Some form of PanX postprocessing paralog split. Partially done.
+      Some form of PanX postprocessing paralog split. Build trees for every pancontig and define some split criterion.
+      Partially done in marginalize (\texttt{--reduce-paralog} flag)
 }
@@
 % - [ ] add in SI the frequency-size distribution of blocks for the different species? To discuss fragmentation?
+% - [ ] extend the code to map annotations on pancontigs?

### 2023-03-02 reviewer_response.tex
update

@@
 However \textit{PanGraph} in its current state does not explicity distinguish between orthologs and paralogs.
+      This is partially a design choice, as better explained in the answer to the last question.
+In short, the possibility of grouping duplicated sequences in the same pancontig will also group information on their context, and the fragmentation introduced in the graph would reflect some real (potentially ancient) structural change in the genome.\\
 
-      Currently two different ways exist of performing some effective form of paralog splitting. The first one consist in tuning
+      However, as the reviewer points out, in the case of distantly related paralogs, such a grouping might be undesirable as it introduces unneeded complications in the graph.
+Currently \textit{PanGraph} supports two different ways of partially controlling paralog splitting.\\
+      The first one is based on sequence divergence.
+By tuning the value of the energy parameter $\beta$ the user can set a divergence limit.
+This would prevent the merging of paralogs that have sequence divergence higher than $1/\beta$.\\
+      The second method relies instead on context.
+The \texttt{marginalize} command possesses the optional \texttt{--reduce-paralog} flag.
+When set, this flag will result in the removal of artificial splits caused by duplicated pancontigs that always occurr in the same non-duplicated context.
+These duplicated pancontigs are then split and absorbed in their context.\\
 
-      In the future we plan on providing an option to perform a more robust form of paralag splitting.
-This could be done using both topological information on the neighborhood of a duplicated pancontig, as currently done in the \texttt{marginalize} command, and phylogenetic information, by building a phylogenetic tree from the alignment of each pancontig and splitting diverged branches.
-This functionality is not however present in the current version of \textit{PanGraph}.
+      In the future we plan on providing a dedicated command to perform a more robust form of paralag splitting, which combines both sequence divergence information and context information.
+This command would build a phylogenetic tree from the alignment of each duplicated pancontig, to detect exceptionally long branches, and check whether the split correlates with a split in the duplication context.
+This functionality is not however not yet implemented in the current version of \textit{PanGraph}.
 
-      We agree with the reviewer's
-      Different design choice.
-      Use of beta parameter for partial control.
-      Some form of PanX postprocessing paralog split. Build trees for every pancontig and define some split criterion.
-      Partially done in marginalize (\texttt{--reduce-paralog} flag)
+      % TODO: if this is sensible add a paragraph in SI about paralog splitting
+      % TODO: confirm with Nick the paralog split algorithm.
 }

### 2023-03-03 reviewer_response.tex
chore: update

@@
 
-\responsedraft{We appreciate the reviewer's remarks and apologize fot our summary discussion of the literature.
+\responsedraft{
+      We appreciate the reviewer's remarks and apologize fot our summary discussion of the literature.
 To this end we modified the introduction with a more in-depth analysis of the existing methods, and how pangraph compares to them.\\
-      In summary...}
+      We modified the introduction adding citations to more methods that use pangenome graphs for either improvement of annotations...
+      % SURVEY OF THE LITERATURE
+      % PGG have been used to:
+      % - improve annotations [sutton + ggcaller]
+      % MERTIS OF OUR METHOD
+      % - de Bruijn graphs represent SNPs and large scale variation in the same way, as edges in the graph.
+We instead have this scale separation and encode nucleotide-level variation in alignments, and structural changes in the graph
+      % - PGG based on genes are sensitive to annotation errors. We are based on homology.
+      % - Scalable algorithm.
+}
 
@@
 This command would build a phylogenetic tree from the alignment of each duplicated pancontig, to detect exceptionally long branches, and check whether the split correlates with a split in the duplication context.
-This functionality is not however not yet implemented in the current version of \textit{PanGraph}.
+This functionality is not however not yet implemented in the current version of \textit{PanGraph}.\\
+
+We now make this explicit in SI sect.~{\SIalgo}D, that now contains these considerations.
 
@@
 
-% meeting notes:
-% SIAS -> (new) Supplementary Information Algorithm Section
-% make changes in the text traceable in the reviewer answers.
-% Reviewer 1:
-% - for blocks, mention that consensus is considered as the centroid of a star-shaped phylogeny?
-% - 2) add line to fig. on compression
-% - 5) the effect comes from the fact that core genes are in long syntenic stretches, compared to accessory that are often in messy islands (cite Rocha?). Just add half sentence to the paper.
-% - 6) annotate the figure with numbers
-% - 8) reference: graph alighner, ggCaller? Many take GFA as input.\
-% - 10) this is something that could be explained in the SIAS: one starts with long and shallow blocks and as graphs are merged and propagated down the tree the blocks become shorter and deeper.
-Removing energy threshold just causes more merging and deeper but shorter blocks.
-(figure?)
-% - test pangraph with 500 E.coli.
-% Reviewer 2:
-% - 1) do not mention blocks, only pancontigs
-% - specify better the meaning of null energy parameters and why more merging.
-% Reviewer 3:
-% - paralog merging is partially controlled by beta. As future direction one should build trees for every blok and split paralogs in a PanX fashion. 
-% - more nuanced treatment of the literature.
-% - update marginalize figure legend
-
 %  TODO: before submission
@@
 % - [ ] add in SI the frequency-size distribution of blocks for the different species? To discuss fragmentation?
+% - [ ] make Liam's repo public
 % - [ ] extend the code to map annotations on pancontigs?

### 2023-03-03 supplementary.tex
update
chore: update

@@
 \usepackage{enumerate}
-\usepackage{xr} % reference to other external latex documents
-\usepackage[normalem]{ulem}
 \usepackage{xcolor}
@@
 
-\subsection*{guide tree construction and distance matrix estimation}
+\subsection{guide tree construction and distance matrix estimation}
 
@@
 
-\subsection*{graph merging}
+\subsection{graph merging}
 
@@
 
-\subsection*{PanGraph parameters effect}
+\subsection{PanGraph parameters effect}
 
@@
 % temporal horizon set with beta.
+
+\subsection{Paralog splitting}
+
+By design, in \textit{PanGraph} duplicated sequences (of size larger than $\Lthr$) will be merged in the same pancontig.
+This is done by design, so that positions in which these duplications are present will correspond to a structural breakpoint in the graph, and information on all of the contexts in which a duplicated region appears will be grouped together in the edges connected to this duplicated pancontig.\\
+This is however not always desirable.
+For example in the case of distantly related paralogs the user might want to avoid unnecessary structural complication in the graph, and prevent merging these sequences together if they are generated by an ancient duplication event.\\
+
+This paralog splitting can be done using two different sources of information: either sequence diversity or local context in which the duplication is inserted.
+Currently \textit{PanGraph} does not provide a command to perform paralog splitting.
+This is a feature that we plan on implementing in the future.
+However partial control of paralog split can currently be achieved in two ways.\\
+The first one is based on sequence divergence.
+By tuning the value of the energy parameter $\beta$ the user can set a divergence limit.
+This would prevent the merging of paralogs that have sequence divergence higher than $1/\beta$ during graph construction.\\
+The second method relies instead on context.
+The \texttt{marginalize} command possesses the optional \texttt{--reduce-paralog} flag.
+When set, this flag will result in the removal of artificial splits caused by duplicated pancontigs that always occurr in the same non-duplicated context.
+These duplicated pancontigs are then split and absorbed in their context, thus simplifying the graph and loosing trace of the duplication.\\
 

### 2023-03-03 reviewer_response.tex
update

@@
 \responsedraft{
-      We appreciate the reviewer's remarks and apologize fot our summary discussion of the literature.
-To this end we modified the introduction with a more in-depth analysis of the existing methods, and how pangraph compares to them.\\
-      We modified the introduction adding citations to more methods that use pangenome graphs for either improvement of annotations...
+      We apologize if our previous discussion of the literature was incomplete, it was not our intention to downplay prior work.
+      We thank the reviewer for highlighting additional important work and aspects that we previously didn't discuss.
+      We have rewritten the part in the introduction to reflect these the work by these and other authors and added a few sentences to the discussion to further highlight previous and alternative approaches.
+We hope that these edits address the concerns of the reviewer.
       % SURVEY OF THE LITERATURE
       % PGG have been used to:
-      % - improve annotations [sutton + ggcaller]
+      % - improve annotations [sutton 2021] + [ggcaller]
       % MERTIS OF OUR METHOD

### 2023-03-03 main.tex
update

@@
 
-% TODO: SI Section IIA: superfluous?
-% TODO: minimap2 asm10 superfluous?
-
 %\section{Introduction}
@@
 While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging.
-Colored generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets, although the underlying efficiency derives from a fixed k-mer size which prevents modelling long-range homology \cite{iqbal2012novo,muggli2017succinct}.
+Colored and compacted generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets \cite{iqbal2012novo,muggli2017succinct}.
+Currently tools exist to build graphs containing thousands of isolates \cite{holley2020bifrost}, and have proven useful in problems such as the detection of core pangenome \cite{schulz2022sequence} or the improvement of gene prediction, annotation and clustering \cite{horsfield2023accurate}.
+However in these methods the graph structure encodes at the same time nucleotide-level variation and large-scale structural rearrangements in the genome.
+
 An orthogonal approach has been to formulate the inference of the pangenome graph as a multiple genome alignment.

### 2023-03-03 reviewer_response.tex
update

@@
 
-\newcommand{\Marco}[1]{{\color{gray}Marco: #1}}
+\newcommand{\Marco}[1]{{\color{orange}Marco: #1}}
 \newcommand{\Liam}[1]{{\color{teal}Liam: #1}}
@@
 
-\responsedraft{
+\response{
       We apologize if our previous discussion of the literature was incomplete, it was not our intention to downplay prior work.
@@
 
+\Marco{
+      Finish literature review.
+      Add part on graph-based methods.
+      In the discussion add edge cases.
+      Cite all the ones suggested by the reviewer.
+}
+
 \reviewer{R3}{An obvious lack in PanGraph is an attempt to map existing annotations back onto the generated PGG which is a necessity for usefulness to biologists.
@@
 % \Marco{Yeah good idea. Maybe something that just map each annotation into a block, without solving the problem of harmonizing annotations which might be some more work...}\\
-\responsedraft{We thank the reviewer for this important suggestion.
+\response{We thank the reviewer for this important suggestion.
 We appreciate the need to include annotation in our pangenome graph structure.
@@
 \responsedraft{}
+\Marco{Copy form above and extend the answer}
 
@@
 }
+\Marco{Summarize, make more to the point. Say that this is now better discussed in SI.}
 
@@
 Maybe one could devise a proper test set and see how it degenerates over the iterations.}\\
-\responsedraft{This point was raised by other reviewers, and we appreciate the need to make this part of the alorithm clearer.
-To this end we added section \SIalgo in the SI (see ``graph merging''), with a detailed explanation of how the merging procedure works.\\
+\response{This point was raised by other reviewers, and we appreciate the need to make this part of the alorithm clearer.
+To this end we added section {\SIalgo} in the SI (see ``graph merging''), with a detailed explanation of how the merging procedure works.\\
       In essence, during graph merging alignment is only performed between the consensus sequences of the merged pancontigs.
@@
 }
+\Marco{Make more to the point.
+Data are from PanX (archived run May 2018, not the version used in the paper).
+Add the date.
+First point at the reason: core genome and not soft core.
+Similar problems in other pipelines.
+Also other, but not dramatic.
+Give numbers}
+
+%     "all core": 778,
+%     "all softcore": 2081,
+%     "all genes": 23128,
+%     "all nondupl": 18798,
+%     "drop core": 822,
+%     "drop softcore": 2069,
+%     "drop genes": 23054,
+%     "drop nondupl": 18749
 
@@
 % \Marco{if I understand correctly this is hinting at some form of post-processing step in which artifactual small blocks - generated at the edges of long block merges - could be absorbed in the flanking block alignments and removed?}\\
-\responsedraft{
+\response{
       We appreciate the importance of the point raised by the reviewer.
@@
 }
+\Marco{It is an important concern. We have investigated it in the paper, good scaling... Can be detected if wrong...}
 
@@
 }
+\Marco{Future work, requires further research. Will consider adding specific procedures for IS (e.g. give a list of IS sequences that could be excluded...)}
 

### 2023-03-04 reviewer_response.tex
updated panx data answer

@@
 % The easiest thing to do might be to just add the size of soft-core for all species? Maybe these values might also be added as dashed lines in fig. 4G for comparison with pangraph's estimation?}\\
-\responsedraft{
-We thank the reviewer for pointing out the inconsistency in the E.~coli dataset and we apologize for not having investigated this further before submission.\\
-The isolates analyzed in our papers were downloaded from RefSeq.
+\response{
+We thank the reviewer for pointing out this issue with the E.~coli dataset.\\
+
+Concerning the origin the data, our datasets are taken from the PanX pipeline (\url{https://pangenome.org/}).
+They were not included in the original publication but are part of a later update, dated March 2018.
+Values displayed in table I are extracted from the PanX analysis.
+This is now made more explicit in the text.\\
+
+Concerning the core genome size, the unusually small value of 0.7 Mbp is due to the fact that we used a strict definition of core genes as single-copy genes that are present in every isolate.
+This definition is very sensitive to annotation errors.\footnote{
+      For example in \cite{sutton2021pan} using a similar strict definition the authors report finding $\sim$700 core orthologous gene clusters for an E.~coli dataset of $\sim$1000 genomes (see table 2 in their publication), compatible with our estimate of 0.7 Mbp.
+They impute this to misannotations, since after refining the annotations using structural infromation from the pangenome this number increases to $\sim$1500 genes.
+} If we measure the size of the soft-core genome with a frequency threshold of 95\% we find a more reasonable size of 2 Mbp. We now report both of these values in table~I.\\
+A careful investigation of the dataset also revealed that a pair of records have been subsequently removed from RefSeq, namely \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_020518.1?report=genbank}{NC\_020518} (an artificial genome) and  \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_017663.1?report=genbank}{NC\_017663}.
+However these are not at the origin of the small size of the core genome.
+In fact by removing these isolates the number of core genes only shows a very limited increase from 778 to 822, and the number of soft-core genes decreases from 2081 to 2069.\\
+
+To make this more clear we updated part of the text in section IIB, the caption of table~I and SI sect.~{\SIdataBenchmark}A.
+}
+% The isolates analyzed in our papers were downloaded from RefSeq.
 The accession numbers are now explicitly included in the sumbission (file \texttt{accession\_numbers.csv}).
@@
 
-As noted by the reviewer, the 0.7 Mbp size for E.~coli core genome found by PanX is smaller than what usually reported in the literature.
+% As noted by the reviewer, the 0.7 Mbp size for E.~coli core genome found by PanX is smaller than what usually reported in the literature.
 Estimates of E.~coli pangenome size are usually around 2000-2500 genes \cite{park2019large,horsfield2023accurate,sutton2021pan}.
 However these estimates often either use a soft-core definition for the core genome \cite{park2019large} or use some form of annotation refinement to mitigate misannotations \cite{horsfield2023accurate,sutton2021pan}.\\
-In our case since we are using a strict definition for core genome the discrepancy could have different origins: problematic assemblies in our datasets, misannotations or problems with gene clustering.\\
-As a first step we checked the size of the soft-core genome, defined as the set of gene clusters that are shared by at least 95\% of the strains.
+% In our case since we are using a strict definition for core genome the discrepancy could have different origins: problematic assemblies in our datasets, misannotations or problems with gene clustering.\\
+% As a first step we checked the size of the soft-core genome, defined as the set of gene clusters that are shared by at least 95\% of the strains.
 For our E.~coli dataset this has size of 2 Mbp, which is closer to our expectation.\\
-We then ranked strains by the amount of soft-core genes that they include, and found some problematic assemblies.
+% We then ranked strains by the amount of soft-core genes that they include, and found some problematic assemblies.
 Some of the isolates with the lowest fraction of soft-core genes were subsequently removed from RefSeq, for example \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_020518.1?report=genbank}{NC\_020518} (an artificial genome) and  \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_017663.1?report=genbank}{NC\_017663}.\\
-These problematic strains however do not explain the small core-genome size that we find.
+% These problematic strains however do not explain the small core-genome size that we find.
 For instance, over the 307 strains in our dataset, 87 are missing at least one of the most common 1000 genes.
@@
 
-To amend this we added the size of soft-core genome for each species in table~1 in the main text, and added part of this discussion in SI sect.~{\SIdataBenchmark}A.
+% To amend this we added the size of soft-core genome for each species in table~1 in the main text, and added part of this discussion in SI sect.~{\SIdataBenchmark}A.
 % TODO: core genome for simulated data if there is time?
-}
-\Marco{Make more to the point.
-Data are from PanX (archived run May 2018, not the version used in the paper).
-Add the date.
-First point at the reason: core genome and not soft core.
-Similar problems in other pipelines.
-Also other, but not dramatic.
-Give numbers}
 

### 2023-03-04 main.tex
updated panx data answer

@@
 We downloaded from RefSeq \cite{o2016reference} completely assembled chromosomes from 5 different bacterial species: \textit{Escherichia coli} (EC), \textit{Klebsiella pneumoniae} (KP), \textit{Helicobacter pylori} (HP), \textit{Mycobacterium tuberculosis} (MT) and \textit{Prochlorococcus marinus} (PM).
+These data had been previously analyzed using PanX \cite{ding2018panx}.
+Using the PanX analysis we estimated the size of the pangenome and core genome, and the average pairwise divergence on core genes (cf.
+Table~\ref{table:panx-dataset} and SI Section~{\SIdataBenchmark}A).
+
+% We downloaded from RefSeq \cite{o2016reference} completely assembled chromosomes from 5 different bacterial species: \textit{Escherichia coli} (EC), \textit{Klebsiella pneumoniae} (KP), \textit{Helicobacter pylori} (HP), \textit{Mycobacterium tuberculosis} (MT) and \textit{Prochlorococcus marinus} (PM).
 These data had been previously analyzed using PanX \cite{ding2018panx}, which allowed us to estimate the size of the pangenome and core genome, and the average pairwise divergence on core genes (cf.

### 2023-03-04 supplementary.tex
updated panx data answer

@@
 In Table~I in the main text we report some summary statistics for these datasets.
-The number of isolates per species ranges from several hundreds for EC to 10 for PM.
-These data were previously analyzed using PanX \cite{ding2018panx}, a pipeline for the analysis of bacterial pangenome.
+The number of isolates per species ranges from several hundreds for EC to 10 for PM.\\
+These data has been chosen as part of a PanX \cite{ding2018panx} analysis dated March 2018.\footnote{
+    We report for completeness that two of the isolates from the \textit{E.~coli} dataset from this analysis, namely \texttt{NC\_020518} and  \texttt{NC\_017663}, were subsequently removed from RefSeq.
+We verified that their removal from the dataset does not significantly impact core-genome and soft-core genome size, but decided to utilize the complete dataset as per original PanX analysis.
+}
+PanX is a pipeline for the analysis of bacterial pangenome.
 The pipeline takes as input a set of annotated genomes, and it clusters and aligns homologous genes.
-This allowed us to identify core genes (i.e.
-genes present in all isolates), and evaluate their average pairwise sequence divergence $\avg{d_\text{core}}$.
-We also evaluated the average genome length $\avg{\Lgen}$, the total pangenome length $\Lpang$, the total core genome length $\Lcore$, and the total soft-core genome length $\Lsoftcore$, where soft-core genes are defined as genes being present in at least 95\% of the isolates.
-These quantities are reported in Table~I in the main text.\\
+We used the PanX analysis to estimate the dataset diversity.\\
+From this analysis we evaluated the average pairwise sequence divergence $\avg{d_\text{core}}$ of core genes, the average genome length $\avg{\Lgen}$, the total pangenome length $\Lpang$, the total core genome length $\Lcore$, and the total soft-core genome length $\Lsoftcore$, where soft-core genes are defined as genes being present in at least 95\% of the isolates.
+These quantities are reported in Table~I in the main text.\footnote{
+    $\Lcore$ is the length of the alignment used to estimate sequence divergence $\avg{d_\text{core}}$.
+However this quantity is sensitive to annotation errors.
+This is particularly evident in the \textit{E.~coli} dataset, for which $\Lcore = 0.7$ Mbp is likelt to be a subestimation of the real core genome size.
+$\Lsoftcore$ is instead less sensitive to annotation errors and usually considered a better core-genome size estimate.
+For example the \textit{E.~coli} core genome is usually considered to be aroud 1500-2500 genes in size, depending on the definition used \cite{park2019large,sutton2021pan,horsfield2023accurate}.
+}\\
 EC is the dataset with the highest number of isolates (307) and is also the one with the biggest pangenome.
@@
 Fig.~\ref{fig:benchmark-accuracy-suppl}), but is merged correctly by the \textit{mmseqs2} kernel.
-The sequence divergence of PM sits instead beyond the capabilities of all alignment kernels.\\
-We point out that the core genome size of the EC dataset found by PanX is significantly smaller than typical values seen in the literature, which are closer to 2 Mbp (see for example \cite{horsfield2023accurate}).\\
-We point out that the value of $\Lcore = 0.7$ Mbp for EC is likely to be biased by annotation errors, to which the PanX pipeline is sensitive.
-Values for the core genome size of E.~coli reported in the literature are usually between 1500 and 2500 genes \cite{sutton2021pan,horsfield2023accurate,park2019large}, depending on the core genome definition used and the method used to mitigate annotation errors.
-For example using a strict definition for core genes the authors in \cite{sutton2021pan} initially find a core genome size of $\sim 700$ genes on a dataset of $\sim 1000$ isolates.
-However correcting for annotation errors this estimate increases to $\sim 1500$ genes.
-In our case we do not correct for these errors, since we only use core gene alignment to estimate the diversity of our dataset and since PanGraph is not sensitive to annotation errors.
-But we point out that for the EC dataset the value of $\Lsoftcore$ is likely to provide a better estimate of the core genome size.
+The sequence divergence of PM sits instead beyond the capabilities of all alignment kernels.
 

### 2023-03-04 reviewer_response.tex
update

@@
 \response{The observation that core pancontigs are longer than average is in agreement with our prior the expectation since multiple core genes are often syntenic even in diverse species, compared to accessory genes that are often found in different contexts.
-This effect seems to outweigh the possibility of increased fragmentation due to the large number of mergers that common (core) blocks undergo.
+This effect seems to outweigh the possibility of increased fragmentation due to the large number of mergers that common (core) pancontigs undergo.
       We modified the line to: ``Core pancontigs having higher-than-average length. This is expected given that core genes tend to stay synteny over large evolutionary distances.''}
@@
 \response{The minimal size parameter $\Lthr$ is set to 100 bp in PanGraph.
-This parameter controls the minimal size of blocks.
+This parameter controls the minimal size of pancontigs.
 In general, this parameter sets a scale separation between structural variation that is encoded in the graph structure or in the pancontig alignment.
@@
 However we do update the consensus of the pangontig at each merge using the majority rule on the merged alignment.\\
-      The idea behind this approach is that the consensus sequence is needed to find long-range homology, and resolving the fine-grained structure of the alignment is not needed at this step and would require an eccessive amount of resources.
+      The idea behind this approach is that the consensus sequence is needed to find large-scale homology, and resolving the fine-grained structure of the alignment is not needed at this step and would require an eccessive amount of resources.
 However after the graph has been built alignments can be refined with \textit{PanGraph}'s \texttt{polish} command.
@@
 ISs that really move around and split long blocks) this might be an intrinsic limit that derives from the pangenome, I wouldn't consider it a problem with PanGraph.}\\
-\responsedraft{
+\response{
       We thank the reviewer for their appreciation of the paper and of our methodology.\\
-      We agree with their concern that the level of fragmentation of the graph constitutes a limitation in the number and diversity of genomes that can meaningfully be added in the pangenome graph.\\
-      An increase in the number or diversity of isolates included in the graph will cause an increase in structural breakpoints in the graph.
-If subsequent structural changes occurr at a scale smaller than $\Lthr$, these will not correctly be captured by the graph topology.
-An excessive number of pancontigs would also increase the amount of computational resources used in the alignment.\\
+      We agree with the reviewer that fragmentation of the graph is an important concern, and that it can constitute a limitation in the number and diversity of genomes that can meaningfully be added in the pangenome graph.
+For example datasets in which subsequent structural changes occurr at a scale comparable with $\Lthr$ cannot be correctly be transformed into a graph.
+Moreover an excessive number of pancontigs will also increase the amount of computational resources required.\\
+      Within the size limit of our datasets, in the paper we investigated fragmentation in sect.~2B (``\textit{scaling with dataset size}'').
+In particular in fig.~5 we quantify how the size and number of pancontigs vary as a function of the number of isolates included in the graph.
+Here we observed good scaling properties, with pancontig size decreasing slowly as more isolates are added.
+We expect the threshold level of fragmentation to be reached around the point where the lines intercept the threshold length $\Lthr = 100$bp.\\
+
       % Other than being introduced by the diversity of the dataset, fragmentation could also be introduced artificially in the merging procedure.
@@
 We found that fragmentation was present only when the diversity of the sample was too high for the sensitivity of the aligner.\\
-      Irrespective of its origin, a possible way to detect excessive fragmentation and check for consistency in the graph structure is to visualize the distribution of block size in the graph (see for instance fig.~\ref{fig:kp-vs-hp} lower right panel).
-Excessive fragmentation in the graph would be detectable as most of the sequence being condensed in blocks having size close to $\Lthr$.\\
-      (As soon as they are done I could put the time and block size distribution for graphs with 500/1000 E.coli.)
+      Irrespective of its origin, a possible way to detect excessive fragmentation and check for consistency in the graph structure is to visualize the distribution of pancontig size in the graph (see for instance fig.~\ref{fig:kp-vs-hp} lower right panel).
+Excessive fragmentation in the graph would be detectable as most of the sequence being condensed in pancontigs having size close to $\Lthr$.\\
       % TODO: show distributions for pangenome graphs with 500/1k isolates? Mention times?
@@
 }
-\Marco{It is an important concern. We have investigated it in the paper, good scaling... Can be detected if wrong...}
 
@@
 % \Marco{agree. I think this is a structural property that we want to capture, the fact that these pancontigs are split.}
-\responsedraft{
+\response{
       We agree with the reviewer's observation that events such as the presence of an IS in an orthologous gene cluster would split the cluster in multiple pancontigs.
-For datasets in which isolates have a high amount of IS (larger than $\Lthr$), this would indeed result in multiple splits in the graph structure.\\
+For datasets in which isolates have a high amount of IS (longer than $\Lthr$), this would indeed result in multiple splits in the graph structure.\\
       This is in part an explicit design choice in \textit{PanGraph}.
@@
 By condensing all occurrences of such an IS in a single pancontig we also condense the information on all of the different contexts in which the said IS appears, at the cost of fragmenting the graph.
-We are already using this property in a work in preparation \cite{shaw_preparation}, in we compare the similarity between different contexts to obtain insights in the evolution of such sequences.
-This is especially important given the fact that IS often carry limited phylogenetic information, due to their limited size.
-      % TODO: maybe a bit off the mark? Not necessary to talk about the work in preparation?
+We are already using this property in a work in preparation \cite{shaw_preparation}, in which we compare the similarity between different contexts to obtain insights in the evolution of such sequences.
+This is especially important given the fact that IS often carry limited phylogenetic information, due to their limited size.\\
+      We agree however with the reviewer in that this is not always a desirable feature.
+When not desired the user could be given the option to specify a list of IS sequences, that could be used to filter out mergers involving exclusively the IS.
+We consider adding this as a future improvement, however this requires further research and is not included in the current version of \textit{PanGraph}.
 }
-\Marco{Future work, requires further research. Will consider adding specific procedures for IS (e.g. give a list of IS sequences that could be excluded...)}
 

### 2023-03-04 reviewer_response.tex
chore

@@
 One can a posteriori decide whether one wants to split paralogs using structural information (flanking blocks) or one can even prevent the merging using diversity and the $\beta$ parameter.}\\
-\responsedraft{
+\response{
 We thank the reviewer for raising an important point that necessitates of more discussion in the paper.
-As stated by the reviewer, we perform merges prioritizing longer and less diverged alignments.
-However \textit{PanGraph} in its current state does not explicity distinguish between orthologs and paralogs.
+\textit{PanGraph} in its current state does not explicity distinguish between orthologs and paralogs.
 This is partially a design choice, as better explained in the answer to the last question.
@@
 However, as the reviewer points out, in the case of distantly related paralogs, such a grouping might be undesirable as it introduces unneeded complications in the graph.
-Currently \textit{PanGraph} supports two different ways of partially controlling paralog splitting.\\
+Currently \textit{PanGraph} supports two different ways of controlling paralog splits.\\
 The first one is based on sequence divergence.
@@
 This would prevent the merging of paralogs that have sequence divergence higher than $1/\beta$.\\
-The second method relies instead on context.
+The second method relies instead on structural information.
 The \texttt{marginalize} command possesses the optional \texttt{--reduce-paralog} flag.
@@
 
-% TODO: if this is sensible add a paragraph in SI about paralog splitting
 % TODO: confirm with Nick the paralog split algorithm.
 }
-\Marco{Summarize, make more to the point. Say that this is now better discussed in SI.}
 

### 2023-03-05 reviewer_response.tex
update

@@
 
-\responsedraft{We thank the reviewer for their appreciation of our contribution and for their critical review of our work.}
+\response{We thank the reviewer for their appreciation of our contribution and for their critical review of our work.}
 
@@
 \reviewer{R3}{The article ``PanGraph: scalable bacterial pan-genome graph construction'' reports a new method for constructing a Pan-Genome Graph (PGG) in an efficient manner with an open and extractable grah structure.
-Unfortunately, the authors appear to either be unaware of or downplay the existence of previous methods for the same purpose.\textbf{ The authors need to survey the existing literature and compare and contrast with their method to show the utility of their approach.}\\
+Unfortunately, the authors appear to either be unaware of or downplay the existence of previous methods for the same purpose.
+The authors need to survey the existing literature and compare and contrast with their method to show the utility of their approach.\\
       For instance the authors state: ``Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.''.
@@
 We hope that these edits address the concerns of the reviewer.
-      % SURVEY OF THE LITERATURE
-      % PGG have been used to:
-      % - improve annotations [sutton 2021] + [ggcaller]
-      % MERTIS OF OUR METHOD
-      % - de Bruijn graphs represent SNPs and large scale variation in the same way, as edges in the graph.
-We instead have this scale separation and encode nucleotide-level variation in alignments, and structural changes in the graph
-      % - PGG based on genes are sensitive to annotation errors. We are based on homology.
-      % - Scalable algorithm.
-}
-
-\Marco{
-      Finish literature review.
-      Add part on graph-based methods.
-      In the discussion add edge cases.
-      Cite all the ones suggested by the reviewer.
 }
@@
 
-\reviewer{R3}{This statement: ``While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging.
+\reviewer{R3}{
+      This statement: ``While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging.
 Colored generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets, although the underlying efficiency derives from a fixed k-mer size which prevents modelling long-range homology (Iqbal et al., 2012; Muggli et al., 2017).'' appears to be incomplete at best.
-De Bruijn graph-based assemblers routinely collapse nonbranching portions of the graph into longer sequences than the k-mer size used for construction.
-      % \Marco{from what I understand the structure in de Bruijn graphs is very fine-grained.
-Every SNP can generate a branching event.
-While in our case we have this intermediate corse-graining where homologous parts are collapsed together and the remaining structure of the graph represents large-scale structural variation in the genomes.
-I don't think this is solved by collapsing nonbranching portions in dB graphs (but I might not know enough about this...)}\\
+De Bruijn graph-based assemblers routinely collapse nonbranching portions of the graph into longer sequences than the k-mer size used for construction.\\
       One early approach that does not explicitly generate a pan-genome graph but deals with some of the rearrangement issues cited by the authors is:
@@
       \end{itemize}}
-\responsedraft{}
-\Marco{Copy form above and extend the answer}
+\response{
+      We apologize and we modified the sentence in question. We also extendend the discussion of de Bruijn graphs methods in the introduction.\\
+      We thank the reviewer for pointing out these important contributions that were not previously discussed in the paper.
+}
+% \Marco{from what I understand the structure in de Bruijn graphs is very fine-grained.
+Every SNP can generate a branching event.
+While in our case we have this intermediate corse-graining where homologous parts are collapsed together and the remaining structure of the graph represents large-scale structural variation in the genomes.
+I don't think this is solved by collapsing nonbranching portions in dB graphs (but I might not know enough about this...)}\\
 
 \reviewer{R3}{I believe this statement: ``It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order.'' is the essential step in the process and perhaps deserves more focus.
-It captures two important issues: \textbf{aligning orthologs not paralogs and keeping the fracturing of the PGG nodes to a minimum}.
+It captures two important issues: aligning orthologs not paralogs and keeping the fracturing of the PGG nodes to a minimum.
 The pseudo-energy captures the alignment length (longer is better for orthologs versus paralogs) but does not appear to capture the surrounding context of the iterative PGGs.
@@
 When choosing between competing mergers of paralogs/orthologs the context of the PGG and other possible mergers is important if the pancontigs have been reduced so that they contain only repetitive sequence without unique flaning sequence.
-I think this is a very important point as \textbf{distinguishing orthologs from paralogs is essential for correct PGG construction and the authors need to explicitly address it}.
+I think this is a very important point as distinguishing orthologs from paralogs is essential for correct PGG construction and the authors need to explicitly address it.
 This could be measured as part of the simulated data where identical paralog sequences of various lengths and copy number could be modeled.}
@@
 
-We now make this explicit in SI sect.~{\SIalgo}D, that now contains these considerations.
+We added SI sect.~{\SIalgo}D to include these considerations.
 

### 2023-03-05 main.tex
chore
update

@@
     With the wide-spread usage of long-read sequencing, the number of high-quality, complete genome assemblies has increased dramatically.
+    % Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.
     Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.
@@
 However in these methods the graph structure encodes at the same time nucleotide-level variation and large-scale structural rearrangements in the genome.
+Separating these two scales is often difficult, and specific techniques have to be employed to detect large-scale homology in these graphs \cite{angiuoli2011mugsy,minkin2013sibelia}.
 
@@
 Here we present \emph{PanGraph}, a Julia \cite{bezanson2017julia} library and command line interface, designed to efficiently align large sets of closely related genomes into a pangenome graph on personal computers.
+Alignment is performed on the base of sequence homology and does not rely on gene annotations.
 The resulting graph both compresses the input sequence set and succinctly captures the population diversity at multiple scales: from nucleotide mutations and indels to structural polymorphisms driven by inversions, rearrangements, and gene gain/loss.
+Importantly, these two scales are naturally separated in our representation.
 The underlying graph data structure can be exported into numerous formats for downstream analysis and visualization in software such as Bandage \cite{wick2015bandage}.
@@
     \end{tabular}
-    \caption{{\bf Dataset properties}. We use the PanX pipeline to cluster homologous genes and characterize core genome size and diversity of our datasets. Columns represent:
+    \caption{{\bf Dataset properties}. Using the PanX analysis results we characterize the size and diversity of our dataset. Columns represent:
         number of isolates (N.),

### 2023-03-06 main.tex
Update on Overleaf.

@@
     % Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.
-    Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.
+    This provides access to very rich information on the variation present in the population, both at the nucleotide level and as large-scale structural changes, that traditional computational approaches struggle to capture and represent.
     Here, we present \emph{PanGraph}, a Julia-based library and command line interface for aligning whole genomes into a graph.
@@
 While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging.
-Colored and compacted generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets \cite{iqbal2012novo,muggli2017succinct}.
-Currently tools exist to build graphs containing thousands of isolates \cite{holley2020bifrost}, and have proven useful in problems such as the detection of core pangenome \cite{schulz2022sequence} or the improvement of gene prediction, annotation and clustering \cite{horsfield2023accurate}.
+Colored and compacted generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets \cite{iqbal2012novo,muggli2017succinct}, with tools existing to build graphs containing thousands of isolates \cite{holley2020bifrost}.
+These have proven useful in problems such as the detection of core pangenome \cite{schulz2022sequence} or the improvement of gene prediction, annotation and clustering \cite{horsfield2023accurate}.
 However in these methods the graph structure encodes at the same time nucleotide-level variation and large-scale structural rearrangements in the genome.
@@
 An orthogonal approach has been to formulate the inference of the pangenome graph as a multiple genome alignment.
-However, current methods either scale poorly to large sets of genomes \cite{darling2010progressivemauve}, focus on comparisons across diverse sets of species across the tree of life at the cost of memory \cite{armstrong2020progressive}, or utilize a reference-guided approach by partitioning genomes first into annotated genes \cite{gautreau2020ppanggolin,colquhoun2021pandora}.
+Some early methods relied on homology searches to locate syntenic regions, termed \textit{locally colinear blocks} \cite{angiuoli2011mugsy,darling2010progressivemauve}, and proved useful in various problems such as improving gene annotations \cite{angiuoli2011improving}.
+These early methods however usually scale poorly to large sets of genomes.
+More recently techniques based on homology have been used to compare genomes across diverse sets of species across the tree of life, at the cost of memory \cite{armstrong2020progressive}.
+Alternatively, homology can be searched directly on genes.
+Many methods rely first on grouping genes in orthologous clusters, representing nodes in the graph, with edges linking clusters that are adjacent on some of the genomes.
+These graphs are usually characterized by a syntenic backbone of core gene clusters, with interspersed regions of structural diversity \cite{chan2015novel,oliveira2017chromosomal}.
+The structural information provided in these graphs has been used to categorize core/accessory regions \cite{sutton2021pan,gautreau2020ppanggolin}, and to improve variants detection \cite{colquhoun2021pandora} or gene annotations \cite{tonkin2020producing,zhou2020accurate}.
+By using genes as nodes, these graphs are much smaller in size and homology detection is facilitated, but many of these methods are sensitive to gene annotation errors.
 
@@
 
+We believe that this feature, i.e. the fact that our representation naturally separates nucleotide level diversity from larger structural variation, is one of the main merits of our tool.\\
+De Bruijn graphs types of representation can be built quickly since they do not necessitate alignment, and even directly from reads without the need of high-quality assemblies.
+However they do not posses this scale separation, with every mutation generating a structural split in the graph.
+Some methods however exist to detect longer-range homology.
+These involve some form of smoothing of the graph to extend homology matches over small-scale variation, and have been used for the detection of core regions with applications to genetic engineering \cite{schulz2022sequence} or to find synteny blocks \cite{minkin2013sibelia}.
+These are particularly effective on very similar sequences with very low divergence.\\
+A second widespread type of representation uses orthologous gene clusters as node in the graphs, and introduces links between clusters that are adjacent on some of the isolates.
+These have proven useful in many different tasks including the categorization of core/accessory regions \cite{gautreau2020ppanggolin,sutton2021pan}, reducing reference-bias in variant detection \cite{colquhoun2021pandora} and the improvement of gene annotations \cite{tonkin2020producing}.
+In these representation the graph usually has a smaller number of nodes, and can include more diverged sequences by detecting homology on protein sequences.
+However this representation is naturally sensitive to gene annotation errors.\\
+
 In our analysis we demonstrated the capabilities and limits of this tool, both on synthetic and real sequences.

### 2023-03-08 reviewer_response.tex
Update on Overleaf.

@@
 \newcommand{\SImarginalize}{VI}
+
 \newcommand{\SIfigMutrate}{4}
@@
 In line 54 you say "Pairwise graph alignment is performed by an all-to-all alignment of the pancontigs" which I interpreted to mean all sequences in the alignment vs all sequences in another alignment.
-You do later add that "To align two graphs, the consensus sequences of all pancontigs in both graphs are searched for homologies and aligned".
+You do later add that ``To align two graphs, the consensus sequences of all pancontigs in both graphs are searched for homologies and aligned".
 Adding the word consensus in the earlier sentence might avoid confusion.}
@@
 % \Liam{that's clear to me.}
-\response{We agree with the reviewer's suggestion. The wording in Line 54 has been updated to ``all-to-all alignment of the consensus sequences of all \textit{pancontigs}''.}
+\response{We agree with the reviewer's suggestion. The wording has been updated to ``all-to-all alignment of the consensus sequences of all \textit{pancontigs}''.}
 
@@
 I agree with you about synteny and think this is a nice point to add here to explain the observation.}
-\response{The observation that core pancontigs are longer than average is in agreement with our prior the expectation since multiple core genes are often syntenic even in diverse species, compared to accessory genes that are often found in different contexts.
+\response{The observation that core pancontigs are longer than average is in agreement with our expectations.
+Multiple core genes are often syntenic even in diverse species, compared to accessory genes that are often found in different contexts.
 This effect seems to outweigh the possibility of increased fragmentation due to the large number of mergers that common (core) pancontigs undergo.
-      We modified the line to: ``Core pancontigs having higher-than-average length. This is expected given that core genes tend to stay synteny over large evolutionary distances.''}
+      We modified the line to: ``Core pancontigs having higher-than-average length. This is expected given that core genes tend to stay syntenic over large evolutionary distances.''}
 
@@
 (3) For genome structure, exploring `hotspots' in the graph as \href{https://www.nature.com/articles/s41467-017-00808-w}{this paper} does could be done using in/out degree of pancontigs.}
-\response{We have so far not explored the use of pangraph as input for graph based mapping approaches like Pandora or GramTools.
-But this is indeed an obvious downstream application of pangraph.
-We have included a brief discussion of these approaches, but we feel that testing and implementation of such approaches is beyond the scope of this article.}
+\response{We have so far not explored the use of pangraph as input for graph based mapping approaches like Pandora or GramTools, but we agree this is an interesting downstream application.
+We have included a discussion of these approaches, although we feel that testing and implementation of them is beyond the scope of this article.}
 
@@
 Moreover we simulate sequence evolution without a notion of genes.
-This is now specified in SI section II.
+This is now specified in SI section \SIsynthdata.
 To avoid confusion we also changed ``Horizontal Gene Transfer'' to ``Horizontal Sequence Transfer''.}
@@
 Therefore we expected that the choice of null energy parameters would lead to shorter pancontigs, and alignments containing more diverged sequences.
-      To make this reasoning more explicit we added section I in the SI to better explain the merging procedure and the effect of tuning the pseudo-energy parameters.}
+      To make this reasoning more explicit we added section {\SIalgo} in the SI to better explain the merging procedure and the effect of tuning the pseudo-energy parameters.}
 
@@
 When i dug around on github , i found this helpful page \url{https://neherlab.github.io/pangraph/tutorials/tutorial_2/} Which answers and raises questions.
-So, my question: for each pancontig (which i think is the same as a block), i think what you do is infer a consensus (what is this?
+So, my question: for each pancontig (which I think is the same as a block), I think what you do is infer a consensus (what is this?
 majority base at each position but somehow also choose an allele when there is an indel?), and then you describe the other alleles in the MSA as edits from this?
@@
 % \Liam{I think a `narrative section' in the supplementary describing this in brief sounds great.}
-\response{This point was also raised by other reviewers and we appreciate the need to make this pointer clearer.
+\response{This point was also raised by other reviewers and we appreciate the need to make this point clearer.
       In our pangenome graph representation each pancontig encodes an alignment using a consensus sequence and list of edits required to obtain each individual sequence from the consensus sequence.
@@
 
-      To make this more clear we introduced section I in SI, in which we explain in more detail the merging procedure and how the consensus sequence is updated.
+      To make this clearer we have added section {\SIgraph} in SI, in which we explain in more details our pangenome graph representation, and section {\SIalgo}B in SI, in which we explain in more detail the merging procedure and how the consensus sequence is updated.
 We also refer to the documentation page mentioned by the reviewer.
@@
 I don't follow how you calculate all pairwise intersections through sorting the list of minimizers, nor do i see how this ends up being log-linear.
-For each genome, you can sort it's minimizers in nlog(n) (ie log linear), where n ~genome length.
-How do you get all pairwise distances from that without another multiplier of n?}
+For each genome, you can sort it's minimizers in $n \log (n)$ (ie log linear), where $n \sim$ genome length.
+How do you get all pairwise distances from that without another multiplier of $n$?}
 % \Marco{in the text we write that ``the pairwise distance matrix is estimated in a time that increases log-linearly in the \textbf{total length} of sequence that is analyzed.'' This might generate some confusion since the scaling depends both on the number of genomes $N$ and their length $l$.
@@
 % \Liam{clarifying $N$ and $l$ makes sense.}
-\response{We agree with the reviewer that the formulation in the main text is too concise and a bit inaccurate.
+\response{We agree with the reviewer that the formulation in the main text was too concise and a bit inaccurate.
       There are two main variables at play: the length of a genome $l$ and the number of isolates $n$.
 As the reviewer points out, the set of minimizers for each genome can be created and sorted in $l \log l$.
-Finding the set of shared minimizers would require comparing $\sim n^2$ sets of ordered lists of minimizers, which scales as $\sim n^2 l$.
+Finding the set of shared minimizers would require comparing $\sim n^2$ pairs of ordered lists of minimizers, which scales as $\sim n^2 l$.
 This would provide a scaling $O(n l \log l + n^2 l)$.\\
@@
 Therefore depending on the diversity of the sample the scaling will be between $O(L \log L + n L)$ and $O(L \log L)$.\\
-      In practice for typical use cases of pangraph ($n \sim 100$ and $L \sim 10^6$) the computationally challenging part is the minimizer sketching, and not the distance matrix computation.
+      In practice, for typical use cases of pangraph ($n \sim 100$ and $L \sim 10^6$) the computationally challenging part is the minimizer sketching, and not the distance matrix computation.
 This is partly due to some prefactors (e.g.
@@
 We also stress that the computational time to build the guide tree is usually negligible compared to the time required for the graph merging procedure.\\
-      We slightly elaborated this point in the main text and added a some of the above explanation to the supplement.
+      We slightly elaborated this point in the main text and added a version of the above explanation to SI sect. {\SIalgo}A.
 }
@@
       \end{itemize}}
-\response{We thank the reviewer for point out that this aspects require more explanation (the point was also raised by other reviewers).
-      We followed the advice of the reviewer and added Section I in the SI, in which we expand on the explanation of how we merge two graphs.\\
+\response{We thank the reviewer for pointing out that we should have provided more explanation (the point was also raised by other reviewers).\\
+      We followed the advice of the reviewer and added Section {\SIalgo}B in the SI, in which we expand on the explanation of how we merge two graphs.\\
       In short, we perform the merging procedure in two steps.
 In the first we use the aligner (minimap2 or mmseqs2) to scan for homology matches between the consensus sequences of all pancontigs of one graph and of the other.
-      This is achieved by simply passing all all these consensi in a single fasta file into mmseqs2 or via the API into minimap2.
+      This is achieved by simply passing all these consensus sequences as a single fasta file into mmseqs2 or via the API into minimap2.
       The resulting matches are ranked by pseudo-energy and mergers performed in order.
       Each pancontig can be subject to only one merge at this stage.
-      For precise ties in this ranking of of matches, the behavior is not explicitly defined but depends on the order in which the matches appear in the output of the homology search tool.
+      For exact ties in this ranking of of matches, the behavior is not explicitly defined but depends on the order in which the matches appear in the output of the homology search tool.
+% Liam: replaced 'precise ties' > 'exact ties'
 
       After all of the matches have been exhausted, the two graphs have undergone a preliminary merging, with some new merged pancontigs being created.
-At this point we repeat the procedure looking for matches on the consensus sequence of all the pancontigs in an all-against-all comparison.
+At this point we repeat the procedure looking for matches between the consensus sequences of all the pancontigs in an all-against-all comparison.
 This self-mapping on the graph will detect any homology that could not have been merged at the previous stage, and will perform the corresponding merging.
@@
 \response{
-      Apologies for these slighly confusion statements.
+      Apologies for these confusing statements.\\
       In this case the second criterion is a consequence of the first.
@@
 
-      To make this more explicit we modified lines 93-95 in ``$\beta$ controls the maximal sequence divergence between the two merger-candidates.
+      To make this more explicit we modified the previous formulation into: ``$\beta$ controls the maximal sequence divergence between the two merger-candidates.
 The negative pseudo-energy requirement means that no merger will be performed if the Hamming distance exceeds $1/\beta$...
@@
       On top of this, only matches with negative pseudo-energy are accepted as mergers.
-      The parameter $\beta$ thus allows to explicitly set a maximal divergence ($1/\beta$) allowed in mergers, rather than relying on the implicit cut-off that comes from the sensitiviy of the aligner.
+      The parameter $\beta$ thus allows to explicitly set a maximal divergence ($1/\beta$) allowed in mergers, rather than relying on the implicit cut-off that comes from the sensitivity of the aligner.
 \\
-      This is now made more explicit in SI section I (PanGraph parameters effect).\\
+      This is now made more explicit in SI sect.~{\SIalgo}C.\\
       When setting $\beta = 0$ the maximal divergence is determined by the aligner sensitivity.
@@
 % \Liam{makes sense.}
-\response{L50 measures the minimum number of pancontigs needed to cover at least 50\% of the pangenome graph sequence.
+\response{This is an interesting observation.
+L50 measures the minimum number of pancontigs needed to cover at least 50\% of the pangenome graph sequence.
 This number is closely related to the total number of pancontigs.
 In general the use of null energy parameters result in more fragmented graphs, with a higher number of pancontigs (cf.
-fig.~4D in SI).
-Helicobacter is an exception with the opposite trend where the use of null energy parameters results in less fragmentation of the graph (both in the number of pancontigs and for the value of L50).
-      This might be explained by that fact that the divergence of some parts of the core genome in this dataset is close to $1/\beta$ for $
-            \beta=10$ used in the figure.
-      Hence some more diverse parts of the genome might remain split in different pancontigs and removing this threshold imposed by $\beta$ results in a larger longer consecutive merger and a less fragmented graph.
-      For a visual verification of this interpretation we display in fig.~\ref{fig:kp-vs-hp} how the distributions of pancontig length and frequency change with the use of different alignment kernels and pseudo-energy parameters.
-We compare the changes observed on the Hp dataset to the ones in Kp, used as a reference.
-In the Kp dataset increasing the aligner sensitivity results in shorter pancontigs (lower-right panel) and a small increase in core genome fraction.
-For Hp instead a more sensitive aligner results in the loss of small and intermediate size pancontigs, but not of long ones.
-It also results in a loss of low-frequency pancontigs, indicating that these had an homologous counterpart that could not be merged due to high-divergence, and the core pangenome fraction increases considerably.
+fig.~{\SIfigBenchmark}D in SI) so L50 increases.
+However, the reviewer is correct that \textit{Helicobacter} is an exception and shows the opposite trend: the use of null energy parameters results in less fragmentation of the graph (both in the number of pancontigs and for the value of L50).\\
+      Having investigated, we believe this might be explained by that fact that the divergence of some parts of the \textit{Helicobacter} core genome in this dataset is close to $1/\beta$ for $\beta=10$ used in the figure.
+      Hence, parts of the core genome where the divergence reaches $>1/\beta$ might be split into different pancontigs.
+Removing the threshold imposed by $\beta$ by setting $\beta=0$ results in a longer consecutive merger and therefore a less fragmented graph with fewer overall pancontigs.\\
+      For a visual verification of this interpretation we display in fig.~\ref{fig:kp-vs-hp} how the distributions of pancontig length and frequency change with the use of different alignment kernels and pseudo-energy parameters: minimap (asm10) with default parameters vs.
+mmseqs with null pseudo-energy.
+We compare the changes observed on the Hp dataset to the ones in Kp.
+In the Kp dataset the null pseudo-energies result in shorter pancontigs (lower-right panel) and a small increase in core genome fraction (lower-left panel).
+For Hp, the effect on pancontigs is quite different.
+More sensitive alignment with null pseudo-energies means small and intermediate pancontigs are lost, but there are slightly more long ones (top-right panel).
+It also results in a loss of low-frequency pancontigs (top-left panel) indicating that these had an homologous counterpart that with default parameters and minimap could not be merged due to high-divergence.
+Using a more sensitive aligner with null pseudo-energies means the core pangenome fraction increases considerably.
       % TODO: add mention in SI
       % QUESTION: this is more or less explicit in SI? Should I expand more?
+      % Liam: I think there are slightly more long pancontigs with mmseqs/null in Hp from looking at figure so seems worth saying
 }
@@
 In the particular case in which merging two pancontigs would leave overhangs shorter than $\Lthr$, these are included in the alignment as flanking insertions.
-This is now explained in SI section I, and mentioned in the main text in section IB (line 103).
+This is now explained in SI section {\SIgraph}, and mentioned in the main text in section IB (line 103).
       In fact, we used the fact that genomes can be reconstructed exactly from the pangraph to test and debug the tool. }
@@
       Feel free to replace with a better reference, this si just meant to be helpful.}
-% \Marco{I think this is a good reference, fig. 2 is familiar to us. And also fig.2 in \cite{doolittle2009origin} which is a bit older. But not sure which other Rocha paper he is referring to.}\\
+% \Marco{I think this is a good reference, fig. 2 is familiar to us. And also fig.2 in \cite{doolittle2009origin} which is a bit older. But not sure which other Rocha paper they are referring to.}\\
 % \Liam{I think they mean Figure 5 of \cite{touchon2009organised} (\href{https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1000344}{here}).
@@
 In practice for each block it collects all of the sequences and uses MAFFT \cite{katoh2002mafft} to perform a multiple sequence alignment, obtaining an improved consensus and more consistent insertion pattern.\\
-      This command was not mentioned in the paper. This has now been amended, and we briefly described in in section IB in the main text (lines ...) and at in section I in the SI (graph merging) }
+      This command was not mentioned in the paper. This has now been amended, and we briefly described in in section IB in the main text and in SI sect.~{\SIalgo}B.}
 
@@
 % \Liam{adjacent sounds good.}
-\response{we agree with the reviewer's observation. We modified ``syntenic on'' to ``adjacent in'' in line 48.}
+\response{We agree with the reviewer's observation. We have changed ``syntenic on'' to ``adjacent in''.}
 
@@
 \response{The value $N=100$ is the standard value used in the rest of the simulation, but as the reviewer pointed out this value is varied for the results showed in fig.~2 and the statement was confusing.
-We modified line 131 removing the value of $N$ and referring to SI section 1 for the standard value of simulation parameters.}
+We modified it removing the value of $N$ and referring to SI section {\SIsimBench} for the standard value of simulation parameters.}
 
@@
 \response{
-      We apologize if our previous discussion of the literature was incomplete, it was not our intention to downplay prior work.
-      We thank the reviewer for highlighting additional important work and aspects that we previously didn't discuss.
-      We have rewritten the part in the introduction to reflect these the work by these and other authors and added a few sentences to the discussion to further highlight previous and alternative approaches.
+      We apologize if our previous discussion of the literature was incomplete. It was not our intention to downplay prior work.
+      We thank the reviewer for highlighting additional important work and aspects that we have now added to better contextualise our work.
+      We have rewritten the introduction to reflect previous work in this field and added  to the discussion to further highlight previous and alternative approaches.
 We hope that these edits address the concerns of the reviewer.
+      % Liam: I think good to say in more detail what has been added here. 
 }
@@
 % \Marco{Yeah good idea. Maybe something that just map each annotation into a block, without solving the problem of harmonizing annotations which might be some more work...}\\
-\response{We thank the reviewer for this important suggestion.
-We appreciate the need to include annotation in our pangenome graph structure.
-We created a script to decorate each annotation in a \texttt{gff} file with information on the pancontig location.
+\response{We thank the reviewer for this important suggestion and appreciate the need to include a way to map annotations back onto the pangenome graph.
+We created a helper script that allows a mapping of pancontig information onto a set of annotations in a \texttt{gff} file.
+In default usage the pancontig location(s) of each feature gets added as a new `attribute', allowing a quick connection between the generated pangenome graph and existing annotations.
 The script and instructions on its usage can be found at \url{https://github.com/liampshaw/pangraph\_annotations}.
-We plan on extending this in the future to use directly the pangenome graph as a reference and add annotations as features of pancontigs.
+In the future, we hope to extend this further for visualization purposes.
       % TODO: mention this in the text.
-      % TODO: make the repository public.
+      % Liam: not sure if we need last sentence
 }
@@
 \response{
-      We apologize and we modified the sentence in question. We also extendend the discussion of de Bruijn graphs methods in the introduction.\\
+      We apologize and we modified the sentence in question. We also extended the discussion of de Bruijn graphs methods in the introduction.\\
       We thank the reviewer for pointing out these important contributions that were not previously discussed in the paper.
@@
 \response{
-We thank the reviewer for raising an important point that necessitates of more discussion in the paper.
-\textit{PanGraph} in its current state does not explicity distinguish between orthologs and paralogs.
-This is partially a design choice, as better explained in the answer to the last question.
-In short, the possibility of grouping duplicated sequences in the same pancontig will also group information on their context, and the fragmentation introduced in the graph would reflect some real (potentially ancient) structural change in the genome.\\
+This is an important point that needed more discussion in the paper than we originally provided.
+To be clear: \textit{PanGraph} in its current state does not explicitly distinguish between orthologs and paralogs when constructing the graph.
+This is a deliberate design choice.
+% Liam: don't understand this part of sentence so deleted 'as better explained in the answer to the last question.'.
+I think it's better to say that it's deliberate than to equivocate
+The possibility of grouping duplicated sequences in the same pancontig will also group information on their context, and the fragmentation introduced in the graph would reflect some real (potentially ancient) structural change in the genome.\\
+% Liam: not sure I quite understand above sentence...
 
-However, as the reviewer points out, in the case of distantly related paralogs, such a grouping might be undesirable as it introduces unneeded complications in the graph.
+However, as the reviewer points out, in the case of distantly related paralogs such a grouping might be undesirable if it introduces `complications' in the graph.
+We agree from experience that these can sometimes make understanding the pangenome more complicated.
 Currently \textit{PanGraph} supports two different ways of controlling paralog splits.\\
@@
 This would prevent the merging of paralogs that have sequence divergence higher than $1/\beta$.\\
-The second method relies instead on structural information.
+The second method relies instead on information on genomic context.
 The \texttt{marginalize} command possesses the optional \texttt{--reduce-paralog} flag.
@@
 
-In the future we plan on providing a dedicated command to perform a more robust form of paralag splitting, which combines both sequence divergence information and context information.
-This command would build a phylogenetic tree from the alignment of each duplicated pancontig, to detect exceptionally long branches, and check whether the split correlates with a split in the duplication context.
-This functionality is not however not yet implemented in the current version of \textit{PanGraph}.\\
+In the future we plan on providing a dedicated command to perform a more robust form of paralog splitting, which combines both sequence divergence information and context information.
+This command would build a phylogenetic tree from the alignment of each duplicated pancontig to detect exceptionally long branches, and check whether the split correlates with a split in the duplication context.
+This functionality is not however not yet implemented in the current version of \textit{PanGraph} and we see it as substantial extra work.\\
 
-We added SI sect.~{\SIalgo}D to include these considerations.
+We appreciate the reviewer's perspective and have added SI sect.~{\SIalgo}D to include discussion of these considerations.
 
@@
 We avoid performing a multiple sequence alignment during merging to maintain the algorithm scalable.
-However we do update the consensus of the pangontig at each merge using the majority rule on the merged alignment.\\
-      The idea behind this approach is that the consensus sequence is needed to find large-scale homology, and resolving the fine-grained structure of the alignment is not needed at this step and would require an eccessive amount of resources.
+However we do update the consensus of the pancontig at each merge using the majority rule on the merged alignment.\\
+      The idea behind this approach is that the consensus sequence is needed to find large-scale homology, and resolving the fine-grained structure of the alignment is not needed at this step and would require an exccessive amount of resources.
 However after the graph has been built alignments can be refined with \textit{PanGraph}'s \texttt{polish} command.
@@
       % TODO: quantify this in simulations if there is time?
+      % Liam: Think the last sentence is potentially risky - I don't get what `consistent' means?
+My rewrite (not confident in this): In our experience, the alignments appear reliable throughout this process, although we have not explicitly quantified it from our simulations.
+However, for cases where the alignments are of interest we recommend running `polish' to perform MSA for pancontig consensus sequences (as in the documentation) to ensure that the final alignments are as accurate as possible.
 }
@@
 But might be unnecessary detail when explaining the algorithm the first time, we might want to move it to the SI section where we explain merging in more detail?}\\
-\response{Transitive pancontigs are an artifact of periodic boundary conditions when merging circular genomes, such as bacterial chromosomes.
+\response{Transitive pancontigs are a small but problematic artifact of periodic boundary conditions when merging circular genomes, such as bacterial chromosomes.
 At the beginning of the procedure the circular genome is saved as a single linear pancontig.
@@
 When during merging this unique pancontig is further split, the flanking pancontigs could be crossing the artificial cut in a co-linear way on all paths.
-This step of the procedure resolves the artificial cut buy merging these pancontigs together.\\
+This step of the procedure resolves the artificial cut by merging these pancontigs together.\\
       We think that the mention of this step of the algorithm in this part of the paper might be a bit confusing and needs more clarification.
-We moved and expanded the statement from section IB of the main text to SI section I (point 5 of the merging algorithm explanation).}
+We moved and expanded the statement from section IB of the main text to SI section {\SIalgo} (point 5 of the merging algorithm explanation).}
 
@@
       We appreciate the importance of the point raised by the reviewer.
-      In it's current version \textit{PanGraph} does not support a post-processing refinement step in which alignments of adjacent edges of pancontigs are harmonized, potentially also removing very small pancontigs when these could be appropriately absorbed in the alignment of larger ones.
-      We agree with the reviewer that this post-processing smoothing of the graph would be an important improvement, and we consider adding it in subsequent versions of \textit{PanGraph}.
+      We agree with the reviewer that this post-processing smoothing of the graph produced by of \textit{PanGraph} would be a valuable addition, and we hope to such further post-processing refinements in subsequent versions.
+      This sort of refinement would allow alignments of adjacent edges of pancontigs to be harmonized, potentially also removing very small pancontigs when these could be appropriately absorbed in the alignment of larger ones.
+We hope to continue to expand the downstream post-processing of \textit{PanGraph} data structures in future work.
+      % Liam: tried to make it sound more explicitly 'future' and 'further'
 }
@@
       This is in part an explicit design choice in \textit{PanGraph}.
-We plan to use this tool for the study of the evolution of bacterial pangenome, and we wanted this feature of the pangenome - namely the fragmentation due to presence of the same IS in multiple contexts - to be explcitly encoded in the graph.
-By condensing all occurrences of such an IS in a single pancontig we also condense the information on all of the different contexts in which the said IS appears, at the cost of fragmenting the graph.
-We are already using this property in a work in preparation \cite{shaw_preparation}, in which we compare the similarity between different contexts to obtain insights in the evolution of such sequences.
-This is especially important given the fact that IS often carry limited phylogenetic information, due to their limited size.\\
-      We agree however with the reviewer in that this is not always a desirable feature.
-When not desired the user could be given the option to specify a list of IS sequences, that could be used to filter out mergers involving exclusively the IS.
-We consider adding this as a future improvement, however this requires further research and is not included in the current version of \textit{PanGraph}.
+We developed this tool for the study of the evolution of bacterial pangenome, and we wanted this feature of the pangenome - namely the fragmentation due to presence of the same IS in multiple contexts - to be encoded in the graph.
+By condensing all occurrences of such an IS in a single pancontig we also condense the information on all of the different contexts in which the said IS appears (at the cost of fragmenting the graph).
+We agree that analysing these kinds of regions is challenging.
+In ongoing work we have been investigating the similarity of IS contexts to obtain insights into the evolution of IS-rich regions.
+This is especially important given the fact that the sequences of the IS themselves have very little phylogenetic information.\\
+      However, we agree with the reviewer that this representation is not always desirable.
+One option to deal with it would be to allow the user to specify a set of `protected sequences' that could be used to filter out mergers involving e.g.
+exclusively ISs.
+Though this is in contrast to the agnostic approach of \textit{PanGraph} to sequence annotation, we are considering adding something like this in the future.
+However, this will require further research and potentially database curation at the level of species, so is not included in the current version.
 }

### 2023-03-08 main.tex
Update on Overleaf.

@@
     % Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.
-    This provides access to very rich information on the variation present in the population, both at the nucleotide level and as large-scale structural changes, that traditional computational approaches struggle to capture and represent.
+    In addition to pangenomic approaches that focus on the variation in the sets of genes present in different genomes, complete assemblies allow investigations of the evolution of genome structure and gene order.
+    This latter problem, however, is computationally demanding and conceptually challenging with few tools available that allow to shed light on this dynamics. 
     Here, we present \emph{PanGraph}, a Julia-based library and command line interface for aligning whole genomes into a graph.
@@
 Colored and compacted generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets \cite{iqbal2012novo,muggli2017succinct}, with tools existing to build graphs containing thousands of isolates \cite{holley2020bifrost}.
-These have proven useful in problems such as the detection of core pangenome \cite{schulz2022sequence} or the improvement of gene prediction, annotation and clustering \cite{horsfield2023accurate}.
-However in these methods the graph structure encodes at the same time nucleotide-level variation and large-scale structural rearrangements in the genome.
+These graphs have been shown to be useful in problems such as the detection of the core genome \cite{schulz2022sequence} or the improvement of gene prediction, annotation and clustering \cite{horsfield2023accurate}.
+However, in these methods the graph structure encodes at the same time nucleotide-level variation and large-scale structural rearrangements.
 Separating these two scales is often difficult, and specific techniques have to be employed to detect large-scale homology in these graphs \cite{angiuoli2011mugsy,minkin2013sibelia}.
@@
 Alternatively, homology can be searched directly on genes.
-Many methods rely first on grouping genes in orthologous clusters, representing nodes in the graph, with edges linking clusters that are adjacent on some of the genomes.
+Many methods rely first on grouping genes into orthologous clusters.
+These clusters are then treated as nodes in the graph, with edges linking clusters that are adjacent on at least one of the genomes.
 These graphs are usually characterized by a syntenic backbone of core gene clusters, with interspersed regions of structural diversity \cite{chan2015novel,oliveira2017chromosomal}.
-The structural information provided in these graphs has been used to categorize core/accessory regions \cite{sutton2021pan,gautreau2020ppanggolin}, and to improve variants detection \cite{colquhoun2021pandora} or gene annotations \cite{tonkin2020producing,zhou2020accurate}.
-By using genes as nodes, these graphs are much smaller in size and homology detection is facilitated, but many of these methods are sensitive to gene annotation errors.
+The structural information contained in these graphs has been used to categorize core/accessory regions \cite{sutton2021pan,gautreau2020ppanggolin}, and to improve variant detection \cite{colquhoun2021pandora} or gene annotations \cite{tonkin2020producing,zhou2020accurate}.
+By using genes as nodes, these graphs are much smaller in size.
+However, relying on gene annotations means that these methods are often sensitive to any errors in this initial annotation.
 
 Here we present \emph{PanGraph}, a Julia \cite{bezanson2017julia} library and command line interface, designed to efficiently align large sets of closely related genomes into a pangenome graph on personal computers.
-Alignment is performed on the base of sequence homology and does not rely on gene annotations.
+The \emph{PanGraph} algorithm creates nodes of homologous sequences through alignments alone, without requiring gene annotation.
 The resulting graph both compresses the input sequence set and succinctly captures the population diversity at multiple scales: from nucleotide mutations and indels to structural polymorphisms driven by inversions, rearrangements, and gene gain/loss.
@@
 This shortens considerably the time needed, but might introduce minor inconsistencies and artifacts in the alignments.
-After building the graph, these can be removed by performing a multiple sequence alignment on each pancontig  (see ``polish'' command in documentation).
+After building the graph, these can be removed by performing a multiple sequence alignment on each pancontig  (see ``polish'' command in documentation and SI sect.{\SIalgo}B).
 
@@
 We simulated populations of size $N$ and sequence length $L$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
-In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (respective rates $0.01$ and $0.05$ per generation), and horizontal sequence transfer that occurs with tunable rate $h$ per generation per genome (see SI sect.~\SIsynthdata for details on the simulation and standard parameter values).
+In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (respective rates $0.01$ and $0.05$ per generation), and horizontal sequence transfer that occurs with tunable rate $h$ per generation per genome (see SI sect.~{\SIsynthdata} for details on the simulation and standard parameter values).
 The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
@@
 We quantified PanGraph's ability to accurately reconstruct the true graph by computing the displacement of the inferred breakpoints relative to their known locus stored from the evolutionary simulation (cf.
-SI Section IIB for details).
+SI Section {\SIsimBench}B for details).
 We evaluate this displacement by generating datasets with different rates of mutation $\mu$ and horizontal sequence transfer $h$.
@@
 
-Fig.~\ref{fig:panx-benchmark}B-C and SI fig.~\SIfigBenchmark quantify the size of the core genome, i.e.~the sum of all pancontigs present once in every genome, identified by PanGraph and the compression of the dataset, i.e.~the ratio for the sum of the lengths of all pancontigs and the sum of the lengths of all input genomes, for different dataset and the different parameters.
+Fig.~\ref{fig:panx-benchmark}B-C and SI fig.~{\SIfigBenchmark} quantify the size of the core genome, i.e.~the sum of all pancontigs present once in every genome, identified by PanGraph and the compression of the dataset, i.e.~the ratio for the sum of the lengths of all pancontigs and the sum of the lengths of all input genomes, for different dataset and the different parameters.
 As expected, the very homogenous \textit{M.~tuberculosis} dataset has a large core genome with and a compression ratio that is almost equal to the inverse number of input genomes, independent of the aligner or parameters used.
@@
 In this case the use of \textit{mmseqs2} and null energy parameters increases core pangenome fraction, decreases the total pangenome size and and does not compromise the fragmentation of the graph (cf.
-Fig.~\ref{fig:panx-benchmark}B-C and SI Fig.~4).
+Fig.~\ref{fig:panx-benchmark}B-C and SI Fig.~{\SIfigBenchmark}).
 The divergence of the \textit{P.~marinus} dataset sits instead beyond the capabilities of \textit{Pangraph}.
@@
 The number and size of pancontigs scale sub-linearly with the number of isolates, with more than 50\% of the pangenome being included in the 10\% longest pancontigs.
-Core pancontigs having higher-than-average length.
-This is expected given that core genes tend to stay synteny over large evolutionary distances.
+Core pancontigs have higher-than-average length.
+This is expected given that core genes tend to stay syntenic over large evolutionary distances.
 Adding more strains does not generate excessive fragmentation of these pancontigs, and their size remains of several kbps even as hundreds of strains are added.
@@
 We performed the same test on species from the other datasets, obtaining similar results on all species whose divergence is compatible with \textit{PanGraph} capabilities (cf.
-SI Section \SImarginalize and SI Fig.~\SIfigMarg).
+SI Section {\SImarginalize} and SI Fig.~\SIfigMarg).
 
@@
 In these representation the graph usually has a smaller number of nodes, and can include more diverged sequences by detecting homology on protein sequences.
-However this representation is naturally sensitive to gene annotation errors.\\
+However this representation is naturally sensitive to gene annotation errors.
+\textit{PanGraph} instead operates by construction in a gene-agnostic way, being only based on sequence homology, and is thus robust to this type of errors.\\
 
@@
 The efficient implementation of \textit{PanGraph} allows it to create pangenome graphs containing hundreds of isolates in a few hours on a 8-core machine.
-The size of pancontigs and the fraction of core sequence have good scaling properties with the number of isolates in the graph, indicating that \textit{PanGraph} is able to successfully capture pangenome properties.
-By construction \textit{PanGraph} operates in a gene-agnostic way, being only based on sequence homology, and is thus robust to gene annotation errors.\\
+The size of pancontigs and the fraction of core sequence have good scaling properties with the number of isolates in the graph, indicating that \textit{PanGraph} is able to successfully capture pangenome properties.\\
 One of the main limitations is the diversity of the input sequences.

### 2023-03-08 supplementary.tex
chore
Update on Overleaf.

@@
     \caption{{\bf Pangenome graph representation}.
-Given a set of genomes \textit{PanGraph} will construct a pangenome graph by condensing all stretches of homologous sequences in \textit{pancontigs}, containing the alignment of these sequences.
+Given a set of genomes \textit{PanGraph} will construct a pangenome graph by condensing all homologous sequences in \textit{pancontigs}, containing the alignment of these sequences.
 Genomes can then be represented as \textit{paths}, i.e.
@@
 In our pancontig representation, each sequence in the alignment is encoded as variations over the pancontig consensus.
-In practice in the pangraph output \texttt{json} file each alignment sequence is saved as a list of gaps to be added to the consenus, sites to be mutated, insertions to be performed in gaps, and deletions.
-By performing these changes in order on the consensus each sequence in the alignment can be retrieved.
+In practice in the pangraph output \texttt{json} file each alignment sequence is saved as a list of gaps to be added to the consensus, sites to be mutated, insertions to be performed in gaps, and deletions.
+By performing these changes in order on the consensus each sequence in the alignment can be reconstructed.
 In this example we display how each of these variations are encoded and performed.
@@
 Each \textbf{path} is an oriented collection of pancontigs that represents a genome.
-For each item in the sequence along with the pacontig id PanGraph also stores the pancontig orientation (whether homology with the consensus is on the forward the reverse strain of the genome) and which variations on the pancontig consesus (i.e.
-which sequence in the pancontig alignment) correspond to this precise location in the genome.\\
+For each item in the sequence along with the pancontig id, PanGraph also stores the pancontig orientation (whether homology with the consensus is on the forward the reverse strand of the genome) and which variations on the pancontig consensus (i.e.
+which sequence in the pancontig alignment) are present in this precise location in the genome of the isolate.\\
 
-From knowledge of all the pancontigs and pahts, every single genome can be exactly reconstructed.
+From knowledge of all the pancontigs and paths, every single genome can be exactly reconstructed.
 Every part of the sequence is always encoded in a pancontig.
@@
 These genomes are initially considered as elementary graphs composed of a single \textit{pancontig}.
-These graphs are then propagated along the tree towards the root, and at each intersection a merge operation is performed.
-The full pangenome graph is then collected from the root.
+These are then propagated along the tree towards the root, and at each intersection a merge operation is performed.
+The full pangenome graph is then collected from the root.\\
 
-We focus on three aspects:
+We focus on four aspects:
 \begin{itemize}
@@
     \item The effect of the main parameters of the algorithm on the resulting graph structure.
+    \item How pangraph processes duplicated regions and possible strategies for paralog splitting.
 \end{itemize}
@@
 
-At each merge operation two pangenome graphs need to be compressed into a single one. Three of pangraph's parameters play a role in the outcome of this procedure:
+At each merge operation two pangenome graphs need to be joined into a single one. Three of \textit{PanGraph}'s parameters play a role in the outcome of this procedure:
 \begin{itemize}
     \item the minimal pancontig length $\Lthr$.
-    \item pseudo-energy parameters $\alpha$ and $\beta$ (cf. eq.~(1)in the main text).
+    \item pseudo-energy parameters $\alpha$ and $\beta$ (cf. eq.~(1) in the main text).
     \item the choice of alignment kernel (minimap2 or mmseqs) and aligner precision (asm5, asm10 or asm20 for minimap2).
 \end{itemize}
-In the merging, these parameters control how much of the diversity between sequence must be stored as large-scare variation (difference in paths) and how much as fine-grained variation (differences in alignments, stored in pancontigs).
+In the merging, these parameters control how much of the diversity between sequences must be stored as ``large-scale'' variation (difference in paths) and how much as ``fine-grained'' variation (differences in alignments, stored in pancontigs).
 This is done for example by setting a divergence limit for sequences to be stored in the same alignment.\\
-An implicit assumption of our procedure is that the phylogeny of sequences stored inside the same panconting is roughly star-like, with each sequence containing independent variations from the consensus.
+An implicit assumption of our procedure is that the phylogeny of sequences stored inside the same pancontig is roughly star-like, with each sequence containing independent variations from the consensus.
 In such cases the consensus is a good ``average'' representation for the set of sequences in the alignment.
-In the next section we briefly discussion how PanGraph behaves when this assumption is not satisfied, such as in the case of merging ortholog and paralog sequences in the same pancontig.\\
-% move this paragraph or change?
+In the next section we briefly discussion how PanGraph behaves when this assumption is not satisfied, such as in the case of merging paralogs in the same pancontig.\\
 
@@
 It will however be performed later on in the self-map phase of the algorithm (see step 4).
-              \item Flanking regions of the consensus sequences that are not part of the matches are cut and not merged and stored in separate pancontigs.
+              \item Flanking regions of the consensus sequences that are not part of the match are cut and not merged and stored in separate pancontigs (see fig.~1 right panel in the main text).
 One exceptions is when any of these flanking regions is shorter than the threshold length $\Lthr$.
 In this case instead of creating a very short pancontig the region is stored as an indel in the merged pancontig alignment.
-              \item The remaining homolous region is then a candidate for the merging.
-However the merging might result in more than a single panconting, depending on the underlying alignment.
-If the alignment contains insertions or deletions longer thant $\Lthr$, these need to appear as separate pancontigs.
+              \item The remaining homologous region is then a candidate for the merging.
+However the merging might result in more than a single pancontig, depending on the underlying alignment.
+If the alignment contains insertions or deletions longer that $\Lthr$, these need to appear as separate pancontigs.
 To this end the CIGAR string is parsed, and the long match is potentially split in multiple sub-matches, with interposed long insertions or deletions that will be saved as separate pancontigs.
@@
 However at the previous stage only one merge could be performed per pancontig, which could be insufficient to collapse all of the homology contained in the graph.
-To this end we repeat the procedure described in step 3, but instead of looking for matches between pancontigs of two different graphs, we search for matches between all pancontigs in the same graph.
+To this end we repeat the procedure described in step 3, but instead of looking for matches between pancontigs of two different graphs, we perform an all-against-all homology search between all pancontigs in the same graph.
 Merges are performed following the same rules, and this procedure is repeated until no new merges can be performed.
@@
 As graphs are propagated along the guide tree more and more merges are performed. This makes pancontigs progressively shorter but their alignments deeper.\\
-We also point out that in our merging procedure only the consensus sequence of any two pancontigs is used in their alignment.
+
+We point out that in our merging procedure only the consensus sequence of any two pancontigs is used in their alignment.
 This makes this procedure fast and is sufficient to capture the large-scale structural variation of the pangenome, but might introduce minor inconsistencies and artifacts in the alignment.
@@
 
-After explaining the procedure we use for merging, we conclude this section with some comments on the effect of PanGraph's parameters on the resulting pangenome graphs.\\
+After explaining the procedure we use for merging, we comment briefly on the effect of \textit{PanGraph}'s parameters on the resulting pangenome graphs.\\
 
@@
 From the definition of the pseudo-energy (cf.
-eq.~(1) in the main text) it follows that matches between homologous sequence with divergence higher than $1/\beta$ will have negative pseudo-energy and will thus not result in a merge.
+eq.~(1) in the main text) it follows that matches between homologous sequence with divergence higher than $1/\beta$ will have positive pseudo-energy and will thus not result in a merge.
 If the molecular clock hypothesis holds, tuning this parameter is equivalent to setting an effective time limit beyond which sequences are not to be merged in the same alignment.
 This parameter can be tuned based on the diversity of the sample in study.
-It can for example be useful in resolving situations in which the hypothesis of a star-like phylogeny in a pancontig alignment is broken, such as to separate loosely related paralogs.
+It can for example be useful in resolving situations in which the hypothesis of a star-like phylogeny in a pancontig alignment is broken, such as to separate loosely related paralogs, as discussed in the next section.
 
@@
 This paralog splitting can be done using two different sources of information: either sequence diversity or local context in which the duplication is inserted.
-Currently \textit{PanGraph} does not provide a command to perform paralog splitting.
+Currently \textit{PanGraph} does not provide a specific command to perform paralog splitting.
 This is a feature that we plan on implementing in the future.
@@
 This is due to the fact that homologous sequences in our simulations are generated by an evolutionary process, and they can have a wide distribution of diversities and hierarchical population structure.
-Inability to merge sequences above a threshold divergence in a single panconting is expected to result in multiple pancontigs containing sub-clades with higher similarity, and average diversity that is appreciably lower than the aligner threshold.\\
+Inability to merge sequences above a threshold divergence in a single pancontig is expected to result in multiple pancontigs containing sub-clades with higher similarity, and average diversity that is appreciably lower than the aligner threshold.\\
 By performing a linear fit on the datapoints with $\mu \leq 0.002$ we are able to recover the conversion factor between the mutation rate and average pairwise divergence of pancontigs in our simulations: $\avg{d} \sim 20.8 \, \mu$.
@@
 For each of our generated datasets we evaluate the average displacement of inferred pancontig breakpoints.
-The figure represents the cumulative distribution of average misplacements, stratified by the average pairwise divergence of sequences in pancontigs (see colorbar on the right).
+The figure represents the cumulative distribution of average misplacements, stratified by the average pairwise divergence of sequences in pancontigs (see legend on the right).
 Each panel corresponds to a different alignment kernel used to infer the pangenome graph (see panel title).
@@
 These data has been chosen as part of a PanX \cite{ding2018panx} analysis dated March 2018.\footnote{
+    The list of accession numbers is available on the repository: \url{https://github.com/neherlab/pangraph/blob/master/script/config/accession_numbers.csv}.
     We report for completeness that two of the isolates from the \textit{E.~coli} dataset from this analysis, namely \texttt{NC\_020518} and  \texttt{NC\_017663}, were subsequently removed from RefSeq.
@@
 However this quantity is sensitive to annotation errors.
-This is particularly evident in the \textit{E.~coli} dataset, for which $\Lcore = 0.7$ Mbp is likelt to be a subestimation of the real core genome size.
+This is particularly evident in the \textit{E.~coli} dataset, for which $\Lcore = 0.7$ Mbp is likely to be an underestimation of the real core genome size.
 $\Lsoftcore$ is instead less sensitive to annotation errors and usually considered a better core-genome size estimate.
-For example the \textit{E.~coli} core genome is usually considered to be aroud 1500-2500 genes in size, depending on the definition used \cite{park2019large,sutton2021pan,horsfield2023accurate}.
+For reference, the \textit{E.~coli} core genome is usually considered to be around 1500-2500 genes in size, depending on the definition used \cite{park2019large,sutton2021pan,horsfield2023accurate}.
 }\\
@@
 For each species we selected 50 different isolates (10 for PM) and built pangenome graphs using standard parameter values and the \textit{minimap2} kernel with \textit{asm20} option.
-For each species we then randomly picked 50 different pairs of isolates (all 45 for PM).\footnote{The list of selected accession numbers and selected pairs for each species is available on the repository: \url{https://github.com/neherlab/pangraph/blob/master/script/config/projection_strains.json}} For each pair we then built pairwise graphs using the same parameters, and compared them to the corresponding marginalized graph, obtained by projecting the full pangenome graph for the 50 isolates on the selected pair (cf.
+For each species we then randomly picked 50 different pairs of isolates (all 45 for PM).\footnote{
+    The list of selected pairs for each species is available on the repository: \url{https://github.com/neherlab/pangraph/blob/master/script/config/marginalize_pairs.csv}
+}
+For each pair we then built pairwise graphs using the same parameters, and compared them to the corresponding marginalized graph, obtained by projecting the full pangenome graph for the 50 isolates on the selected pair (cf.
 Fig.~6A, main text).\\
@@
 As described in the previous section, the divergence of the HP dataset sits on the limit of what the \textit{minimap2} alignment kernel can successfully handle.
-In this case the results of merging different graphs together can depend on the order of merging, and pairwise and marginalized graphs can potentially show some disagreement.
+As such, the results of merging different graphs together can depend on the order of merging, and pairwise and marginalized graphs can potentially show some disagreement.
 In this case the disagreement remains minor and only extends to few percents of the genome size, with disagreement blocks having size only slightly bigger than the 100 bp threshold of \textit{PanGraph}.

### 2023-03-08 reviewer_response.tex
Update on Overleaf.

@@
 \reviewer{R1.1}{A pancontig is defined as a multiple sequence (sub)alignment.
-In line 54 you say "Pairwise graph alignment is performed by an all-to-all alignment of the pancontigs" which I interpreted to mean all sequences in the alignment vs all sequences in another alignment.
-You do later add that ``To align two graphs, the consensus sequences of all pancontigs in both graphs are searched for homologies and aligned".
+In line 54 you say ``Pairwise graph alignment is performed by an all-to-all alignment of the pancontigs'' which I interpreted to mean all sequences in the alignment vs all sequences in another alignment.
+You do later add that ``To align two graphs, the consensus sequences of all pancontigs in both graphs are searched for homologies and aligned''.
 Adding the word consensus in the earlier sentence might avoid confusion.}
@@
 
-\reviewer{R1.10}{Supplementary Information C: You comment that the use of null energy parameters results on average in ``more and shorter pancontigs" but also that they result in more diverged sequences being merged into the same pancontig.
+\reviewer{R1.10}{Supplementary Information C: You comment that the use of null energy parameters results on average in ``more and shorter pancontigs'' but also that they result in more diverged sequences being merged into the same pancontig.
 Why is a more diverse pancontig more likely to become fragmented?}
@@
 
-\reviewer{R2.1}{Line 46: when you say ``PanGraph transforms an arbitrary set of genomes into a graph that simultaneously compresses the collection of sequences and exhaustively summarizes both the structural and nucleotide-level polymorphisms." - I am a bit unclear on how the nucleotide level variation is summarised.
+\reviewer{R2.1}{Line 46: when you say ``PanGraph transforms an arbitrary set of genomes into a graph that simultaneously compresses the collection of sequences and exhaustively summarizes both the structural and nucleotide-level polymorphisms.'' - I am a bit unclear on how the nucleotide level variation is summarised.
 The paper is a bit hand-wavey about this.
@@
 For each panel we report the results for pangenome graphs built using minimap2 aligner with standard parameter values and asm20 sensitivity (blue) and with mmseqs2 and null pseudo-energy parameters (orange).
-Distributions are weighed by pancontig length, so that the area under the curve represents the total pangenome graph size.}
+Distributions are weighed by pancontig length, so that the sum of all bins heights is equal to the total pangenome graph size.}
       \label{fig:kp-vs-hp}
@@
 In the particular case in which merging two pancontigs would leave overhangs shorter than $\Lthr$, these are included in the alignment as flanking insertions.
-This is now explained in SI section {\SIgraph}, and mentioned in the main text in section IB (line 103).
+This is now explained in SI section {\SIgraph}, and mentioned in the main text in section IB (line 125).
 In fact, we used the fact that genomes can be reconstructed exactly from the pangraph to test and debug the tool. }
@@
 
-\reviewer{R2.3}{line 48: ``Pancontigs are connected by an edge if they are syntenic on at least one input sequence;" - i think this is a bit of an abuse of modern language.
+\reviewer{R2.3}{line 48: ``Pancontigs are connected by an edge if they are syntenic on at least one input sequence;'' - i think this is a bit of an abuse of modern language.
 The word syntenic is generally used to mean two genes/whatever have order conserved on multiple chromosomes (although some googling tells me that in classical genetics it meant things were on the same chromosome).
@@
 I did a tentative correction: removed from the main the specification on population size and mentioned to check the SI for details on standard parameter values.}
-\response{The value $N=100$ is the standard value used in the rest of the simulation, but as the reviewer pointed out this value is varied for the results showed in fig.~2 and the statement was confusing.
-We modified it removing the value of $N$ and referring to SI section {\SIsimBench} for the standard value of simulation parameters.}
+\response{The value $N=100$ is the standard value used in the rest of the simulations (e.g.
+fig.~3), but as the reviewer pointed out this value is varied for the results showed in fig.~2 and the statement was confusing.
+We modified it removing the value of $N$ and referring to SI table~I for the standard value of simulation parameters.}
 
-\reviewer{R2.5}{Figure 5 legend, lower case C in "coli" of "E. coli". Also italics.}
-% \Marco{fixed.}
+\reviewer{R2.5}{Figure 5 legend, lower case C in ``coli'' of ``E. coli''. Also italics.}
 \response{The legend has been corrected.}
@@
 }
+\Marco{Tentative extension - to be updated depending on how we modify the intro/discussion:\\
+
+To amend this we performed the following changes:
+\begin{itemize}
+    \item The sentence in the abstract mentioned by the reviewer has been removed and modified.
+
+    \item We extended the discussion of other methods in the introduction.\\
+    In particular we extended the paragraph on methods based on de Bruijn graphs, mentioning also compacted generalization and the fact that tools exist that are able to build graphs containing thousands of isolates.
+We now cite more tools and put more emphasis on their applications, restricting it to tools specific for microbial genomics (e.g.
+we did not include PGGB).
+We removed the previous claim that homology detection was limited by the k-mer size, and cite tools that can be used to extend homology detection through small sequence modifications.\\
+    We expanded in the same way the paragraph on homology-based and gene-based approaches, in which we now discuss more tools putting emphasis on their applications.
+
+    \item We added a paragraph in the discussion in which we tried to emphasize how \textit{PanGraph} compares to gene-based and k-mer based PGG methods.\\
+    In essence, methods based on de Bruijn graph have the advantage of being fast since they do not require alignments.
+They also do not necessarily require fully assembled genomes.
+However in these models the graph encodes in the same way nucleotide-level variation and large-scale structural variation, and specific post-processing of the graph is needed to distinguish the two.
+Conversely, PanGraph requires fully-assembled genomes and is potentially slower, but its pangenome graph representation provides a natural separation between small sequence differences (mutations or small indels) and big structural changes (gene gain/loss, inversions, rearrangements).\\
+    Gene-based pangenome graphs have instead the advantage of being smaller in size (a node corresponds to a gene) and they can include more divergent sequences, since homology detection can be performed on protein sequences.
+Differently from homology-based methods like PanGraph however, they are often sensitive to annotation errors.
+\end{itemize}
+}
 
@@
 In the future, we hope to extend this further for visualization purposes.
-      % TODO: mention this in the text.
+      % TODO: mention this in the text?
       % Liam: not sure if we need last sentence
@@
 
-\reviewer{R3}{In the pseudo-energy discussion a further parameter is mentioned: "In addition, there is parameter that controls the minimal size of contigs produced by the algorithm.".
+\reviewer{R3}{In the pseudo-energy discussion a further parameter is mentioned: ``In addition, there is parameter that controls the minimal size of contigs produced by the algorithm.''.
 The value chosen for this parameter should be explicitly stated along with a rationale for the choice.
@@
 This might be explained in the SI?}\\
-\response{The minimal size parameter $\Lthr$ is set to 100 bp in PanGraph.
+\response{The minimal size parameter, now named $\Lthr$, is set to 100 bp in PanGraph.
 This parameter controls the minimal size of pancontigs.
@@
       We appreciate the need to better explain the role of this parameter in the merging procedure and in the structure of the resulting graph.
-To this end we extended its definition in section IB in the main text, and give a more extended discussion in SI section~\SIalgo.\\
+To this end we extended its definition in section IB in the main text, and give a more extended discussion in SI section~\SIalgo B and \SIalgo C.\\
       % TODO: Homogenization in the main text with discussion on precision. Extend.
@@
 \response{This point was raised by other reviewers, and we appreciate the need to make this part of the alorithm clearer.
-To this end we added section {\SIalgo} in the SI (see ``graph merging''), with a detailed explanation of how the merging procedure works.\\
+To this end we added section {\SIalgo}B in the SI, with a detailed explanation of how the merging procedure works.\\
       In essence, during graph merging alignment is only performed between the consensus sequences of the merged pancontigs.
@@
 
-\reviewer{R3}{The statement: "Upon completion, transitive 108 edges within the graph, i.e., edges along which all path are co-linear, are removed by merging adjacent pancontigs." needs clarification.
+\reviewer{R3}{The statement: ``Upon completion, transitive 108 edges within the graph, i.e., edges along which all path are co-linear, are removed by merging adjacent pancontigs.'' needs clarification.
 Why would these pancontigs have been split in the first place?
@@
       We appreciate the importance of the point raised by the reviewer.
-      We agree with the reviewer that this post-processing smoothing of the graph produced by of \textit{PanGraph} would be a valuable addition, and we hope to such further post-processing refinements in subsequent versions.
+      We agree with the reviewer that this post-processing smoothing of the graph produced by of \textit{PanGraph} would be a valuable addition, and we hope to include such further post-processing refinements in subsequent versions.
       This sort of refinement would allow alignments of adjacent edges of pancontigs to be harmonized, potentially also removing very small pancontigs when these could be appropriately absorbed in the alignment of larger ones.
@@
 % - [ ] add files with accession numbers and pairs to the submission
-% - [ ] update SI section numbers and line references in the main text and in the response.
-% - [ ] update SI figure numbers.
 % - [x] add mention of ``polish'' in main text.
 % - [ ] for the future: add sequence reconstruction to pangraph automatic testing.
-% - [ ] check that all SI sections have been mentioned in the main text
-% - [ ] add in SI the frequency-size distribution of blocks for the different species? To discuss fragmentation?
-% - [ ] make Liam's repo public
-% - [ ] extend the code to map annotations on pancontigs?
+% - [x] check that all SI sections have been mentioned in the main text
+% - [x] make Liam's repo public

### 2023-03-08 main.tex
Update on Overleaf.

@@
 We simulated populations of size $N$ and sequence length $L$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
-In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (respective rates $0.01$ and $0.05$ per generation), and horizontal sequence transfer that occurs with tunable rate $h$ per generation per genome (see SI sect.~{\SIsynthdata} for details on the simulation and standard parameter values).
+In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (respective rates $0.01$ and $0.05$ per generation), and horizontal sequence transfer that occurs with tunable rate $h$ per generation per genome (see SI sect.~{\SIsynthdata} and SI table~I for details on the simulation and standard parameter values).
 The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.

### 2023-03-10 reviewer_response.tex
Update on Overleaf.

@@
 
-\response{We thank the reviewer for their appreciation of our contribution and for their critical review of our work.}
+\response{We thank the reviewer for their appreciation of our contribution and for their constructive review of our work.}
 
@@
 Adding the word consensus in the earlier sentence might avoid confusion.}
-% \Marco{changed to ``Pairwise graph alignment is performed by an all-to-all alignment of the \emph{pancontigs} consensus sequences between both graphs, and the order of pairwise alignments is determined by a guide tree." Too compressed?
-I could also break in two sentences and expand a bit more.}\\
-% \Liam{that's clear to me.}
 \response{We agree with the reviewer's suggestion. The wording has been updated to ``all-to-all alignment of the consensus sequences of all \textit{pancontigs}''.}
 
-\reviewer{R1.2}{Figure 4C: The main text talks about how to interpret this, as ~1/N for TB and a bit less compression for E.
+\reviewer{R1.2}{Figure 4C: The main text talks about how to interpret this, as $\sim 1/N$ for TB and a bit less compression for E.
 coli and K.
@@
 Or in another way aid this reading?}
-\response{We agree with the reviewer that it is important to be able visually compare the compression with the number $N$ of isolates for each species, and we thank them for the suggestion.
+\response{This is a very useful suggestion -- thank you!
 We modified fig.~4C in the main text and fig.~{\SIfigBenchmark}I in SI marking the value of $1/N$ for each species with horizontal lines, and updated the captions accordingly.}
-% \Marco{maybe try visualizing it as deviation from "one strain" compression (1/N)?
-But the "maximal" compression depends not only on the sample size N but also on the gene frequency distribution.
-Another possible comparison might be to pangraph's estimation of the pangenome size, but again we might have different effects coming into play.}\\
-% \Liam{yes, I think just adding a dashed line for 1/N for each species to the plot would aid that reading and satisfy this point.}
 
@@
 \reviewer{R1.5}{Line 213: Core pancontigs have higher than average size - is this what you would expect given that if they are core genes then they will be present in more/all of the samples so there could be more diversity due to being more sampled?}
-% \Marco{here the pancontig size is the consensus sequence length.
-I would say that the fact that they have higher-than-average size is mostly due to synteny.
-Long blocks (several kbps) contain core genes that always appear in the same order, without any accessory sequence ever occurring in these regions.
-As opposed to accessory regions with a lot of structural diversity, that generates short fragments.
-But here the main message is that core blocks are well-behaved (close to N50 line in the figure), as opposed to fine-grained structural diversity, part of which contains artifacts generated by the block size limit (many short 100bp blocks).
-Should we show block frequency distributions in the supplementary?}\\
-% \Liam{I don't understand their point on diversity - maybe they misunderstood the meaning of size.
-Maybe pancontig `length' is clearer?
-I agree with you about synteny and think this is a nice point to add here to explain the observation.}
-\response{The observation that core pancontigs are longer than average is in agreement with our expectations.
+
+\response{It is true that there are multiple factors at play that could influenze the size of core pancontigs in different ways, but the observation that core pancontigs are longer than average is in agreement with our expectations.
 Multiple core genes are often syntenic even in diverse species, compared to accessory genes that are often found in different contexts.
@@
 I can't read off the fraction agree (shared+private) or disagree which I would like to (although at a glance this is exactly the result you want to see, so happy for it to stand if it is the best projection).}
-% \Marco{maybe add mean and standard deviation as numbers on the side?}\\
-% \Liam{that seems ideal.}
 \response{We considered different projections and in the end decided to maintain the linear projection for ease of interpretation, but we added next to each distribution its mean and standard deviation.
@@
 \reviewer{R1.7}{Have you tried running PanGraph on more than 300 references? I'd love to see another order of magnitude (not necessary for the paper, just curiosity)}
-% \Marco{not really, I never tried on ~500 or ~1000 strains. I could try but would not consider this a priority.}\\
-% \Liam{agree that it's not necessary.}
 \response{The biggest dataset we tried PanGraph on is a collection of 500 E.coli isolates.
@@
 Or how it can be used to analyze genome structure.}
-% \Marco{other than explicitly exporting block sequences and mapping to those, we do not really using the pangenome graph as a reference against which to map new strains...
-But we could expand the discussion with more ideas on applications to study pangenome organization and evolution.
-For software for further analysis and exploration I have \href{https://github.com/mmolari/pypangraph}{this package} to explore and interact with graphs in python, but not sure it's worth mentioning.}\\
-% \Liam{Some ideas that could be mentioned.
-(1) I think one obvious (bad!) use would be to take the core blocks and map reads to those as a `core reference'.
-(2) I know pandora has \href{https://github.com/iqbal-lab-org/make_prg/}{\texttt{make\_prg}} which will construct a `pangenome reference graph' from a set of aligned sequences - I didn't try it on pangraph output.
-(3) For genome structure, exploring `hotspots' in the graph as \href{https://www.nature.com/articles/s41467-017-00808-w}{this paper} does could be done using in/out degree of pancontigs.}
 \response{We have so far not explored the use of pangraph as input for graph based mapping approaches like Pandora or GramTools, but we agree this is an interesting downstream application.
@@
 \reviewer{R1.9}{Supplementary Information: simulation - could you comment on how/if homologous recombination is included in this model e.g. within a gene.}
-% \Marco{homologous recombination is not implemented directly in the model. Only sequence transfer between strains. Genes are not explicitly modeled in the simulation.}\\
-% \Liam{think just adding a comment in the SI saying this is required.}
 \response{Homologous recombination is not directly included in our simulated data generation procedure.
@@
 This is now specified in SI section \SIsynthdata.
-To avoid confusion we also changed ``Horizontal Gene Transfer'' to ``Horizontal Sequence Transfer''.}
+To avoid confusion we also changed ``Horizontal Gene Transfer'' to ``Horizontal Sequence Transfer''.
+Homologous recombination would  ``reshuffle'' fine scale diversity within the contigs.
+We don't expect this to have a large effect on graph reconstruction, though it will affect how fine-scale diversity within the contigs is represented.
+}
 
@@
 Why is a more diverse pancontig more likely to become fragmented?}
-% \Marco{I think that basically the reason is that merging more distant paralogs/orthologs can fragment more the graph. Maybe we could explain this more explicitly in the text.}\\
-% \Liam{makes sense and explaining this reasoning sounds good.}
 \response{We use the pseudo-energy defined in eq.~(1) as an heuristic to decide whether a homologous match detected by our aligner should result in a merge between two pancontigs, depending on the match length and on sequence divergence.
-In particular any choice of $\beta$ would prevent mergers between sequences with divergence higher than $1/\beta$.
-Setting $\beta=0$ instead does not impose a divergence limit on mergers other than the one implicitly set by the aligner sensitivity.
+In particular any choice of $\beta$ would prevent merges between sequences with divergence higher than $1/\beta$.
+Setting $\beta=0$ instead does not impose a divergence limit on merges other than the one implicitly set by the aligner sensitivity.
 In this sense the choice of null energy parameter results in more diverged sequences being merged into the same pancontig.
@@
 
-\response{We thank the reviewer for this succinct summary of the field and the essence of our contribution.}
+\response{We thank the reviewer for this succinct summary of the field and the essence of our contribution and the constructive feedback. We hope we have addressed all points below. }
 
@@
 In our pangenome graph representation each pancontig encodes an alignment using a consensus sequence and list of edits required to obtain each individual sequence from the consensus sequence.
-      These edits consist of single nucleotide changes (mutations) as well as insertions and deletions.
+These edits consist of single nucleotide changes (mutations) as well as short insertions and deletions.
 The consensus sequence is defined using the majority rule on the alignment and is updated after each merger.
 
-      To make this clearer we have added section {\SIgraph} in SI, in which we explain in more details our pangenome graph representation, and section {\SIalgo}B in SI, in which we explain in more detail the merging procedure and how the consensus sequence is updated.
-We also refer to the documentation page mentioned by the reviewer.
-Moreover, to avoid ambiguity we also avoided referring to pancontigs as ``blocks'' in the main text and use pancontigs throughout.}
+Following the suggestion by the reviewer, we have added section {\SIgraph} in SI, in which we explain in more details our pangenome graph representation, and section {\SIalgo}B in SI, in which we explain in more detail the merging procedure and how the consensus sequence is updated.
+We also refer to the documentation page mentioned by the reviewer. Moreover, to avoid ambiguity we also avoided referring to pancontigs as ``blocks'' in the main text and use pancontigs throughout.}
 
@@
 % \Liam{clarifying $N$ and $l$ makes sense.}
-\response{We agree with the reviewer that the formulation in the main text was too concise and a bit inaccurate.
-      There are two main variables at play: the length of a genome $l$ and the number of isolates $n$.
-As the reviewer points out, the set of minimizers for each genome can be created and sorted in $l \log l$.
-Finding the set of shared minimizers would require comparing $\sim n^2$ pairs of ordered lists of minimizers, which scales as $\sim n^2 l$.
-This would provide a scaling $O(n l \log l + n^2 l)$.\\
-      In our implementation we achieve a slightly better scaling as a function of $n$ by sketching and sorting all the genomes together, which can be done in time $L \log L$, where $L= n \times l$ is the total length of all genomes.
+\response{Are sorry that the formulation in the main text was too concise and a bit inaccurate.
+There are two main variables at play: the length of a genome $L$ and the number of isolates $n$. 
+As the reviewer points out, the set of minimizers for each genome can be created and sorted in $L \log L$. 
+Finding the set of shared minimizers would require comparing $\sim n^2$ pairs of ordered lists of minimizers, which scales as $\sim n^2 L$. This would provide a scaling $O(n L \log L + n^2 L)$.\\
+In our implementation we achieve a slightly better scaling as a function of $n$ by sketching and sorting all the genomes together, which can be done in time $\mathcal{L} \log \mathcal{L}$, where $\mathcal{L}= n \times l$ is the total length of all genomes.
 We keep the information on each minimizer origin in the list.
 We then scroll the list and for each chunk of equal minimizers we update the corresponding entries of a $n^2$ matrix of minimizer hit counts.
-Depending on the amount of shared minimizers, this scales as $n^2 l$ in the worst case (all minimizers shared) and as $n l$ in the best case (no minimizer shared).
-Therefore depending on the diversity of the sample the scaling will be between $O(L \log L + n L)$ and $O(L \log L)$.\\
+Depending on the amount of shared minimizers, this scales as $n^2 L$ in the worst case (all minimizers shared) and as $n L$ in the best case (no minimizer shared). 
+Therefore depending on the diversity of the sample the scaling will be between $O(\mathcal{L} \log \mathcal{L} + n^2 L)$ and $O(\mathcal{L} \log \mathcal{L})$.\\
 In practice, for typical use cases of pangraph ($n \sim 100$ and $L \sim 10^6$) the computationally challenging part is the minimizer sketching, and not the distance matrix computation.
@@
 \response{We thank the reviewer for pointing out that we should have provided more explanation (the point was also raised by other reviewers).\\
-      We followed the advice of the reviewer and added Section {\SIalgo}B in the SI, in which we expand on the explanation of how we merge two graphs.\\
+We added Section {\SIalgo}B in the SI, in which we expand on the explanation of how we merge two graphs.\\
 In short, we perform the merging procedure in two steps.
-In the first we use the aligner (minimap2 or mmseqs2) to scan for homology matches between the consensus sequences of all pancontigs of one graph and of the other.
+In the first we use the aligner (minimap2 or mmseqs2) to scan for homologuos matches between the consensus sequences of all pancontigs of one graph and of the other.
 This is achieved by simply passing all these consensus sequences as a single fasta file into mmseqs2 or via the API into minimap2.
@@
 For exact ties in this ranking of of matches, the behavior is not explicitly defined but depends on the order in which the matches appear in the output of the homology search tool.
-% Liam: replaced 'precise ties' > 'exact ties'
 
 After all of the matches have been exhausted, the two graphs have undergone a preliminary merging, with some new merged pancontigs being created. 
-At this point we repeat the procedure looking for matches between the consensus sequences of all the pancontigs in an all-against-all comparison.
+At this point we repeat the procedure looking for matches in an all-against-all comparison of consensus sequences of pancontigs of the newlyt generated graph to itself. 
 This self-mapping on the graph will detect any homology that could not have been merged at the previous stage, and will perform the corresponding merging. 
@@
 Not sure if that tells us anything.}
-% \Marco{the second is actually a consequence of the first, but except for that it is correct.}\\
-% \Liam{maybe say `$\beta$ controls the maximal sequence divergence between the two merger-candidates.
-The negative pseudo-energy requirement means that no merger will be performed if the Hamming distance exceeds $1/\beta$.'}\\
-% \Marco{indeed, the parameter $\beta$ could control how far in the past one wants to accept merges. Should we mention this explicilty in the text?}\\
-% \Liam{could be mentioned as an `intuitively, this should correspond to merging of homologous stretches of sequence in roughly reverse evolutionary order'.}\\
 \response{
@@
 The pseudo-energy ranking also effectively results in merges being performed in reverse evolutionary order, with less diverged sequences being merged first.''
-      % QUESTION: is there ambiguity in the main text on the value of N_m? As SNPs in the alignment? Misleading figure?
 }
@@
 It will also mean you make more false merges, right?}
-% \Marco{setting $\beta$ to zero will indeed remove the 10\% divergence limit set by the pseudo-energy.
-There is another limit that depends on the aligner sensitivity (20 to 30\%, depending on the aligner).
-This limit remains, and in this sense no ``false merges'' are performed?
-Not sure I interpret this correctly.}\\
-% \Liam{I think your interpretation is right. Important to clarify that there are those two limits and that $\beta$ only controls the divergence limit \textit{within} what one can actually align.}
 \response{There are two factors limiting the divergence of mergers that are performed.
@@
 Minimap2 is in general able to find matches up to roughly 20\% sequence divergence, while mmseqs2 can reach divergence up to roughly 30\%.
-Only homology matches of up to this divergence will be suggested by the aligners.
+Only homologous matches of up to this divergence will be suggested by the aligners.
 On top of this, only matches with negative pseudo-energy are accepted as mergers.
-      The parameter $\beta$ thus allows to explicitly set a maximal divergence ($1/\beta$) allowed in mergers, rather than relying on the implicit cut-off that comes from the sensitivity of the aligner.
-\\
+The parameter $\beta$ thus allows to explicitly set a maximal divergence ($1/\beta$) allowed in mergers, rather than relying on the implicit cut-off that comes from the sensitivity of the aligner. \\
 This is now made more explicit in SI sect.~{\SIalgo}C.\\
 When setting $\beta = 0$ the maximal divergence is determined by the aligner sensitivity.
-In this sense more merges are allowed, but all mergers will have some implicit maximal divergence threshold (20\% or 30\%, depending on the aligner).}
+In this sense more merges are allowed, but all merges will have some implicit maximal divergence threshold (20\% or 30\%, depending on the aligner).}
 
@@
 % \Liam{makes sense.}
-\response{This is an interesting observation.
-L50 measures the minimum number of pancontigs needed to cover at least 50\% of the pangenome graph sequence.
+\response{This is an interesting observation. L50 measures the minimum number of pancontigs needed to cover at least 50\% of the pangenome graph sequence. 
 This number is closely related to the total number of pancontigs. 
-In general the use of null energy parameters result in more fragmented graphs, with a higher number of pancontigs (cf.
-fig.~{\SIfigBenchmark}D in SI) so L50 increases.
+In general the use of null energy parameters ($\beta=\alpha=0$) results in more fragmented graphs, with a higher number of pancontigs (cf. fig.~{\SIfigBenchmark}D in SI) so L50 increases. 
 However, the reviewer is correct that \textit{Helicobacter} is an exception and shows the opposite trend: the use of null energy parameters results in less fragmentation of the graph (both in the number of pancontigs and for the value of L50).\\
@@
 
-\begin{figure}[htb]
+\begin{figure}[tb]
       \includegraphics[width=.7\textwidth]{figs_response/hp_vs_kp.pdf}
@@
 \reviewer{R2.8}{Just to make sure i understand - is it possible for sequence to be ``left out'' in between pancontigs?}
-% \Marco{no, every part of the genome must be contained in a pancontig. It can be left out from the consensus if it is considered an insertion.}\\
-% \Liam{if they possibly misunderstood, suggests that some statement that all sequence ends up in a pancontig should be added to manuscript.
-I suspect they are thinking of case of short ($<$100bp) stretches `in between' pancontigs, which (if I understand correctly) will be incorporated as insertions at the end of the consensus alignment of the adjacent pancontig.}
 \response{No, every part of the sequence of each genome must be included in some pancontig. 
@@
 Feel free to replace with a better reference, this si just meant to be helpful.}
-% \Marco{I think this is a good reference, fig. 2 is familiar to us. And also fig.2 in \cite{doolittle2009origin} which is a bit older. But not sure which other Rocha paper they are referring to.}\\
-% \Liam{I think they mean Figure 5 of \cite{touchon2009organised} (\href{https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1000344}{here}).
-Citing Touchon 2009 and 2020 and the Doolittle paper seems fine to me.}
 \response{We thank the reviewer for the suggestion. We now cite the suggested article, along with \cite{touchon2009organised,doolittle2009origin}.}
@@
 \reviewer{R2.2}{Why on earth is the ``get an MSA of a block'' option in pangraph called ``polish''?}
-% \Marco{i don't think this comes from the paper (I used grep on the tex file and we never mention polish).
-I think it comes from the documentation but not sure where.
-We explain polish \href{https://neherlab.github.io/pangraph/tutorials/tutorial_3/\#Polishing-the-alignments}{in this part} of the tutorial and I don't think it's ambiguous.
-If in the paper we explain how we propagate the consensus sequence and related problems we will also mention the polish command in the same part and make explicit that it can be used to polish alignments, this should solve the ambiguity.}\\
-% \Liam{this sounds good.
-I think they misunderstood that polish is not just `extract the MSA' but actually produces a pangraph with the same structure but redone alignments.
-I think what is in the documentation there is great and could almost go straight in the paper, something like: `At each stage pancontigs are compared and aligned using their consensus sequences.
-This shortens considerably the time needed, but might introduce minor inconsistencies and artifacts in the alignments.
-These can be removed at the end of the process by reconstructing the sequences in the block and performing a multiple sequence alignment (see Polish command in documentation).')}
-\response{In our method pancontig mergers are performed using only the consensus sequence as a guide.
+\response{In our method pancontig merges are performed using only the consensus sequence as a guide.
 This makes the merging procedure very fast, but can introduce minor inconsistencies and artifacts with insertions in the alignment as insertions are not aligned to each other in the procedure.
@@
 So it's about adjacency rather than synteny.}
-% \Marco{I agree, we could use ``if they are ajacent/contiguous on at least one input sequence''?}\\
-% \Liam{adjacent sounds good.}
 \response{We agree with the reviewer's observation. We have changed ``syntenic on'' to ``adjacent in''.}
@@
 \reviewer{R2.4}{Line 131: was N really=100? on the plot in figure 2, you have n.isolates going up to 1000}
-% \Marco{$N=100$ was the standard value for the rest of the simulations (e.g.
-later for accuracy).
-I did a tentative correction: removed from the main the specification on population size and mentioned to check the SI for details on standard parameter values.}
 \response{The value $N=100$ is the standard value used in the rest of the simulations (e.g.
@@
       % Liam: I think good to say in more detail what has been added here. 
-}
-\Marco{Tentative extension - to be updated depending on how we modify the intro/discussion:\\
-
-To amend this we performed the following changes:
+      Specifically, we have
 \begin{itemize}
-    \item The sentence in the abstract mentioned by the reviewer has been removed and modified.
-
-    \item We extended the discussion of other methods in the introduction.\\
+    \item The sentence in the abstract mentioned by the reviewer has been replaced by a general summary of the need to and challenges in understanding evolution of genome structure.
+    \item We extended the discussion of other methods in the introduction.
     In particular we extended the paragraph on methods based on de Bruijn graphs, mentioning also compacted generalization and the fact that tools exist that are able to build graphs containing thousands of isolates.
@@
 we did not include PGGB).
+We removed the previous claim that homology detection was limited by the k-mer size, and cite tools that can be used to extend homology detection through small sequence modifications.
+    \item We expanded in the same way the paragraph on homology-based and gene-based approaches, in which we now discuss more tools putting emphasis on their applications.
+    \item We added a paragraph in the discussion in which we tried to emphasize how \textit{PanGraph}, as the reviewer remarks, treats nucleotide-level variation and large-scale structural variation in different ways, thereby facilitating separate analyses of these aspects.
+\end{itemize}
+}
+% \Marco{Tentative extension - to be updated depending on how we modify the intro/discussion:\\
+
+% To amend this we performed the following changes:
+% \begin{itemize}
+%     \item The sentence in the abstract mentioned by the reviewer has been removed and modified.
+
+%     \item We extended the discussion of other methods in the introduction.\\
+%     In particular we extended the paragraph on methods based on de Bruijn graphs, mentioning also compacted generalization and the fact that tools exist that are able to build graphs containing thousands of isolates.
+We now cite more tools and put more emphasis on their applications, restricting it to tools specific for microbial genomics (e.g.
+we did not include PGGB).
 We removed the previous claim that homology detection was limited by the k-mer size, and cite tools that can be used to extend homology detection through small sequence modifications.\\
-    We expanded in the same way the paragraph on homology-based and gene-based approaches, in which we now discuss more tools putting emphasis on their applications.
+%     We expanded in the same way the paragraph on homology-based and gene-based approaches, in which we now discuss more tools putting emphasis on their applications.
 
-    \item We added a paragraph in the discussion in which we tried to emphasize how \textit{PanGraph} compares to gene-based and k-mer based PGG methods.\\
-    In essence, methods based on de Bruijn graph have the advantage of being fast since they do not require alignments.
+%     \item We added a paragraph in the discussion in which we tried to emphasize how \textit{PanGraph} compares to gene-based and k-mer based PGG methods.\\
+%     In essence, methods based on de Bruijn graph have the advantage of being fast since they do not require alignments.
 They also do not necessarily require fully assembled genomes.
@@
 Conversely, PanGraph requires fully-assembled genomes and is potentially slower, but its pangenome graph representation provides a natural separation between small sequence differences (mutations or small indels) and big structural changes (gene gain/loss, inversions, rearrangements).\\
-    Gene-based pangenome graphs have instead the advantage of being smaller in size (a node corresponds to a gene) and they can include more divergent sequences, since homology detection can be performed on protein sequences.
+%     Gene-based pangenome graphs have instead the advantage of being smaller in size (a node corresponds to a gene) and they can include more divergent sequences, since homology detection can be performed on protein sequences.
 Differently from homology-based methods like PanGraph however, they are often sensitive to annotation errors.
-\end{itemize}
-}
+% \end{itemize}
+% }
 
@@
 The mechanism for this is obvious but providing a useful user interface might be tedious.}
-% \Liam{I think perhaps producing a helper python script that converts a genbank annotation file for a genome within a pangraph to an output gff/csv with the block locations and coordinates would please this reviewer.}\\
-% \Marco{Yeah good idea. Maybe something that just map each annotation into a block, without solving the problem of harmonizing annotations which might be some more work...}\\
 \response{We thank the reviewer for this important suggestion and appreciate the need to include a way to map annotations back onto the pangenome graph. 
@@
 The script and instructions on its usage can be found at \url{https://github.com/liampshaw/pangraph\_annotations}. 
-In the future, we hope to extend this further for visualization purposes.
-      % TODO: mention this in the text?
-      % Liam: not sure if we need last sentence
+In the future, we hope to extend this further to enable more direct biological interpretation of what the pancontigs code for. 
+Integrating a fully featured annotation into pangraph was not feasible during the 30 day revision window, but is on our list of future milestones. 
 }
@@
 \response{
-      We apologize and we modified the sentence in question. We also extended the discussion of de Bruijn graphs methods in the introduction.\\
+We apologize and we modified the sentence in question. We also extended the discussion of de Bruijn graphs methods in the introduction (see answer above).\\
 We thank the reviewer for pointing out these important contributions that were not previously discussed in the paper.
 }
-% \Marco{from what I understand the structure in de Bruijn graphs is very fine-grained.
-Every SNP can generate a branching event.
-While in our case we have this intermediate corse-graining where homologous parts are collapsed together and the remaining structure of the graph represents large-scale structural variation in the genomes.
-I don't think this is solved by collapsing nonbranching portions in dB graphs (but I might not know enough about this...)}\\
 
@@
 This is an important point that needed more discussion in the paper than we originally provided. 
-To be clear: \textit{PanGraph} in its current state does not explicitly distinguish between orthologs and paralogs when constructing the graph.
-This is a deliberate design choice.
-% Liam: don't understand this part of sentence so deleted 'as better explained in the answer to the last question.'.
-I think it's better to say that it's deliberate than to equivocate
-The possibility of grouping duplicated sequences in the same pancontig will also group information on their context, and the fragmentation introduced in the graph would reflect some real (potentially ancient) structural change in the genome.\\
-% Liam: not sure I quite understand above sentence...
+\textit{PanGraph} in its current state does not explicitly distinguish between orthologs and paralogs when constructing the graph. 
+This is a deliberate design choice: PanGraph is intended to facilitate the investigation of genome evolution and two very similar paralogs represent a recent duplication or acquisition that we want to capture in the graph structure.
+Whether paralogs are merged depends on their similarity and the parameters of pangraph allow tuning this thresholds.
 
-However, as the reviewer points out, in the case of distantly related paralogs such a grouping might be undesirable if it introduces `complications' in the graph.
-We agree from experience that these can sometimes make understanding the pangenome more complicated.
-Currently \textit{PanGraph} supports two different ways of controlling paralog splits.\\
-The first one is based on sequence divergence.
-By tuning the value of the energy parameter $\beta$ the user can set a divergence limit.
-This would prevent the merging of paralogs that have sequence divergence higher than $1/\beta$.\\
-The second method relies instead on information on genomic context.
-The \texttt{marginalize} command possesses the optional \texttt{--reduce-paralog} flag.
+But we agree from experience that these can sometimes complicate the interpretion of the pangenome graph, in particular in presence of high copy-number mobile elements.  
+Currently \textit{PanGraph} supports two different ways of controlling paralog splits.
+The first one is based on sequence divergence as mentioned above.
+By tuning the value of the energy parameter $\beta$ the user can set a divergence limit. This would prevent the merging of paralogs that have sequence divergence higher than $1/\beta$.\\
+The second method relies instead on information on genomic context. The \texttt{marginalize} command possesses the optional \texttt{--reduce-paralog} flag. 
 When set, this flag will result in the removal of artificial splits caused by duplicated pancontigs that always occurr in the same non-duplicated context. 
-These duplicated pancontigs are then split and absorbed in their context.\\
+These duplicated pancontigs are then split and absorbed in their context.
+In addition, pangraph can export a graph omiting all duplicated blocks, which is a simplified version of the graph often useful for visualization.
 
@@
 It is also not clear if any of the splitting pancontigs are less than this value if the entire merger is dropped or just massaged to not have the too small new pancontig.}
-% \Liam{this relates to R2's point above about all sequence ending up in pancontigs so I think is just clarification.}\\
-% \Marco{I agree.
-We set a minimal cutoff of 100bp.
-This is stated later in line 106.
-Maybe we could expand the sentence in line 99 and give the value and few more words on the meaning/effect.
-This controls in some sense the ``corse graining'' of the graph: what is the size of an event that we want to encode as structural variation in the graph, rather than as an indel in the alignment.
-In the algorithm this minimal cutoff has different effects depending on the context.
-When aligning, if the match lenght is less than this threshold the merge is refused.
-For long alignments, if the CIGAR string contains indels that are longer than this threshold, these are saved as separate blocks and not as indels in the block.
-The same goes for the flanking sequences, if merging two long blocks leaves overhangs that are less than this threshold size then these overhangs are saved as part of the merged block and not as separate small blocks.
-This might be explained in the SI?}\\
 \response{The minimal size parameter, now named $\Lthr$, is set to 100 bp in PanGraph.
@@
       \begin{itemize}
-            \item In addition to the pseudo-energy costraint, merges are refused if they have alignment length shorter than $\Lthr$.
+            \item In addition to the pseudo-energy constraint, merges are refused if they have alignment length shorter than $\Lthr$.
             \item During merging, if the alignment contains long ($> \Lthr$) indels, these will cause a split in the merged pancontigs and will be treated as separate pancontigs.
@@
 For the simulated data how accurate are the final pancontig consensus sequences?}
-% \Liam{expanding discussion of doing an MSA at end as for R2's comments would partially address.
-Quantifying how the consensus sequence degenerates iteratively is an interesting point that seems possible?}\\
-% \Marco{I agree, this could be addressed by adding a more detailed algorithm explanation of the merging algorithm in the SI.
-The short answer is that at each merging the full alignment is constructed using the CIGAR of the alignment of the consensus of the two blocks.
-Then consensus of the full alignment is re-evaluated.
-This does not solve the multiple alignment problem during constructions and there can be degeneration on where indels are mapped, but the final graph can be eventually polished.
-I don't think we ever quantified how much the consensus sequence degenerates during the iterations.
-Maybe one could devise a proper test set and see how it degenerates over the iterations.}\\
-\response{This point was raised by other reviewers, and we appreciate the need to make this part of the alorithm clearer.
+
+\response{This point was raised by other reviewers, and we appreciate the need to make this part of the algorithm clearer. 
 To this end we added section {\SIalgo}B in the SI, with a detailed explanation of how the merging procedure works.\\
 In essence, during graph merging alignment is only performed between the consensus sequences of the merged pancontigs. 
-We avoid performing a multiple sequence alignment during merging to maintain the algorithm scalable.
+We avoid performing a multiple sequence alignment during merging to keep the algorithm scalable. 
 However we do update the consensus of the pancontig at each merge using the majority rule on the merged alignment.\\
-      The idea behind this approach is that the consensus sequence is needed to find large-scale homology, and resolving the fine-grained structure of the alignment is not needed at this step and would require an exccessive amount of resources.
-However after the graph has been built alignments can be refined with \textit{PanGraph}'s \texttt{polish} command.
+The idea behind this approach is that the consensus sequence is needed to find large-scale homology, and resolving the fine-grained structure of the alignment is not needed at this step and would require an excessive amount of resources.
+This separates efficiently the problem of resolving the large-scale and nucleotide-level diversity between sequences.\\
+However after the graph has been built, alignments can be refined with \textit{PanGraph}'s \texttt{polish} command. 
 This command performs a multiple sequence alignment on all the sequences in each pancontig and updates the consensus sequence accordingly.
-This separates efficiently the problem of resolving the large-scale and nucleotide-level diversity between sequences.\\
-      We did not quantify explicitly the divergence of consensus sequences during graph building, but we observed that, even without ``polishing'', alignments of up to hundreds of sequences look consistent.
-      % TODO: quantify this in simulations if there is time?
-      % Liam: Think the last sentence is potentially risky - I don't get what `consistent' means?
-My rewrite (not confident in this): In our experience, the alignments appear reliable throughout this process, although we have not explicitly quantified it from our simulations.
+In our experience, the alignments appear reliable throughout this process, although we have not explicitly quantified it from our simulations.
 However, for cases where the alignments are of interest we recommend running `polish' to perform MSA for pancontig consensus sequences (as in the documentation) to ensure that the final alignments are as accurate as possible.
@@
 The authors should also include analysis on the calculated core genome length versus actual core genome length for the simulated data.}
-% \Liam{0.7Mbp does seem lower than I would expect - I guess that they expect $>$2 Mbp - but it shows that obsessing about core is a bit unwise.
-With slightly relaxed definitions (95\% of genomes) does the core get sensibly bigger?
-For comparison e.g.
-ggCaller gets ~2,500 core genes in 100\% of n=162 \textit{E.
-coli} (Table 1 of \href{https://www.biorxiv.org/content/10.1101/2023.01.24.524926v1.full.pdf}{here}).}\\
-% \Marco{I checked and soft core (gene in 95\% of strains) is 2Mbp in our dataset.
-Some of our problems might come from weird strains.
-E.g.
-the second strain with the least amount of soft-core genes is \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_020518.1?report=genbank}{NC\_020518}, an artificial genome that was removed from RefSeq.
-The strain that misses the most of the 1000 most frequent genes \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_017663.1?report=genbank}{NC\_017663} was also removed from RefSeq.
-The rest might be due to misannotations?
-At least one of the most common 1000 genes are missed by 87/307 strains.
-Interestingly in \cite{sutton2021pan} they report $\sim$700 core orthologous gene clusters for an E.coli dataset of $\sim$1000 genomes, if they do not harmonize annotations.
-Those increase to $\sim$1500 after refining their annotations.
-% The easiest thing to do might be to just add the size of soft-core for all species? Maybe these values might also be added as dashed lines in fig. 4G for comparison with pangraph's estimation?}\\
 \response{
@@
       For example in \cite{sutton2021pan} using a similar strict definition the authors report finding $\sim$700 core orthologous gene clusters for an E.~coli dataset of $\sim$1000 genomes (see table 2 in their publication), compatible with our estimate of 0.7 Mbp.
-They impute this to misannotations, since after refining the annotations using structural infromation from the pangenome this number increases to $\sim$1500 genes.
+They impute this to misannotations, since after refining the annotations using structural information from the pangenome this number increases to $\sim$1500 genes.
 } If we measure the size of the soft-core genome with a frequency threshold of 95\% we find a more reasonable size of 2 Mbp. We now report both of these values in table~I.\\
@@
 }
-% The isolates analyzed in our papers were downloaded from RefSeq.
-The accession numbers are now explicitly included in the sumbission (file \texttt{accession\_numbers.csv}).
-These data were previously analyzed in our group with the PanX pipeline \cite{ding2018panx}.
-We rely on this pipeline for the clustering of homologous genes, and the values shown in table~1 results from this analysis.
-This is now made explicit in the caption.
-This table is only meant to provide some general characterization of the size and diversity of the various datasets, and PanX analysis results are not used in PanGraph.\\
-
-% As noted by the reviewer, the 0.7 Mbp size for E.~coli core genome found by PanX is smaller than what usually reported in the literature.
-Estimates of E.~coli pangenome size are usually around 2000-2500 genes \cite{park2019large,horsfield2023accurate,sutton2021pan}.
-However these estimates often either use a soft-core definition for the core genome \cite{park2019large} or use some form of annotation refinement to mitigate misannotations \cite{horsfield2023accurate,sutton2021pan}.\\
-% In our case since we are using a strict definition for core genome the discrepancy could have different origins: problematic assemblies in our datasets, misannotations or problems with gene clustering.\\
-% As a first step we checked the size of the soft-core genome, defined as the set of gene clusters that are shared by at least 95\% of the strains.
-For our E.~coli dataset this has size of 2 Mbp, which is closer to our expectation.\\
-% We then ranked strains by the amount of soft-core genes that they include, and found some problematic assemblies.
-Some of the isolates with the lowest fraction of soft-core genes were subsequently removed from RefSeq, for example \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_020518.1?report=genbank}{NC\_020518} (an artificial genome) and  \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_017663.1?report=genbank}{NC\_017663}.\\
-% These problematic strains however do not explain the small core-genome size that we find.
-For instance, over the 307 strains in our dataset, 87 are missing at least one of the most common 1000 genes.
-This discrepancy is however compatible with misannotations.
-For instance by using a strict core definition in \cite{sutton2021pan} the authors report finding $\sim$700 core orthologous gene clusters for an E.~coli dataset of $\sim$1000 genomes (see table 2 in their publication).
-This would be compatible with our estimate of 0.7 Mbp.
-After refining the annotations using structural infromation from a pangenome graph, the number of strict core gene clusters increases to $\sim$1500.
-In our PanX analysis we do not perform this harmonization and our esitmate remains biased by annotation errors.\\
-
-% To amend this we added the size of soft-core genome for each species in table~1 in the main text, and added part of this discussion in SI sect.~{\SIdataBenchmark}A.
-% TODO: core genome for simulated data if there is time?
-
-%     "all core": 778,
-%     "all softcore": 2081,
-%     "all genes": 23128,
-%     "all nondupl": 18798,
-%     "drop core": 822,
-%     "drop softcore": 2069,
-%     "drop genes": 23054,
-%     "drop nondupl": 18749
 
@@
 We hope to continue to expand the downstream post-processing of \textit{PanGraph} data structures in future work.
-      % Liam: tried to make it sound more explicitly 'future' and 'further'
+      For small scale variation within pancontigs, we already have the {\it polish} command. 
 }
@@
       We agree with the reviewer that fragmentation of the graph is an important concern, and that it can constitute a limitation in the number and diversity of genomes that can meaningfully be added in the pangenome graph.
-For example datasets in which subsequent structural changes occurr at a scale comparable with $\Lthr$ cannot be correctly be transformed into a graph.
-Moreover an excessive number of pancontigs will also increase the amount of computational resources required.\\
       Within the size limit of our datasets, in the paper we investigated fragmentation in sect.~2B (``\textit{scaling with dataset size}''). 
       In particular in fig.~5 we quantify how the size and number of pancontigs vary as a function of the number of isolates included in the graph. 
-Here we observed good scaling properties, with pancontig size decreasing slowly as more isolates are added.
-We expect the threshold level of fragmentation to be reached around the point where the lines intercept the threshold length $\Lthr = 100$bp.\\
+      Here we observed good scaling properties of the pancontig size as more isolates are added (much slower than $1/n$).
+      This is true in particular for core pancontigs. 
+      However, if the size of contigs decreases to a degree that it is comparable to the minimal length threshold $\Lthr$, we expect the graph to be no longer very useful.
 
-      % Other than being introduced by the diversity of the dataset, fragmentation could also be introduced artificially in the merging procedure.
-We tested for excessive fragmentation in simulated data using partition entropy, comparing the real underlying structural partition to the one reconstructed by pangraph (not shown in the paper).
-We found that fragmentation was present only when the diversity of the sample was too high for the sensitivity of the aligner.\\
-      Irrespective of its origin, a possible way to detect excessive fragmentation and check for consistency in the graph structure is to visualize the distribution of pancontig size in the graph (see for instance fig.~\ref{fig:kp-vs-hp} lower right panel).
+      %Irrespective of its origin, a possible way to detect excessive fragmentation and check for consistency in the graph structure is to visualize the distribution of pancontig size in the graph (see for instance fig.~\ref{fig:kp-vs-hp} lower right panel).
 Excessive fragmentation in the graph would be detectable as most of the sequence being condensed in pancontigs having size close to $\Lthr$.\\

### 2023-03-10 main.tex
Update on Overleaf.

@@
     In addition to pangenomic approaches that focus on the variation in the sets of genes present in different genomes, complete assemblies allow investigations of the evolution of genome structure and gene order.
-    This latter problem, however, is computationally demanding and conceptually challenging with few tools available that allow to shed light on this dynamics. 
+    This latter problem, however, is computationally demanding and conceptually challenging with few tools available that allow to shed light on these dynamics. 
     Here, we present \emph{PanGraph}, a Julia-based library and command line interface for aligning whole genomes into a graph.
@@
 During evolution, microbial genomes change by both local mutations and large-scale alterations \cite{arnold2021horizontal}.
-Local mutations only change a few nucleotides by substitution, insertion or deletion.
+Local mutations only change a few nucleotides by substitution or small insertions and deletions.
 Conversely, large-scale alterations reorganize the sequence, and involve either the homologous recombination of large segments, gene loss or gain, inversions, or mobilisation of genetic elements.
@@
 Homologous recombination is rapid enough that most genes in many bacterial core genomes have distinct phylogenies \cite{sakoparnig2021whole} and even closely related genomes differ dramatically in gene content \cite{touchon2020phylogenetic,touchon2009organised,doolittle2009origin}.
-% \cite{touchon2020phylogenetic,doolittle2009origin,touchon2009organised}.
-% comment: rewritten to use consistent terminology (I think 'mutation' and 'polymorphism' for most people make them think of local changes only)
-%Scalable inference techniques of this complicated relatedness structure would, in part, spur the advance of critical epidemiological surveillance tools, analogous to NextStrain, for bacterial pathogens.
 
@@
 Since reference-based approaches only partially capture microbial diversity \cite{tettelin2008comparative}, the concept of the \emph{pangenome} has motivated the development of methods that account for substantial variation in gene content.
-However, though such approaches can accurately capture nucleotide polymorphisms within genes, they approximate structural polymorphisms as gene presence-absence relationships \cite{page2015roary,ding2018panx} irrespective of gene order or orientation.
+However, though such approaches can accurately capture nucleotide polymorphisms within genes, they usually approximate structural polymorphisms as gene presence-absence relationships \cite{page2015roary,ding2018panx} irrespective of gene order or orientation.
 This new era of pangenomics demands novel data structures to encapsulate the \emph{complete} diversity of a given genomic sample set.
@@
 In recent years, efforts have focused on generalizing the \emph{pangenome} framework of microbial diversity to graphical models \cite{eizenga2020pangenome}.
-At a high level, pangenome graphs generalize the reference sequence coordinate system conventionally used and encode genomes as directed paths through the graph consisting of node that represent sequence.
+At a high level, pangenome graphs generalize the reference sequence coordinate system conventionally used and encode genomes as directed paths through the graph consisting of nodes that represent sequence.
+Method development for pangenome graph inference of humans and other eukaryotes is a very active and broad area of research \cite{eizenga2020pangenome}. 
 In multicellular eukaryotes, pangenome graphs usually have an overall linear structure in which structural diversity can be encoded as short range excursions.
@@
 However, in these methods the graph structure encodes at the same time nucleotide-level variation and large-scale structural rearrangements.
-Separating these two scales is often difficult, and specific techniques have to be employed to detect large-scale homology in these graphs \cite{angiuoli2011mugsy,minkin2013sibelia}.
+For investigations of the evolutionary dynamics of these genomes, separating these two scales if often of interest. 
+%Separating these two scales is often difficult, and specific techniques have to be employed to detect large-scale homology in these graphs \cite{angiuoli2011mugsy,minkin2013sibelia}.
 
@@
 These early methods however usually scale poorly to large sets of genomes.
-More recently techniques based on homology have been used to compare genomes across diverse sets of species across the tree of life, at the cost of memory \cite{armstrong2020progressive}.
-Alternatively, homology can be searched directly on genes.
-Many methods rely first on grouping genes into orthologous clusters.
+%More recently techniques based on homology have been used to compare genomes across diverse sets of species across the tree of life, at the cost of memory \cite{armstrong2020progressive}.
+Other methods rely on first grouping genes into orthologous clusters.
 These clusters are then treated as nodes in the graph, with edges linking clusters that are adjacent on at least one of the genomes.
@@
 Here we present \emph{PanGraph}, a Julia \cite{bezanson2017julia} library and command line interface, designed to efficiently align large sets of closely related genomes into a pangenome graph on personal computers.
-The \emph{PanGraph} algorithm creates nodes of homologous sequences through alignments alone, without requiring gene annotation.
+The \emph{PanGraph} algorithm creates nodes of homologous sequences through nucleotide alignments alone, without requiring gene annotation.
 The resulting graph both compresses the input sequence set and succinctly captures the population diversity at multiple scales: from nucleotide mutations and indels to structural polymorphisms driven by inversions, rearrangements, and gene gain/loss.
@@
 See Fig \ref{fig:visualization}B for a graphical definition for each term.
-Additionally, $\alpha$ and $\beta$ are hyperparameters of the algorithm.
-Only mergers whose alignment has negative pseudo-energy are performed.
+$E$ thus compares the compactification of the graph by the length of the match $\ell$ with the increase in complexity of encoding the graph in terms of the number of additional blocks $N_c$ and the diversity within the block $N_m$.
+The smaller $E$, the more favorable the match and only mergers whose alignment has negative pseudo-energy are performed.
+
 The parameter $\beta$ controls the maximal sequence divergence between the two merger-candidates.
@@
 Importantly, this divergence threshold is applied to the comparison of the consensus sequences of the merger candidates.
-The pseudo-energy ranking also effectively results in merges being performed in reverse evolutionary order, with less diverged sequences being merged first.
+The pseudo-energy ranking also effectively results in merges being performed in reverse evolutionary order, with less diverged sequences with long-range synteny being merged first.
+
 The parameter $\alpha$ controls the fragmentation of the graph and imposes a minimal length on the pan-contigs resulting from the merger.
-The merger is always performed if it doesn't increase the number of contigs, but mergers that introduce 1, 2, 3 or 4 cuts are only performed for sufficiently long homologous stretches as parameterized by $\alpha$.
-In addition to the fragmentation parameter $\alpha$, there is a further parameter $\Lthr$ that controls the minimal length of pancontigs.
-Structural variation at a scale smaller than this threshold (i.e.
-short indels) is stored in pancontig alignments, rather than in the graph structure, and no merges are performed if they have length shorter than this threshold.
+This parameter has not effect if merging does not increase the number of contigs, but mergers that introduce 1, 2, 3 or 4 cuts are only performed for sufficiently long homologous stretches as parameterized by $\alpha$.
+In addition to the fragmentation parameter $\alpha$, there is a further parameter $\Lthr$ that explicitly controls the minimal length of pancontigs.
+Structural variation at a scale smaller than this threshold (i.e.~short indels) is stored in pancontig alignments, rather than in the graph structure, and no mergers are performed if they have length shorter than this threshold.
 For instance, when a merge would generate flanking contigs shorter than this threshold, the merging is performed but the alignment is extended to include these regions.
-We set a default value of 100 bp for this parameter.
+The default value for this parameter is 100bp.
 
@@
 
-At the graph level, the merger of two \emph{pancontigs} defines a new \emph{pancontig}, connected on both sides by edges to the neighboring \emph{pancontigs} of both inputs, and thus locally collapses the two graphs under consideration.
-At the nucleotide level, the pairwise alignment of two \emph{pancontigs} maps the consensus of one onto the other; the merger of two \emph{pancontigs} requires the application of the map onto the underlying multiple-sequence alignment.
+At the graph level, the merge of two \emph{pancontigs} defines a new \emph{pancontig}, connected on both sides by edges to the neighboring \emph{pancontigs} of both inputs, and thus locally collapses the two graphs under consideration.
+At the nucleotide level, the pairwise alignment of two \emph{pancontigs} maps the consensus of one onto the other; the merge of two \emph{pancontigs} requires the application of the map onto the nucleotide level diversity of the underlying multiple-sequence alignment.
 Once both sets of sequences are placed onto a common coordinate system, the resultant consensus sequence, and thus polymorphisms, are recomputed.
@@
 This shortens considerably the time needed, but might introduce minor inconsistencies and artifacts in the alignments.
-After building the graph, these can be removed by performing a multiple sequence alignment on each pancontig  (see ``polish'' command in documentation and SI sect.{\SIalgo}B).
+After building the graph, these can be removed by performing a standard multiple alignment of sequences within each pancontig (see ``polish'' command in documentation and SI sect.{\SIalgo}B).
 
@@
 The algorithmic complexity was measured empirically by constructing pangenome graphs from data generated by simulating populations of varying size and sequence length.
-The mean and standard deviation of the run-time obtained from 50 iterations are shown in Fig. \ref{fig:toy-performance}.
+The mean and standard deviation of the run-time obtained from 50 iterations are shown in Fig.~\ref{fig:toy-performance}.
 Importantly, PanGraph's computational complexity grows \emph{linearly} with the number of input genomes once the number of genomes exceeds a certain threshold and parallelism can be exploited.
@@
 
-% We downloaded from RefSeq \cite{o2016reference} completely assembled chromosomes from 5 different bacterial species: \textit{Escherichia coli} (EC), \textit{Klebsiella pneumoniae} (KP), \textit{Helicobacter pylori} (HP), \textit{Mycobacterium tuberculosis} (MT) and \textit{Prochlorococcus marinus} (PM).
-These data had been previously analyzed using PanX \cite{ding2018panx}, which allowed us to estimate the size of the pangenome and core genome, and the average pairwise divergence on core genes (cf.
-Table~\ref{table:panx-dataset} and SI Section~{\SIdataBenchmark}A).
-
-Among these 5 data sets, the \textit{E.~coli}, \textit{K.~pneumoniae}, \textit{M.~tuberculosis} genomes have core genome diversity below the thresholds of the minimap2 aligner, while \textit{H.~pylori} is diverse enough that we don't expect minimap2 to find all relevant matches, while mmseq2 should process this set without problems.
+Among these 5 data sets, the \textit{E.~coli}, \textit{K.~pneumoniae}, \textit{M.~tuberculosis} genomes have core genome diversity below the thresholds of the minimap2 aligner, while \textit{H.~pylori} is diverse enough that we don't expect minimap2 to find all relevant matches, while mmseqs2 should process this set without problems.
 \textit{P.~marinus}, on the other hand, is very diverse with an average core genome diversity of 26\% and is beyond what we expect PanGraph to handle.
-
-% The \textit{E.~coli} (EC) dataset is the one with the highest number of isolates (307 chromosomes) and the largest pangenome.
-The species with the highest core genome divergence are \textit{H.~pylori} (HP) and \textit{P.~marinus} (PM).
-For the former, the divergce sits at the limit of what the \textit{minimap2} alignment kernel is able to accurately merge (cf.
-Fig.~\ref{fig:toy-accuracy}), while for the latter the diversity is beyond even the limits of \textit{mmseqs2}.
 
@@
         and average pairwise divergence of core genes ($\dcore$).
+Note that the $\Lcore$ of might be artificially low due to missing annotations in some of the input genomes and $\Lsoftcore$ is a more robust measure of the core-genome size.
     }
@@
 While single nucleotide differences in the core genome are straightforward to analyze with existing tools, these analyses miss the great majority of genetic diversity.
-The ability to rapidly align large sets of complete genomes of a bacterial species is crucial for the investigation of the processes governing microbial evolution.
-We developed \textit{PanGraph} as a tool that can address this need, by being able to capture the structural and nucleotide diversity in both the core and accessory genome in a scalable way.\\
-Importantly, this diversity is encoded in different ways depending on the scale, with large-scale variation affecting the graph structure, while nucleotide-level variation being included in pancontig alignments.\\
+The ability to rapidly align large sets of complete genomes of a bacterial species is crucial for the investigation of the processes governing microbial genome evolution and structure.
+\textit{PanGraph}, the tool presented here, is able to capture the structural and nucleotide diversity in both the core and accessory genome in a scalable way.
+The resulting data structure captures large-scale structural variation in the connectivity of the graph, while nucleotide-level variation being included in pancontig alignments.
 

### 2023-03-10 supplementary.tex
Update on Overleaf.

@@
 \newcommand{\red}[1]{{\color{red}#1}}
+\renewcommand{\figurename}{Fig.~S}
 
@@
 This section contains a short explanation of the pangenome graph representation used in \textit{PanGraph}.
-A pangenome graph is essentially composed of two elements: pancontigs, which encode for alignment of homologous sequences, and paths, that correspond to representation for genomes as a sequence of pancontigs (see fig.~\ref{fig:graph-representation}).\\
+A pangenome graph is essentially composed of two elements: pancontigs, which encode the alignment of homologous sequences, and paths, that correspond to representation for genomes as a sequence of pancontigs (see fig.~S\ref{fig:graph-representation}).\\
 
-\begin{figure}[htb]
-    \includegraphics[width=0.5\textwidth]{figs_suppl/graph_representation.pdf}
+\begin{figure}[tb]
+    \includegraphics[width=0.7\textwidth]{figs_suppl/graph_representation.pdf}
     \caption{{\bf Pangenome graph representation}.
 Given a set of genomes \textit{PanGraph} will construct a pangenome graph by condensing all homologous sequences in \textit{pancontigs}, containing the alignment of these sequences.
-Genomes can then be represented as \textit{paths}, i.e.
-lists of pancontigs.
+Genomes can then be represented as \textit{paths} through the graph, i.e.~lists of pancontigs that are traversed when moving along the genome.
     In our representation a pangenome graph is in essence a collection of \textit{pancontigs} and \textit{paths}.
@@
 
-\begin{figure}[htb]
-    \includegraphics[width=0.5\textwidth]{figs_suppl/pancontig_alignment.pdf}
+\begin{figure}[tb]
+    \includegraphics[width=0.7\textwidth]{figs_suppl/pancontig_alignment.pdf}
     \caption{{\bf From pancontig consensus to alignment}.
@@
 Each \textbf{path} is an oriented collection of pancontigs that represents a genome.
-For each item in the sequence along with the pancontig id, PanGraph also stores the pancontig orientation (whether homology with the consensus is on the forward the reverse strand of the genome) and which variations on the pancontig consensus (i.e.
-which sequence in the pancontig alignment) are present in this precise location in the genome of the isolate.\\
+For each item in the sequence along with the pancontig id, PanGraph also stores the pancontig orientation (whether homology with the consensus is on the forward the reverse strand of the genome).
+Since a sequence can traverse a particular pancontig multiple times (e.g.~paralogs), an additional index labels possible repeated occurrences in the pancontig. \\
 
 From knowledge of all the pancontigs and paths, every single genome can be exactly reconstructed.
-Every part of the sequence is always encoded in a pancontig.
-``Small-scale'' difference between genomes (i.e.
+Every part of the sequence is encoded in a pancontig.
+``Small-scale'' differences between genomes (i.e.
 SNPs, small insertions and deletions) are encoded in pancontigs as variations over the consensus.
@@
 In this section we describe in more detail the algorithmic procedure used to build a pangenome graph.
-In essence, the full pangenome graph is built by placing genomes on terminal leaves of a guide tree, built from a sequence minimizer distance matrix.
+In essence, the full pangenome graph is built by placing genomes on terminal leaves of a guide tree built from a sequence minimizer distance matrix.
 These genomes are initially considered as elementary graphs composed of a single \textit{pancontig}.
@@
 
-We focus on four aspects:
+Below, we discuss the following points in greater detail:
 \begin{itemize}
@@
 
-\subsection{guide tree construction and distance matrix estimation}
+\subsection{Guide tree construction and distance matrix estimation}
 
@@
 To maximize parallelization, the final tree is a balanced binary tree with the same ordering of leaves as the one found by neighbour-joining.
-Here we briefly discuss how the complexity of this distance matrix estimation scales with the number $n$ and length $l$ of input sequences.\\
+Here we briefly discuss how the complexity of this distance matrix estimation scales with the number $n$ and length $L$ of input sequences.\\
 As a first step of the distance estimation, we sketch all input sequences together and build a unique sorted list of minimizers.
-This can be done in time $O(L \log L)$, where $L = n \times l$ is the total length of input sequences.
+This can be done in time $O(\mathcal{L} \log \mathcal{L})$, where $\mathcal{L} = n \times L$ is the total length of input sequences.
 With each minimizer in the list we also keep the information of the genome of origin.\\
 As a second step we count the number of minimizers shared by each pair of sequences.
-This can be done efficiently by scrolling the sorted list of minimizers (having size proportional to $L$) only once.
-For each chunk of equal minimizers in the list (having size between 1 and $n$, depending on how shared is the original k-mer) we increment by one the corresponding pairs in a minimizer count matrix.
-Depending on how similar the input sequences are, this operation has complexity between $O(n^2 l) = O(nL)$ (all minimizers shared) and $O(nl) = O(L)$ (no minimizer shared).\\
-In conclusion, depending on the similarity of the input sequences, evaluation of the pairwise distance matrix has complexity between $O(L \log L)$ and $O(L \log L + nL)$.\\
+This can be done efficiently by scrolling the sorted list of minimizers (having size at most proportional to $\mathcal{L}$) only once.
+For each chunk of equal minimizers in the list (having size between 1 and $n$, depending on common the original k-mer is) we increment by one the corresponding pairs in a minimizer count matrix.
+Depending on how similar the input sequences are, this operation has complexity between $O(n^2 L) = O(n\mathcal{L})$ (all minimizers shared) and $O(nL) = O(\mathcal{L})$ (no minimizer shared).\\
+In conclusion, depending on the similarity of the input sequences, evaluation of the pairwise distance matrix has complexity between $O(\mathcal{L} \log \mathcal{L})$ and $O(\mathcal{L} \log \mathcal{L} + n\mathcal{L})$.\\
 
@@
 
-\subsection{graph merging}
+\subsection{Graph merging}
 
@@
     \item pseudo-energy parameters $\alpha$ and $\beta$ (cf. eq.~(1) in the main text).
-    \item the choice of alignment kernel (minimap2 or mmseqs) and aligner precision (asm5, asm10 or asm20 for minimap2).
+    \item the choice of alignment kernel (minimap2 or mmseqs2) and aligner precision (asm5, asm10 or asm20 for minimap2).
 \end{itemize}
@@
 
-We point out that in our merging procedure only the consensus sequence of any two pancontigs is used in their alignment.
+During the graph merge procedure, only consensus sequences from each graph are used to find homologies between pancontigs, not the individual sequences represented by the pancontig. 
 This makes this procedure fast and is sufficient to capture the large-scale structural variation of the pangenome, but might introduce minor inconsistencies and artifacts in the alignment.
 To this end \textit{PanGraph} has the command \verb|polish|\footnote{See \url{https://neherlab.github.io/pangraph/tutorials/tutorial_3/}.}.
-This command ``polishes'' alignments by solving on each pancontig the multiple sequence alignment problem using MAFFT \cite{katoh2002mafft}.
-It has a simple interface: it takes as input a pangenome graph and produces another pangenome graph with the same structure but improved pancontig alignments.
-This efficiently separates the problem of finding large-scale structural variation, and nucleotide-level variation.
-When precise knowledge of the latter is needed, a graph can be \textit{polished} restricting the multiple sequence alignment problem only on the homologous sequences contained in each pancontig.
+This command ``polishes'' alignments by aligning all sequence represented by the pancontig using MAFFT \cite{katoh2002mafft}.
+The subcommand has a simple interface: it takes as input a pangenome graph and produces another pangenome graph with the same structure but improved pancontig alignments.
+This efficiently separates the problem of finding large-scale structural variation, and nucleotide-level variation by using established algoritms for multiple sequence alignment only on parts of the input genomes with established homology.
 
@@
 
-\subsection{PanGraph parameters effect}
+\subsection{PanGraph parameters}
 
@@
 This is done to capture structural variation at the gene scale, while ignoring small indels.
-The value can hoewever be tuned by the user depending on the specific use-case.\\
+The value can, however, be tuned by the user depending on the specific use-case.\\
 A similar role is played by the parameter $\alpha$.
@@
 The first intrinsic limit is set by the choice of alignment kernel and its sensitivity: minimap2 is significantly faster than mmseqs2, but it is in general unable to find matches between homologous sequences with more than 20\% divergence (with asm20 sensitivity).\\
-On top of this intrinsic limit of the aligner, a more fine-tuned setting of the acceptable amount of divergence can be achieved by tuning the value of $\beta$.
+On top of this intrinsic limit of the aligner, more explicit control of the acceptable amount of divergence can be achieved by tuning the value of $\beta$.
 From the definition of the pseudo-energy (cf.
 eq.~(1) in the main text) it follows that matches between homologous sequence with divergence higher than $1/\beta$ will have positive pseudo-energy and will thus not result in a merge.
-If the molecular clock hypothesis holds, tuning this parameter is equivalent to setting an effective time limit beyond which sequences are not to be merged in the same alignment.
-This parameter can be tuned based on the diversity of the sample in study.
-It can for example be useful in resolving situations in which the hypothesis of a star-like phylogeny in a pancontig alignment is broken, such as to separate loosely related paralogs, as discussed in the next section.
+If the molecular clock hypothesis holds, tuning this parameter is equivalent to setting an effective time limit beyond which contigs are not to be merged.
+%This parameter can be tuned based on the diversity of the sample in study.
 
@@
 The \texttt{marginalize} command possesses the optional \texttt{--reduce-paralog} flag.
-When set, this flag will result in the removal of artificial splits caused by duplicated pancontigs that always occurr in the same non-duplicated context.
+When set, this flag will result in the removal of artificial splits caused by duplicated pancontigs that always occur in the same non-duplicated context.
 These duplicated pancontigs are then split and absorbed in their context, thus simplifying the graph and loosing trace of the duplication.\\

### 2023-03-13 reviewer_response.tex
Update on Overleaf.

@@
 
-\response{It is true that there are multiple factors at play that could influenze the size of core pancontigs in different ways, but the observation that core pancontigs are longer than average is in agreement with our expectations.
+\response{It is true that there are multiple factors at play that could influence the size of core pancontigs in different ways, but the observation that core pancontigs are longer than average is in agreement with our expectations.
 Multiple core genes are often syntenic even in diverse species, compared to accessory genes that are often found in different contexts.
@@
 % \Liam{clarifying $N$ and $l$ makes sense.}
-\response{Are sorry that the formulation in the main text was too concise and a bit inaccurate.
+\response{ We are sorry that the formulation in the main text was too concise and a bit inaccurate.
 There are two main variables at play: the length of a genome $L$ and the number of isolates $n$. 
@@
 After all of the matches have been exhausted, the two graphs have undergone a preliminary merging, with some new merged pancontigs being created. 
-At this point we repeat the procedure looking for matches in an all-against-all comparison of consensus sequences of pancontigs of the newlyt generated graph to itself. 
+At this point we repeat the procedure looking for matches in an all-against-all comparison of consensus sequences of pancontigs of the newly generated graph to itself. 
 This self-mapping on the graph will detect any homology that could not have been merged at the previous stage, and will perform the corresponding merging. 
@@
       \caption{frequency and length distribution (respectively left and right panels) of pancontigs for the \textit{Helicobacter pylori} and \textit{Klebsiella pneumoniae} datasets (top and bottom).
-For each panel we report the results for pangenome graphs built using minimap2 aligner with standard parameter values and asm20 sensitivity (blue) and with mmseqs2 and null pseudo-energy parameters (orange).
-Distributions are weighed by pancontig length, so that the sum of all bins heights is equal to the total pangenome graph size.}
+For each panel we report the results for pangenome graphs built using minimap2 aligner with standard parameter values and asm10 sensitivity (blue) and with mmseqs2 and null pseudo-energy parameters (orange).
+Distributions are weighed by pancontig length, so that the sum of all bin heights is equal to the total pangenome graph size.}
       \label{fig:kp-vs-hp}

### 2023-03-16 reviewer_response.tex
Update on Overleaf.

@@
 \definecolor{response}{rgb}{0.2, 0.2, 0.2}
-\newcommand{\reviewer}[2]{\textbf{#1:} #2\vskip 5mm}
-\newcommand{\response}[1]{{\it {\color{response}\textbf{Response:} #1}}\vskip 5mm}
-\newcommand{\responsedraft}[1]{{\it {\color{purple}\textbf{ResponseDraft:} #1}}\vskip 5mm}
+\newcommand{\reviewer}[2]{{\it \textbf{#1:} #2\vskip 5mm}}
+\newcommand{\response}[1]{{{\color{response}\textbf{Response:} #1}}\vskip 5mm}
+\newcommand{\responsedraft}[1]{{{\color{purple}\textbf{ResponseDraft:} #1}}\vskip 5mm}
 
@@
 
+{\bf Dear Dr.~Rasko,}
+
+thank you for handling our manuscript ``PanGraph: scalable bacterial pan-genome graph construction" at {\it Microbial Genomics} and providing us with three expert reviews.
+These reviews have helped us to improve the manuscript and we are grateful for the constructive suggestions, comments, and criticism.
+
+Specifically, we have
+\begin{itemize}
+    \item rewritten parts of the introduction to reflect existing tools and prior work in a more accurate way,
+    \item extended the discussion of the graph merging and consensus sequence generation. In particular, we have greatly extended the SI with a detailed discussion of the methodology,
+    \item made various adjustments to text and figures as suggested by the reviewers,
+    \item archived the github repository via Zenodo \href{https://doi.org/10.5281/zenodo.7740393}{DOI: 10.5281/zenodo.7740393}
+\end{itemize}
+
+We hope that these revisions address the comments by the reviewers.
+
 \subsection*{Reviewer 1}
@@
 
-      There was no comparison with other tools - would be appropriate if e.g. downstream mapping performance were evaluated but that is outside the scope here.\\
-
-      On reading I had a few queries, which could perhaps be clarified further in the text.}
+      There was no comparison with other tools - would be appropriate if e.g. downstream mapping performance were evaluated but that is outside the scope here.}
 
@@
 We used minimap2 as alignment kernel, asm20 sensitivity and values of $\alpha = 100$, $\beta=20$.
-Running on 8 cores this took 18 hours and 12 minutes to complete and had maximum memory usage of 17Gb.}
+Running on 8 cores this took 18 hours and 12 minutes to complete and had maximum memory usage of 17Gb.
+But we have not explored a full order of  magnitude -- excessive fragmentation of the pangenome might slow down the pangraph construction beyond the expected linear scaling.
+}
 
@@
 Why is a more diverse pancontig more likely to become fragmented?}
-\response{We use the pseudo-energy defined in eq.~(1) as an heuristic to decide whether a homologous match detected by our aligner should result in a merge between two pancontigs, depending on the match length and on sequence divergence.
-In particular any choice of $\beta$ would prevent merges between sequences with divergence higher than $1/\beta$.
-Setting $\beta=0$ instead does not impose a divergence limit on merges other than the one implicitly set by the aligner sensitivity.
+\response{We use the pseudo-energy defined in eq.~(1) as an heuristic to decide whether a homologous match detected by the aligner should result in a merger between two pancontigs, depending on the match length and on sequence divergence.
+In particular any choice of $\beta$ would prevent mergers between sequences with divergence higher than $1/\beta$.
+Setting $\beta=0$ instead does not impose a divergence limit on mergers other than the one implicitly set by the aligner sensitivity.
 In this sense the choice of null energy parameter results in more diverged sequences being merged into the same pancontig.
@@
 This is all valuable information and it would be great to have it in the supplementary or main methods somewhere.}
-% \Marco{this point was also raised by R3.
-Mutations are always represented as variation on the consensus.
-The way consensus is propagated during graph merging might require more explanation.
-Maybe we could add a section in the supplementary with a somewhat more detailed description of the algorithmic steps we perform when merging two graphs?
-In the code \href{https://github.com/neherlab/pangraph/blob/e1c11012a3d19975d4b1dc61b1af3eb3949f7cd2/src/block.jl\#L1274}{this part} takes care of merging alignments between two blocks with only small (i.e.
-below cutoff) insertions and deletions.
-In short what we do is stitch the two alignments together using the cigar string as a guide, and re-compute the consensus on the newly-obtained alignment.
-This ensures that the consensus agrees with the majority of sequences in the alignment, but does not solve the full multiple sequence alignment problem.
-However we also provide the ``polish'' step for this, that can be executed after graph creation.
-This will take care of refining the block alignments.}\\
-% \Liam{I think a `narrative section' in the supplementary describing this in brief sounds great.}
-\response{This point was also raised by other reviewers and we appreciate the need to make this point clearer.
+
+\response{This point was also raised by other reviewers and we appreciate the need to make the representation of fine scale diversity clearer.
 In our pangenome graph representation each pancontig encodes an alignment using a consensus sequence and list of edits required to obtain each individual sequence from the consensus sequence.
@@
 How do you get all pairwise distances from that without another multiplier of $n$?}
-% \Marco{in the text we write that ``the pairwise distance matrix is estimated in a time that increases log-linearly in the \textbf{total length} of sequence that is analyzed.'' This might generate some confusion since the scaling depends both on the number of genomes $N$ and their length $l$.
-The algorithmic time for all the pairwise distances should scale as $\sim N^2 l \log l$.
-We should be in a regime where $N$ is not a limiting factor, it's just a multiplicative constant, and thus I think our statement is fundamentally correct.
-But we can make it more precise.
-This scaling is at any rate not a limiting factor in pangraph.}\\
-% \Liam{clarifying $N$ and $l$ makes sense.}
 \response{ We are sorry that the formulation in the main text was too concise and a bit inaccurate.
@@
 random choice?)
-                  % \Marco{the ranking would be the same and we perform only one of the two merges in the first phase, when comparing the two graphs.
-After this we align the graph on itself and the second merge should be performed at this point.
-This comment also suggests we could add a section in the supplementary with a more detailed description of the algorithm?}
             \item as you move on, you are aligning graphs, and you really skim over the issue of how you do that.
@@
 I would really appreciate it if you could clarify this and even ideally add some example figures in the supplement showing with some toy genomes how you align two graphs.
-                  % \Marco{another comment pointing at the fact that we could add a more detailed explanation about graph merging on the supplementary.
-What we do is using the aligner using as reference and queries the consensus sequences of each block in the two different graphs, and create a first version of the merged graph.
-We then repeat this procedure in an all-vs-all comparison of the consensus sequences of the blocks, and proceed with further merges.
-This is necessary because when two alignments overlap only one of the two is performed in the first alignment round, but the second will be performed in further rounds.}\\
-                  % \Liam{agree, sounds like a section in supplementary is a good idea to address these comments.}
+
       \end{itemize}}
-\response{We thank the reviewer for pointing out that we should have provided more explanation (the point was also raised by other reviewers).\\
+\response{We thank the reviewer for pointing out that we should have provided more explanation (the point was also raised by other reviewers).
 We added Section {\SIalgo}B in the SI, in which we expand on the explanation of how we merge two graphs.\\
-In short, we perform the merging procedure in two steps.
-In the first we use the aligner (minimap2 or mmseqs2) to scan for homologuos matches between the consensus sequences of all pancontigs of one graph and of the other.
+In short, the merging procedure involves two steps.
+In the first we use the aligner (minimap2 or mmseqs2) to scan for homologous matches between the consensus sequences of all pancontigs of one graph and of the other.
 This is achieved by simply passing all these consensus sequences as a single fasta file into mmseqs2 or via the API into minimap2.
 The resulting matches are ranked by pseudo-energy and mergers performed in order.
-Each pancontig can be subject to only one merge at this stage.
+Each pancontig can be subject to only one merger at this stage.
 For exact ties in this ranking of of matches, the behavior is not explicitly defined but depends on the order in which the matches appear in the output of the homology search tool.
@@
 This self-mapping on the graph will detect any homology that could not have been merged at the previous stage, and will perform the corresponding merging. 
-The self-mapping is repeated iteratively until no new merges can be performed.
+The self-mapping is repeated iteratively until no new mergers can be performed.
 Exact duplicate matches would thus be merged in successive iterations of this merge-process.
@@
 eq.~(1) in the main text) it follows that for an alignment of pancontig consensus sequences to have negative pseudo-energy must be that the number of SNPs divided by the alignment length $N_m/l$ must be less than $1/\beta$.
-Moreover, as the reviewer pointed out, the pseudo-energy ranking of merges effectively results in merges being performed in order of similarity, effectively in reverse evolutionary order.
+Moreover, as the reviewer pointed out, the pseudo-energy ranking of mergers effectively results in mergers being performed in order of similarity, effectively in reverse evolutionary order.
 \\
@@
 The negative pseudo-energy requirement means that no merger will be performed if the Hamming distance exceeds $1/\beta$...
-The pseudo-energy ranking also effectively results in merges being performed in reverse evolutionary order, with less diverged sequences being merged first.''
+The pseudo-energy ranking also effectively results in mergers being performed in reverse evolutionary order, with less diverged sequences being merged first.''
 }
@@
 When setting $\beta = 0$ the maximal divergence is determined by the aligner sensitivity.
-In this sense more merges are allowed, but all merges will have some implicit maximal divergence threshold (20\% or 30\%, depending on the aligner).}
+In this sense more mergers are allowed, but all mergers will have some implicit maximal divergence threshold (20\% or 30\%, depending on the aligner).}
 
@@
 Using a more sensitive aligner with null pseudo-energies means the core pangenome fraction increases considerably.
-      % TODO: add mention in SI
-      % QUESTION: this is more or less explicit in SI? Should I expand more?
-      % Liam: I think there are slightly more long pancontigs with mmseqs/null in Hp from looking at figure so seems worth saying
 }
@@
       \end{enumerate}}
-% \Marco{on the repo we have \href{https://github.com/neherlab/pangraph/blob/master/script/config/accnums.json}{this file} for the list of all accession numbers we use, and \href{https://github.com/neherlab/pangraph/blob/master/script/config/projection_strains.json}{this file} for the marginalization pairs.
-I could create some excel spreadsheet with accession numbers that we could add to the submission?}\\
-% \Liam{for sure that's all they want (most biologists don't like json!)}
 \response{We now included in the submission two csv files.
@@
 \texttt{marginalize\_pairs.csv} contains instead a list of all the pairs used in the graph marginalization section.
-      % TODO: archive repository state.
+Additionally, to guarantee persistence we archived the repository via Zenodo (\href{https://doi.org/10.5281/zenodo.7740393}{DOI: 10.5281/zenodo.7740393})
 }
@@
 \reviewer{R2.2}{Why on earth is the ``get an MSA of a block'' option in pangraph called ``polish''?}
-\response{In our method pancontig merges are performed using only the consensus sequence as a guide.
+\response{In our method pancontig mergers are performed using only the consensus sequence as a guide.
 This makes the merging procedure very fast, but can introduce minor inconsistencies and artifacts with insertions in the alignment as insertions are not aligned to each other in the procedure.
@@
 \response{
-      We apologize if our previous discussion of the literature was incomplete. It was not our intention to downplay prior work.
-      We thank the reviewer for highlighting additional important work and aspects that we have now added to better contextualise our work.
+      We apologize if our previous discussion of the literature was inaccurate and incomplete. It was not our intention to downplay prior work.
+      We thank the reviewer for highlighting additional important papers and aspects that we have now added to better contextualise our work.
       We have rewritten the introduction to reflect previous work in this field and added to the discussion to further highlight previous and alternative approaches.
@@
       % Liam: I think good to say in more detail what has been added here. 
-      Specifically, we have
+      Specifically, 
 \begin{itemize}
-    \item The sentence in the abstract mentioned by the reviewer has been replaced by a general summary of the need to and challenges in understanding evolution of genome structure.
+    \item We replaced the sentence in the abstract mentioned by the reviewer by a general summary of the need to and challenges in understanding evolution of genome structure.
     \item We extended the discussion of other methods in the introduction.
     In particular we extended the paragraph on methods based on de Bruijn graphs, mentioning also compacted generalization and the fact that tools exist that are able to build graphs containing thousands of isolates.
-We now cite more tools and put more emphasis on their applications, restricting it to tools specific for microbial genomics (e.g.
-we did not include PGGB).
-We removed the previous claim that homology detection was limited by the k-mer size, and cite tools that can be used to extend homology detection through small sequence modifications.
+We now cite more tools and put more emphasis on their applications, focusing on tools specific for microbial genomics.
     \item We expanded in the same way the paragraph on homology-based and gene-based approaches, in which we now discuss more tools putting emphasis on their applications.
-    \item We added a paragraph in the discussion in which we tried to emphasize how \textit{PanGraph}, as the reviewer remarks, treats nucleotide-level variation and large-scale structural variation in different ways, thereby facilitating separate analyses of these aspects.
+    \item We now discuss how \textit{PanGraph} treats nucleotide-level variation and large-scale structural variation in different ways, thereby facilitating separate analyses of these aspects. 
 \end{itemize}
 }
-% \Marco{Tentative extension - to be updated depending on how we modify the intro/discussion:\\
-
-% To amend this we performed the following changes:
-% \begin{itemize}
-%     \item The sentence in the abstract mentioned by the reviewer has been removed and modified.
-
-%     \item We extended the discussion of other methods in the introduction.\\
-%     In particular we extended the paragraph on methods based on de Bruijn graphs, mentioning also compacted generalization and the fact that tools exist that are able to build graphs containing thousands of isolates.
-We now cite more tools and put more emphasis on their applications, restricting it to tools specific for microbial genomics (e.g.
-we did not include PGGB).
-We removed the previous claim that homology detection was limited by the k-mer size, and cite tools that can be used to extend homology detection through small sequence modifications.\\
-%     We expanded in the same way the paragraph on homology-based and gene-based approaches, in which we now discuss more tools putting emphasis on their applications.
-
-%     \item We added a paragraph in the discussion in which we tried to emphasize how \textit{PanGraph} compares to gene-based and k-mer based PGG methods.\\
-%     In essence, methods based on de Bruijn graph have the advantage of being fast since they do not require alignments.
-They also do not necessarily require fully assembled genomes.
-However in these models the graph encodes in the same way nucleotide-level variation and large-scale structural variation, and specific post-processing of the graph is needed to distinguish the two.
-Conversely, PanGraph requires fully-assembled genomes and is potentially slower, but its pangenome graph representation provides a natural separation between small sequence differences (mutations or small indels) and big structural changes (gene gain/loss, inversions, rearrangements).\\
-%     Gene-based pangenome graphs have instead the advantage of being smaller in size (a node corresponds to a gene) and they can include more divergent sequences, since homology detection can be performed on protein sequences.
-Differently from homology-based methods like PanGraph however, they are often sensitive to annotation errors.
-% \end{itemize}
-% }
 
@@
 \response{
-We apologize and we modified the sentence in question. We also extended the discussion of de Bruijn graphs methods in the introduction (see answer above).\\
 We thank the reviewer for pointing out these important contributions that were not previously discussed in the paper.
+We agree that these are important context for our work and we modified the sentence in question. We also extended the discussion of de Bruijn graphs methods in the introduction (see answer above).\\
 }
@@
 This could be measured as part of the simulated data where identical paralog sequences of various lengths and copy number could be modeled.}
-% \Liam{I am not sure I agree that constructing a pangenome graph means one must distinguish orthologs from paralogs.
-The structure is still useful even if this isn't done.
-Maybe there is a way of doing something of what they propose with simulations, but I suspect this isn't necessary for the editor to accept the paper\ldots}\\
-% \Marco{I agree, I think it's a different design choice.
-We could also mention that the marginalize command has an option to project out paralogs, but should be a post-processing step rather than something that we want to do while building the graph.
-One can a posteriori decide whether one wants to split paralogs using structural information (flanking blocks) or one can even prevent the merging using diversity and the $\beta$ parameter.}\\
 \response{
@@
 
-But we agree from experience that these can sometimes complicate the interpretion of the pangenome graph, in particular in presence of high copy-number mobile elements.  
+But we agree from experience that these can sometimes complicate the interpretation of the pangenome graph, in particular in presence of high copy-number mobile elements.  
 Currently \textit{PanGraph} supports two different ways of controlling paralog splits.
 The first one is based on sequence divergence as mentioned above.
-By tuning the value of the energy parameter $\beta$ the user can set a divergence limit. This would prevent the merging of paralogs that have sequence divergence higher than $1/\beta$.\\
-The second method relies instead on information on genomic context. The \texttt{marginalize} command possesses the optional \texttt{--reduce-paralog} flag. 
-When set, this flag will result in the removal of artificial splits caused by duplicated pancontigs that always occurr in the same non-duplicated context. 
-These duplicated pancontigs are then split and absorbed in their context.
-In addition, pangraph can export a graph omiting all duplicated blocks, which is a simplified version of the graph often useful for visualization.
+By tuning the value of the energy parameter $\beta$ the user can set a divergence limit. This would prevent the merging of paralogs that have sequence divergence higher than $1/\beta$.
+In addition, pangraph can export a graph omitting all duplicated blocks, which is a simplified version of the graph often useful for visualization.
 
@@
 We appreciate the reviewer's perspective and have added SI sect.~{\SIalgo}D to include discussion of these considerations.
-
-% TODO: confirm with Nick the paralog split algorithm.
 }
@@
       \begin{itemize}
-            \item In addition to the pseudo-energy constraint, merges are refused if they have alignment length shorter than $\Lthr$.
+            \item In addition to the pseudo-energy constraint, mergers are refused if they have alignment length shorter than $\Lthr$.
             \item During merging, if the alignment contains long ($> \Lthr$) indels, these will cause a split in the merged pancontigs and will be treated as separate pancontigs.
@@
 We avoid performing a multiple sequence alignment during merging to keep the algorithm scalable. 
-However we do update the consensus of the pancontig at each merge using the majority rule on the merged alignment.\\
+However we do update the consensus of the pancontig at each merger using the majority rule on the merged alignment.\\
 The idea behind this approach is that the consensus sequence is needed to find large-scale homology, and resolving the fine-grained structure of the alignment is not needed at this step and would require an excessive amount of resources.
@@
 %  TODO: before submission
-% - [ ] archive repository state and add in the response
+% - [x] archive repository state and add in the response
 % - [ ] add files with accession numbers and pairs to the submission

### 2023-03-16 main.tex
Update on Overleaf.

@@
     With the wide-spread usage of long-read sequencing, the number of high-quality, complete genome assemblies has increased dramatically.
-    % Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.
     In addition to pangenomic approaches that focus on the variation in the sets of genes present in different genomes, complete assemblies allow investigations of the evolution of genome structure and gene order.
-    This latter problem, however, is computationally demanding and conceptually challenging with few tools available that allow to shed light on these dynamics. 
+    This latter problem, however, is computationally demanding with few tools available that allow to shed light on these dynamics. 
     Here, we present \emph{PanGraph}, a Julia-based library and command line interface for aligning whole genomes into a graph.
@@
 However, in these methods the graph structure encodes at the same time nucleotide-level variation and large-scale structural rearrangements. 
-For investigations of the evolutionary dynamics of these genomes, separating these two scales if often of interest. 
-%Separating these two scales is often difficult, and specific techniques have to be employed to detect large-scale homology in these graphs \cite{angiuoli2011mugsy,minkin2013sibelia}.
+For investigations of the evolutionary dynamics of these genomes, separating these two scales is often of interest.
 
@@
 These early methods however usually scale poorly to large sets of genomes.
-%More recently techniques based on homology have been used to compare genomes across diverse sets of species across the tree of life, at the cost of memory \cite{armstrong2020progressive}.
 Other methods rely on first grouping genes into orthologous clusters.
@@
 
-Here we present \emph{PanGraph}, a Julia \cite{bezanson2017julia} library and command line interface, designed to efficiently align large sets of closely related genomes into a pangenome graph on personal computers.
-The \emph{PanGraph} algorithm creates nodes of homologous sequences through nucleotide alignments alone, without requiring gene annotation.
+Here we present \emph{PanGraph}, a Julia \cite{bezanson2017julia} library and command line interface, designed to efficiently align large sets of closely related microbial genomes into a pangenome graph on personal computers.
+The \emph{PanGraph} algorithm groups homologous sequences through nucleotide alignments alone, without relying on gene annotation.
 The resulting graph both compresses the input sequence set and succinctly captures the population diversity at multiple scales: from nucleotide mutations and indels to structural polymorphisms driven by inversions, rearrangements, and gene gain/loss.
@@
 The underlying graph data structure can be exported into numerous formats for downstream analysis and visualization in software such as Bandage \cite{wick2015bandage}.
-%The algorithm, along with extensive documentation, examples, and instructions for installation, are open source and can be found on GitHub.
-%We anticipate PanGraph to be immediately useful in understanding microbial evolution in natural populations.
 
@@
 \end{equation}
-where $\ell$, $N_c$, and $N_m$ denote the alignment length, number of \emph{pancontigs} created by the merger, and number of polymorphisms per genome in the newly created {pancontig} respectively.
+where $\ell$, $N_c$, and $N_m$ denote the alignment length, number of additional \emph{pancontigs} created by the merger, and number of polymorphisms in the alignment of the consensus sequences, respectively.
 See Fig \ref{fig:visualization}B for a graphical definition for each term.
@@
 Importantly, this divergence threshold is applied to the comparison of the consensus sequences of the merger candidates.
-The pseudo-energy ranking also effectively results in merges being performed in reverse evolutionary order, with less diverged sequences with long-range synteny being merged first.
+The pseudo-energy ranking also effectively results in mergers being performed in reverse evolutionary order, with less diverged sequences with long-range synteny being merged first.
 
@@
 Structural variation at a scale smaller than this threshold (i.e.~short indels) is stored in pancontig alignments, rather than in the graph structure, and no mergers are performed if they have length shorter than this threshold.
-For instance, when a merge would generate flanking contigs shorter than this threshold, the merging is performed but the alignment is extended to include these regions.
+For instance, when a merger would generate flanking contigs shorter than this threshold, the merging is performed but the alignment is extended to include these regions.
 The default value for this parameter is 100bp.
@@
 
-At the graph level, the merge of two \emph{pancontigs} defines a new \emph{pancontig}, connected on both sides by edges to the neighboring \emph{pancontigs} of both inputs, and thus locally collapses the two graphs under consideration.
-At the nucleotide level, the pairwise alignment of two \emph{pancontigs} maps the consensus of one onto the other; the merge of two \emph{pancontigs} requires the application of the map onto the nucleotide level diversity of the underlying multiple-sequence alignment.
+At the graph level, the merger of two \emph{pancontigs} defines a new \emph{pancontig}, connected on both sides by edges to the neighboring \emph{pancontigs} of both inputs, and thus locally collapses the two graphs under consideration.
+At the nucleotide level, the pairwise alignment of two \emph{pancontigs} maps the consensus of one onto the other; merging two \emph{pancontigs} requires the application of the map onto the nucleotide level diversity of the underlying multiple-sequence alignment.
 Once both sets of sequences are placed onto a common coordinate system, the resultant consensus sequence, and thus polymorphisms, are recomputed.
@@
 \textit{PanGraph}, the tool presented here, is able to capture the structural and nucleotide diversity in both the core and accessory genome in a scalable way.
-The resulting data structure captures large-scale structural variation in the connectivity of the graph, while nucleotide-level variation being included in pancontig alignments.
-
-We believe that this feature, i.e. the fact that our representation naturally separates nucleotide level diversity from larger structural variation, is one of the main merits of our tool.\\
-De Bruijn graphs types of representation can be built quickly since they do not necessitate alignment, and even directly from reads without the need of high-quality assemblies.
-However they do not posses this scale separation, with every mutation generating a structural split in the graph.
-Some methods however exist to detect longer-range homology.
-These involve some form of smoothing of the graph to extend homology matches over small-scale variation, and have been used for the detection of core regions with applications to genetic engineering \cite{schulz2022sequence} or to find synteny blocks \cite{minkin2013sibelia}.
-These are particularly effective on very similar sequences with very low divergence.\\
-A second widespread type of representation uses orthologous gene clusters as node in the graphs, and introduces links between clusters that are adjacent on some of the isolates.
-These have proven useful in many different tasks including the categorization of core/accessory regions \cite{gautreau2020ppanggolin,sutton2021pan}, reducing reference-bias in variant detection \cite{colquhoun2021pandora} and the improvement of gene annotations \cite{tonkin2020producing}.
-In these representation the graph usually has a smaller number of nodes, and can include more diverged sequences by detecting homology on protein sequences.
-However this representation is naturally sensitive to gene annotation errors.
-\textit{PanGraph} instead operates by construction in a gene-agnostic way, being only based on sequence homology, and is thus robust to this type of errors.\\
+The resulting data structure captures large-scale structural variation in the connectivity of the graph, while nucleotide-level variation is included in pancontig alignments.
 
@@
 This study was funded by the University of Basel and the NSF.
-L.P.S. is a Sir Henry Wellcome Postdoctoral Fellow funded by Wellcome (Grant 220422/Z/20/Z).
+L.P.S.~is a Sir Henry Wellcome Postdoctoral Fellow funded by Wellcome (Grant 220422/Z/20/Z).
 

### 2023-03-16 supplementary.tex
Update on Overleaf.

@@
 \begin{figure}[tb]
-    \includegraphics[width=0.7\textwidth]{figs_suppl/pancontig_alignment.pdf}
+    \includegraphics[width=0.6\textwidth]{figs_suppl/pancontig_alignment.pdf}
     \caption{{\bf From pancontig consensus to alignment}.
@@
     \item Once the list of all matches has been exhausted the two graphs have been merged into one.
-However at the previous stage only one merge could be performed per pancontig, which could be insufficient to collapse all of the homology contained in the graph.
+However at the previous stage only one merger could be performed per pancontig, which could be insufficient to collapse all of the homology contained in the graph.
 To this end we repeat the procedure described in step 3, but instead of looking for matches between pancontigs of two different graphs, we perform an all-against-all homology search between all pancontigs in the same graph.
@@
 
-During the graph merge procedure, only consensus sequences from each graph are used to find homologies between pancontigs, not the individual sequences represented by the pancontig. 
+During the graph merging procedure, only consensus sequences from each graph are used to find homologies between pancontigs, not the individual sequences represented by the pancontig. 
 This makes this procedure fast and is sufficient to capture the large-scale structural variation of the pangenome, but might introduce minor inconsistencies and artifacts in the alignment.
@@
 From the definition of the pseudo-energy (cf.
-eq.~(1) in the main text) it follows that matches between homologous sequence with divergence higher than $1/\beta$ will have positive pseudo-energy and will thus not result in a merge.
+eq.~(1) in the main text) it follows that matches between homologous sequence with divergence higher than $1/\beta$ will have positive pseudo-energy and will thus not result in a merger.
 If the molecular clock hypothesis holds, tuning this parameter is equivalent to setting an effective time limit beyond which contigs are not to be merged.

### 2023-03-16 supplementary.tex
Update on Overleaf.

@@
           \begin{enumerate}
-              \item We check that no prior merges have been performed on the same two pancontigs.
+              \item We check that no prior mergers have been performed on the same two pancontigs.
 Sometimes multiple matches are found on the same pancontig, and in this case we prioritize the ones with more negative pseudo-energy.
@@
 This completes the creation of a new pancontig.
-              \item The original starting pancontigs are then removed from the pangenome graph, and no new merges can be performed on them at this step of the algorithm.
+              \item The original starting pancontigs are then removed from the pangenome graph, and no new mergers can be performed on them at this step of the algorithm.
 The newly-created pancontigs are added to the graph and the corresponding paths are updated.
@@
 To this end we repeat the procedure described in step 3, but instead of looking for matches between pancontigs of two different graphs, we perform an all-against-all homology search between all pancontigs in the same graph.
-Merges are performed following the same rules, and this procedure is repeated until no new merges can be performed.
+Mergers are performed following the same rules, and this procedure is repeated until no new mergers can be performed.
     \item In the final step of the procedure we perform some final grooming of the graph.
@@
 
-As graphs are propagated along the guide tree more and more merges are performed. This makes pancontigs progressively shorter but their alignments deeper.\\
+As graphs are propagated along the guide tree more and more mergers are performed. This makes pancontigs progressively shorter but their alignments deeper.\\
 

### 2023-03-16 main.tex
Update on Overleaf.

@@
 The efficient implementation of \textit{PanGraph} allows it to create pangenome graphs containing hundreds of isolates in a few hours on a 8-core machine.
-The size of pancontigs and the fraction of core sequence have good scaling properties with the number of isolates in the graph, indicating that \textit{PanGraph} is able to successfully capture pangenome properties.\\
+The size of pancontigs and the fraction of core sequence have good scaling properties with the number of isolates in the graph, indicating that \textit{PanGraph} is able to successfully capture pangenome properties.
+By construction \textit{PanGraph} operates in a gene-agnostic way, being only based on sequence homology, and is thus robust to gene annotation errors.\\
 One of the main limitations is the diversity of the input sequences.

### 2023-03-16 main.tex
resubmitted

@@
     Each genome is represented as an undirected path along vertices, which in turn, encapsulate homologous multiple sequence alignments.
-    The resultant data structure succinctly summarizes population-level nucleotide and structural polymorphisms and can be exported into a several common formats for either downstream analysis or immediate visualization.
+    The resultant data structure succinctly summarizes population-level nucleotide and structural polymorphisms and can be exported into several common formats for either downstream analysis or immediate visualization.
 \end{abstract}
@@
 While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging.
-Colored and compacted generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets \cite{iqbal2012novo,muggli2017succinct}, with tools existing to build graphs containing thousands of isolates \cite{holley2020bifrost}.
+Colored and compacted generalizations of the de Bruijn graph-based assemblers have been successfully used to build graphs from large sequence sets \cite{iqbal2012novo,muggli2017succinct}, with tools existing to build graphs containing thousands of isolates \cite{holley2020bifrost}.
 These graphs have been shown to be useful in problems such as the detection of the core genome \cite{schulz2022sequence} or the improvement of gene prediction, annotation and clustering \cite{horsfield2023accurate}.
@@
 An orthogonal approach has been to formulate the inference of the pangenome graph as a multiple genome alignment.
-Some early methods relied on homology searches to locate syntenic regions, termed \textit{locally colinear blocks} \cite{angiuoli2011mugsy,darling2010progressivemauve}, and proved useful in various problems such as improving gene annotations \cite{angiuoli2011improving}.
+Some early methods relied on homology searches to locate syntenic regions, termed \textit{locally collinear blocks} \cite{angiuoli2011mugsy,darling2010progressivemauve}, and proved useful in various problems such as improving gene annotations \cite{angiuoli2011improving}.
 These early methods however usually scale poorly to large sets of genomes.
@@
 \subsection{Iterative graph alignment}
-The full pangraph representing all genomes is constructed by aligning/merging graphs that represent subsets of the genomes in an iterative manner illustrated in Fig.~\ref{fig:visualization}.
+The full pangraph representing all genomes is constructed by aligning/merging graphs that represent subsets of the genomes in an iterative manner illustrated in fig.~\ref{fig:visualization}.
 The iteration starts with one subgraph per input genome, each representing its respective genome as a single \emph{pancontig}.
 Pairs of subgraphs are aligned in a postorder traversal of the guide tree.
-The identified homologous intervals of the pancontig are then merged, thereby creating shorter contigs that represent homologous sections of multiple input genomes, see Fig.
+The identified homologous intervals of the pancontig are then merged, thereby creating shorter contigs that represent homologous sections of multiple input genomes, see fig.
 \ref{fig:visualization}B for a graphical depiction.
@@
 If the above alignment step detected homologous stretches between the consensus sequences of two pancontigs, these pancontigs, or parts of them, can be merged.
-It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order.
+It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make a choice on which potential mergers are performed and in which order.
 We rank each alignment between two pancontigs according to the pseudo-energy
@@
 
-The parameter $\alpha$ controls the fragmentation of the graph and imposes a minimal length on the pan-contigs resulting from the merger.
-This parameter has not effect if merging does not increase the number of contigs, but mergers that introduce 1, 2, 3 or 4 cuts are only performed for sufficiently long homologous stretches as parameterized by $\alpha$.
+The parameter $\alpha$ controls the fragmentation of the graph and imposes a minimal length on the pancontigs resulting from the merger.
+This parameter has no effect if merging does not increase the number of contigs, but mergers that introduce 1, 2, 3 or 4 cuts are only performed for sufficiently long homologous stretches as parameterized by $\alpha$.
 In addition to the fragmentation parameter $\alpha$, there is a further parameter $\Lthr$ that explicitly controls the minimal length of pancontigs.
@@
 This shortens considerably the time needed, but might introduce minor inconsistencies and artifacts in the alignments.
-After building the graph, these can be removed by performing a standard multiple alignment of sequences within each pancontig (see ``polish'' command in documentation and SI sect.{\SIalgo}B).
+After building the graph, these can be removed by performing a standard multiple alignment of sequences within each pancontig (see ``polish'' command in documentation and SI sect.~{\SIalgo}B).
 
@@
 \subsection{Parallelism}
-\emph{PanGraph} guide trees are balanced binary trees that break the task into many independent sub-problems and thereby enable scalable parallelism, as shown in Fig \ref{fig:visualization}A for a cartoon example.
+\emph{PanGraph} guide trees are balanced binary trees that break the task into many independent sub-problems and thereby enable scalable parallelism, as shown in fig.~\ref{fig:visualization}A for a cartoon example.
 Each internal node of the guide tree represents a job that performs a single pairwise graph alignment between its two children.
@@
 PanGraph is published under an MIT license with source code, extensive documentation, examples, and instructions for installation available at \url{github.com/neherlab/pangraph}.
-Additionally, pre-built binaries are available as GitHub releases.
 All data and scripts used to validate PanGraph are available within the same repository.\footnote{Instructions on how to run the analysis scripts are available at \url{https://github.com/neherlab/pangraph/blob/master/script/README.md}}
@@
 The algorithmic complexity was measured empirically by constructing pangenome graphs from data generated by simulating populations of varying size and sequence length.
-The mean and standard deviation of the run-time obtained from 50 iterations are shown in Fig.~\ref{fig:toy-performance}.
+The mean and standard deviation of the run-time obtained from 50 iterations are shown in fig.~\ref{fig:toy-performance}.
 Importantly, PanGraph's computational complexity grows \emph{linearly} with the number of input genomes once the number of genomes exceeds a certain threshold and parallelism can be exploited.
@@
 For each alignment kernel and value of average sequence divergence in the population we evaluated the fraction of breakpoints that have displacement greater than the default minimal pancontig size for Pangraph of 100 bp (cf.
-Fig.~\ref{fig:toy-accuracy}).
+fig.~\ref{fig:toy-accuracy}).
 On the sequences generated by our simulations, \textit{minimap2} with option \textit{asm20} show a loss of accuracy at an average pairwise sequence divergence of around 5\%, while \textit{mmseqs2} is accurate up to around 12\%, at the cost of higher computational time.
-These threshold are lower than the nominal sensitivity of the aligners, which are 10\% and 20\% for minimap2 with settings \textit{asm10} and \textit{asm20}, respectively, and around 30\% for \textit{mmseqs2}.
+These thresholds are lower than the nominal sensitivity of the aligners, which are 10\% and 20\% for minimap2 with settings \textit{asm10} and \textit{asm20}, respectively, and around 30\% for \textit{mmseqs2}.
 This discrepancy is due to the fact that for accurate graph constructions, the largest divergence, rather than the average divergence, is the relevant quantity.
@@
 Different alignment kernels are expected to reach different accuracy on datasets with different diversities (cf.
-Fig.~\ref{fig:toy-accuracy}).
+fig.~\ref{fig:toy-accuracy}).
 Moreover the use of null energy parameters ($\alpha = \beta = 0$) is expected to remove the threshold divergence of 10\% associated with the standard values of parameters ($\alpha=100$, $\beta=10$) at the cost of more fragmented pangenome graphs.\\
@@
 In this case the use of \textit{mmseqs2} and null energy parameters increases core pangenome fraction, decreases the total pangenome size and and does not compromise the fragmentation of the graph (cf.
-Fig.~\ref{fig:panx-benchmark}B-C and SI Fig.~{\SIfigBenchmark}).
+fig.~\ref{fig:panx-benchmark}B-C and SI fig.~{\SIfigBenchmark}).
 The divergence of the \textit{P.~marinus} dataset sits instead beyond the capabilities of \textit{Pangraph}.
@@
 
-For species other than \textit{P.~marinus}, the pangraphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in pancontigs spanning several kbps (see SI Fig.~\SIfigBenchmark).
+For species other than \textit{P.~marinus}, the pangraphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in pancontigs spanning several kbps (see SI fig.~\SIfigBenchmark).
 For these species the use of null energy parameters slightly increases merging and sequence compression, at the cost of slightly more fragmented graphs (cf.
@@
 To do so, we build pangenome graphs from randomized subsets of isolates of increasing size from the \textit{E.~coli} dataset.
-Results are reported in Fig.~\ref{fig:panx-size}.\\
+Results are reported in fig.~\ref{fig:panx-size}.\\
 The number and size of pancontigs scale sub-linearly with the number of isolates, with more than 50\% of the pangenome being included in the 10\% longest pancontigs.
@@
 We then picked 50 random pairs of strains from the same set and built two graphs for each pair: the graph obtained by marginalizing the complete graph on the pair (i.e.
-marginalized graph in Fig.~\ref{fig:marginalization}A top), and the graph built directly from the pair (i.e.
-pairwise graph in Fig.~\ref{fig:marginalization}A top).
+marginalized graph in fig.~\ref{fig:marginalization}A top), and the graph built directly from the pair (i.e.
+pairwise graph in fig.~\ref{fig:marginalization}A top).
 In order to compare both graphs, we compute the partition they generated on the genome of the isolates they include.
 Each genome is partitioned in pancontigs that are either shared on the pair, or private to one isolate (cf.
-Fig.~\ref{fig:marginalization}A bottom).
+fig.~\ref{fig:marginalization}A bottom).
 We can classify the segments in the intersection of the two partitions in different categories, depending on whether the two partitions agree on whether the segment is shared or private.
@@
 Indeed, we verified that over the pairs we picked, these segments cover a very small fraction of the genome ($<1\%$) and have average size compatible with the default 100 bp precision threshold of \textit{PanGraph} set by the value of $\Lthr$ (cf.
-Fig.~\ref{fig:marginalization}B-C).
+fig.~\ref{fig:marginalization}B-C).
 Conversely, segments on which the partitions agree have average size of several kbps.
@@
 The resulting distribution of shared sequence is in very good agreement with the fraction of segments that are shared on both graphs (cf.
-Fig.~\ref{fig:marginalization}B), suggesting that homologous sequences are correctly merged on the complete, marginalized and pairwise graphs.\\
+fig.~\ref{fig:marginalization}B), suggesting that homologous sequences are correctly merged on the complete, marginalized and pairwise graphs.\\
 We performed the same test on species from the other datasets, obtaining similar results on all species whose divergence is compatible with \textit{PanGraph} capabilities (cf.
-SI Section {\SImarginalize} and SI Fig.~\SIfigMarg).
+SI Section {\SImarginalize} and SI fig.~\SIfigMarg).
 
